### 2023-10-24 analysis/paper/paper.Rmd
chipping away at paper

---
title: "National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou ranges in Canada, both survival and recruitment decrease with the percentage of the range that is disturbed. However, there is also variation in demographic rates among ranges, particularly where anthropogenic disturbance is low, and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships. Demographic projections derived from local range-specific data are also uncertain because few places have enough long-term monitoring data to characterize local demographic-disturbance relationships. We present a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections. We combine this model with methods for simulating local population dynamics and monitoring to show how monitoring requirements depend on landscape condition. Where anthropogenic disturbance is very high or very low, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At intermediate disturbance levels where uncertainty is higher, deploying few collars for more years may be a lower cost option for improving the reliability of status projections relative to more intensive short-term monitoring. We provide an R Shiny interface to enable others to explore the implications of alternative monitoring scenarios, disturbance trajectories, priors, and other model parameters for boreal caribou demographic projections. This tool can help users assess monitoring needs and strategies.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

<!-- Link to data here: https://drive.google.com/file/d/179lw6MPK4l3xGmded9kofzJkB6VlJVb9/view?usp=sharing -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figs/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

NOTE: find working version of intro and discussion [here:]
(<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit?usp=sharing&ouid=111541730342945675365&rtpof=true&sd=true>).

```{=html}
<!--
Motivating example - forecasting impacts of disturbance - WBI and RoF example. In order to reduce uncertainty, need to integrate local information on current state of the population into forecasts.

Lots of variability among caribou ranges in monitoring data: recency, duration, types of data. E.g. 27+ consecutive years of both telemetry-survival and calf/cow-ratios in most Alberta ranges, vs. 1-4 years of data in most Ontario ranges (and that's old) *note we may want to be careful about inter-province comparisons for political reasons, but the variation is super relevant when to-date we've had the same 'rules' apply across the board...

In Missisa (and other places, esp Ontario) available local data are old and limited. Not clear how indicative older data is of state of population. Need new data to resolve questions, but what type, how much is enough. Can't monitor in great detail everywhere all the time. More generally, need for tools to help assess the value of additional monitoring and the best way to invest additional monitoring resources, given the info that is already available. 
-->
```
# Methods

```{r setup2}
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(ggpubr)
library(MetBrewer)
library(here)
library(tmap)
library(sf)

devtools::load_all(here::here())

#set paths
data_path_raw <- "analysis/data/raw_data/"
data_path_drvd <- "analysis/data/derived_data"

pal_nm <- "Redon"

fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)

runScenarios = F 
doPriorCalibration=F #Note the prior calibration is time consuming. 
```

## Caribou population monitoring: survival and recruitment metrics

Demographic data for caribou populations, including abundance, vital rates and population growth, are most often obtained through the tracking of caribou cows marked with telemetry collars [e.g. @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@johnson_science_2020]. Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with collars carrying  Very High Frequency (VHF) and/or Global Positioning Satellite (GPS) transmitters allowing remote detection and data retrieval [see @nbckc_nbckc_2021]. Because these collars send out a mortality signal when no longer active after a predetermined period of time, adult survival can be estimated by tracking the fate of marked individuals. See @kaplan_nonparametric_1958; @pollock_survival_1989 for original approach and @mcloughlin_declines_2003; @hervieux_widespread_2013; @eacker_web-based_2019 for examples of application to caribou. Male caribou are not frequently collared.

Collared females are also used in most estimates of recruitment. Recruitment surveys involve locating collared individuals by aircraft and then surveying the entire group of caribou to estimate the ratio of calves to cows in the group [@nbckc_nbckc_2022]. Because this estimate is based on more than just the collared sample of cows, the sample size for recruitment estimates is typically larger than for survival (see multiplier parameter $w$, cowMult below). Note that occasionally young bulls can be misidentified as cows (parameter $u$ below; refs? ), and there is also a possibility that some calves may be missed during aerial composition surveys (parameter $z$ below). We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (parameter $c$; see Supplement B and below for details). Telemetry and aerial survey data from these methods can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for monitoring population change of many caribou herds [e.g. @decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019].

## Anthropogenic disturbance and monitoring scenarios

```{r monitoringScenarios, echo=FALSE,message=FALSE}

source(here::here("analysis/scripts/formatScenarios.R"))

knitr::kable(tbl,caption="Monitoring scenario parameters.")
```

```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))

```

For this investigation we assume collars are deployed in January, and
last for up to 6 years. The target number of collared females $n$ varies
among scenarios, as does the total duration of the monitoring program $d$ and the ratio of uncollared to collared adults in the recruitment survey $w$ 
(Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be
replaced each year ($o=1$), or there may be a 3 year gap between
deployments ($o=4$). We focus on exploring a set of scenarios in which
anthropogenic disturbance, measured as the percentage of area within 500
m of disturbance [@johnson_science_2020; @eccc_scientific_2011],
increases by 1% per year from a variety of starting points (Figure \@ref(fig:figDisturbance)).
Scenarios in which anthropogenic disturbance does not increase over time
are included in Supplement D.

```{=html}
<!--
*Monitoring WG input from BC, ON, NL, (MB pending): lots of detail provided but generally those numbers suggest we could try values of ~3-10 [e.g. 3,6,9?] for the cow multiplier, 5-50 for collars [e.g. 5,10,20,40,60?], [not sure about renewal interval, but see SaraM email for some insights]. We know (e.g. from the MonitoringWG perspectives report/update to Rettie summary) that number of years of monitoring can be as low as 2 or as many as 20+

*from Anna: withouot access to actual estimates of R and S for different ranges, some potentially interesting other ranges to compare to could be an Alberta range or two, where there are ++years of R and S estimates but really low lambda (E.g. ESAR and WSAR have 27+ yrs consecutive monitoring but declining numbers and high disturbance), some sections of NWT1 that are thought to be self-sustaining and have pretty good data sets (e.g. Dehcho North and South, Hay River Lowlands with 13+yrs of S and R) -could also use other MB ranges as examples for fairly sparse data (though not as bad as ON), and QC or LAB ranges as examples of good data (e.g. how about Charlevoix for a very well monitored but tiny pop, northern QC for really broad-expanse pop, or Labrador pops have long term monitoring data but uncertain status...)

*feedback from MonitoringWG contacts includes details on collar numbers for BC, ON, NL, (MB pending); e.g. in Ontario they aim for 20-30 collars in smaller pops, 35+ in bigger pops; Labrador aims to have between 5-25 collars monitored at any one time, depending on pop size; in BC, real numbers of collared animals have ranged been between 3-59 per pop/year, varying both across-yrs-within-pop and across-pop-within-yr [with COVID yrs by far the lowest].

relevant refs: [@rettie_summary_2017];?

TO DO: Fire scenarios? Add high fire scenario in appendix of one by one sensitivity checks.

TO DO: run low initial population size scenario, put in appendix of one by one sensitivity checks? 
-->
```
## Simulation of local population dynamics and monitoring

To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a method for simulating local population dynamics and monitoring. This method produces example population trajectories for which the true state is known, allowing us to assess the performance of our proposed method for integrating local demographic observations with national disturbance-demographic relationships in a Bayesian population model.

### Population dynamics informed by national disturbance-demographic relationships

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction. White dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}

knitr::include_graphics(here::here("analysis/paper/figs/DemographicRatesDelay.png"))

```

To simulate demographic trajectories for example populations
we used a modified version of Johnson et al's [-@johnson_science_2020]
two-stage demographic model described by Dyson et al.
[-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Maximum
potential recruitment rate is adjusted for sex ratio, misidentification
biases (Supplement B) and (optionally) delayed age at first reproduction
[@eacker_web-based_2019; @decesare_estimating_2012]:
\begin{equation}
  \dot{X}_t=\frac{c\dot{R}_t/2}{1+c\dot{R}_t/2};c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)},
  (\#eq:dotX)
\end{equation}
where $w$ is a multiplier that defines the apparent number of adult
females in the composition survey as a function of the number of
collared animals, $q$ is the ratio of young bulls to adult females, $u$
is the probability of misidentifying young bulls as adult females and
vice versa, and $z$ is the probability of missing a calf.

Realized recruitment rate varies with population density
[@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census is a binomially distributed function
of the number of surviving post-juvenile females and the adjusted
recruitment rate:
\begin{equation}
  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{N_0k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
  (\#eq:dotJ)
\end{equation}

Given the values of parameters $p_0,p_k,a, and b$ used by Johnson et al. (2020) (Table A1), recruitment rate is lowest
$(0.5\dot{X}_t)$ when $\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$
at intermediate population sizes, and declines to $0.6\dot{X}_t$ as the
population reaches carrying capacity of $k=100$ times the initial
population size. The post-juvenile female population in the next year
includes both survivors and new recruits:
\begin{equation}
\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t).
  (\#eq:dotN)
\end{equation}

Interannual variation in survival and recruitment is modelled using
truncated beta distributions [rtrunc function;
@novomestky_package_2016]:
$\dot{R}_t \sim \text{TruncatedBeta}(\bar{R}_t,\nu_R,l_R,h_R); \dot{S}_t \sim \text{TruncatedBeta}(\bar{S}_t,\nu_S,l_S,h_S)$.
Coefficients of variation among years $(\nu_R,\nu_S)$ and
maximum/minimum values $l_R,h_R,l_S,h_S$ for recruitment and survival
are given in Table A1. Expected recruitment ($\bar{R}_t$) and survival
($\bar{S}_t$) vary with disturbance according the beta regression models
estimated by Johnson et al. [-@johnson_science_2020]:

\begin{equation}
\bar{R}_t \sim Beta(\mu^R_t,\phi^R); 
log(\mu^R_t) = \dot{\beta}^R_0+\dot{\beta}^R_a A_t+\dot{\beta}^R_f F_t,
  (\#eq:barR)
\end{equation}

\begin{equation}
\bar{S}_t \sim (46 \times Beta(\mu^S_t,\phi^S)-0.5)/45;
log(\mu^S_t) = \dot{\beta}^S_0+\dot{\beta}^S_a A_t.
  (\#eq:barS)
\end{equation}

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example among populations
(Figure \@ref(fig:exampleTrajectories)).

### Observed survival of collared animals

In our simulated populations, mortality risk does not vary over time or
among individuals, so the observed state (dead or alive) of a collared
animal $i$ in month $m$ and year $t$ depends on the true survival rate
$\dot{S}_t$ as follows:
\begin{equation}
\hat{I}_{i,t,m} \sim \text{Bernoulli}(\dot{S}^{1/12}_t\hat{I}_{i,t,m-1}).
  (\#eq:hatI)
\end{equation}

### Observed recruitment from composition surveys

The number of cows in the composition survey $\hat{W}_t$ is given by the
number of collared cows $\hat{T}_t$ and the apparent number of adult
females per collared female observed in the composition survey $w$:
\begin{equation}
\hat{W}_t = w\hat{T}_t.
  (\#eq:hatW)
\end{equation}
The number of observed calves $\hat{J}_t$
also depends on the unadjusted apparent recruitment for the population
$\dot{R}_t$: 
\begin{equation}
\hat{J}_t \sim \text{Binomial}(\hat{W}_t,\dot{R}_t).
  (\#eq:hatJ)
\end{equation}

## Integration of local demographic data and national disturbance-demographic relationships in a Bayesian population model

Following Eacker et al. [-@eacker_web-based_2019], the observed number
of calves $\hat{J}_t$ is a binomially distributed function of the
observed number of adult female caribou $\hat{W}_t$ and the estimated
recruitment rate $R_t$:
\begin{equation}
\hat{J}_t \sim \text{Binomial}(\hat{W}_t,R_t).
  (\#eq:hatJEst)
\end{equation}

We model recruitment probability as a function of anthropogenic
disturbance and fire with a log link [@johnson_science_2020;
@stewart_climate-informed_2023; @dyson_existing_2022] and Gaussian
distributed random year effect [@eacker_web-based_2019]:
\begin{equation}
\log(R_t)=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t+\epsilon^R_t; \epsilon^R_t \sim \text{Normal}(0,\sigma^2_{R}).
  (\#eq:R)
\end{equation}

Estimated recruitment probability is adjusted for sex ratio,
misidentification biases (Supplement B), and age of first reproduction
[@eacker_web-based_2019; @decesare_estimating_2012] to get expected
recruits per female:
\begin{equation}
X_t=\frac{cR_t/2}{1+cR_t/2}; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
  (\#eq:X)
\end{equation}

The composition survey bias correction term $c$ depends on the apparent
number of adult females per collared animal $w$, the ratio of young
bulls to adult females $q$, the probability of misidentifying young
bulls as adult females and vice versa $u$, and the probability of
missing a calf $z$ (Supplement B). To maintain consistency with Johnson
et al. [-@johnson_science_2020] we also provide an option to only adjust
for sex ratio and misidentification biases $X_t=cR_t/2$.

In the Bayesian model, prior uncertainty about the value of the bias
correction term $c$ is approximated with a Log-normal distribution
(Supplement B). Given the apparent number of adult females per collared
animal $w$ in each monitoring scenario (Table
\@ref(tab:monitoringScenarios)), we calculate the mean and standard
deviation of 10,000 samples of $\log{c}$ assuming the ratio of young
bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult
misidentification probability $u$ varies uniformly between 0 and 0.2,
and the probability of missing a calf $z$ varies uniformly between 0 and
0.2.

Also following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the survival R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a Gaussian
distributed function of the true survival probability:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).
  (\#eq:hatSEst)
\end{equation}

As in the national demographic model, we model survival probability as an adjusted function
of buffered anthropogenic disturbance $A_t$ with a log link
[@johnson_science_2020; @stewart_climate-informed_2023;
@dyson_existing_2022] and Gaussian distributed random year effect
[@eacker_web-based_2019]:
\begin{equation}
S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).
  (\#eq:S)
\end{equation}

To assess sensitivity of results to the survival estimation method and
handle cases in which there is little or no survival data, we also
implemented an alternative parametric exponential survival model,
wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed
[@schaub_integrated_2021]:
\begin{equation}
\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).
  (\#eq:hatIEst)
\end{equation}

All regression coefficients are assumed to be Gaussian distributed, with
priors derived from the expected values and 95% confidence intervals
estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We
calibrated the prior distributions so that the 95% prior prediction
intervals for survival and recruitment from the Bayesian model match the
range between the 2.5% and 97.5% quantiles of 1000 simulated survival
and recruitment trajectories (Supplement C, Figure
\@ref(fig:examplePredictions)).

```{r priors, echo=FALSE,message=FALSE}

source(here::here("analysis/scripts/formatPriors.R"))

knitr::kable(tbl,caption="Prior means and standard errors of Gaussian distributed survival and reproduction model parameters. Standard errors are calculated by dividing Johnson et al's (2020) reported confidence intervals by 3.92. Multipliers on the standard errors (*) are calibrated so that the 95\\% prior prediction intervals for survival and recruitment from the Bayesian model match the range between the 2.5\\% and 97.5\\% quantiles of 1000 simulated survival
and recruitment trajectories (Supplement C).")
#TO DO: indicate which of these are from Johnson, and which are calibrated.
```

```{r examplePredictions, fig.cap="Prior (top row) and posterior (bottom row) predictions from the Bayesian model for an example population trajectory. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 15 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty (bottom row), but when disturbance is very high local data is not required to predict viability (population growth < 1).", cache=TRUE}

knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamples.png"))

```

We assume a simple two-class demographic model, with a census in late
winter of surviving females and new recruits to the post-juvenile (aged
1+) class. The number of post-juvenile females that survive from year
$t$ to the census $W_t$ is binomially distributed with survival
probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The number of
juveniles recruiting to the post-juvenile class at the census is a
Poisson distributed function of the number of surviving post-juvenile
females and the adjusted recruitment rate $X_t$:
$J_{t} \sim \text{Poisson}(X_tW_t)$. The post-juvenile female population
in the next year includes both survivors and new recruits:
$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
stable estimates of population status, we report average population
growth over a period of $y$ years, calculated as
$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
post-juvenile population size $N_0$ is 1000 animals. For reference, all
variables and parameters are defined in Table A1.

\begin{equation}
J_{t} \sim \text{Poisson}(W_tR_t/2)
  (\#eq:J)
\end{equation}

<!--
Annotated equations for poster.
$$\underbrace{N_{t+1}}_{\text{Adult females in year t+1}}=\underbrace{W_t}_{\text{Surviving adults}}+\underbrace{J_t}_{\text{New recruits to adult age class}}$$
$$W_{t} \sim \text{Binomial}(N_t,\underbrace{S_t}_{\text{Adult survival probability}})$$
$$S_t=\underbrace{\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})}_{\text{Adjustment as in in Johnson et al. 2020}}=\beta^S_0+ \underbrace{\beta^S_aA_t}_{\text{Anthropogenic disturbance effect}}+\underbrace{\epsilon^S_t}_{\text{Random effect of year}}.$$
$$J_{t} \sim \text{Poisson}(W_t\underbrace{R_t/2}_{\text{Recruitment rate}})$$
$$\log(R_t)=\beta^R_0+\underbrace{\beta^R_a A_t}_{\text{Anthropogenic disturbance effect}}+ \underbrace{\beta^R_fF_t}_{\text{Fire effect}}+\underbrace{\epsilon^R_t}_{\text{Random effect of year}}.$$
$$\underbrace{\hat{C}_t}_{\text{Observed calves}} \sim \text{Binomial}(\underbrace{\hat{W}_t}_{\text{Observed cows}},R_t)$$
$$\underbrace{\hat{S_t}}_{\text{Observed survival (Kaplan-Meier estimate)}} \sim \text{Normal}(S_t,1/\tau_t).$$ -->

TO DO: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
Ensure notation consistency amongst manuscript, UI, documentaiton and
code where appropriate.

## Probability of correct status assessment

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, it is important to know whether a
population is declining or not. If uncertainty about population status
is low, additional investment in reducing uncertainty about the
population growth rate may not be necessary or warranted. Thus, in this
analysis our outcome of interest is how often the Bayesian posterior
mean correctly predicts the true population status (increasing or
decreasing). To calculate this outcome for each disturbance and
monitoring scenario (Figure \@ref(fig:figDisturbance)) we fit the
Bayesian model to 200 example population trajectories (e.g. Figure
\@ref(fig:exampleTrajectories)). The probability of correct status
assessment for that scenario is the percentage of these 200 examples in
which the Bayesian posterior mean population growth rate and the true
population growth rate are both above or below 1.

## User interface and open, flexible, reuseable implementation

To make this work more user friendly and accessible to decision makers
we built a user interface that allows for the exploration of different
monitoring and anthropogenic disturbance scenarios. We built the user
interface as an R Shiny app [@chang_shiny_2023], building off and expanding an
app published by Eacker et al. [-@eacker_web-based_2019]. The app allows
users to provide existing local data or simulate data based on national
models as shown above, define an anthropogenic disturbance scenario and
make projections. The results shown are similar to Figure
\@ref(fig:examplePredictions) and allow for comparisons to the national
model. Many parameters controlling the Bayesian model, local data
simulation and antropogenic scenario are exposed in the app with
sensible defaults set, such that novice users can get reasonable results
quickly while advanced user can customise the projections to suite their
needs. We have also made the R code needed to run the app and carry out
the above analysis available as an R package, caribouMetrics. Making our
code available as an R package ensures that it is flexible and can be
re-used or expanded upon in future projects or combined with other tools
and packages (eg SpaDES or SyncroSym). The caribouMetrics package is
available on GitHub at <https://github.com/LandSciTech/caribouMetrics>
and a package website with detailed documentation is also available
(<https://landscitech.github.io/caribouMetrics/>).

# Results

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows, starting in 2023), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). Probabilities are calculated for 200 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)). For each example population, the status assessment is correct if the Bayesian posterior mean and the true population growth rate are both less than or greater than one. When disturbance is very high the probability of correct status assessment is high even without local monitoring. Where disturbance is moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

# Discussion

Will need a section on the fundamental assumptions of these
models/aspatial nature of them, e.g. that collared cows and cows
observed in surveys are a random representative sample of the larger
population, and that there is full mixing in the population so that
estimated survival/recruitment rates apply equally to all animals in the
population. We can discuss this with some contrasting examples, e.g.
collaring 30 animals from a very small/restricted population (like
Little Smoky, 308,606ha, pop size estimate 78 as of COSEWIC 2014) would
likely be much more representative of the population vs. collaring 30
animals from a relatively large/expansive population (northern Quebec
range Q6, 62,156,186 ha, pop size 9000 as of COSEWIC 2014). But because
of the way this model is built, we assume that the sample is equally
representative in either of those cases, so changing the initial
starting N won't affect model performance/predictions much here...

On a related topic to population size; Note that the stuff we got from
ArtR regarding simulations for sample size does acknowledge that pops of
different sizes have different monitoring requirements, and Sara's
Labrador numbers also show targets relative to pop size... Since this
kind of aspatial model will mostly be using small population sizes in
order to see any effects, we'll jsut have to acknowledge that a caveat
of this is that we're not fully able to explore all the population size
effects with this model, but generally Josie has found little impact of
initial population size to the results.

Include some commentary on addressing limitations of aspatial
demographic models with spatial models. Not a panacea. It is possible
(cite examples) to develop spatially explicit models of movement and
adult mortality risk - though sample size is a challenge. Recruitment,
which is more variable among ranges and with anthropogenic disturbance,
is difficult to characterize at finer spatial scales because it depends
on the conditions experienced by mothers over years as they move across
the range, and on spatial variation in calf mortality risk that is not
measured directly.

We should likely also have a pgph about costs of monitoring - it'll be
pretty short, cause basically we have no idea what it costs since nobody
will tell us :) Through the NBCKC Monitoring Working Group work, we
tried over and over to get some basic cost estimates from P/T reps and
practitioners, and hit a wall - nobody wanted to tell us a thing. Don't
know if we'd now find anyone is more open to sharing, but don't think we
should get our hopes up. But at the very least we'll want to acknowledge
that we haven't accounted for relative costs of monitoring... [e.g.
costs outlined in the Telemetry guidance report from the MonitoringWG:
"Expenses to consider in radio-collared monitoring will vary, based on:
helicopter (or other aircraft) use, purchase of tracking collars, data
fees associated with the collars (the cost of which varies by operating
system, recording frequency and data transmission frequency), and
live-capture equipment (3 or more sets of capture nets and associated
buckets, a netting gun, and ankle hobbles)"] --\>perhaps the keen
MonitoringWG contacts would be willing to provide even really rough cost
estimates, and/or even jsut a relative sense of how different it would
be to e.g. monitor a few pops for many yrs vs lots of pops for a short
time...?

Re paragraph on costs of monitoring. The reality is that costs vary with
remoteness and the difficulty of finding animals to collar (which also
depends on how many collars are deployed). Since costs are so
context-specific, there isn't a good way to incorporate them into this
general analysis. It would be interesting to work through a case study
(i.e. Newfoundland) informed by context-specific costs, and consider how
incorporation of those costs may change the relative value of
alternative monitoring scenarios.

Biases in composition survey estimates also vary with vegetation cover
(i.e. the difficulty of seeing animals) and the care taken to correctly
identify the sex of unknown animals. Misidentification rates could be
estimated from discrepancies among results from multiple independent
observers analyzing photos of animal groups.

The tool we have developed is a proof of concept, designed to
demonstrate possibilities and help users develop a general understanding
of the relative value of alternative monitoring strategies for boreal
caribou. In current form the tool will have limited value for managers
who have some local data and wish to know the value of collecting more,
because it is designed to either integrate real local data or simulated
local data, but not both. Development of a tool designed to help
decision makers assess the relative value of strategies for additional
monitoring given their existing local data and national-demographic
disturbance relationships would be a useful followup to this work.

A tool that integrates real local data with national demographic
disturbance relationships can also be used to provide consistent and
transparent assessments of the strength of evidence for local deviations
from the expected mean of national demographic-disturbance
relationships. Help distinguish real variation among populations from
apparent variation that can be attributed to lack of local data.

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

### Colophon

This report was generated on `r Sys.time()` using the following
computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
#if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
#if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2023-10-24 analysis/paper/supplement.Rmd
chipping away at paper

---
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      number_sections: yes
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style

editor_options: 
  chunk_output_type: inline
---

# (APPENDIX) Supplementary Material for National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions {-}

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figs/",
  dpi = 300
)
```

```{r setup2}
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(ggpubr)
library(MetBrewer)
library(here)
library(tmap)
library(sf)
library(caribouMetrics)

devtools::load_all(here::here())

#set paths
data_path_raw <- "analysis/data/raw_data/"
data_path_drvd <- "analysis/data/derived_data"

pal_nm <- "Redon"

fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)

runScenarios = F 
doPriorCalibration=F #Note the prior calibration is time consuming.

resultScenario = "S6"
```

# Definitions of variables and parameters in the Bayesian population model

Table A1. Variables and parameters in the Bayesian population model and the national demographic model.

|**Symbol**|**Units**|**Description**|
|:-----|-----|:-----------|
|**Indices & constants**|
|$t$|year|Year.|
|$m$|month|Month.|
|$i$|ID|Individual animal identifier.|
|**Observed**|
|$A_t$|%|Buffered anthropogenic disturbance.|
|$F_t$|%|Fire disturbance.| 
|$\hat{J}_t$|animals|Calves observed in calf:cow survey.|
|$\hat{W}_t$|animals|Cows observed in calf:cow survey.|
|$\hat{T}_t$|animals|Number of collared cows.|
|$\hat{S}_t$|probability|Kaplan-Meier estimate of observed adult female survival.|
|$\tau_t$||Precision of Kaplan-Meier survival estimate.|
|$\hat{I}_{i,t,m}$|0 or 1|Indicator of whether animal $i$ is dead or alive.|
|**Estimated state**|
|$R_t$|rate|Recruitment (calves per cow).|
|$S_t$|probability|Adult female survival.|
|$N_t$|animals|Adult female population size.|
|$\bar{\lambda}_{y,t}$|rate|Average population growth over $y$ years.|
|$X_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|
|$W_t$|animals|Surviving adult females before recruitment.|
|$J_t$|animals|New recruits to adult age class.|
|$\lambda_t$|rate|Annual population growth.|
|**Estimated Parameters**|
|$\beta^R_0$||Recruitment intercept.|
|$\beta^R_a$||Anthropogenic disturbance slope for recruitment.|
|$\beta^R_f$||Fire slope for recruitment.|
|$\sigma^2_{R}$||Variance in random effect of year on recruitment.|
|$\beta^S_0$||Survival intercept.|
|$\beta^S_a$||Anthropogenic disturbance slope for survival.|
|$\sigma^2_{S}$||Variance in random effect of year on survival.|
|**True state**|
|$\dot{R}_t$|rate|Recruitment (calves per cow).|
|$\dot{S}_t$|probability|Adult female survival.|
|$\dot{N}_t$|animals|Adult female population size.|
|$\dot{X}_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|
|$\dot{W}_t$|animals|Surviving adult females before recruitment.|
|$\dot{J}_t$|animals|New recruits to adult age class.|
|**True demographic parameters**|
|$N_0$|animals|Initial adult population size.|
|$\bar{R}_t$|rate|Expected recruitment (calves per cow).|
|$\bar{S}_t$|probability|Expected adult female survival.|
|$\dot{\beta}^R_0$||Recruitment intercept.|
|$\dot{\beta}^R_a$||Anthropogenic disturbance slope for recruitment.|
|$\dot{\beta}^R_f$||Fire slope for recruitment.|
|$\dot{\beta}^S_0$||Survival intercept.|
|$\dot{\beta}^S_a$||Anthropogenic disturbance slope for survival.|
|$p_0=1$||Maximum recruitment multiplier.|
|$p_k=0.6$||Recruitment multiplier at carrying capacity.|
|$k=100$||Carrying capacity multiplier.|
|$b=4$||Density dependence shape parameter.|
|$a=1$||Allee effect parameter.|
|$r_{max}=1.3$||Maximum population growth rate.|
|$\nu_R=0.46$||Coefficient of variation in $\dot{R}_t$ among years.|
|$\nu_S=0.08696$||Coefficient of variation in $\dot{S}_t$ among years.|
|$l_R=0$||Minimum value for $\dot{R}_t$.|
|$l_S=0$||Minimum value for $\dot{S}_t$.|
|$h_R=0$||Maximum value for $\dot{R}_t$.|
|$h_S=0$||Maximum value for $\dot{S}_t$.|
|**Observation model parameters**|
|$d=1,2,4,8,16, \text{ or } 24$|years|Monitoring duration.|
|$n=15,30, \text{ or } 60$|animals|Target number of collared females.|
|$o=1 \text{ or } 4$|years|Years between collar deployments.|
|$y=3$|years|Assessment period for population growth rate.|
|$w=3,6,\text{ or } 9$|ratio|The apparent number of adult females per collared animal in composition survey.|
|$q=\text{uniform}(0,0.6)$|ratio|True ratio of young bulls to adult females in composition survey.|
|$u=\text{uniform}(0,0.2)$|probability|Probability of misidentifying young bulls as adult females and vice versa in composition survey.|
|$z=\text{uniform}(0,0.2)$|probability|Probability of missing calves in composition survey.|
|$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}$||Bias correction term (Appendix X).|

# Model of bias in recruitment estimates from calf:cow surveys

TO DO: get photo of calf:cow survey group. Label classes of animals to illustrate meaning of T, U, V, W, O, P, J, K.

We assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$). The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is a multiplier that defines the apparent number of adult females as a function of the number of collared animals. The ratio of young bulls to uncollared adult females in the group is:
$$q = \frac{P+O}{U+V}.$$
Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$. Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$.

Our objective is to model the sex and bias-corrected recruitment rate $X=\frac{J+K}{2(T+U+V)}$ as a function of the observed calf:cow ratio $R=J/(T+V+P)$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$. We start by solving for $T+U+V$ as a function of $q,w,u$ and $T$. Recognize that $P=Tw-T-V$, $U+V=V/(1-u)$, and $P+O=P/u$ to write $q$ as 
$$q=\frac{Tw-T-V}{uV/(1-u)}.$$
Rearrange to get 
$$V=\frac{T(w-1)(1-u)}{qu+1-u}.$$
Recognize that $U=Vu/(1-u)$ to write $T+U+V$ as a function of $q,w,u$ and $T$: 
$$T+U+V=T\frac{qu+w-u}{qu+1-u}.$$
Recognize that the number of observed calves $J$ is the product of the apparent recruitment rate and the apparent number of adult females $J=RTw$, and that therefore $J+K=RTw/(1-z)$ to rewrite the bias corrected recruitment rate $X=\frac{J+K}{2(T+U+V)}$ as a function of $w,u,z$ and $R$:
$$X=R\frac{w(1+qu-u)}{2(w+qu-u)(1-z)}.$$
For simplicity, we write $X$ as a function of a bias correction term $c$: 
$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$

If we also adjust for delayed age at first reproduction (DeCesare et al. 2012; Eacker et al. 2019), the adjusted recruitment rate becomes $$X=\frac{cR/2}{1+cR/2}.$$

```{r biasPlot, fig.cap="Variation in the sex and bias corrected recruitment rate $X$ with the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the adult misidentification probability $u$, and the probability of missing a calf $z$. If only collared females are counted ($w=1$) and there is a risk of missing calves $z>0$ then the apparent recruitment rate $R$ will be lower than $X$ (first column). When uncollared animals are counted ($w>1$), the risk of misidentifying adults $u$ inflates the apparent recruitment rate $R$ as on average more cows are misidentified as bulls than vice versa, but this effect is smaller than the effect of missing calves $z$. Increasing the proportion of bulls in the group $q$ reduces the effect of adult misidentification $u$.", fig.height=6,cache=TRUE}
library(ggplot2)
#Notes
#q: assume more females than males in calf:cow group, so q<1

p=expand.grid(delay=c(T,F),q=c(0,0.3,0.6),w=c(1,3,6),
              u=c(0,0.1,0.2),z=seq(0,0.2,length.out=3),R=seq(0.15,0.6,length.out=10))

getX<-function(R,delay,q,w,u,z){
  c = compositionBiasCorrection(w,q,u,z)
  X = 0.5*R*c
  X=X*(1-delay)+delay*X/(1+X)
  return(X)
}

p$X = getX(p$R,p$delay,p$q,p$w,p$u,p$z)

p$grp = paste(p$u,p$z)
p$u=as.factor(p$u)
p$z=as.factor(p$z)

base = ggplot(subset(p,!delay),aes(x=R,y=2*X/R,col=z,linetype=u,group=grp))+
  geom_hline(yintercept=1)+geom_line()+facet_grid(q~w,labeller=label_both)+theme_bw()+ylab("2X/R=c, without delayed reproduction")

base2 = ggplot(subset(p,delay),aes(x=R,y=2*X/R,col=z,linetype=u,group=grp))+
  geom_hline(yintercept=1)+geom_line()+facet_grid(q~w,labeller=label_both)+theme_bw()+ylab("2X/R, with delayed reproduction")

ggpubr::ggarrange(base, base2, labels = "",
                  ncol = 1, vjust = 1)

```

```{r biasDistribution,fig.cap="In the Bayesian model, uncertainty about the value of the bias correction term $c$ (grey bars) is approximated with a Log-normal distribution (red). In these examples, the ratio of young bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult misidentification probability $u$ varies uniformly between 0 and 0.2, and the probability of missing a calf $z$ varies uniformly between 0 and 0.2. We assume that the apparent number of adult females per collared animal $w$ is known.", cache=TRUE}

cr=expand.grid(w = c(1,2,3,4,5,6,7,8,9),rep=seq(1,10000))
nr=nrow(cr)
cr$c = compositionBiasCorrection(q=runif(nr,0,0.6),w=cr$w,u=runif(nr,0,0.2),z=runif(nr,0,0.2))
cs = compositionBiasCorrection(q=runif(nr,0,0.6),w=cr$w,u=runif(nr,0,0.2),z=runif(nr,0,0.2),approx=T)

cr = merge(cr,cs)
cr$ca= exp(rnorm(nrow(cr),mean=cr$mu,sd=cr$sig2^0.5))

base = ggplot(cr,aes(x=c))+geom_histogram(aes(y=..density..))+
  facet_wrap(~w,nrow=3,labeller=label_both)+theme_bw()+
  geom_density(aes(x=ca),alpha=.2, fill="#FF6666",col = "#FF6666")
plot(base)

```

# Calibration of Priors

```{r priorCalibration, cache=TRUE,echo=FALSE, message=FALSE,results='hide'}
if(doPriorCalibration){

  source(here::here("analysis/scripts/calibratePriors.R"))
}else{
  KS<-read.csv(here::here("tabs/SurvKSAll.csv"))
  KSRec<-read.csv(here::here("tabs/RecKSAll.csv"))

}
  #library(ggplot2);fig_widths=data.frame(two=7)
  KS=subset(KS,scn=="All")
  png(here::here("figs/PriorKSDists.png"),
    height = 3, width = fig_widths["two"], units = "in",res=600)
    base = ggplot(KS,aes(x=sIntSEMod,y=KSDistance,color=sSigmaMean))+
    geom_point()+facet_wrap(~sSigmaSD,labeller="label_both")+theme_bw()+ylab("Kolmogorov-Smirnov Distance")+xlab("sIntSEMod: multiplier on SD of survival intercept")
    print(base)
  dev.off()

  KSRec=subset(KSRec,scn=="All")
  png(here::here("figs/PriorKSDistsRec.png"),
    height = 3, width = fig_widths["two"], units = "in",res=600)
    base = ggplot(KSRec,aes(x=rIntSEMod,y=KSDistance,color=rSigmaMean))+
    geom_point()+facet_wrap(~rSigmaSD,labeller="label_both")+theme_bw()+ylab("Kolmogorov-Smirnov Distance")+xlab("rIntSEMod: multiplier on SD of recruitment intercept")
    print(base)
  dev.off()
```

```{r PriorKSDists, fig.cap="Similarity of the prior survival distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022]. The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\\sigma^2_S$ in Table 2). Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
knitr::include_graphics(here::here("figs/PriorKSDists.png"))

```

```{r PriorAllFits, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, the standard deviation of the random effect prior (SigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}

knitr::include_graphics(here::here("figs/SurvInterceptAllsSigmaSD0.03.png"))
```

```{r PriorLowFits, fig.cap="Posterior predictive survival distributions from the Bayesian model ('local') assuming intensive monitoring (15 years of 60 collars per year) of populations from the 2.5 \\% quantile of the national model. In these examples, the standard deviation of the random effect prior (SigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and  anthropogenic disturbance is 0. Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates. When the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is too low (<5) the insufficiently vague priors substantially influence the posterior distributions, yielding poor local parameter estimates.", cache=TRUE}

knitr::include_graphics(here::here("figs/SurvsurvInterceptLowsSigmaSD0.03.png"))
```

```{r PriorAllFitsAnthro90, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, anthropogenic disturbance is 90\\%. When the multiplier on the prior standard deviation of the survival slope (sSlopeSEMod: $\\beta^S_a$ in Table 2) is 3, the posterior survival distribution is similar to the distribution of simulated outcomes from the national model. Bands show 95\\% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.", cache=TRUE}
knitr::include_graphics(here::here("figs/SurvSlopeAll.png"))
```

```{r PriorHighFitsAnthro90, fig.cap="Posterior predictive survival distributions from the Bayesian model, assuming intensive monitoring (15 years of 60 collars per year) of populations from the 97.5 \\% quantile of the national model. In these examples, anthropogenic disturbance is 90\\%, and the effect of anthropogenic disturbance varies from none (sSlopeMod=0) through expected (sSlopeMod=1) to double the expected slope (sSlopeMod=2). All considered multipliers on the prior standard deviation of the survival slope (sSlopeSEMod: $\\beta^S_a$ in Table 2) are sufficiently vague to reflect local variation among populations. Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates.", cache=TRUE}
knitr::include_graphics(here::here("figs/SurvSlopeHigh.png"))
```

Prior parameter values should be set so that, given no monitoring data, the prior survival distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance. This condition is met when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\sigma^2_S$ in Table 2) (Figs \@ref(fig:PriorKSDists) and \@ref(fig:PriorAllFits)). Priors should also be vague enough so that, given intensive monitoring of populations where survival is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 60 collars per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local survival, rather than priors from the national model. Setting the mean and standard deviations of the prior distribution of the random effect of year on survival to moderately low values (SigmaMean=0.034784, SigmaSD=0.03, $\sigma^2_S$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDists), \@ref(fig:PriorAllFits) and \@ref(fig:PriorLowFits)).

```{r PriorKSDistsRec, fig.cap="Similarity of the prior recruitment distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022]. The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (rIntSEMod: $\\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\\sigma^2_R$ in Table 2). Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
knitr::include_graphics(here::here("figs/PriorKSDistsRec.png"))
```

```{r PriorAllFitsRec, fig.cap="Differences between prior predictive recruitment distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, the standard deviation of the random effect prior (SigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}
knitr::include_graphics(here::here("figs/RecInterceptAllrSigmaSD0.22.png"))
```

```{r PriorHighFitsRec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model ('local') assuming intensive monitoring (9*60 cows per year for 15 years) of populations from the 97.5% quantile of the national model.  In these examples, the standard deviation the random effect prior (SigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and  anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals, black dots show true recruitment rates from the simulated population, and open triangles show observed recruitment rates.", cache=TRUE}
knitr::include_graphics(here::here("figs/RecInterceptHighrSigmaSD0.22.png"))
```

```{r PriorAllFitsAnthro90Rec, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, anthropogenic disturbance is 90\\%. When the multiplier on the prior standard deviation of the recruitment slope (rSlopeSEMod: $\\beta^R_a$ in Table 2) is 4, the posterior survival distribution is similar to the distribution of simulated outcomes from the national model. Bands show 95\\% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.", cache=TRUE}
knitr::include_graphics(here::here("figs/RecSlopeAll.png"))
```

```{r PriorHighFitsAnthro90Rec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model, assuming intensive monitoring (15 years of 60 collars per year) of populations from the 97.5 \\% quantile of the national model. In these examples, anthropogenic disturbance is 90\\%, and the effect of anthropogenic disturbance varies from none (rSlopeMod=0) through expected (rSlopeMod=1) to double the expected slope (rSlopeMod=2). All considered multipliers on the prior standard deviation of the recruitment slope >1 (rSlopeSEMod: $\\beta^R_a$ in Table 2) are sufficiently vague to reflect local variation among populations. Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates.", cache=TRUE}
knitr::include_graphics(here::here("figs/RecSlopeHigh.png"))
```

Similarly, prior parameter values should be set so that, given no monitoring data, the prior recruitment distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance. This condition is met when the multiplier on the prior standard deviation of the recruitment intercept (rIntSEMod: $\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\sigma^2_R$ in Table 2) (Figs \@ref(fig:PriorKSDistsRec) and \@ref(fig:PriorAllFitsRec)). Priors should also be vague enough so that, given intensive monitoring of populations where recruitment is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 9*60 cows per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local recruitment, rather than priors from the national model. Setting the mean and standard deviations of the prior distribution of the random effect of year on recruitment to moderately low values (SigmaMean=0.23, SigmaSD=0.22, $\sigma^2_R$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDistsRec), \@ref(fig:PriorAllFitsRec) and \@ref(fig:PriorHighFitsRec)).

\newpage

# Alternative disturbance and monitoring scenarios

```{r figDisturbanceAnthroStatic, fig.cap="Anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult6ay3aSf0.png")))

```

```{r powercmult3ay3aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for for static anthropogenic disturbance scenarios). In these examples there are 3 cows per collared cow in recruitment surveys. See Figure 4 for additional details. ", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult3ay3aSf0.png")))
```

```{r powercmult6ay3aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios). In these examples there are 6 cows per collared cow in recruitment surveys. See Figure 4 for additional details. ", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult6ay3aSf0.png")))
```

```{r powercmult9ay3aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios). In these examples there are 9 cows per collared cow in recruitment surveys. See Figure 4 for additional details. ", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult9ay3aSf0.png")))
```

```{r powercmult3ay3aSf1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure 1 for scenarios in which anthropogenic disturbance increases over time). In these examples there are 3 cows per collared cow in recruitment surveys. See Figure 4 for additional details. ", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult3ay3aSf1.png")))
```

```{r powercmult9ay3aSf1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure 1 for scenarios in which anthropogenic disturbance increases over time). In these examples there are 9 cows per collared cow in recruitment surveys. See Figure 4 for additional details. ", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult9ay3aSf1.png")))
```

# Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
#if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
#if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2023-10-24 analysis/scripts/calibratePriors.R

@@ -18,20 +18,20 @@ scResults = runScenario(scns,survAnalysisMethod = "Exponential")
 KSAll = subset(scResults$ksDists,(Year==2023)&(Parameter=="Adult female survival"))
 addBit = paste0("InterceptAll")
 scResults$obs.all = NULL
-makeInterceptPlots(scResults,addBit=addBit,facetVars=c("SigmaMean","sIntSEMod"),loopVars="sSigmaSD",whichPlots=c("Adult female survival"))
+makeInterceptPlots(scResults,addBit=addBit,facetVars=c("sSigmaMean","sIntSEMod"),loopVars="sSigmaSD",whichPlots=c("Adult female survival"))
 
 scns=merge(scnsSInt,scnsApplyAll);scns=merge(scns,scnsLow)
 scResultsLow = runScenario(scns,quants=c(0.025,0.025),Anthro=0,survAnalysisMethod = "KaplanMeier")
 addBit = paste0("InterceptLow")
 KSLow = subset(scResultsLow$ksDists,(Year==2023)&(Parameter=="Adult female survival"))
-makeInterceptPlots(scResultsLow,addBit=addBit,facetVars=c("SigmaMean","sIntSEMod"),loopVars="sSigmaSD",
+makeInterceptPlots(scResultsLow,addBit=addBit,facetVars=c("sSigmaMean","sIntSEMod"),loopVars="sSigmaSD",
                    whichPlots=c("Adult female survival"),useNational=F)
 
 scns=merge(scnsSInt,scnsApplyAll);scns=merge(scns,scnsHigh)
 scResultsHigh = runScenario(scns,quants=c(0.975,0.975),Anthro=0,survAnalysisMethod = "KaplanMeier")
 addBit = paste0("InterceptHigh")
 KSHigh = subset(scResultsHigh$ksDists,(Year==2023)&(Parameter=="Adult female survival"))
-makeInterceptPlots(scResultsHigh,addBit=addBit,facetVars=c("SigmaMean","sIntSEMod"),loopVars="sSigmaSD",
+makeInterceptPlots(scResultsHigh,addBit=addBit,facetVars=c("sSigmaMean","sIntSEMod"),loopVars="sSigmaSD",
                    whichPlots=c("Adult female survival"),survLow=0.8,useNational=F)
 
 KSAll$scn = "All"
[…]

### 2023-10-24 analysis/scripts/formatPriors.R

@@ -12,7 +12,7 @@ nameSet <- gsub(".Prior1","",names(rawPriors)[grepl("Prior1",names(rawPriors))],
 
 start=T
 for(n in nameSet){
-  crow = data.frame(name=n,Mean=rawPriors[[paste0(n,".Prior1")]],SD=rawPriors[[paste0(n,".Prior2")]])
+  crow = data.frame(name=n,Mean=rawPriors[[paste0(n,".Prior1")]],SE=rawPriors[[paste0(n,".Prior2")]])
 
   if(start){
     tbl = crow

### 2023-10-24 analysis/scripts/formatScenarios.R

@@ -1,21 +1,21 @@
 
 library(caribouMetrics)
 
-rawScns<-read.csv(here::here("tabs/s5.csv"))
+rawScns<-read.csv(here::here("tabs/s6.csv"))
 nrow(rawScns)
 rawScns = subset(rawScns,collarCount!=1)
 
 monitoringPars = list(obsYears=list(symbol= "d", description = "Monitoring duration.",unit="years"),
                       collarCount=list(symbol="n",description="Target # of collars.", unit="cows"),
                       collarInterval = list(symbol = "o",description="Years between collar deployments.", unit="years"),
-                      assessmentYrs = list(symbol="y",description="Assessment period for population growth rate.",unit="years"),
                       cowMult=list(symbol="w",description="Cows per collared animal in composition survey.",unit = "ratio")
                       )
 
+#assessmentYrs = list(symbol="y",description="Assessment period for population growth rate.",unit="years"),
 start=T
 for(par in names(monitoringPars)){
   #par="st"
-  crow = data.frame(name=paste(monitoringPars[[par]]$symbol,par,sep=", "),description=monitoringPars[[par]]$description,
+  crow = data.frame(name=paste("*",monitoringPars[[par]]$symbol,"*, ",par,sep=""),description=monitoringPars[[par]]$description,
                     units=monitoringPars[[par]]$unit,
                     values = paste(unique(rawScns[,par]),collapse=","))
 

### 2023-10-24 analysis/scripts/sensitivityMinimalTest.R

@@ -43,6 +43,7 @@ scns$uMin = 0; scns$uMax = 0
 scns$qMin = 0; scns$qMax = 0
 message("batch ", cpageId, " started")
 
+str(scns)
 
 if(n_reps=="all"){
   scResults = caribouMetrics:::runScnSet(scns,eParsIn,simBig,getKSDists=F,printProgress=F)

### 2023-10-25 analysis/paper/paper.Rmd
chipping away at manuscript

@@
 
-Demographic data for caribou populations, including abundance, vital rates and population growth, are most often obtained through the tracking of caribou cows marked with telemetry collars [e.g.
+Demographic data for caribou populations, including abundance, vital rates and population growth, are often obtained through the tracking of caribou cows marked with telemetry collars [e.g.
 @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@johnson_science_2020].
 Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with collars carrying  Very High Frequency (VHF) and/or Global Positioning Satellite (GPS) transmitters allowing remote detection and data retrieval [see @nbckc_nbckc_2021].
-Because these collars send out a mortality signal when no longer active after a predetermined period of time, adult survival can be estimated by tracking the fate of marked individuals.
-See @kaplan_nonparametric_1958; @pollock_survival_1989 for original approach and @mcloughlin_declines_2003; @hervieux_widespread_2013; @eacker_web-based_2019 for examples of application to caribou.
+Because these collars send out a mortality signal when no longer active after a predetermined period of time, adult survival can be estimated by tracking the fate of marked individuals (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou).
 Male caribou are not frequently collared.
@@
 Because this estimate is based on more than just the collared sample of cows, the sample size for recruitment estimates is typically larger than for survival (see multiplier parameter $w$, cowMult below).
-Note that occasionally young bulls can be misidentified as cows (parameter $u$ below; refs?
-), and there is also a possibility that some calves may be missed during aerial composition surveys (parameter $z$ below).
-We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (parameter $c$; see Supplement B and below for details).
+Note that occasionally young bulls can be misidentified as cows (parameters $\dot{u}$ and $u$ below; refs?
+), and there is also a possibility that some calves may be missed during aerial composition surveys (parameters $\dot{z}$ and $z$ below).
+We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details).
 Telemetry and aerial survey data from these methods can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for monitoring population change of many caribou herds [e.g.
@@
 
-To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a method for simulating local population dynamics and monitoring.
-This method produces example population trajectories for which the true state is known, allowing us to assess the performance of our proposed method for integrating local demographic observations with national disturbance-demographic relationships in a Bayesian population model.
+To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics.
+We refer to this as the "national demographic model" because it is informed by national demographic-disturbance relationships.
+We use simulated trajectories to assess the performance of our proposed method for integrating local demographic observations with national disturbance-demographic relationships in a Bayesian population model.
+For reference, all variables and parameters are defined in supplemental Table A1.
 
@@
 \begin{equation}
-  \dot{X}_t=\frac{c\dot{R}_t/2}{1+c\dot{R}_t/2};c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)},
+  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
   (\#eq:dotX)
@@
 females in the composition survey as a function of the number of
-collared animals, $q$ is the ratio of young bulls to adult females, $u$
+collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
 is the probability of misidentifying young bulls as adult females and
-vice versa, and $z$ is the probability of missing a calf.
+vice versa, and $\dot{z}$ is the probability of missing a calf.
+We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6.
 
@@
 
-Given the values of parameters $p_0,p_k,a, and b$ used by Johnson et al. (2020) (Table A1), recruitment rate is lowest
+Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et al. (2020) (Table A1), recruitment rate is lowest
 $(0.5\dot{X}_t)$ when $\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$
@@
 
-Following Eacker et al. [-@eacker_web-based_2019], the observed number
-of calves $\hat{J}_t$ is a binomially distributed function of the
-observed number of adult female caribou $\hat{W}_t$ and the estimated
-recruitment rate $R_t$:
+Simulated local demographic and covariate data are provided as inputs to the Bayesian population model, including the observed state (dead or alive) of each collared animal in each month ($\hat{I}_{i,t,m}$), the observed number of adult females and calves in the recruitment surveys ($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per collared female in the composition survey $w$, and the measures of anthropogenic disturbance and fire ($A_t, F_t$).
+Following Eacker et al.
+[-@eacker_web-based_2019], recruitment rate $R_t$ is estimated from the observed number of calves $\hat{J}_t$ and adult female caribou $\hat{W}_t$ assuming a binomial distribution:
 \begin{equation}
@@
 
-We model recruitment probability as a function of anthropogenic
+Where appropriate, the Bayesian population model and national demographic model share the same structure, and we have selected notation to highlight the similarities.
+The difference between equations \@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to simulate observations.
+In Bayesian model observations are provided, and used to estimate the unobserved recruitment rate $R_t$.
+
+In both models recruitment probability is a function of anthropogenic
 disturbance and fire with a log link [@johnson_science_2020;
-@stewart_climate-informed_2023; @dyson_existing_2022] and Gaussian
+@stewart_climate-informed_2023; @dyson_existing_2022]. In the Bayesian model we add a Gaussian
 distributed random year effect [@eacker_web-based_2019]:
@@
 
-Estimated recruitment probability is adjusted for sex ratio,
-misidentification biases (Supplement B), and age of first reproduction
-[@eacker_web-based_2019; @decesare_estimating_2012] to get expected
-recruits per female:
+In the simulation model (eq \@ref(eq:barR)), regression coefficient values ($\dot{\beta}^R_0,\dot{\beta}^R_a,\dot{\beta}^R_f$) are selected for each example population and used to calculate recruitment rate.
+In the Bayesian model (eq \@ref(eq:R)), unobserved regression coefficients ($\beta^R_0,\beta^R_a,\beta^R_f$) are estimated from local recruitment survey and disturbance observations.
+
+To account for bias and additional uncertainty caused by unknown misidentification biases (Supplement B) and delayed age of age of first reproduction
+[@eacker_web-based_2019; @decesare_estimating_2012] the estimated recruitment probability is adjusted to get expected recruits per female:
 \begin{equation}
@@
 
-The composition survey bias correction term $c$ depends on the apparent
-number of adult females per collared animal $w$, the ratio of young
-bulls to adult females $q$, the probability of misidentifying young
-bulls as adult females and vice versa $u$, and the probability of
-missing a calf $z$ (Supplement B). To maintain consistency with Johnson
-et al. [-@johnson_science_2020] we also provide an option to only adjust
-for sex ratio and misidentification biases $X_t=cR_t/2$.
-
-In the Bayesian model, prior uncertainty about the value of the bias
+The composition survey bias correction term $c$ depends on the known apparent
+number of adult females per collared animal $w$, the unknown ratio of young
+bulls to adult females $q$, the unknown probability of misidentifying young
+bulls as adult females and vice versa $u$, and the unknown probability of
+missing a calf $z$ (Supplement B). Prior uncertainty about the value of thebias
 correction term $c$ is approximated with a Log-normal distribution
@@
 and the probability of missing a calf $z$ varies uniformly between 0 and
-0.2.
+0.2. To maintain consistency with Johnson
+et al. [-@johnson_science_2020] we also provide an option to only adjust
+for sex ratio and misidentification biases $X_t=cR_t/2$.
 
-Also following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
+Following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
 estimates of observed adult female survival probability $\hat{S}_t$ and
@@
 
-As in the national demographic model, we model survival probability as an adjusted function
-of buffered anthropogenic disturbance $A_t$ with a log link
-[@johnson_science_2020; @stewart_climate-informed_2023;
-@dyson_existing_2022] and Gaussian distributed random year effect
-[@eacker_web-based_2019]:
+As in the simulations (eq \@ref(eq:barS)), we model survival probability as an adjusted function of buffered anthropogenic disturbance $A_t$ with a log link
+[@johnson_science_2020; @stewart_climate-informed_2023; @dyson_existing_2022]. We add a Gaussian distributed random year effect [@eacker_web-based_2019]:
 \begin{equation}
@@
 
-We assume a simple two-class demographic model, with a census in late
+As in the simulations, we assume a simple two-class demographic model, with a census in late
 winter of surviving females and new recruits to the post-juvenile (aged
@@
 stable estimates of population status, we report average population
-growth over a period of $y$ years, calculated as
+growth over a period of $y=3$ years, calculated as
 $\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
 post-juvenile population size $N_0$ is 1000 animals. For reference, all
-variables and parameters are defined in Table A1.
+variables and parameters are defined in supplemental Table A1.
 
-\begin{equation}
-J_{t} \sim \text{Poisson}(W_tR_t/2)
-  (\#eq:J)
-\end{equation}
-
-<!--
-Annotated equations for poster.
+<!-- Annotated equations for poster.
 $$\underbrace{N_{t+1}}_{\text{Adult females in year t+1}}=\underbrace{W_t}_{\text{Surviving adults}}+\underbrace{J_t}_{\text{New recruits to adult age class}}$$
@@
 
-TO DO: \* note distribution of random effect parameters are truncated to
+<!--
+Gritty details for the record: \* note distribution of random effect parameters are truncated to
 ensure R and S remain within reasonable bounds (not guaranteed given log
@@
 example in all particulars, change some aspects, or do comparisons? \*
-Ensure notation consistency amongst manuscript, UI, documentaiton and
-code where appropriate.
+-->
 
@@
 relevant to decision making [@gregory_structured_2012]. For boreal
-caribou and other endangered species, it is important to know whether a
+caribou and other endangered species, a relevant outcome is whether a
 population is declining or not. If uncertainty about population status
@@
 
-To make this work more user friendly and accessible to decision makers
-we built a user interface that allows for the exploration of different
-monitoring and anthropogenic disturbance scenarios. We built the user
-interface as an R Shiny app [@chang_shiny_2023], building off and expanding an
-app published by Eacker et al. [-@eacker_web-based_2019]. The app allows
+To make this work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios.
+Where appropropriate, we borrowed and built on code from an app developed by Eacker et al.
+[-@eacker_web-based_2019].
+Our app allows
 users to provide existing local data or simulate data based on national
@@
 \@ref(fig:examplePredictions) and allow for comparisons to the national
-model. Many parameters controlling the Bayesian model, local data
-simulation and antropogenic scenario are exposed in the app with
-sensible defaults set, such that novice users can get reasonable results
-quickly while advanced user can customise the projections to suite their
-needs. We have also made the R code needed to run the app and carry out
-the above analysis available as an R package, caribouMetrics. Making our
-code available as an R package ensures that it is flexible and can be
-re-used or expanded upon in future projects or combined with other tools
-and packages (eg SpaDES or SyncroSym). The caribouMetrics package is
-available on GitHub at <https://github.com/LandSciTech/caribouMetrics>
-and a package website with detailed documentation is also available
-(<https://landscitech.github.io/caribouMetrics/>).
+model.
+The app allows advanced users to adjust most model parameters (supplemental Table A1).
+Default parameter values are set to allow novice users to get reasonable results quickly.
+Most of the code needed to run the app is in an R package, caribouMetrics, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of contexts (e.g.
+Frances SpaDES paper, Dyson preprint, syncrosim reference).
+The caribouMetrics package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.
+Find documentation and vignettes on the package website
+<https://landscitech.github.io/caribouMetrics/>.
+The shiny app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>, and code required to run analyses and produce figures in the paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.
 

### 2023-10-25 analysis/paper/supplement.Rmd
chipping away at manuscript

@@
 |$w=3,6,\text{ or } 9$|ratio|The apparent number of adult females per collared animal in composition survey.|
-|$q=\text{uniform}(0,0.6)$|ratio|True ratio of young bulls to adult females in composition survey.|
-|$u=\text{uniform}(0,0.2)$|probability|Probability of misidentifying young bulls as adult females and vice versa in composition survey.|
-|$z=\text{uniform}(0,0.2)$|probability|Probability of missing calves in composition survey.|
-|$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}$||Bias correction term (Appendix X).|
+|$\dot{q}=\text{uniform}(0,0.6)$|ratio|True ratio of young bulls to adult females in composition survey.|
+|$\dot{u}=\text{uniform}(0,0.2)$|probability|True probability of misidentifying young bulls as adult females and vice versa in composition survey.|
+|$\dot{z}=\text{uniform}(0,0.2)$|probability|True probability of missing calves in composition survey.|
+|$\dot{c}$||True bias correction term in simulation model (Supplement B).|
+|$q=\text{uniform}(0,0.6)$|ratio|Unknown ratio of young bulls to adult females in composition survey.|
+|$u=\text{uniform}(0,0.2)$|probability|Unknown probability of misidentifying young bulls as adult females and vice versa in composition survey.|
+|$z=\text{uniform}(0,0.2)$|probability|Unknown probability of missing calves in composition survey.|
+|$c$||Unknown bias correction term in Bayesian model (Supplement B).|
 

### 2023-11-21 analysis/paper/paper.Rmd
ongoing work on paper

@@
 Because this estimate is based on more than just the collared sample of cows, the sample size for recruitment estimates is typically larger than for survival (see multiplier parameter $w$, cowMult below).
-Note that occasionally young bulls can be misidentified as cows (parameters $\dot{u}$ and $u$ below; refs?
-), and there is also a possibility that some calves may be missed during aerial composition surveys (parameters $\dot{z}$ and $z$ below).
+However, a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020].
+Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa (parameters $\dot{u}$ and $u$ below; refs?
+), and failing to detect calves (parameters $\dot{z}$ and $z$ below) [@ellington_divergent_2020].
 We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details).
@@
 model.
-The app allows advanced users to adjust most model parameters (supplemental Table A1).
+The app allows advanced users to adjust many model parameters (supplemental Table A1).
 Default parameter values are set to allow novice users to get reasonable results quickly.
-Most of the code needed to run the app is in an R package, caribouMetrics, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of contexts (e.g.
-Frances SpaDES paper, Dyson preprint, syncrosim reference).
+Most of the code needed to run the app is in an R package, caribouMetrics, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g.
+@dyson_existing_2022; @stewart_climate-informed_2023].
 The caribouMetrics package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.
@@
 <https://landscitech.github.io/caribouMetrics/>.
-The shiny app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>, and code required to run analyses and produce figures in the paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.
+The Shiny user interface app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>, and code required to run analyses and produce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.
 

### 2023-11-21 analysis/paper/supplement.Rmd
ongoing work on paper

@@
 ```{r PriorKSDists, fig.cap="Similarity of the prior survival distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022].
-The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\\sigma^2_S$ in Table 2).
+The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (sSigmaMean & sSigmaSD, $\\sigma^2_S$ in Table 2).
 Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
@@
 ```{r PriorAllFits, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring.
-In these examples, the standard deviation of the random effect prior (SigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and anthropogenic disturbance is 0.
+In these examples, the standard deviation of the random effect prior (sSigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and anthropogenic disturbance is 0.
 Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}
@@
 ```{r PriorLowFits, fig.cap="Posterior predictive survival distributions from the Bayesian model ('local') assuming intensive monitoring (15 years of 60 collars per year) of populations from the 2.5 \\% quantile of the national model.
-In these examples, the standard deviation of the random effect prior (SigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and  anthropogenic disturbance is 0.
+In these examples, the standard deviation of the random effect prior (sSigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and  anthropogenic disturbance is 0.
 Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates.
@@
 
-knitr::include_graphics(here::here("figs/SurvsurvInterceptLowsSigmaSD0.03.png"))
+knitr::include_graphics(here::here("figs/SurvInterceptLowsSigmaSD0.03.png"))
 ```
@@
 Prior parameter values should be set so that, given no monitoring data, the prior survival distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance.
-This condition is met when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\sigma^2_S$ in Table 2) (Figs \@ref(fig:PriorKSDists) and \@ref(fig:PriorAllFits)).
+This condition is met when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (sSigmaMean & sSigmaSD, $\sigma^2_S$ in Table 2) (Figs \@ref(fig:PriorKSDists) and \@ref(fig:PriorAllFits)).
 Priors should also be vague enough so that, given intensive monitoring of populations where survival is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 60 collars per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local survival, rather than priors from the national model.
-Setting the mean and standard deviations of the prior distribution of the random effect of year on survival to moderately low values (SigmaMean=0.034784, SigmaSD=0.03, $\sigma^2_S$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDists), \@ref(fig:PriorAllFits) and \@ref(fig:PriorLowFits)).
+Setting the mean and standard deviations of the prior distribution of the random effect of year on survival to moderately low values (sSigmaMean=0.034784, sSigmaSD=0.03, $\sigma^2_S$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDists), \@ref(fig:PriorAllFits) and \@ref(fig:PriorLowFits)).
 
 ```{r PriorKSDistsRec, fig.cap="Similarity of the prior recruitment distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022].
-The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (rIntSEMod: $\\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\\sigma^2_R$ in Table 2).
+The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (rIntSEMod: $\\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (rSigmaMean & rSigmaSD, $\\sigma^2_R$ in Table 2).
 Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
@@
 ```{r PriorAllFitsRec, fig.cap="Differences between prior predictive recruitment distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring.
-In these examples, the standard deviation of the random effect prior (SigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and anthropogenic disturbance is 0.
+In these examples, the standard deviation of the random effect prior (rSigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and anthropogenic disturbance is 0.
 Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}
@@
 ```{r PriorHighFitsRec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model ('local') assuming intensive monitoring (9*60 cows per year for 15 years) of populations from the 97.5% quantile of the national model.
-In these examples, the standard deviation the random effect prior (SigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and  anthropogenic disturbance is 0.
+In these examples, the standard deviation the random effect prior (rSigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and  anthropogenic disturbance is 0.
 Bands show 95% posterior predictive intervals, black dots show true recruitment rates from the simulated population, and open triangles show observed recruitment rates.", cache=TRUE}
@@
 
-```{r PriorHighFitsAnthro90Rec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model, assuming intensive monitoring (15 years of 60 collars per year) of populations from the 97.5 \\% quantile of the national model.
+```{r PriorHighFitsAnthro90Rec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model, assuming intensive monitoring (9*60 cows per year for 15 years) of populations from the 97.5 \\% quantile of the national model.
 In these examples, anthropogenic disturbance is 90\\%, and the effect of anthropogenic disturbance varies from none (rSlopeMod=0) through expected (rSlopeMod=1) to double the expected slope (rSlopeMod=2).
@@
 Similarly, prior parameter values should be set so that, given no monitoring data, the prior recruitment distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance.
-This condition is met when the multiplier on the prior standard deviation of the recruitment intercept (rIntSEMod: $\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\sigma^2_R$ in Table 2) (Figs \@ref(fig:PriorKSDistsRec) and \@ref(fig:PriorAllFitsRec)).
+This condition is met when the multiplier on the prior standard deviation of the recruitment intercept (rIntSEMod: $\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (rSigmaMean & rSigmaSD, $\sigma^2_R$ in Table 2) (Figs \@ref(fig:PriorKSDistsRec) and \@ref(fig:PriorAllFitsRec)).
 Priors should also be vague enough so that, given intensive monitoring of populations where recruitment is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 9*60 cows per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local recruitment, rather than priors from the national model.
-Setting the mean and standard deviations of the prior distribution of the random effect of year on recruitment to moderately low values (SigmaMean=0.23, SigmaSD=0.22, $\sigma^2_R$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDistsRec), \@ref(fig:PriorAllFitsRec) and \@ref(fig:PriorHighFitsRec)).
+Setting the mean and standard deviations of the prior distribution of the random effect of year on recruitment to moderately low values (rSigmaMean=0.23, rSigmaSD=0.22, $\sigma^2_R$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDistsRec), \@ref(fig:PriorAllFitsRec) and \@ref(fig:PriorHighFitsRec)).
 

### 2023-11-21 analysis/scripts/sensitivityMinimalTest.R

@@ -45,6 +45,8 @@ message("batch ", cpageId, " started")
 
 str(scns)
 
+scns$collarCount = 1100
+
 if(n_reps=="all"){
   scResults = caribouMetrics:::runScnSet(scns,eParsIn,simBig,getKSDists=F,printProgress=F)
 }else{

### 2023-12-13 analysis/paper/paper.Rmd
in progress

@@
     correspondence: false
-  - name: Cheryl A Johnson
-    email: cheryl-ann.johnson@ec.gc.ca
-    institute: [ECCC]
-    correspondence: false
   - name: Anna M Calvert
@@
 institute:
+  - name: Cheryl A Johnson
+    email: cheryl-ann.johnson@ec.gc.ca
+    institute: [ECCC]
+    correspondence: false
   - ECCC: Environment and Climate Change Canada 
@@
 abstract: |
-  Across fifty-eight boreal caribou ranges in Canada, both survival and recruitment decrease with the percentage of the range that is disturbed.
-However, there is also variation in demographic rates among ranges, particularly where anthropogenic disturbance is low, and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships.
-Demographic projections derived from local range-specific data are also uncertain because few places have enough long-term monitoring data to characterize local demographic-disturbance relationships.
+  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed.
+Disturbance is the primary driver of caribou population declines.
+However, there is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships.
+Demographic projections derived from local population-specific data are also uncertain because few places have enough long-term monitoring data to characterize local demographic-disturbance relationships.
 We present a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections.
@@
 @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@johnson_science_2020].
+Although genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g.
+@hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021], there is still heavy reliance on collaring.
 Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with collars carrying  Very High Frequency (VHF) and/or Global Positioning Satellite (GPS) transmitters allowing remote detection and data retrieval [see @nbckc_nbckc_2021].
@@
 Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa (parameters $\dot{u}$ and $u$ below; refs?
-), and failing to detect calves (parameters $\dot{z}$ and $z$ below) [@ellington_divergent_2020].
+), and failing to detect calves (parameters $\dot{z}$ and $z$ below) [@ellington_divergent_2020; @decesare_estimating_2012].
 We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details).
@@
 $\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
-post-juvenile population size $N_0$ is 1000 animals. For reference, all
+post-juvenile population size $N_0$ is 2000 animals. For reference, all
 variables and parameters are defined in supplemental Table A1.
@@
 adult mortality risk - though sample size is a challenge. Recruitment,
-which is more variable among ranges and with anthropogenic disturbance,
+which is more variable among study areas and with anthropogenic disturbance,
 is difficult to characterize at finer spatial scales because it depends
 on the conditions experienced by mothers over years as they move across
-the range, and on spatial variation in calf mortality risk that is not
+the study area, and on spatial variation in calf mortality risk that is not
 measured directly.

### 2023-12-13 analysis/scripts/bayesianModelExamples.R

@@ -20,7 +20,7 @@ scns=merge(monitoringScns,stateScns)
 scns$iAnthro = 0
 scns$tA = scns$iAnthro+(scns$obsYears)*scns$obsAnthroSlope
 scns$projYears = 50-scns$obsYears
-scns$N0 = 1000
+scns$N0 = 2000
 scns$adjustR = TRUE
 
 ####################

### 2023-12-13 analysis/scripts/nationalModelTrajectories.R

@@ -97,7 +97,7 @@ johnsonCompare$c = 1
 pal = colorRampPalette(brewer.pal(7, "Blues"))(35)
 
 # demography
-pars <- data.frame(N0 = 1000)
+pars <- data.frame(N0 = 2000)
 # increase to get a better sample size, or set interannualVar to NA
 pars <- merge(pars, data.frame(rrp = 1))
 pars <- merge(pars, rateSamplesLarge)
@@ -107,7 +107,7 @@ pars1 <- cbind(pars, caribouPopGrowth(pars$N0,
   S_bar = pars$S_bar, c=pars$c, probOption = "binomial"
 ))
 
-pars <- data.frame(N0 = 1000)
+pars <- data.frame(N0 = 2000)
 # increase to get a better sample size, or set interannualVar to NA
 pars <- merge(pars, data.frame(rrp = 1))
 pars <- merge(pars, rateSamples)
@@ -137,7 +137,7 @@ plot_lambda <- ggplot(oo,
             alpha = 1) + scale_color_manual(values=pal)+
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
   xlab("Anthropogenic disturbance (%)") +
-  ylab(expression("Population Growth Rate no delay" * lambda)) +
+  ylab(expression("Population Growth Rate " * lambda)) +
   theme(legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
 
 str(pars1)
@@ -154,7 +154,7 @@ plot_recruitment3 <- ggplot(data = rateSummaries,
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
   scale_y_continuous(limits = c(0, 0.4), breaks = c(0, 0.1, 0.2, 0.3, 0.4)) +
   xlab("Anthropogenic disturbance (%)") +
-  ylab(expression("Adjusted Recruitment no delay " * dot(X)[t])) +
+  ylab(expression("Adjusted Recruitment " * dot(X)[t])) +
   theme(legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
 plot(plot_recruitment3)
 
@@ -174,7 +174,7 @@ base1 <- ggplot(data = rateSummaries,
 plot(base1)
 
 # combine ggplots to one figure
-ggpubr::ggarrange(plot_recruitment3,base1,plot_lambda, labels = "",
[…]

### 2023-12-13 analysis/scripts/sensitivity1.R

@@ -16,12 +16,16 @@
 #monitoringScns = expand.grid(P=c(1,4,8,16),st=c(30,60),cmult=c(3),ri=c(1,2,4),assessmentYrs=1)
 #stateScns = expand.grid(iA=c(0),rep=seq(1:30))
 
-setName = "s6"
-monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(1,15,30,60),cowMult=c(3,6,9),collarInterval=c(1,4),assessmentYrs=c(3))
+setName = "s7"
+monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
+                             cowMult=c(3,6,9),collarInterval=c(1,4),
+                             assessmentYrs=c(3))
 #TO DO - in next iteration, remove multiple years of one collar
-monitoringScns = subset(monitoringScns, !((obsYears>1)&(collarCount==1))&!((collarCount==1)&(cowMult>3))&!((collarCount==1)&(collarInterval>1)))
-stateScns = data.frame(tA=c(0,20,40,60,0,20,40,60),obsAnthroSlope=c(0,1,1,1,0,0,0,0),projAnthroSlope=c(1,1,1,1,0,0,0,0))
-stateScns = merge(stateScns,data.frame(rep=seq(1:200)))
+monitoringScns = subset(monitoringScns, !((obsYears>1)&(collarCount==0))&!((collarCount==1)&(cowMult>3))&!((collarCount==0)&(collarInterval>1)))
+stateScns = data.frame(tA=c(0,20,40,60,0,20,40,60),
+                       obsAnthroSlope=c(0,1,1,1,0,0,0,0),
+                       projAnthroSlope=c(1,1,1,1,0,0,0,0))
+stateScns = merge(stateScns,data.frame(rep=seq(1:500)))
 
 stateScns$sQuantile=runif(nrow(stateScns),min=0.01,max=0.99)
 stateScns$rQuantile = runif(nrow(stateScns),min=0.01,max=0.99)
@@ -33,8 +37,9 @@ scns$projYears = 20
 unique(scns$iAnthro)
 scns$repBatch = ceiling(scns$rep/50)
 table(scns$repBatch)
-scns$N0 = 1000
+scns$N0 = 2000
 scns$adjustR = TRUE
+scns$preYears = max(scns$obsYears)-scns$obsYears
 #scns=scns[10,]
 #scResults = readRDS("temp.Rds")
 head(scns)

### 2023-12-13 analysis/scripts/sensitivityMinimal.R

@@ -6,7 +6,7 @@ cpageId <- commandArgs(trailingOnly = TRUE)
 #setwd("C:/Users/HughesJo/Documents/gitprojects/Caribou-Demographic-Projection-Paper")
 library(caribouMetrics)
 
-setName = "s6"
+setName = "s7"
 
 #######################
 dir.create(paste0("figs/",setName),recursive=T)

### 2023-12-13 analysis/scripts/sensitivityMinimalTest.R

@@ -9,9 +9,9 @@ cDir = getwd()
 
 library(caribouMetrics)
 
-# cpageId <- 5;n_reps <- 1;cDir = "C:/Users/HughesJo/Documents/gitprojects/Caribou-Demographic-Projection-Paper"
+# cpageId <- 31;n_reps <- 1;cDir = "C:/Users/HughesJo/Documents/gitprojects/Caribou-Demographic-Projection-Paper"
 
-setName = "s6"
+setName = "s7"
 
 #######################
 dir.create(paste0(cDir,"/figs/",setName),recursive=T)
@@ -38,20 +38,17 @@ scns = subset(allScns, pageId==cpageId)
 
 str(scns)
 
-scns$zMin= 0; scns$zMax = 0
-scns$uMin = 0; scns$uMax = 0
-scns$qMin = 0; scns$qMax = 0
+#scns$zMin= 0; scns$zMax = 0
+#scns$uMin = 0; scns$uMax = 0
+#scns$qMin = 0; scns$qMax = 0
 message("batch ", cpageId, " started")
 
 str(scns)
 
-scns$collarCount = 1100
-
 if(n_reps=="all"){
   scResults = caribouMetrics:::runScnSet(scns,eParsIn,simBig,getKSDists=F,printProgress=F)
 }else{
-  unique(scns$rep)
-  scResults = caribouMetrics:::runScnSet(subset(scns,rep<=n_reps)[1,],eParsIn,simBig,getKSDists=F,printProgress=F)
+  scResults = caribouMetrics:::runScnSet(subset(scns,rep<=n_reps),eParsIn,simBig,getKSDists=F,printProgress=F)
 }
 
 saveRDS(scResults,paste0("results/",setName,"/rTest",cpageId,n_reps,cpageId,".Rds"))

### 2023-12-13 analysis/scripts/viewSensitivityResults.R

@@ -102,7 +102,7 @@ for(i in 1:length(pages)){
   probs = subset(scResults$rr.summary.all,(Parameter=="Population growth rate"))
   probs$projectionTime = probs$Year-2023
   probs$Anthro2023 = pmax(0,probs$tA)#pmax(0,probs$tA-2) #correcting for error - change back in round 4
-  probs$AssessmentYear = probs$Year
+  probs$YearsOfProjection = probs$projectionTime
 
   names(probs)
 
@@ -112,8 +112,8 @@ for(i in 1:length(pages)){
   distScns$grp = paste0(distScns$Anthro2023,distScns$projAnthroSlope)
 
   distScns$Timeline[distScns$Year==2023]="Finish monitoring"
-  distScns$Timeline[distScns$projectionTime==5]="Assessment 2028"
-  distScns$Timeline[distScns$projectionTime==20]="Assessment 2043"
+  distScns$Timeline[distScns$projectionTime==5]="End 5 yr projection (2028)"
+  distScns$Timeline[distScns$projectionTime==20]="End 20 yr projection (2043)"
 
   Ps = unique(distScns$obsYears)
   for(s in Ps){
@@ -166,6 +166,8 @@ for(i in 1:length(pages)){
   probs$wrong = ((probs$Mean<=0.99)|(probs$projSize<=10))&probs$viableTrue
   probs$CorrectStatus[probs$wrong]="no"
   probs$CorrectStatus[!probs$wrong]="yes"
+  probs$LambdaDiff = probs$trueMean-probs$Mean
+  #probs$LambdaDiff[(probs$projSize<=10)]=NA
 
   probs$pageLabB = paste0(batchStrip(probs$pageLab),"st",probs$collarCount,"ri",probs$collarInterval)
   pagesB=unique(probs$pageLabB)
@@ -174,20 +176,41 @@ for(i in 1:length(pages)){
   levels(probs$AnthroScn) = c("low","low-med","med-high","high")
 
   for(pp in pagesB){
-    #pp=pagesB[1]
+    #pp=pagesB[3]
     png(here::here(paste0("figs/",setName,"/bands",pp,".png")),
         height = 4, width = 7.48, units = "in",res=600)
     base=ggplot(subset(probs,pageLabB==pp),aes(x=obsYears,y=Mean,col=CorrectStatus))+geom_point(shape="-",size=3)+
-      facet_grid(AssessmentYear~AnthroScn,labeller="label_both")+
[…]

### 2023-12-14 analysis/paper/paper.Rmd
Paper in progress, new batch of sims to run.

@@
 However, a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020].
-Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa (parameters $\dot{u}$ and $u$ below; refs?
-), and failing to detect calves (parameters $\dot{z}$ and $z$ below) [@ellington_divergent_2020; @decesare_estimating_2012].
+Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa (parameters $\dot{u}$ and $u$ below), and failing to detect calves (parameters $\dot{z}$ and $z$ below) [@ellington_divergent_2020; @decesare_estimating_2012].
 We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details).
@@
 
-```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
+```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios.
+In all scenarios, the projection period begins in 2024.
+Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%).
+These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, not to reflect real landscape change.", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))
@@
 replaced each year ($o=1$), or there may be a 3 year gap between
-deployments ($o=4$). We focus on exploring a set of scenarios in which
+deployments ($o=4$).
+
+We focus on exploring a set of scenarios in which
 anthropogenic disturbance, measured as the percentage of area within 500
 m of disturbance [@johnson_science_2020; @eccc_scientific_2011],
-increases by 1% per year from a variety of starting points (Figure \@ref(fig:figDisturbance)).
-Scenarios in which anthropogenic disturbance does not increase over time
-are included in Supplement D.
+increases by 1% per year (Figure \@ref(fig:figDisturbance)).
+The duration of monitoring varies among scenarios from 1 to 24 years.
+In all scenarios, monitoring ends in 2023.
+The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years.
+The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through med-low (20\%) and med-high (40\%) to high (60\%).
+These simple hypothetical scenarios span a large range of values to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change.
+Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.
 
@@
 
-To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics.
+To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics which is described in the following paragraphs.
 We refer to this as the "national demographic model" because it is informed by national demographic-disturbance relationships.
@@
 
-### Population dynamics informed by national disturbance-demographic relationships
-
-```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction.
-White dots show means of 500 example trajectories.
-For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}
-
-knitr::include_graphics(here::here("analysis/paper/figs/DemographicRatesDelay.png"))
-
-```
+### A model for simulating example population dynamics informed by national disturbance-demographic relationships
 
@@
 distributed with true survival probability $\dot{S}_t$:
-$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Maximum
-potential recruitment rate is adjusted for sex ratio, misidentification
-biases (Supplement B) and (optionally) delayed age at first reproduction
-[@eacker_web-based_2019; @decesare_estimating_2012]:
+$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification
+biases (Supplement B) and (optionally) a delayed reproduction term to account for the reality that yearlings generally do not breed
+[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum potential recruitment of female calves per cow $\dot{X}_t$:
 \begin{equation}
@@
 We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6.
+Other authors have noted these sources of error [@ellington_divergent_2020; @decesare_estimating_2012], but our method of including uncertainty associated with unknown probabilities of missing calves and misclassification in recruitment surveys is novel.
 
@@
 
-### Observed survival of collared animals
+```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction.
+White dots show means of 500 example trajectories.
+For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}
 
-In our simulated populations, mortality risk does not vary over time or
+knitr::include_graphics(here::here("analysis/paper/figs/DemographicRatesDelay.png"))
+
+```
+
+### Simulation of observed survival of collared animals
+
+In our simulated example populations, mortality risk does not vary over time or
 among individuals, so the observed state (dead or alive) of a collared
@@
 
-### Observed recruitment from composition surveys
+### Simulation of observed recruitment from composition surveys
 
@@
 @stewart_climate-informed_2023; @dyson_existing_2022]. In the Bayesian model we add a Gaussian
-distributed random year effect [@eacker_web-based_2019]:
+distributed random year effect to account for stochastic variation among years [@eacker_web-based_2019]:
 \begin{equation}
@@
 To account for bias and additional uncertainty caused by unknown misidentification biases (Supplement B) and delayed age of age of first reproduction
-[@eacker_web-based_2019; @decesare_estimating_2012] the estimated recruitment probability is adjusted to get expected recruits per female:
+[@eacker_web-based_2019; @decesare_estimating_2012] the estimated recruitment probability is adjusted to get expected female recruits per female:
 \begin{equation}
@@
 bulls as adult females and vice versa $u$, and the unknown probability of
-missing a calf $z$ (Supplement B). Prior uncertainty about the value of thebias
+missing a calf $z$ (Supplement B). Prior uncertainty about the value of the bias
 correction term $c$ is approximated with a Log-normal distribution
@@
 As in the simulations (eq \@ref(eq:barS)), we model survival probability as an adjusted function of buffered anthropogenic disturbance $A_t$ with a log link
-[@johnson_science_2020; @stewart_climate-informed_2023; @dyson_existing_2022]. We add a Gaussian distributed random year effect [@eacker_web-based_2019]:
+[@johnson_science_2020; @stewart_climate-informed_2023; @dyson_existing_2022].
+We add a Gaussian distributed random year effect to account for stochastic variation among years [@eacker_web-based_2019]:
 \begin{equation}
@@
 #TO DO: indicate which of these are from Johnson, and which are calibrated.
-```
-
-```{r examplePredictions, fig.cap="Prior (top row) and posterior (bottom row) predictions from the Bayesian model for an example population trajectory.
-In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%.
-The population is monitored for 15 years with 30 collars per year, and 6 cows per collared cow in the composition surveys.
-The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading).
-Local population data (open triangles) reduces uncertainty (bottom row), but when disturbance is very high local data is not required to predict viability (population growth < 1).", cache=TRUE}
-
-knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamples.png"))
-
 ```
@@
 which the Bayesian posterior mean population growth rate and the true
-population growth rate are both above or below 1.
+population growth rate are both greater than 0.99, or both less than or equal to 0.99.
+To aid in interpretation of results and to highlight the advantages of selecting a metric that is directly relevant to the management objective of stable or increasing populations, we also show how integrating local monitoring data alters the distribution of differences between true population growth rates and Bayesian posterior mean estimated growth rates.
 
@@
 [-@eacker_web-based_2019].
-Our app allows
-users to provide existing local data or simulate data based on national
+Our app allows users to provide existing local data or simulate data based on national
 models as shown above, define an anthropogenic disturbance scenario and
@@
 <https://landscitech.github.io/caribouMetrics/>.
-The Shiny user interface app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>, and code required to run analyses and produce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.
+The Shiny user interface app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>.
+Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.
 
@@
 
-```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows, starting in 2023), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
+```{r examplePredictions, fig.cap="Prior (top row) and posterior (bottom row) predictions from the Bayesian model for an example population trajectory.
+In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%.
+The population is monitored for 15 years with 30 collars per year, and 6 cows per collared cow in the composition surveys.
+The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading).
+Local population data (open triangles) reduces uncertainty (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99).", cache=TRUE}
+
+knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamples.png"))
+
+```
+
+```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
 Probabilities are calculated for 200 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)).
-For each example population, the status assessment is correct if the Bayesian posterior mean and the true population growth rate are both less than or greater than one.
-When disturbance is very high the probability of correct status assessment is high even without local monitoring.
-Where disturbance is moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.", cache=TRUE}
+For each example population, the status assessment is correct if the Bayesian posterior mean and the true population growth rate are both less than or equal to 0.99, or both greater than 0.99.
+When disturbance is very high or very low the probability of correct status assessment is high even without local monitoring.
+Where disturbance is moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.
+In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6; varying this parameter does not alter qualitative outcomes (supplement D).", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
+```
+
+```{r figDiffs, fig.cap="The distribution of differences between the true population growth rate and the Bayesian posterior mean growth rate varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
+Differences are calculated for 200 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)).
+Differences decrease as the number of years of monitoring and the number of collars increase.
+Differences are larger when the projection period is longer.
+Differences are also higher when anthropogenic disturbance is high because stochastic variation in population growth rate is higher when population sizes are small.
+In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6, and collars are renewed each year.", cache=TRUE}
+knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
 ```
@@
 initial population size to the results.
+
+Discuss limitations of assumption that example populations remain in their quantile over time which limits ranges from transitioning e.g.
+really low values to really high values.
+If you think of the larger variation at low disturbance values = variation created by climatic events or some other short term factor influence habitat this could be problematic.
+not sure if being clear but trying to say that ranges with poor demography at low disturbance values could experience high demographic values a few years later (as observed in NWT)
+
+Discuss what is gained by including misidentification bias parameters.
+
+Discuss pros and cons of alternative approach to modelling with more age classes...
 

### 2023-12-14 analysis/scripts/sensitivity1.R

@@ -22,9 +22,9 @@ monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60)
                              assessmentYrs=c(3))
 #TO DO - in next iteration, remove multiple years of one collar
 monitoringScns = subset(monitoringScns, !((obsYears>1)&(collarCount==0))&!((collarCount==1)&(cowMult>3))&!((collarCount==0)&(collarInterval>1)))
-stateScns = data.frame(tA=c(0,20,40,60,0,20,40,60),
-                       obsAnthroSlope=c(0,1,1,1,0,0,0,0),
-                       projAnthroSlope=c(1,1,1,1,0,0,0,0))
+stateScns = data.frame(tA=c(0,20,40,60,0,20,40,60,60,40,20),
+                       obsAnthroSlope=c(0,1,1,1,0,0,0,0,0,0,0),
+                       projAnthroSlope=c(1,1,1,1,0,0,0,0,-1,-1,-1))
 stateScns = merge(stateScns,data.frame(rep=seq(1:500)))
 
 stateScns$sQuantile=runif(nrow(stateScns),min=0.01,max=0.99)

### 2023-12-14 analysis/scripts/viewSensitivityResults.R

@@ -16,6 +16,8 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 setName = "s6"
 dir.create(paste0("figs/",setName),recursive=T)
 scns = read.csv(here::here("tabs/s6.csv"))
+
+nrow(scns)
 pagesa=unique(scns$pageId)
 
 pages = pagesa

### 2023-12-15 analysis/paper/paper.Rmd
Tweaks, and adding simulation set without composition bias.

@@
 Where disturbance is moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.
-In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6; varying this parameter does not alter qualitative outcomes (supplement D).", cache=TRUE}
+In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6; varying this parameter does not qualitatively alter outcomes (supplement D).", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
@@
 
-```{r figDiffs, fig.cap="The distribution of differences between the true population growth rate and the Bayesian posterior mean growth rate varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
+```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
 Differences are calculated for 200 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)).
@@
 Differences are larger when the projection period is longer.
-Differences are also higher when anthropogenic disturbance is high because stochastic variation in population growth rate is higher when population sizes are small.
+Differences are also larger when anthropogenic disturbance is high because stochastic variation in population growth rate is high for small populations.
 In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6, and collars are renewed each year.", cache=TRUE}

### 2023-12-15 analysis/scripts/bayesianModelExamples.R

@@ -7,8 +7,6 @@ library(RColorBrewer)
 
 baseDir <- "."
 
-simBig<-getSimsNational(replicates=3000,forceUpdate=T)
-
 monitoringScns = data.frame(obsYears=c(1,15),collarCount=c(0,30),cowMult=c(6),collarInterval=c(1),assessmentYrs=c(3))
 stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
 stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
@@ -33,8 +31,11 @@ eParsIn$freqStartsByYear <- data.frame(Year = 1981:2023,
 eParsIn$collarOnTime=1
 eParsIn$collarOffTime=12
 eParsIn$collarNumYears=1
-
 labFontSize = 10; breakInterval=5
+
+###############################
+#With bias adjustment
+simBig<-getSimsNational(replicates=3000,forceUpdate=T)
 
 priorResult = caribouMetrics:::runScnSet(scns[1,],eParsIn,simBig,getKSDists=F,printProgress=F)
 priorResult$obs.all=NULL
@@ -90,6 +91,106 @@ plts <- ggpubr::ggarrange(recPrior, survPrior, lambdaPrior,
 ggpubr::ggarrange(plts, leg, ncol = 2, widths = c(6,1))+bgcolor("white")
 
 ggsave(paste0(baseDir,"/analysis/paper/figs/bayesianModelExamples.png"),
+       width = 12*0.8, height = 3.6*2, units = "in",
+       dpi = 1200)
+
+###############################
+#Without bias adjustment
+
+biasPars = data.frame(qMax=0,uMax=0,zMax=0)
+simBig<-getSimsNational(replicates=3000,forceUpdate=T,cPars=getScenarioDefaults(paramTable=biasPars))
+
+scns=merge(subset(scns,select=setdiff(names(scns),names(biasPars))),biasPars)
+priorResult = caribouMetrics:::runScnSet(scns[1,],eParsIn,simBig,getKSDists=F,printProgress=F)
+head(priorResult$obs.all)
+
+priorResult$obs.all=NULL
+
+yr_scale1 <- scale_x_continuous(breaks = 2023:2072 %>% .[0:9*5+1],
+                                labels = 2009:2058 %>% .[0:9*5+1] )
+yr_scale2 <- scale_x_continuous(breaks = 2009:2058 %>% .[0:9*5+1],
[…]

### 2023-12-15 analysis/scripts/sensitivity1.R

@@ -3,19 +3,6 @@
 
 ########################
 #sensitivity
-#setName = "s1"
-#minimalScn = expand.grid(P=1,st=1,cmult=0,ri=c(1,4),assessmentYrs=3)
-#monitoringScns = expand.grid(P=c(1,2,4,8,16),st=c(30,60),cmult=c(2,3,4),ri=c(1,2,4),assessmentYrs=c(1,3))
-#stateScns = data.frame(tA=c(0,20,40,60,0,20,40,60),aS=c(0,1,1,1,0,0,0,0),aSf=c(1,1,1,1,0,0,0,0))
-#stateScns = merge(stateScns,data.frame(rep=seq(1:100)))
-
-#monitoringScns = expand.grid(P=c(1,4,16),st=c(60),cmult=c(3),ri=c(1),assessmentYrs=c(3))
-#stateScns = data.frame(tA=c(0,20,40,60),aS=c(0,1,1,1),aSf=c(1,1,1,1))
-#stateScns = merge(stateScns,data.frame(rep=seq(1:10)))
-
-#monitoringScns = expand.grid(P=c(1,4,8,16),st=c(30,60),cmult=c(3),ri=c(1,2,4),assessmentYrs=1)
-#stateScns = expand.grid(iA=c(0),rep=seq(1:30))
-
 setName = "s7"
 monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
                              cowMult=c(3,6,9),collarInterval=c(1,4),
@@ -40,13 +27,36 @@ table(scns$repBatch)
 scns$N0 = 2000
 scns$adjustR = TRUE
 scns$preYears = max(scns$obsYears)-scns$obsYears
-#scns=scns[10,]
-#scResults = readRDS("temp.Rds")
-head(scns)
 
 scns$pageLab = paste0("cmult",scns$cowMult,"ay",scns$assessmentYrs,"aSf",scns$projAnthroSlope,"repBatch",scns$repBatch)
 scns$pageId = as.numeric(as.factor(scns$pageLab))
-unique(scns$pageId)
+
+write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
+pages=unique(scns$pageLab)
+
+
+setName = "s8"
+stateScns = data.frame(tA=c(0,20,40,60),
+                       obsAnthroSlope=c(0,1,1,1),
+                       projAnthroSlope=c(1,1,1,1))
[…]

### 2024-02-12 analysis/paper/paper.Rmd
adding EVPI and EVSI

@@
 
-runScenarios = F 
+runScenarios = F #To do: set up to run full analysis from within this document.
 doPriorCalibration=F #Note the prior calibration is time consuming.
+
+resultScenario = "S7"
+resultSet ="cmult6ay3aSf1"
+figSet = c("diffs","distScns","EVSI","power","EVPIall")
+
+for(f in figSet){
+  currentFile = here::here(paste0("figs/",resultScenario,"/",f,resultSet,".png"))
+  newFile = here::here(paste0("figs/forPaper/",f,resultSet,".png"))
+  file.copy(currentFile, newFile,overwrite=T)
+}
+
 ```
@@
 
-## Probability of correct status assessment
+## Probability of correct status assessment and expected value of sample information
 
@@
 monitoring scenario (Figure \@ref(fig:figDisturbance)) we fit the
-Bayesian model to 200 example population trajectories (e.g. Figure
+Bayesian model to 500 example population trajectories (e.g. Figure
 \@ref(fig:exampleTrajectories)). The probability of correct status
-assessment for that scenario is the percentage of these 200 examples in
+assessment for that scenario is the percentage of these 500 examples in
 which the Bayesian posterior mean population growth rate and the true
 population growth rate are both greater than 0.99, or both less than or equal to 0.99.
+
+Following Dunham et al.
+[-@dunham_strategic_2023] we also calculate the expected value of sample information (EVSI) and the expected value of perfect information (EVPI) [@runge_which_2011;@yokota_value_2004], assigning a value of 1 to correct status assessments, and a value of 0 to incorrect assessments ($\nu_{\alpha,\psi}=1 \text{when} \alpha=\psi, 0 \text{otherwise}$).
+The expected value with perfect knowledge is $EV_{certainty}=1$.
+The difference between this and the expected score using knowledge of disturbance the disturbance scenario $\tau$ (Figure \@ref(fig:figDisturbance)) and priors from the national model gives EVPI, a measure of the maximum possible benefit that can be gained from local sampling:
+\begin{equation}
+EVPI_{\tau} = 1-max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
+  (\#eq:EVPI)
+\end{equation}
+P(\psi|\tau) is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior.
+EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information:
+\begin{equation}
+EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - (1-EVPI_{\tau}),
+  (\#eq:EVSI)
+\end{equation}
+where $x$ is the sample information from a particular example trajectory.
+Simulated observations from 500 example trajectories for each disturbance & monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) represent the distribution of possible values of $x$.
+$P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.
+
 To aid in interpretation of results and to highlight the advantages of selecting a metric that is directly relevant to the management objective of stable or increasing populations, we also show how integrating local monitoring data alters the distribution of differences between true population growth rates and Bayesian posterior mean estimated growth rates.
@@
 
+Analysis of simulated example demographic trajectories confirms that our Bayesian model can effectively combine local demographic data with priors from national demographic-disturbance relationships to reduce uncertainty in boreal caribou demographic projections (Figs \@ref(fig:examplePredictions) and \@ref(fig:figDiffs)).
+As expected, uncertainty about the population growth rate decreases as sampling effort in each year (i.e.
+number of collars) or the number of years of monitoring (Fig \@ref(fig:figDiffs)) increase.
+Uncertainty is higher when the forecast horizon is long (compare 5 and 20 year projections in Fig \@ref(fig:figDiffs)).
+Uncertainty is also higher in our high anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because stochastic variation in population growth rates increase as populations become very small (as in top right panel of Fig \@ref(fig:examplePredictions)).
+These results confirm the Bayesian population model is working as it should, and none of them are surprising.
+
+Our analysis becomes more interesting and informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of sample information (Fig \@ref(fig:figEVSI)).
+When anthropogenic disturbance is high (right panels of Fig \@ref(fig:figStatusErrors) and Fig \@ref(fig:figEVSI)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy.
+At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Fig \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).
+The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (supplement D).
+Adding unknown unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (supplement E).
+
 ```{r examplePredictions, fig.cap="Prior (top row) and posterior (bottom row) predictions from the Bayesian model for an example population trajectory.
@@
 The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading).
-Local population data (open triangles) reduces uncertainty (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99).", cache=TRUE}
+Local population data (open triangles) reduces uncertainty (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99).
+When projected populations become very small, stochastic variation in projected population growth rate increases (as in the top right panel).", cache=TRUE}
 
 knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamples.png"))
+```
 
+```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
+Differences are calculated for 500 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)).
+Differences decrease as the number of years of monitoring and the number of collars increase.
+Differences are larger when the projection period is longer.
+Differences are also larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations.
+In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6, and collars are renewed each year.
+To ensure results are comparable across monitoring scenarios and that populations remain extant through the monitoring period, all simulations are started with $N_0=5000$ animals 24 years prior to the start of the projection period.", cache=TRUE}
+knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
 ```
@@
 ```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
-Probabilities are calculated for 200 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)).
+Probabilities are calculated for 500 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)).
+To ensure results are comparable across monitoring scenarios and that populations remain extant through the monitoring period, all simulations are started with $N_0=5000$ animals 24 years prior to the start of the projection period.
 For each example population, the status assessment is correct if the Bayesian posterior mean and the true population growth rate are both less than or equal to 0.99, or both greater than 0.99.
-When disturbance is very high or very low the probability of correct status assessment is high even without local monitoring.
-Where disturbance is moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.
-In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6; varying this parameter does not qualitatively alter outcomes (supplement D).", cache=TRUE}
+When disturbance is very high the probability of correct status assessment is high even without local monitoring.
+Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.
+In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6; varying this parameter does alter qualitative conclusions (supplements D and E).", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
@@
 
-```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
-Differences are calculated for 200 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)).
-Differences decrease as the number of years of monitoring and the number of collars increase.
-Differences are larger when the projection period is longer.
-Differences are also larger when anthropogenic disturbance is high because stochastic variation in population growth rate is high for small populations.
-In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6, and collars are renewed each year.", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
+```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
+When disturbance is very high EVSI is low, indicating that local sampling does not alter status assessments.
+Where disturbance is low or moderate, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as sample size increases.
+See Figure \\@ref(fig:figStatusErrors) for additional scenario details.", cache=TRUE}
+knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay3aSf1.png"))
+```
+
+```{r figEVPI, fig.cap="The expected value of perfect information (EVSI) varies with among anthropogenic disturbance scenarios (see increasing, decreasing and stable scenarios in Figs \\@ref(fig:figDisturbance), D.1 and D.5 respectively) and the number of years projected (x axis).
+When disturbance is high EVPI is low, indicating that local sampling will not alter status assessments.
+Increasing the forecast horizon can either increase or decrease the EVPI, depending on whether the anthropogenic disturbance trend makes outcomes more or less predictable.", cache=TRUE}
+knitr::include_graphics(here::here("figs/forPaper/EVPIall.png"))
 ```
@@
 
-Will need a section on the fundamental assumptions of these
-models/aspatial nature of them, e.g. that collared cows and cows
-observed in surveys are a random representative sample of the larger
-population, and that there is full mixing in the population so that
-estimated survival/recruitment rates apply equally to all animals in the
-population. We can discuss this with some contrasting examples, e.g.
-collaring 30 animals from a very small/restricted population (like
-Little Smoky, 308,606ha, pop size estimate 78 as of COSEWIC 2014) would
-likely be much more representative of the population vs. collaring 30
-animals from a relatively large/expansive population (northern Quebec
-range Q6, 62,156,186 ha, pop size 9000 as of COSEWIC 2014). But because
-of the way this model is built, we assume that the sample is equally
-representative in either of those cases, so changing the initial
-starting N won't affect model performance/predictions much here...
-
-On a related topic to population size; Note that the stuff we got from
-ArtR regarding simulations for sample size does acknowledge that pops of
-different sizes have different monitoring requirements, and Sara's
-Labrador numbers also show targets relative to pop size... Since this
-kind of aspatial model will mostly be using small population sizes in
-order to see any effects, we'll jsut have to acknowledge that a caveat
-of this is that we're not fully able to explore all the population size
-effects with this model, but generally Josie has found little impact of
-initial population size to the results.
-
-Discuss limitations of assumption that example populations remain in their quantile over time which limits ranges from transitioning e.g.
-really low values to really high values.
-If you think of the larger variation at low disturbance values = variation created by climatic events or some other short term factor influence habitat this could be problematic.
-not sure if being clear but trying to say that ranges with poor demography at low disturbance values could experience high demographic values a few years later (as observed in NWT)
-
-Discuss what is gained by including misidentification bias parameters.
-
-Discuss pros and cons of alternative approach to modelling with more age classes...
-
-Include some commentary on addressing limitations of aspatial
-demographic models with spatial models. Not a panacea. It is possible
-(cite examples) to develop spatially explicit models of movement and
-adult mortality risk - though sample size is a challenge. Recruitment,
-which is more variable among study areas and with anthropogenic disturbance,
-is difficult to characterize at finer spatial scales because it depends
-on the conditions experienced by mothers over years as they move across
-the study area, and on spatial variation in calf mortality risk that is not
-measured directly.
-
-We should likely also have a pgph about costs of monitoring - it'll be
-pretty short, cause basically we have no idea what it costs since nobody
-will tell us :) Through the NBCKC Monitoring Working Group work, we
-tried over and over to get some basic cost estimates from P/T reps and
-practitioners, and hit a wall - nobody wanted to tell us a thing. Don't
-know if we'd now find anyone is more open to sharing, but don't think we
-should get our hopes up. But at the very least we'll want to acknowledge
-that we haven't accounted for relative costs of monitoring... [e.g.
-costs outlined in the Telemetry guidance report from the MonitoringWG:
-"Expenses to consider in radio-collared monitoring will vary, based on:
-helicopter (or other aircraft) use, purchase of tracking collars, data
-fees associated with the collars (the cost of which varies by operating
-system, recording frequency and data transmission frequency), and
-live-capture equipment (3 or more sets of capture nets and associated
-buckets, a netting gun, and ankle hobbles)"] --\>perhaps the keen
-MonitoringWG contacts would be willing to provide even really rough cost
-estimates, and/or even jsut a relative sense of how different it would
-be to e.g. monitor a few pops for many yrs vs lots of pops for a short
-time...?
-
-Re paragraph on costs of monitoring. The reality is that costs vary with
-remoteness and the difficulty of finding animals to collar (which also
-depends on how many collars are deployed). Since costs are so
-context-specific, there isn't a good way to incorporate them into this
-general analysis. It would be interesting to work through a case study
-(i.e. Newfoundland) informed by context-specific costs, and consider how
-incorporation of those costs may change the relative value of
-alternative monitoring scenarios.
-
-Biases in composition survey estimates also vary with vegetation cover
-(i.e. the difficulty of seeing animals) and the care taken to correctly
-identify the sex of unknown animals. Misidentification rates could be
-estimated from discrepancies among results from multiple independent
-observers analyzing photos of animal groups.
-
-The tool we have developed is a proof of concept, designed to
-demonstrate possibilities and help users develop a general understanding
-of the relative value of alternative monitoring strategies for boreal
-caribou. In current form the tool will have limited value for managers
-who have some local data and wish to know the value of collecting more,
-because it is designed to either integrate real local data or simulated
-local data, but not both. Development of a tool designed to help
-decision makers assess the relative value of strategies for additional
-monitoring given their existing local data and national-demographic
-disturbance relationships would be a useful followup to this work.
-
-A tool that integrates real local data with national demographic
-disturbance relationships can also be used to provide consistent and
-transparent assessments of the strength of evidence for local deviations
-from the expected mean of national demographic-disturbance
-relationships. Help distinguish real variation among populations from
-apparent variation that can be attributed to lack of local data.
+See <https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.
 

### 2024-02-12 analysis/scripts/bayesianModelExamples.R

@@ -18,7 +18,7 @@ scns=merge(monitoringScns,stateScns)
 scns$iAnthro = 0
 scns$tA = scns$iAnthro+(scns$obsYears)*scns$obsAnthroSlope
 scns$projYears = 50-scns$obsYears
-scns$N0 = 2000
+scns$N0 = 5000
 scns$adjustR = TRUE
 
 ####################

### 2024-02-12 analysis/scripts/nationalModelTrajectories.R

@@ -97,21 +97,21 @@ johnsonCompare$c = 1
 pal = colorRampPalette(brewer.pal(7, "Blues"))(35)
 
 # demography
-pars <- data.frame(N0 = 2000)
+pars <- data.frame(N0 = 5000)
 # increase to get a better sample size, or set interannualVar to NA
 pars <- merge(pars, data.frame(rrp = 1))
 pars <- merge(pars, rateSamplesLarge)
-numSteps <- 20
+numSteps <- 1
 pars1 <- cbind(pars, caribouPopGrowth(pars$N0,
   numSteps = numSteps, R_bar = pars$R_bar,
   S_bar = pars$S_bar, c=pars$c, probOption = "binomial"
 ))
 
-pars <- data.frame(N0 = 2000)
+pars <- data.frame(N0 = 5000)
 # increase to get a better sample size, or set interannualVar to NA
 pars <- merge(pars, data.frame(rrp = 1))
 pars <- merge(pars, rateSamples)
-numSteps <- 20
+numSteps <- 1
 pars2 <- cbind(pars, caribouPopGrowth(pars$N0,
   numSteps = numSteps, R_bar = pars$R_bar,
   S_bar = pars$S_bar,c=pars$c, probOption = "binomial"
@@ -183,21 +183,21 @@ ggsave(paste0(baseDir,"/analysis/paper/figs/DemographicRates.png"), width = 12*0
 ##################
 # demography with delayed reproduction
 
-pars <- data.frame(N0 = 2000)
+pars <- data.frame(N0 = 5000)
 # increase to get a better sample size, or set interannualVar to NA
 pars <- merge(pars, data.frame(rrp = 1))
 pars <- merge(pars, rateSamplesLarge)
-numSteps <- 20
+numSteps <- 1
 pars1 <- cbind(pars, caribouPopGrowth(pars$N0,
   numSteps = numSteps, R_bar = pars$R_bar,
   S_bar = pars$S_bar, c=pars$c, probOption = "binomial",adjustR=T
 ))
 
-pars <- data.frame(N0 = 2000)
+pars <- data.frame(N0 = 5000)
 # increase to get a better sample size, or set interannualVar to NA
 pars <- merge(pars, data.frame(rrp = 1))
 pars <- merge(pars, rateSamples)
-numSteps <- 20
+numSteps <- 1
 pars2 <- cbind(pars, caribouPopGrowth(pars$N0,
[…]

### 2024-02-12 analysis/scripts/viewSensitivityResults.R

@@ -13,9 +13,12 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 
 ########################
 #sensitivity
-setName = "s9"
+setName = "s8"
 dir.create(paste0("figs/",setName),recursive=T)
 scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
+
+
+table(scns$pageId)
 
 nrow(scns)
 pagesa=unique(scns$pageId)
@@ -26,7 +29,7 @@ for (p in pagesa){
   if(!file.exists(paste0("results/",setName,"/rTest",p,".Rds"))){pages=pages[pages!=p]}
 }
 
-batchStrip<-function(l,batches=seq(1:10)){
+batchStrip<-function(l,batches=c(10,seq(1:9))){
   for(b in batches){
     l=gsub(paste0("repBatch",b),"",l,fixed=T)
   }
@@ -36,8 +39,10 @@ batchStrip<-function(l,batches=seq(1:10)){
 pages=sort(pages)
 
 combine=T
+addEV = F
+addProbs=F
 for(i in 1:length(pages)){
-  #combine=F;i=1
+  #combine=F;i=50
   cpageId=pages[i]
 
   if(i==length(pages)){
@@ -63,9 +68,10 @@ for(i in 1:length(pages)){
     scResults = readRDS(paste0("results/",setName,"/rTest",cpageId,".Rds"))
   }
 
-  if(as.numeric(strsplit(nextP,"repBatch")[[1]][2])<=as.numeric(strsplit(p,"repBatch")[[1]][2])){
+  if((as.numeric(strsplit(p,"repBatch")[[1]][2])!=10)&(as.numeric(strsplit(nextP,"repBatch")[[1]][2])<=as.numeric(strsplit(p,"repBatch")[[1]][2]))){
     combine=F
   }else{combine=T;next}
+
 
   #figure out how to count out errors.
   head(scResults$rr.summary.all)
@@ -103,6 +109,7 @@ for(i in 1:length(pages)){
     geom_hline(yintercept=1, color = "black",size=0.7)+scale_fill_discrete(type=pal2)+scale_color_discrete(type=pal2)
   print(base)
   dev.off()
+
 
   probs = subset(scResults$rr.summary.all,(Parameter=="Population growth rate"))
   probs$startYear = probs$startYear+probs$preYears
@@ -166,11 +173,12 @@ for(i in 1:length(pages)){
   probs = merge(probs,sizeTrue)
   probs = merge(probs,sizeProj)
 
-  head(probs)
[…]

### 2024-02-23 analysis/paper/paper.Rmd
Writing in progress.

@@
     correspondence: false
-institute:
   - name: Cheryl A Johnson
@@
     correspondence: false
+
+institute:
   - ECCC: Environment and Climate Change Canada 
@@
 abstract: |
-  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed.
-Disturbance is the primary driver of caribou population declines.
-However, there is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships.
-Demographic projections derived from local population-specific data are also uncertain because few places have enough long-term monitoring data to characterize local demographic-disturbance relationships.
-We present a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections.
-We combine this model with methods for simulating local population dynamics and monitoring to show how monitoring requirements depend on landscape condition.
-Where anthropogenic disturbance is very high or very low, reasonably accurate status projections can be made using only national demographic-disturbance relationships.
-At intermediate disturbance levels where uncertainty is higher, deploying few collars for more years may be a lower cost option for improving the reliability of status projections relative to more intensive short-term monitoring.
-We provide an R Shiny interface to enable others to explore the implications of alternative monitoring scenarios, disturbance trajectories, priors, and other model parameters for boreal caribou demographic projections.
-This tool can help users assess monitoring needs and strategies.
 keywords: |
@@
 for(f in figSet){
-  currentFile = here::here(paste0("figs/",resultScenario,"/",f,resultSet,".png"))
-  newFile = here::here(paste0("figs/forPaper/",f,resultSet,".png"))
+  if(f =="EVPIall"){addBit=""}else{addBit=resultSet}
+  currentFile = here::here(paste0("figs/",resultScenario,"/",f,addBit,".png"))
+  newFile = here::here(paste0("figs/forPaper/",f,addBit,".png"))
   file.copy(currentFile, newFile,overwrite=T)
@@
 (Figure \@ref(fig:exampleTrajectories)).
+To ensure results are comparable across monitoring scenarios and that populations remain extant through the monitoring period, all simulations are started with $N_0=5000$ animals 24 years prior to the start of the projection period.
 
@@
 females and the adjusted recruitment rate $X_t$:
-$J_{t} \sim \text{Poisson}(X_tW_t)$. The post-juvenile female population
+$J_{t} \sim \text{Poisson}(W_tX_t)$. The post-juvenile female population
 in the next year includes both survivors and new recruits:
@@
 <!-- Annotated equations for poster.
+$$J_{t} \sim \text{Poisson}(W_tX_t)$$
 $$\underbrace{N_{t+1}}_{\text{Adult females in year t+1}}=\underbrace{W_t}_{\text{Surviving adults}}+\underbrace{J_t}_{\text{New recruits to adult age class}}$$
@@
 
-## Probability of correct status assessment and expected value of sample information
+## Probability of correct status assessment and expected value of information
 
@@
 caribou and other endangered species, a relevant outcome is whether a
-population is declining or not. If uncertainty about population status
+population is declining or not [@dunham_strategic_2023]. If uncertainty about population status
 is low, additional investment in reducing uncertainty about the
-population growth rate may not be necessary or warranted. Thus, in this
-analysis our outcome of interest is how often the Bayesian posterior
+population growth rate may not be necessary or warranted. Thus, our outcome of interest is how often the Bayesian posterior
 mean correctly predicts the true population status (increasing or
 decreasing). To calculate this outcome for each disturbance and
-monitoring scenario (Figure \@ref(fig:figDisturbance)) we fit the
-Bayesian model to 500 example population trajectories (e.g. Figure
+monitoring scenario (Fig \@ref(fig:figDisturbance)) we fit the
+Bayesian model to simulated observations from 500 example population trajectories (e.g. Fig
 \@ref(fig:exampleTrajectories)). The probability of correct status
-assessment for that scenario is the percentage of these 500 examples in
+assessment is the percentage of these 500 examples in
 which the Bayesian posterior mean population growth rate and the true
@@
 Following Dunham et al.
-[-@dunham_strategic_2023] we also calculate the expected value of sample information (EVSI) and the expected value of perfect information (EVPI) [@runge_which_2011;@yokota_value_2004], assigning a value of 1 to correct status assessments, and a value of 0 to incorrect assessments ($\nu_{\alpha,\psi}=1 \text{when} \alpha=\psi, 0 \text{otherwise}$).
+[-@dunham_strategic_2023] we also calculate the expected value of sample information (EVSI) and the expected value of perfect information (EVPI) [@runge_which_2011;@yokota_value_2004], assigning a value of 1 to correct status assessments, and a value of 0 to incorrect assessments ($\nu_{\alpha,\psi}=1 \text{ when } \alpha=\psi, 0 \text{ otherwise}$).
 The expected value with perfect knowledge is $EV_{certainty}=1$.
-The difference between this and the expected score using knowledge of disturbance the disturbance scenario $\tau$ (Figure \@ref(fig:figDisturbance)) and priors from the national model gives EVPI, a measure of the maximum possible benefit that can be gained from local sampling:
+The difference between this and the expected score using knowledge of disturbance the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model gives EVPI, a measure of the maximum possible benefit that can be gained from local sampling:
 \begin{equation}
@@
 \end{equation}
-P(\psi|\tau) is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior.
+$P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior.
 EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information:
@@
 where $x$ is the sample information from a particular example trajectory.
-Simulated observations from 500 example trajectories for each disturbance & monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) represent the distribution of possible values of $x$.
+Simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) represent the distribution of possible values of $x$.
 $P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.
@@
 
-Our analysis becomes more interesting and informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of sample information (Fig \@ref(fig:figEVSI)).
-When anthropogenic disturbance is high (right panels of Fig \@ref(fig:figStatusErrors) and Fig \@ref(fig:figEVSI)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy.
-At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Fig \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).
+Our analysis becomes more interesting and informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of information (Figs \@ref(fig:figEVSI) and \@ref(fig:figEVPI)).
+When anthropogenic disturbance is high (right panels of Figs \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and \@ref(fig:figEVPI)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy.
+At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Figs \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).
 The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (supplement D).
@@
 Probabilities are calculated for 500 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)).
-To ensure results are comparable across monitoring scenarios and that populations remain extant through the monitoring period, all simulations are started with $N_0=5000$ animals 24 years prior to the start of the projection period.
-For each example population, the status assessment is correct if the Bayesian posterior mean and the true population growth rate are both less than or equal to 0.99, or both greater than 0.99.
 When disturbance is very high the probability of correct status assessment is high even without local monitoring.
@@
 
-```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
-When disturbance is very high EVSI is low, indicating that local sampling does not alter status assessments.
-Where disturbance is low or moderate, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as sample size increases.
+```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort.
+When disturbance is high EVSI is low, indicating that local sampling does not alter status assessments.
+Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as sample size increases.
 See Figure \\@ref(fig:figStatusErrors) for additional scenario details.", cache=TRUE}
+
 knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay3aSf1.png"))
@@
 
-```{r figEVPI, fig.cap="The expected value of perfect information (EVSI) varies with among anthropogenic disturbance scenarios (see increasing, decreasing and stable scenarios in Figs \\@ref(fig:figDisturbance), D.1 and D.5 respectively) and the number of years projected (x axis).
+```{r figEVPI, fig.cap="The expected value of perfect information (EVPI) varies among anthropogenic disturbance scenarios (see increasing, decreasing and stable scenarios in Figs \\@ref(fig:figDisturbance), D.1 and D.5 respectively) and the number of years projected.
 When disturbance is high EVPI is low, indicating that local sampling will not alter status assessments.
-Increasing the forecast horizon can either increase or decrease the EVPI, depending on whether the anthropogenic disturbance trend makes outcomes more or less predictable.", cache=TRUE}
+Increasing the forecast horizon can either increase or decrease the EVPI as the anthropogenic disturbance trend makes outcomes more or less predictable.", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/EVPIall.png"))
@@
 See <https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.
+
+## Bayesian integration of local demographic data and national demographic-disturbance relationships
+
+Very few places in Canada have enough local demographic data to characterize local demographic-disturbance relationships [@johnson_science_2020; @rudolph_demographic_2017;@hervieux_widespread_2013] so in most places it is necessary to make assumptions informed by observed variation in demographic rates with disturbance among populations and other relevant information about impacts of disturbance on caribou.
+However, high observed variation where anthropogenic disturbance is low leads to high uncertainty in projections informed by national demographic-disturbance relationships [@stewart_climate-informed_2023;@dyson_existing_2022].
+In some cases predictions of population growth rate informed by national demographic-disturbance relationships appear to disagree with estimates from local demographic data [@eccc_scientific_2011; @dyson_existing_2022], but without formal methods of accounting for uncertainty it can be difficult to distinguish true discrepancies from stochastic variation.
+To help address these challenges, we further developed a simple two-stage Bayesian population model [@eacker_web-based_2019] to acknowledge uncertainty associated with local demographic data and national demographic-disturbance relationships, and integrate these two sources of information to reduce uncertainty in projections of the impacts of changing landscape conditions.
+This Bayesian approach represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships.
+
+Our model assumes that increasing cumulative anthropogenic disturbance causes a log-linear decrease in caribou demographic rates (equations 4, 5, 10 and 13), reflecting a scientific consensus informed by a number of lines of evidence that cumulative anthropogenic disturbance has a negative impact on boreal caribou [@eccc_scientific_2011; @johnson_science_2020; @gouvernement_du_quebec_revue_2021].
+Local data can be integrated to estimate local demographic-disturbance parameters that differ from the national means, and in the absence of local data appropriate priors lead to a range of outcomes consistent with the observed range of variation across the country [Figures 3 and 5 in @johnson_science_2020].
+Others [@moeller_integrating_2021; @mcnay_demographic_2022; @eacker_web-based_2019] have also used Bayesian models to estimate caribou demographic rates from local monitoring data, but our addition of prior information from national demographic-disturbance relationships is novel.
+
+To help address concerns about potential confounding by large-scale variation in climate and primary productivity [@wilson_causal_2021;@neufeld_primary_2021], another reasonable approach would be to use priors from regional analyses of ecologically similar populations, such as areas of Eastern Canada that are sufficiently productive for industrial forestry.
+However, while acknowledging and allowing for variability, we also need to acknowledge and guard against risks that uncertainty will be inflated and used as an excuse to delay action, a tactic that is used in this and other decision contexts [@bissonnette_les_2020; @boan_climate_2018].
+In order to be informative and credible regional demographic-disturbance analyses must include a sufficient number of observations spanning a range from high to low anthropogenic disturbance.
+Priors that lead to projections of outcomes for caribou that fall outside the observed range of variation across the country, characterized here (Figure 2) by 95% predictive intervals from a national analysis (Johnson et al.
+2020), are implausible.
+In other words, if we have seen no examples of self-sustaining caribou populations at a particularly high level of disturbance, it is unreasonable to project or expect this outcome.
+These premises constrain the plausible demographic-disturbance priors, and clarify that no additional information is needed to predict outcomes for caribou where anthropogenic disturbance is high (Figure 5).
+
+## Allowing for composition survey errors
+
+Another novel aspect of our Bayesian population model is a term that allows for bias associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012].
+Lacking information about the true magnitude of this bias, we examined sensitivity to a range of assumed values, and found that adding unknown bias did increase uncertainty, but these increases were not large (supplement E).
+A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults, and the probability of missing calves.
+Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment.
+Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out.
+We did not investigate sensitivity to the assumption that new recruits are 50% females, but we note that this is another possible source of error that would most likely lead to low recruitment estimates.
+
+Given lack of knowledge about true error rates, the most appropriate interpretation of our results is that better information on composition survey error rates is required in order to assess their importance.
+Fecal pellets are proving to be an interesting source of information on sex and age ratios in surveyed groups [@mcfarlane_genetic_2018;@jones_novel_2023]; for example, in analysis of 14 populations of boreal and southern mountain caribou, Jones et al.
+2023 found sampled groups contained more females than males, and that mothers tend to stay with their daughters and sons, while fathers do not.
+These results are broadly consistent with our assumption that composition survey groups include mothers and younger animals of both sexes that can be difficult to classify visually.
+Documenting animals that are difficult to classify as “unknown” can help analysts minimize and account for classification bias [@ellington_divergent_2020], as can comparing classification results from multiple independent observers.
+The probability of missing calves is more difficult to quantify and account for using aerial survey methods alone, but likely increases with vegetation cover.
+
+Alternatives to aerial composition surveys for estimating recruitment to adult age classes include fecal pellet analysis [@mcfarlane_genetic_2018] and calf collaring [@ellington_divergent_2020].
+Each method has costs and benefits, and the reality is that monitoring everything everywhere all the time in great detail is neither feasible nor necessary.
+We hope that including a bias term in our simple demographic models will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is needed.
+
+## Other model assumptions
+
+Other caribou Bayesian population models have complexities that aren’t included in our simple model, including population size information from aerial or fecal dna surveys, and explicit representation of additional classes [@moeller_integrating_2021; @mcnay_demographic_2022].
+At present, our model is intended as a proof-of-concept that can be customized and extended as necessary to meet the needs of particular monitoring programs.
+In many contexts, population size and fecal dna survey information is not available, and a framework that includes these sources of information will be unnecessarily complicated.
+Rather than further complicate a proof-of-concept model applied to simulated data, we believe the most effective way forward will be to collaborate with interested decision makers to extend and customize the proof-of-concept model to meet their real local needs, recognizing that these real needs will vary among monitoring programs [@nbckc_nbckc_2021;@nbckc_nbckc_2021].
+That said, it would also be interesting to compare models with different numbers of age classes to clarify the benefits and drawbacks of that additional complexity.
+
+We further constrain our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower or higher than average remain lower or higher than average over time.
+Switching from low to high each year is unlikely, but environmental conditions are changing so reality likely lies between these two extremes.
+That said, we are not overly concerned about the implications of this assumption because it is not made in the Bayesian model, which can accommodate evidence of a shift from unexpectedly low to unexpectedly high demographic rates over time.
+We believe our simulated trajectories are sufficient to demonstrate how the Bayesian model could be used to integrate local data with national priors, and to show how monitoring requirements depend on landscape condition.
+
+Other important assumptions of these simple aspatial demographic models are that observed animals are a random and representative sample of a fully mixed population in which survival and recruitment rates apply equally to all animals.
+However, available genetic evidence suggests high variation in fitness among individuals [@mcfarlane_genetic_2018], and spatial structure within populations [@jones_novel_2023;@mcfarlane_spatial_2021;@mcfarlane_identification_2022].
+Risks and benefits for caribou vary spatially [@decesare_linking_2014;@avgar_spaceuse_2015] (more?) and observed variation in habitat preferences and space use among individuals [@avgar_spaceuse_2015] (other habitat selection models with individual variation?) also suggest imperfect mixing and spatial variation in survival and recruitment within populations.
+
+Although there is evidence of spatial structure and high variation in demographic rates among individuals within populations, estimation of spatial variation in demographic rates requires data that is not widely available.
+That said, DeCesare [-@decesare_linking_2014] successfully estimated spatial variation in adult female survival using 13 years of data from 468 collared animals.
+Analysis of in metrics derived from fecal pellet analysis such as reproductive success [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022], age structure and pregnancy rates [@flasko_fecal_2017;@mcfarlane_identification_2022], and mother-offspring co-occurence [@jones_novel_2023] can also provide insight into spatial variation in demographic rates.
+We look forward to a future in which caribou demographic projections are better informed by robust analyses of spatial variation in demographic rates.
+In the meantime, high variability in demographic rates among individuals [@mcfarlane_genetic_2018] could be incorporated into an aspatial Bayesian population model [@armstrong_incorporating_2021].
+
+In both our Bayesian and simulation models stochastic variation in demographic rates increase when population sizes are very small.
+In the simulation model, expected recruitment rate is highest at intermediate population sizes, and declines when populations are small or near carrying capacity.
+Our Bayesian population model does not include this density dependence, and it is not necessary for accurately the viability of our simulated populations (Figure 4).
+That said, real small populations may face additional challenges such as inbreeding depression and difficulty finding mates [@mills_conservation_2012] that are not included in our models, so the real probability of extinction of very small populations is likely to be higher than projected by these models.
+
+## Assessing the value of additional monitoring
+
+Monitoring rare and widely distributed species such as boreal caribou is expensive, resources are limited, and additional monitoring is not always necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012].
+An analysis of action plans for Canadian species at risk found half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022].
+Our demographic modeling tool is intended to help address that need, and our analysis helps clarify that no additional local monitoring information is needed to predict whether boreal caribou populations in highly disturbed landscapes are self-sustaining.
+
+There is more variation in caribou demographic rates among study areas with low to moderate disturbance, and thus more need for local data to inform demographic projections.
+In these places, an important question is how much additional monitoring information is required to inform decisions.
+Our analysis suggests a general pattern of diminishing returns from increasing monitoring effort, but more information on monitoring costs, objectives, and existing local data would be required to assess the adequacy and cost-effectiveness of monitoring scenarios in particular circumstances.
+Monitoring costs vary substantially with technologies used, the remoteness of herds, and the difficulty of collaring animals, which in turn depends on the number of currently collared animals, the population size, and vegetation across the landscape [@nbckc_boreal_2021].
+Additional impacts of monitoring on threatened animals should also be considered [@nbckc_boreal_2019].
+It is also important to recognize that projecting population viability is only one possible objective for monitoring [@nbckc_boreal_2019].
+Since monitoring costs, options and goals are context specific, we believe the most effective next step will be to use our proof-of-concept model and analysis as a starting point for working with local decision makers to build decision support tools suited to their particular circumstances and needs.
 

### 2024-02-23 analysis/paper/supplement.Rmd
in progress
Writing in progress.

@@
 
-resultScenario = "S6"
+resultScenario = "S7"
+noBiasResultScenario = "S8"
+
 ```
@@
 |**True demographic parameters**|
-|$N_0$|animals|Initial adult population size.|
+|$N_0=2000$|animals|Initial adult population size.|
 |$\bar{R}_t$|rate|Expected recruitment (calves per cow).|
@@
 |$c$||Unknown bias correction term in Bayesian model (Supplement B).|
+|**Expected value of perfect and sample information**|
+|$\psi=\text{viable or not}$|binary|True status of a population.|
+|$\alpha=\text{viable or not}$|binary|Assigned population status.|
+|$\nu_{\alpha,\psi}=1 \text{when} \alpha=\psi, 0 \text{otherwise}$|binary|Value of classification decision.|
+|$x$||Simulated observations from an example population.|
+|$\tau$||A disturbance and monitoring scenario.|
+|$P(\psi|\tau)$|probability|Prior probability of state $\psi$.|
+|$P(\psi|x,\tau)$|probability|Posterior probability of state $\psi$.|
+|$EVPI_\tau$||Expected value of perfect information.|
+|$EVSI_\tau$||Expected value of sample information.|
 
@@
 
-```{r figDisturbanceAnthroStatic, fig.cap="Anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
+```{r figDisturbanceAnthroStatic, fig.cap="Static anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
 knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult6ay3aSf0.png")))
@@
 
-```{r powercmult3ay3aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for for static anthropogenic disturbance scenarios).
+```{r powercmult3ay3aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios).
 In these examples there are 3 cows per collared cow in recruitment surveys.
@@
 
-```{r powercmult3ay3aSf1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure 1 for scenarios in which anthropogenic disturbance increases over time).
+```{r figDisturbanceAnthroDeclining, fig.cap="Declining anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult6ay3aSf-1.png")))
+```
+
+```{r powercmult3ay3aSfm1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroDeclining) for declining anthropogenic disturbance scenarios).
 In these examples there are 3 cows per collared cow in recruitment surveys.
+See Figure 4 for additional details.
+", cache=TRUE}
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult3ay3aSf-1.png")))
+```
+
+```{r powercmult6ay3aSfm1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroDeclining) for declining anthropogenic disturbance scenarios).
+In these examples there are 6 cows per collared cow in recruitment surveys.
+See Figure 4 for additional details.
+", cache=TRUE}
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult6ay3aSf-1.png")))
+```
+
+```{r powercmult9ay3aSfm1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroDeclining) for declining anthropogenic disturbance scenarios).
+In these examples there are 9 cows per collared cow in recruitment surveys.
+See Figure 4 for additional details.
+", cache=TRUE}
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult9ay3aSf-1.png")))
+```
+
+\newpage
+
+# Sensitivity to unkown bias in recruitment estimates from calf:cow surveys
+
+Adding unknown unknown bias does increase uncertainty, but these increases are not large (compare Figs \@ref(fig:powercmult3ay3aSf1)/\@ref(fig:powercmult3ay3aSf1NoBias) and \@ref(fig:powercmult9ay3aSf1)/\@ref(fig:powercmult9ay3aSf1NoBias)).
+A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults, and the probability of missing calves (supplement B).
+Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment.
+Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out.
+
+```{r powercmult3ay3aSf1, fig.cap="The probability of a correct status assessment with unknown bias in recruitment estimates from calf:cow surveys.
+In these examples there are 3 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
 See Figure 4 for additional details.
@@
 
-```{r powercmult9ay3aSf1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure 1 for scenarios in which anthropogenic disturbance increases over time).
-In these examples there are 9 cows per collared cow in recruitment surveys.
+```{r powercmult3ay3aSf1NoBias, fig.cap="The probability of a correct status assessment without bias in recruitment estimates from calf:cow surveys.
+In these examples there are 3 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
+See Figure 4 for additional details.
+", cache=TRUE}
+knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powercmult3ay3aSf1.png")))
+```
+
+```{r powercmult9ay3aSf1, fig.cap="The probability of a correct status assessment with unknown bias in recruitment estimates from calf:cow surveys.
+In these examples there are 9 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
 See Figure 4 for additional details.
@@
 knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult9ay3aSf1.png")))
+```
+
+```{r powercmult9ay3aSf1NoBias, fig.cap="The probability of a correct status assessment without bias in recruitment estimates from calf:cow surveys.
+In these examples there are 9 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
+See Figure 4 for additional details.
+", cache=TRUE}
+knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powercmult9ay3aSf1.png")))
 ```

### 2024-03-06 analysis/paper/paper.Rmd
Writing in progress.

@@
 
-NOTE: find working version of intro and discussion [here:]
-(<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit?usp=sharing&ouid=111541730342945675365&rtpof=true&sd=true>).
+NOTE: please provide comments and suggestions for the introduction here:
+<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit?usp=sharing&ouid=111541730342945675365&rtpof=true&sd=true>.
 
-```{=html}
-<!--
-Motivating example - forecasting impacts of disturbance - WBI and RoF example. In order to reduce uncertainty, need to integrate local information on current state of the population into forecasts.
+Across 58 study areas for threatened boreal caribou (_Rangifer tarandus caribou_) in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020].
+The causes of this relationship include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011].
+Decreases in recruitment as disturbance increases are also observed over time in places where there is sufficient historic demographic data [@rudolph_demographic_2017;@hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships  [@johnson_science_2020].
+There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020;@neufeld_primary_2021; @fortin_forest_2017; @gagne_impact_2016], and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships [@stewart_climate-informed_2023;@dyson_existing_2022].
+Boreal caribou are widely distributed across the Canadian boreal region [@eccc_scientific_2011], inhabiting some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022;@ibisch_global_2016;@venter_sixteen_2016].
+Monitoring rare and widely distributed species such as boreal caribou is expensive, and resources are limited.
+Analysis of the potential implications of monitoring outcomes for decisions can clarify that additional monitoring is not always necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012].
+A recent analysis of action plans for Canadian species at risk found half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022].
+For populations of caribou and other species at risk, deciding whether or not management actions will cause declines is important.
+Ecological simulation models can be used to evaluate the ability of alternative monitoring options to reduce the risk of errors in species status classifications and population viability projections [@dunham_strategic_2023].
 
-Lots of variability among caribou ranges in monitoring data: recency, duration, types of data.
-E.g.
-27+ consecutive years of both telemetry-survival and calf/cow-ratios in most Alberta ranges, vs.
-1-4 years of data in most Ontario ranges (and that's old) *note we may want to be careful about inter-province comparisons for political reasons, but the variation is super relevant when to-date we've had the same 'rules' apply across the board...
+Efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019], following broader trends in the analysis of population dynamics [@schaub_integrated_2021].
+In Bayesian population models, all available demographic or population data (e.g.
+abundance surveys, vital rate estimates, trend data) are combined into a single analysis that can permit novel parameter estimation, and improve precision, and account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021].
+These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021].
+However Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships, ignoring an important source of information about outcomes for boreal caribou across their range that can help reduce uncertainty in local demographic projections and status assessments, and limiting their ability to project outcomes on changing landscapes.
+Existing models have also ignored uncertainty associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012].
+Here, we present a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data and accounts for composition survey errors, building on the work of Eacker et al.
+(2019).
+Following the framework proposed by Dunham et al.
+(2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.
+We also offer an open source app that allows users to analyze existing local demographic data, or to explore alternative monitoring scenarios with simulated data.
 
-In Missisa (and other places, esp Ontario) available local data are old and limited.
-Not clear how indicative older data is of state of population.
-Need new data to resolve questions, but what type, how much is enough.
-Can't monitor in great detail everywhere all the time.
-More generally, need for tools to help assess the value of additional monitoring and the best way to invest additional monitoring resources, given the info that is already available.
--->
-```
 # Methods
@@
 
-Demographic data for caribou populations, including abundance, vital rates and population growth, are often obtained through the tracking of caribou cows marked with telemetry collars [e.g.
+Caribou demographic data are often obtained by tracking cows with telemetry collars [e.g.
 @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@johnson_science_2020].
 Although genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g.
-@hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021], there is still heavy reliance on collaring.
-Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with collars carrying  Very High Frequency (VHF) and/or Global Positioning Satellite (GPS) transmitters allowing remote detection and data retrieval [see @nbckc_nbckc_2021].
-Because these collars send out a mortality signal when no longer active after a predetermined period of time, adult survival can be estimated by tracking the fate of marked individuals (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou).
-Male caribou are not frequently collared.
+@hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021], there is still heavy reliance on collaring.
+Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou).
 
-Collared females are also used in most estimates of recruitment.
-Recruitment surveys involve locating collared individuals by aircraft and then surveying the entire group of caribou to estimate the ratio of calves to cows in the group [@nbckc_nbckc_2022].
-Because this estimate is based on more than just the collared sample of cows, the sample size for recruitment estimates is typically larger than for survival (see multiplier parameter $w$, cowMult below).
+To estimate recruitment, collared females are often located and surveyed by aircraft to estimate the ratio of calves to cows in the groups of animals that include collared animals [@nbckc_nbckc_2022].
+Groups typically contain uncollared animals so sample size is typically larger for recruitment than for survival estimates (see multiplier parameter $w$, cowMult below).
 However, a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020].
@@
 We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details).
-Telemetry and aerial survey data from these methods can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for monitoring population change of many caribou herds [e.g.
-@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019].
+Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].
 
@@
 ```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios.
-In all scenarios, the projection period begins in 2024.
 Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%).
-These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, not to reflect real landscape change.", cache=TRUE}
+These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, not to reflect real landscape change.
+In all scenarios, the projection period begins in 2024.", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))
@@
 The duration of monitoring varies among scenarios from 1 to 24 years.
-In all scenarios, monitoring ends in 2023.
 The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years.
 The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through med-low (20\%) and med-high (40\%) to high (60\%).
-These simple hypothetical scenarios span a large range of values to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change.
+These simple hypothetical scenarios span a large range to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change.
 Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.
@@
 ```
-## Simulation of local population dynamics and monitoring
+## Simulation of population dynamics and monitoring
 
-To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics which is described in the following paragraphs.
-We refer to this as the "national demographic model" because it is informed by national demographic-disturbance relationships.
-We use simulated trajectories to assess the performance of our proposed method for integrating local demographic observations with national disturbance-demographic relationships in a Bayesian population model.
+To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics.
+These example population trajectories are informed by prior information about the distribution of possible parameter values and outcomes from the national analysis of demographic-disturbance relationships [@johnson_science_2020] so we call it the "national" model.
+Our method for selecting parameter values for each example trajectory yields a continuous set of alternative models (hypotheses) about the state of a particular caribou population, wherein there are more examples of parameter values that prior analysis suggests are more likely; these alternative scenarios informed by prior beliefs are required for analysis of the expected value of alternative sampling strategies [@runge_which_2011; @dunham_strategic_2023].
+We also use the simulated trajectories to confirm that our Bayesian model successfully integrates local data with priors from the national model.
 For reference, all variables and parameters are defined in supplemental Table A1.
@@
 
-### Simulation of observed survival of collared animals
+### Simulation of observed survival and recruitment
 
@@
 \end{equation}
-
-### Simulation of observed recruitment from composition surveys
 
@@
 \@ref(tab:monitoringScenarios)), we calculate the mean and standard
-deviation of 10,000 samples of $\log{c}$ assuming the ratio of young
+deviation of 10,000 samples of $\log{ c}$ assuming the ratio of young
 bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult
@@
 associated precisions $\tau_t$ are estimated from known-fate radio
-collar data using the survival R package [@therneau_modeling_2000], with
+collar data using the \ital{survival} R package [@therneau_modeling_2000], with
 precisions for years with no observed mortality estimated using a
@@
 $\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
-post-juvenile population size $N_0$ is 2000 animals. For reference, all
+post-juvenile population size $N_0$ is 5000 animals. For reference, all
 variables and parameters are defined in supplemental Table A1.
@@
 -->
+We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the \ital{R2jags} R package [@su_r2jags_2021] to fit the Bayesian model.
+We ran 15,000 iterations of 4 chain, discarded the first 10,000 interations, and used a thinning rate of 2.
+We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the \ital{mcmcplots} R package [@curtis_mcmcplots_2018].
 
@@
 relevant to decision making [@gregory_structured_2012]. For boreal
-caribou and other endangered species, a relevant outcome is whether a
-population is declining or not [@dunham_strategic_2023]. If uncertainty about population status
-is low, additional investment in reducing uncertainty about the
-population growth rate may not be necessary or warranted. Thus, our outcome of interest is how often the Bayesian posterior
-mean correctly predicts the true population status (increasing or
-decreasing). To calculate this outcome for each disturbance and
+caribou and other endangered species, a relevant outcome is the probability of correctly predicting the true population status (increasing or
+decreasing) [@dunham_strategic_2023]. To calculate this outcome for each disturbance and
 monitoring scenario (Fig \@ref(fig:figDisturbance)) we fit the
 Bayesian model to simulated observations from 500 example population trajectories (e.g. Fig
-\@ref(fig:exampleTrajectories)). The probability of correct status
-assessment is the percentage of these 500 examples in
+\@ref(fig:exampleTrajectories)), and calculated the percentage of these 500 examples in
 which the Bayesian posterior mean population growth rate and the true
@@
 
-To make this work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios.
-Where appropropriate, we borrowed and built on code from an app developed by Eacker et al.
-[-@eacker_web-based_2019].
-Our app allows users to provide existing local data or simulate data based on national
-models as shown above, define an anthropogenic disturbance scenario and
-make projections. The results shown are similar to Figure
-\@ref(fig:examplePredictions) and allow for comparisons to the national
-model.
+Analysis was implemented using the R language [@r_core_team_r_2023] and a variety of \ital{tidyverse} packages [@wickham_welcome_2019].
+To make the work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios, borrowing and building on code from an app developed by Eacker et al.
+[-@eacker_web-based_2019] where appropriate.
+Our app allows users to provide existing local data, or to make projections informed by simulated data as we have done in this analysis (similar to Figure
+\@ref(fig:examplePredictions)).
 The app allows advanced users to adjust many model parameters (supplemental Table A1).
 Default parameter values are set to allow novice users to get reasonable results quickly.
-Most of the code needed to run the app is in an R package, caribouMetrics, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g.
+Most of the code needed to run the app is in an R package, \ital{caribouMetrics}, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g.
 @dyson_existing_2022; @stewart_climate-informed_2023].
-The caribouMetrics package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.
+The \ital{caribouMetrics} package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.
 Find documentation and vignettes on the package website
@@
 ```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
-Differences are calculated for 500 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)).
 Differences decrease as the number of years of monitoring and the number of collars increase.
-Differences are larger when the projection period is longer.
-Differences are also larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations.
-In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6, and collars are renewed each year.
-To ensure results are comparable across monitoring scenarios and that populations remain extant through the monitoring period, all simulations are started with $N_0=5000$ animals 24 years prior to the start of the projection period.", cache=TRUE}
+Differences are larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations.
+In these examples the composition survey cow multiplier $w$ is 6, and collars are renewed each year.", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
@@
 ```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
-Probabilities are calculated for 500 example population trajectories (see example trajectories in Figure \\@ref(fig:exampleTrajectories)).
-When disturbance is very high the probability of correct status assessment is high even without local monitoring.
+When disturbance is high the probability of correct status assessment is high even without local monitoring.
 Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.
-In these examples the apparent number of adult females per collared animal in composition survey $w$ is 6; varying this parameter does alter qualitative conclusions (supplements D and E).", cache=TRUE}
+In these examples the composition cow multiplier $w$ is 6; varying $w$ does alter qualitative conclusions (supplements D and E).", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
@@
 
-See <https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.
+NOTE: Please provide comments and suggestions for the discussion here: <https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.
 
@@
 
-Alternatives to aerial composition surveys for estimating recruitment to adult age classes include fecal pellet analysis [@mcfarlane_genetic_2018] and calf collaring [@ellington_divergent_2020].
+Alternatives to aerial composition surveys for estimating recruitment include fecal pellet analysis [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022] and calf collaring [@ellington_divergent_2020].
 Each method has costs and benefits, and the reality is that monitoring everything everywhere all the time in great detail is neither feasible nor necessary.
@@
 
-Other caribou Bayesian population models have complexities that aren’t included in our simple model, including population size information from aerial or fecal dna surveys, and explicit representation of additional classes [@moeller_integrating_2021; @mcnay_demographic_2022].
+Other caribou Bayesian population models have complexities not included in our simple model, including population size information from aerial or fecal dna surveys, and additional age classes [@moeller_integrating_2021; @mcnay_demographic_2022].
 At present, our model is intended as a proof-of-concept that can be customized and extended as necessary to meet the needs of particular monitoring programs.
@@
 Rather than further complicate a proof-of-concept model applied to simulated data, we believe the most effective way forward will be to collaborate with interested decision makers to extend and customize the proof-of-concept model to meet their real local needs, recognizing that these real needs will vary among monitoring programs [@nbckc_nbckc_2021;@nbckc_nbckc_2021].
-That said, it would also be interesting to compare models with different numbers of age classes to clarify the benefits and drawbacks of that additional complexity.
+That said, it would be helpful to clarify the benefits and drawbacks of adding additional age classes.
 
@@
 However, available genetic evidence suggests high variation in fitness among individuals [@mcfarlane_genetic_2018], and spatial structure within populations [@jones_novel_2023;@mcfarlane_spatial_2021;@mcfarlane_identification_2022].
-Risks and benefits for caribou vary spatially [@decesare_linking_2014;@avgar_spaceuse_2015] (more?) and observed variation in habitat preferences and space use among individuals [@avgar_spaceuse_2015] (other habitat selection models with individual variation?) also suggest imperfect mixing and spatial variation in survival and recruitment within populations.
+Risks and benefits for caribou vary spatially [@decesare_linking_2014;@avgar_spaceuse_2015] and observed variation in habitat preferences and space use among individuals [@avgar_spaceuse_2015] (other habitat selection models with individual variation?) also suggest imperfect mixing and spatial variation in survival and recruitment within populations.
 
@@
 That said, DeCesare [-@decesare_linking_2014] successfully estimated spatial variation in adult female survival using 13 years of data from 468 collared animals.
-Analysis of in metrics derived from fecal pellet analysis such as reproductive success [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022], age structure and pregnancy rates [@flasko_fecal_2017;@mcfarlane_identification_2022], and mother-offspring co-occurence [@jones_novel_2023] can also provide insight into spatial variation in demographic rates.
-We look forward to a future in which caribou demographic projections are better informed by robust analyses of spatial variation in demographic rates.
+Metrics derived from fecal pellet analysis such as reproductive success [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022], age structure and pregnancy rates [@flasko_fecal_2017;@mcfarlane_identification_2022], and mother-offspring co-occurence [@jones_novel_2023] can also provide insight into spatial variation in demographic rates.
+We look forward to a future in which caribou demographic projections are better informed by robust analyses of spatial variation.
 In the meantime, high variability in demographic rates among individuals [@mcfarlane_genetic_2018] could be incorporated into an aspatial Bayesian population model [@armstrong_incorporating_2021].
@@
 In both our Bayesian and simulation models stochastic variation in demographic rates increase when population sizes are very small.
-In the simulation model, expected recruitment rate is highest at intermediate population sizes, and declines when populations are small or near carrying capacity.
-Our Bayesian population model does not include this density dependence, and it is not necessary for accurately the viability of our simulated populations (Figure 4).
-That said, real small populations may face additional challenges such as inbreeding depression and difficulty finding mates [@mills_conservation_2012] that are not included in our models, so the real probability of extinction of very small populations is likely to be higher than projected by these models.
+Real small populations may face additional challenges such as inbreeding depression and difficulty finding mates [@mills_conservation_2012] that are not included in our models, so the real probability of extinction of very small populations is likely to be higher than projected by these models.
 
@@
 
-Monitoring rare and widely distributed species such as boreal caribou is expensive, resources are limited, and additional monitoring is not always necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012].
-An analysis of action plans for Canadian species at risk found half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022].
-Our demographic modeling tool is intended to help address that need, and our analysis helps clarify that no additional local monitoring information is needed to predict whether boreal caribou populations in highly disturbed landscapes are self-sustaining.
-
+We applied the framework suggested by Dunham et al.
+(2023) to show that no additional local monitoring information is needed to predict whether boreal caribou populations in highly disturbed landscapes are self-sustaining.
 There is more variation in caribou demographic rates among study areas with low to moderate disturbance, and thus more need for local data to inform demographic projections.
-In these places, an important question is how much additional monitoring information is required to inform decisions.
-Our analysis suggests a general pattern of diminishing returns from increasing monitoring effort, but more information on monitoring costs, objectives, and existing local data would be required to assess the adequacy and cost-effectiveness of monitoring scenarios in particular circumstances.
+In these places, our analysis suggests a general pattern of diminishing returns from increasing monitoring effort, but more information on monitoring costs, objectives, and existing local data would be required to assess the adequacy and cost-effectiveness of monitoring scenarios in particular circumstances.
 Monitoring costs vary substantially with technologies used, the remoteness of herds, and the difficulty of collaring animals, which in turn depends on the number of currently collared animals, the population size, and vegetation across the landscape [@nbckc_boreal_2021].
-Additional impacts of monitoring on threatened animals should also be considered [@nbckc_boreal_2019].
-It is also important to recognize that projecting population viability is only one possible objective for monitoring [@nbckc_boreal_2019].
+It is important to recognize that projecting population viability is only one possible objective for monitoring [@nbckc_boreal_2019], and that the impacts of monitoring on threatened animals should be considered [@nbckc_boreal_2019].
 Since monitoring costs, options and goals are context specific, we believe the most effective next step will be to use our proof-of-concept model and analysis as a starting point for working with local decision makers to build decision support tools suited to their particular circumstances and needs.

### 2024-04-12 analysis/paper/paper.Rmd
Writing ongoing

@@
 
-Across 58 study areas for threatened boreal caribou (_Rangifer tarandus caribou_) in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020].
+The number of species threatened with extinction is high and getting higher [@ipbes_global_2019].
+Reliable information on the status of populations and likely impacts of management are needed to inform conservation action but effective monitoring is challenging and costly [@legg_why_2006; @lindenmayer_effective_2018;@wintle_allocating_2010;@dunham_strategic_2023].
+Resources to support conservation and recovery are limited [@buxton_balancing_2022;@mccarthy_financial_2012;@gerber_conservation_2016;@wintle_spending_2019], delays increase risk [@martin_acting_2012], and some populations have been monitored to extinction [@lindenmayer_counting_2013].
+A recent analysis of action plans for Canadian species at risk found half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022].
+Analysis of the implications of monitoring outcomes for decisions can help clarify whether additional monitoring is necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012].
+
+Analysis methods that can help make better use of available monitoring data to inform decisions include Bayesian population modelling [@schaub_integrated_2021] and value of information analyses [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012].
+To be useful, these methods must be tailored to particular species and decision contexts.
+Here, we offer a Bayesian population model for boreal caribou (_Rangifer tarandus caribou_) that integrates information about the impacts of landscape change with available demographic data to reduce uncertainty in population viability projections for boreal caribou [building on @eacker_web-based_2019].
+Following the value of information analysis framework proposed by Dunham et al.
+(2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.
+To help make the approach more accessible, we offer an open source app that allows users to analyze existing local demographic data, or to explore alternative monitoring scenarios with simulated data.
+Our approach is tailored to boreal caribou, but could be adapted for other species at risk found in multiple populations that are costly to monitor and sensitive to habitat disturbance.
+
+Boreal caribou are widely and sparsely distributed through the Canadian boreal region across a gradient of anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020] that includes some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022;@ibisch_global_2016;@venter_sixteen_2016].
+Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020].
 The causes of this relationship include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011].
@@
 There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020;@neufeld_primary_2021; @fortin_forest_2017; @gagne_impact_2016], and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships [@stewart_climate-informed_2023;@dyson_existing_2022].
-Boreal caribou are widely distributed across the Canadian boreal region [@eccc_scientific_2011], inhabiting some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022;@ibisch_global_2016;@venter_sixteen_2016].
-Monitoring rare and widely distributed species such as boreal caribou is expensive, and resources are limited.
-Analysis of the potential implications of monitoring outcomes for decisions can clarify that additional monitoring is not always necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012].
-A recent analysis of action plans for Canadian species at risk found half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022].
-For populations of caribou and other species at risk, deciding whether or not management actions will cause declines is important.
-Ecological simulation models can be used to evaluate the ability of alternative monitoring options to reduce the risk of errors in species status classifications and population viability projections [@dunham_strategic_2023].
 
-Efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019], following broader trends in the analysis of population dynamics [@schaub_integrated_2021].
+The most widely available source of boreal caribou demographic information is from cows fitted with telemetry collars [e.g.
+@courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@decesare_estimating_2012], though genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g.
+@hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021].
+Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou).
+Recruitment is estimated from aerial surveys of the composition of groups of animals that include collared animals [@nbckc_nbckc_2022].
+Groups typically also contain uncollared animals, and a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020].
+Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa , and failing to detect calves [@ellington_divergent_2020; @decesare_estimating_2012].
+
+Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].
+Some efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019], following broader trends in the analysis of population dynamics [@schaub_integrated_2021].
 In Bayesian population models, all available demographic or population data (e.g.
@@
 These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021].
-However Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships, ignoring an important source of information about outcomes for boreal caribou across their range that can help reduce uncertainty in local demographic projections and status assessments, and limiting their ability to project outcomes on changing landscapes.
-Existing models have also ignored uncertainty associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012].
-Here, we present a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data and accounts for composition survey errors, building on the work of Eacker et al.
-(2019).
-Following the framework proposed by Dunham et al.
-(2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.
-We also offer an open source app that allows users to analyze existing local demographic data, or to explore alternative monitoring scenarios with simulated data.
+However Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships or composition survey errors.
+
+Here, we present a Bayesian population model for boreal caribou [building on @eacker_web-based_2019] designed to reduce uncertainty in projections of the impacts of changing landscape conditions by integrating prior information from national analysis of demographic-disturbance relationships with available local demographic data.
+Information is shared among populations using priors instead of integrating multiple populations into a hierarchical analysis [@leasure_hierarchical_2019] because obtaining permission to use caribou demographic data can be difficult, and most managers only have access to their own data.
+This Bayesian approach represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships.
+We show that integrating the most informative predictor of outcomes for boreal caribou across their range can substantially reduce uncertainty in local demographic projections and status assessments, improve viability projections on changing landscapes, and help clarify where, why and when additional data is required to inform decisions and project outcomes on changing landscapes.
+In cases where available information is sufficient for assessing and projecting population viability, there is a risk of misallocating limited resources to monitoring that could be better spent on conservation action.
+
+While we believe that using available information to reduce uncertainty is important, we also believe it is important to acknowledge remaining uncertainty.
+In general, parametric Bayesian population models include a number of simplifying assumptions, and misspecified models can yield biased or overconfident results [@schaub_integrated_2021].
+More specifically, models that do not account for observation errors can be misleading [@schaub_integrated_2021].
+Several authors have pointed out that  misclassification of adults and missing calves in composition surveys can bias recruitment estimates from those surveys [@ellington_divergent_2020; @decesare_estimating_2012], but none have thus far examined sensitivity of Bayesian caribou population models to these sources of error [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019].
+We address this gap with a new bias correction term derived from a simple model of the recruitment survey observation process.
 
@@
 
-## Caribou population monitoring: survival and recruitment metrics
+## Caribou population monitoring data and composition survey bias correction
 
-Caribou demographic data are often obtained by tracking cows with telemetry collars [e.g.
-@courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@johnson_science_2020].
-Although genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g.
-@hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021], there is still heavy reliance on collaring.
-Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou).
+We focus on analyzing survival of collared females (as in @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019), and the ratio of calves to cows in aerial surveys of groups that include collared females [@nbckc_nbckc_2022], because these remain the most widely available sources of demographic information for boreal caribou populations [@johnson_science_2020].
+Surveyed groups also contain uncollared animals so sample size is typically larger for recruitment than for survival estimates.
+Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa, and failing to detect calves [@ellington_divergent_2020; @decesare_estimating_2012].
 
-To estimate recruitment, collared females are often located and surveyed by aircraft to estimate the ratio of calves to cows in the groups of animals that include collared animals [@nbckc_nbckc_2022].
-Groups typically contain uncollared animals so sample size is typically larger for recruitment than for survival estimates (see multiplier parameter $w$, cowMult below).
-However, a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020].
-Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa (parameters $\dot{u}$ and $u$ below), and failing to detect calves (parameters $\dot{z}$ and $z$ below) [@ellington_divergent_2020; @decesare_estimating_2012].
-We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details).
-Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].
+To derive a composition survey bias correction term we assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$).
+The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is a multiplier that defines the apparent number of adult females as a function of the number of collared animals.
+The ratio of young bulls to uncollared adult females in the group is
+$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$.
+Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$.
+Given these assumptions and definitions, the sex and bias-corrected recruitment rate $X$ can be written as function of the observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$ (see Supplement B for details).
+For simplicity, we combine the probabilities into a bias correction term $c$ that is integrated into the Bayesian population model and simulations (below):
+$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$
 
@@
 Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%).
-These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, not to reflect real landscape change.
-In all scenarios, the projection period begins in 2024.", cache=TRUE}
+These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, and not intended to reflect real landscape change.
+In all scenarios, the projection period begins in 2024, and the monitoring period occurs prior to the projection period.", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))
@@
 increases by 1% per year (Figure \@ref(fig:figDisturbance)).
-The duration of monitoring varies among scenarios from 1 to 24 years.
+The duration of monitoring prior to the projection period varies among scenarios from 1 to 24 years.
 The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years.
@@
 We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6.
-Other authors have noted these sources of error [@ellington_divergent_2020; @decesare_estimating_2012], but our method of including uncertainty associated with unknown probabilities of missing calves and misclassification in recruitment surveys is novel.
 
@@
 ```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction.
-White dots show means of 500 example trajectories.
+Orange dots show means of 500 example trajectories.
 For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}
@@
 The difference between equations \@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to simulate observations.
-In Bayesian model observations are provided, and used to estimate the unobserved recruitment rate $R_t$.
+In the Bayesian model, in contrast, observations are provided, and used to estimate the unobserved recruitment rate $R_t$.
 
@@
 associated precisions $\tau_t$ are estimated from known-fate radio
-collar data using the \ital{survival} R package [@therneau_modeling_2000], with
+collar data using the _survival_ R package [@therneau_modeling_2000], with
 precisions for years with no observed mortality estimated using a
@@
 -->
-We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the \ital{R2jags} R package [@su_r2jags_2021] to fit the Bayesian model.
+We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the _R2jags_ R package [@su_r2jags_2021] to fit the Bayesian model.
 We ran 15,000 iterations of 4 chain, discarded the first 10,000 interations, and used a thinning rate of 2.
-We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the \ital{mcmcplots} R package [@curtis_mcmcplots_2018].
+We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the _mcmcplots_ R package [@curtis_mcmcplots_2018].
 
@@
 \end{equation}
-$P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior.
+where $P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior.
 EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information:
@@
 
-Analysis was implemented using the R language [@r_core_team_r_2023] and a variety of \ital{tidyverse} packages [@wickham_welcome_2019].
+Analysis was implemented using the R language [@r_core_team_r_2023] and a variety of _tidyverse} packages [@wickham_welcome_2019].
 To make the work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios, borrowing and building on code from an app developed by Eacker et al.
@@
 Default parameter values are set to allow novice users to get reasonable results quickly.
-Most of the code needed to run the app is in an R package, \ital{caribouMetrics}, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g.
+Most of the code needed to run the app is in an R package, _caribouMetrics_, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g.
 @dyson_existing_2022; @stewart_climate-informed_2023].
-The \ital{caribouMetrics} package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.
+The _caribouMetrics_ package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.
 Find documentation and vignettes on the package website
@@
 Uncertainty is also higher in our high anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because stochastic variation in population growth rates increase as populations become very small (as in top right panel of Fig \@ref(fig:examplePredictions)).
-These results confirm the Bayesian population model is working as it should, and none of them are surprising.
+These results confirm the Bayesian population model is working as it should.
 
@@
 When anthropogenic disturbance is high (right panels of Figs \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and \@ref(fig:figEVPI)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy.
+Interestingly, when the projection period is long and anthropogenic disturbance is high, we see a negative EVSI (lower right panel of Fig \@ref(fig:figEVSI)), and we do not see an associated decrease in the probability of correctly assessing population status (lower right panel of Fig \@ref(fig:figStatusErrors)).
+This means that the long-term projections informed by local monitoring data are on average less precise but not less accurate, which can happen when there is very small prior probability that population is viable (orange band in the top right panel of Fig \@ref(fig:examplePredictions) falls below 1).
 At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Figs \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).
 The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (supplement D).
-Adding unknown unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (supplement E).
+Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (supplement E).
 
@@
 In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%.
-The population is monitored for 15 years with 30 collars per year, and 6 cows per collared cow in the composition surveys.
+The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys.
 The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading).
@@
 When disturbance is high the probability of correct status assessment is high even without local monitoring.
-Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as sample size increases.
-In these examples the composition cow multiplier $w$ is 6; varying $w$ does alter qualitative conclusions (supplements D and E).", cache=TRUE}
+Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases.
+In these examples the composition cow multiplier $w$ is 6; varying $w$ does not alter qualitative conclusions (supplements D and E).", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
@@
 When disturbance is high EVSI is low, indicating that local sampling does not alter status assessments.
-Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as sample size increases.
+Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases.
 See Figure \\@ref(fig:figStatusErrors) for additional scenario details.", cache=TRUE}
@@
 NOTE: Please provide comments and suggestions for the discussion here: <https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.
-
-## Bayesian integration of local demographic data and national demographic-disturbance relationships
-
-Very few places in Canada have enough local demographic data to characterize local demographic-disturbance relationships [@johnson_science_2020; @rudolph_demographic_2017;@hervieux_widespread_2013] so in most places it is necessary to make assumptions informed by observed variation in demographic rates with disturbance among populations and other relevant information about impacts of disturbance on caribou.
-However, high observed variation where anthropogenic disturbance is low leads to high uncertainty in projections informed by national demographic-disturbance relationships [@stewart_climate-informed_2023;@dyson_existing_2022].
-In some cases predictions of population growth rate informed by national demographic-disturbance relationships appear to disagree with estimates from local demographic data [@eccc_scientific_2011; @dyson_existing_2022], but without formal methods of accounting for uncertainty it can be difficult to distinguish true discrepancies from stochastic variation.
-To help address these challenges, we further developed a simple two-stage Bayesian population model [@eacker_web-based_2019] to acknowledge uncertainty associated with local demographic data and national demographic-disturbance relationships, and integrate these two sources of information to reduce uncertainty in projections of the impacts of changing landscape conditions.
-This Bayesian approach represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships.
-
-Our model assumes that increasing cumulative anthropogenic disturbance causes a log-linear decrease in caribou demographic rates (equations 4, 5, 10 and 13), reflecting a scientific consensus informed by a number of lines of evidence that cumulative anthropogenic disturbance has a negative impact on boreal caribou [@eccc_scientific_2011; @johnson_science_2020; @gouvernement_du_quebec_revue_2021].
-Local data can be integrated to estimate local demographic-disturbance parameters that differ from the national means, and in the absence of local data appropriate priors lead to a range of outcomes consistent with the observed range of variation across the country [Figures 3 and 5 in @johnson_science_2020].
-Others [@moeller_integrating_2021; @mcnay_demographic_2022; @eacker_web-based_2019] have also used Bayesian models to estimate caribou demographic rates from local monitoring data, but our addition of prior information from national demographic-disturbance relationships is novel.
-
-To help address concerns about potential confounding by large-scale variation in climate and primary productivity [@wilson_causal_2021;@neufeld_primary_2021], another reasonable approach would be to use priors from regional analyses of ecologically similar populations, such as areas of Eastern Canada that are sufficiently productive for industrial forestry.
-However, while acknowledging and allowing for variability, we also need to acknowledge and guard against risks that uncertainty will be inflated and used as an excuse to delay action, a tactic that is used in this and other decision contexts [@bissonnette_les_2020; @boan_climate_2018].
-In order to be informative and credible regional demographic-disturbance analyses must include a sufficient number of observations spanning a range from high to low anthropogenic disturbance.
-Priors that lead to projections of outcomes for caribou that fall outside the observed range of variation across the country, characterized here (Figure 2) by 95% predictive intervals from a national analysis (Johnson et al.
-2020), are implausible.
-In other words, if we have seen no examples of self-sustaining caribou populations at a particularly high level of disturbance, it is unreasonable to project or expect this outcome.
-These premises constrain the plausible demographic-disturbance priors, and clarify that no additional information is needed to predict outcomes for caribou where anthropogenic disturbance is high (Figure 5).
-
-## Allowing for composition survey errors
-
-Another novel aspect of our Bayesian population model is a term that allows for bias associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012].
-Lacking information about the true magnitude of this bias, we examined sensitivity to a range of assumed values, and found that adding unknown bias did increase uncertainty, but these increases were not large (supplement E).
-A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults, and the probability of missing calves.
-Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment.
-Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out.
-We did not investigate sensitivity to the assumption that new recruits are 50% females, but we note that this is another possible source of error that would most likely lead to low recruitment estimates.
-
-Given lack of knowledge about true error rates, the most appropriate interpretation of our results is that better information on composition survey error rates is required in order to assess their importance.
-Fecal pellets are proving to be an interesting source of information on sex and age ratios in surveyed groups [@mcfarlane_genetic_2018;@jones_novel_2023]; for example, in analysis of 14 populations of boreal and southern mountain caribou, Jones et al.
-2023 found sampled groups contained more females than males, and that mothers tend to stay with their daughters and sons, while fathers do not.
-These results are broadly consistent with our assumption that composition survey groups include mothers and younger animals of both sexes that can be difficult to classify visually.
-Documenting animals that are difficult to classify as “unknown” can help analysts minimize and account for classification bias [@ellington_divergent_2020], as can comparing classification results from multiple independent observers.
-The probability of missing calves is more difficult to quantify and account for using aerial survey methods alone, but likely increases with vegetation cover.
-
-Alternatives to aerial composition surveys for estimating recruitment include fecal pellet analysis [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022] and calf collaring [@ellington_divergent_2020].
-Each method has costs and benefits, and the reality is that monitoring everything everywhere all the time in great detail is neither feasible nor necessary.
-We hope that including a bias term in our simple demographic models will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is needed.
-
-## Other model assumptions
-
-Other caribou Bayesian population models have complexities not included in our simple model, including population size information from aerial or fecal dna surveys, and additional age classes [@moeller_integrating_2021; @mcnay_demographic_2022].
-At present, our model is intended as a proof-of-concept that can be customized and extended as necessary to meet the needs of particular monitoring programs.
-In many contexts, population size and fecal dna survey information is not available, and a framework that includes these sources of information will be unnecessarily complicated.
-Rather than further complicate a proof-of-concept model applied to simulated data, we believe the most effective way forward will be to collaborate with interested decision makers to extend and customize the proof-of-concept model to meet their real local needs, recognizing that these real needs will vary among monitoring programs [@nbckc_nbckc_2021;@nbckc_nbckc_2021].
-That said, it would be helpful to clarify the benefits and drawbacks of adding additional age classes.
-
-We further constrain our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower or higher than average remain lower or higher than average over time.
-Switching from low to high each year is unlikely, but environmental conditions are changing so reality likely lies between these two extremes.
-That said, we are not overly concerned about the implications of this assumption because it is not made in the Bayesian model, which can accommodate evidence of a shift from unexpectedly low to unexpectedly high demographic rates over time.
-We believe our simulated trajectories are sufficient to demonstrate how the Bayesian model could be used to integrate local data with national priors, and to show how monitoring requirements depend on landscape condition.
-
-Other important assumptions of these simple aspatial demographic models are that observed animals are a random and representative sample of a fully mixed population in which survival and recruitment rates apply equally to all animals.
-However, available genetic evidence suggests high variation in fitness among individuals [@mcfarlane_genetic_2018], and spatial structure within populations [@jones_novel_2023;@mcfarlane_spatial_2021;@mcfarlane_identification_2022].
-Risks and benefits for caribou vary spatially [@decesare_linking_2014;@avgar_spaceuse_2015] and observed variation in habitat preferences and space use among individuals [@avgar_spaceuse_2015] (other habitat selection models with individual variation?) also suggest imperfect mixing and spatial variation in survival and recruitment within populations.
-
-Although there is evidence of spatial structure and high variation in demographic rates among individuals within populations, estimation of spatial variation in demographic rates requires data that is not widely available.
-That said, DeCesare [-@decesare_linking_2014] successfully estimated spatial variation in adult female survival using 13 years of data from 468 collared animals.
-Metrics derived from fecal pellet analysis such as reproductive success [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022], age structure and pregnancy rates [@flasko_fecal_2017;@mcfarlane_identification_2022], and mother-offspring co-occurence [@jones_novel_2023] can also provide insight into spatial variation in demographic rates.
-We look forward to a future in which caribou demographic projections are better informed by robust analyses of spatial variation.
-In the meantime, high variability in demographic rates among individuals [@mcfarlane_genetic_2018] could be incorporated into an aspatial Bayesian population model [@armstrong_incorporating_2021].
-
-In both our Bayesian and simulation models stochastic variation in demographic rates increase when population sizes are very small.
-Real small populations may face additional challenges such as inbreeding depression and difficulty finding mates [@mills_conservation_2012] that are not included in our models, so the real probability of extinction of very small populations is likely to be higher than projected by these models.
-
-## Assessing the value of additional monitoring
-
-We applied the framework suggested by Dunham et al.
-(2023) to show that no additional local monitoring information is needed to predict whether boreal caribou populations in highly disturbed landscapes are self-sustaining.
-There is more variation in caribou demographic rates among study areas with low to moderate disturbance, and thus more need for local data to inform demographic projections.
-In these places, our analysis suggests a general pattern of diminishing returns from increasing monitoring effort, but more information on monitoring costs, objectives, and existing local data would be required to assess the adequacy and cost-effectiveness of monitoring scenarios in particular circumstances.
-Monitoring costs vary substantially with technologies used, the remoteness of herds, and the difficulty of collaring animals, which in turn depends on the number of currently collared animals, the population size, and vegetation across the landscape [@nbckc_boreal_2021].
-It is important to recognize that projecting population viability is only one possible objective for monitoring [@nbckc_boreal_2019], and that the impacts of monitoring on threatened animals should be considered [@nbckc_boreal_2019].
-Since monitoring costs, options and goals are context specific, we believe the most effective next step will be to use our proof-of-concept model and analysis as a starting point for working with local decision makers to build decision support tools suited to their particular circumstances and needs.
 

### 2024-04-12 analysis/paper/supplement.Rmd
Writing ongoing

@@
 |**True demographic parameters**|
-|$N_0=2000$|animals|Initial adult population size.|
+|$N_0=5000$|animals|Initial adult population size.|
 |$\bar{R}_t$|rate|Expected recruitment (calves per cow).|
@@
 |$\alpha=\text{viable or not}$|binary|Assigned population status.|
-|$\nu_{\alpha,\psi}=1 \text{when} \alpha=\psi, 0 \text{otherwise}$|binary|Value of classification decision.|
+|$\nu_{\alpha,\psi}=1 \text{ when } \alpha=\psi, 0 \text{ otherwise}$|binary|Value of classification decision.|
 |$x$||Simulated observations from an example population.|
@@
 
-TO DO: get photo of calf:cow survey group. Label classes of animals to illustrate meaning of T, U, V, W, O, P, J, K.
-
-We assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$).
+Assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$).
 The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is a multiplier that defines the apparent number of adult females as a function of the number of collared animals.
-The ratio of young bulls to uncollared adult females in the group is:
-$$q = \frac{P+O}{U+V}.$$
-Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$. Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$.
-
 Our objective is to model the sex and bias-corrected recruitment rate $X=\frac{J+K}{2(T+U+V)}$ as a function of the observed calf:cow ratio $R=J/(T+V+P)$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$.

### 2024-04-12 analysis/scripts/bayesianModelExamples.R

@@ -7,7 +7,7 @@ library(RColorBrewer)
 
 baseDir <- "."
 
-monitoringScns = data.frame(obsYears=c(1,15),collarCount=c(0,30),cowMult=c(6),collarInterval=c(1),assessmentYrs=c(3))
+monitoringScns = data.frame(obsYears=c(1,16),collarCount=c(0,30),cowMult=c(6),collarInterval=c(1),assessmentYrs=c(3))
 stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
 stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
 

### 2024-04-12 analysis/scripts/nationalModelTrajectories.R

@@ -150,7 +150,7 @@ plot_recruitment3 <- ggplot(data = rateSummaries,
             aes(x = Anthro, y = 0.5*R_t, group = fullGrp, color = fullGrp),
             alpha = 1) +scale_color_manual(values=pal)+
   geom_line(data=johnsonCompare,colour = "black", size = 1.5) +
-  geom_line(colour = "white", size = 2, linetype = "dotted") +
+  geom_line(colour = "#ef8a62", size = 2, linetype = "dotted") +
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
   scale_y_continuous(limits = c(0, 0.4), breaks = c(0, 0.1, 0.2, 0.3, 0.4)) +
   xlab("Anthropogenic disturbance (%)") +
@@ -165,7 +165,7 @@ base1 <- ggplot(data = rateSummaries,
             aes(x = Anthro, y = S_t, group = rep, colour = rep)) +
   scale_color_manual(values=pal)+
   geom_line(data=johnsonCompare, colour = "black", size = 1.5) +
-  geom_line(colour = "white", size = 2, linetype = "dotted") +
+  geom_line(colour = "#ef8a62", size = 2, linetype = "dotted") +
   xlab("Anthropogenic Disturbance (%)") +
   ylab(expression("Adult Female Survival " * dot(S)[t])) +
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
@@ -234,7 +234,7 @@ plot_recruitment4 <- ggplot(data = rateSummaries,
             aes(x = Anthro, y = 0.5*R_t, group = fullGrp, color = fullGrp),
             alpha = 1) +scale_color_manual(values=pal)+
   geom_line(data=johnsonCompare,aes(y=R_bar/2),colour = "black", size = 1.5) +
-  geom_line(colour = "white", size = 2, linetype = "dotted") +
+  geom_line(colour = "#ef8a62", size = 2, linetype = "dotted") +
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
   scale_y_continuous(limits = c(0, 0.4), breaks = c(0, 0.1, 0.2, 0.3, 0.4)) +
   xlab("Anthropogenic disturbance (%)") +

### 2024-05-24 analysis/paper/paper.Rmd
Draft to coauthors.

@@
 ---
-title: "National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions"
+title: "Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions"
 author:
@@
 abstract: |
+  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed.
+There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but where disturbance is high we have no evidence of self-sustaining populations across the country.
+Demographic projections derived from local population-specific data are  uncertain for populations with limited monitoring.
+We propose a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and reduce the risk that lack of local data will be used as a reason to delay conservation action.
+The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process.
+We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.
+Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships.
+At lower disturbance levels where uncertainty is higher, deploying fewer collars for more years may be better than intensive short-term monitoring for improving the reliability of status projections.
 keywords: |
@@
 To be useful, these methods must be tailored to particular species and decision contexts.
-Here, we offer a Bayesian population model for boreal caribou (_Rangifer tarandus caribou_) that integrates information about the impacts of landscape change with available demographic data to reduce uncertainty in population viability projections for boreal caribou [building on @eacker_web-based_2019].
+Here, we offer a simple Bayesian population model for boreal caribou (_Rangifer tarandus caribou_) that integrates information about the impacts of landscape change with available demographic data to reduce uncertainty in population viability projections for boreal caribou [building on @eacker_web-based_2019].
 Following the value of information analysis framework proposed by Dunham et al.
 (2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.
-To help make the approach more accessible, we offer an open source app that allows users to analyze existing local demographic data, or to explore alternative monitoring scenarios with simulated data.
 Our approach is tailored to boreal caribou, but could be adapted for other species at risk found in multiple populations that are costly to monitor and sensitive to habitat disturbance.
@@
 Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020].
-The causes of this relationship include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011].
+The causes of these demographic-disturbance relationships include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011].
 Decreases in recruitment as disturbance increases are also observed over time in places where there is sufficient historic demographic data [@rudolph_demographic_2017;@hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships  [@johnson_science_2020].
 There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020;@neufeld_primary_2021; @fortin_forest_2017; @gagne_impact_2016], and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships [@stewart_climate-informed_2023;@dyson_existing_2022].
+
+Interventions such as wolf culls, maternal penning, and restoration of linear features are intended to improve survival and reproduction of populations that are not otherwise self-sustaining [@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a need to assess and compare effects of boreal caribou conservation actions with careful monitoring.
+Our focus in this paper is assessing and projecting self-sustainability of populations in the absence of these interventions, to clarify when additional information is needed to reduce uncertainty about whether there is a need for action.
+Where actions are taken, outcomes should be monitored; the simple model we propose here does not include effects of management interventions, and cannot be used to project outcomes for managed populations.
 
@@
 These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021].
-However Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships or composition survey errors.
+However, there is very little monitoring data available for some caribou populations [@johnson_science_2020] and there is a risk that lack of local data will be used as a reason to delay conservation action [@boan_climate_2018].
+Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships, and we are concerned that models informed only by local demographic data could be used to inflate uncertainty about the status of poorly monitored populations.
 
-Here, we present a Bayesian population model for boreal caribou [building on @eacker_web-based_2019] designed to reduce uncertainty in projections of the impacts of changing landscape conditions by integrating prior information from national analysis of demographic-disturbance relationships with available local demographic data.
-Information is shared among populations using priors instead of integrating multiple populations into a hierarchical analysis [@leasure_hierarchical_2019] because obtaining permission to use caribou demographic data can be difficult, and most managers only have access to their own data.
-This Bayesian approach represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships.
+We propose to reduce uncertainty in assessments and projections of local caribou population viability by integrating prior information from national analysis of demographic-disturbance relationships with available local demographic data in a simple Bayesian population model for boreal caribou [building on @eacker_web-based_2019].
+In our simple proof-of-concept model, information is shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020].
+A more rigorous statistical approach would be to integrate multiple populations into a hierarchical Bayesian analysis of variation within and among populations [@leasure_hierarchical_2019]; we have not done this because owners have not authorized the use of the data for this purpose, obtaining permission to use caribou demographic data is difficult, and most managers only have access to their own local data.
+Despite limitations, our simple Bayesian model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships.
 We show that integrating the most informative predictor of outcomes for boreal caribou across their range can substantially reduce uncertainty in local demographic projections and status assessments, improve viability projections on changing landscapes, and help clarify where, why and when additional data is required to inform decisions and project outcomes on changing landscapes.
-In cases where available information is sufficient for assessing and projecting population viability, there is a risk of misallocating limited resources to monitoring that could be better spent on conservation action.
+We hope that this approach can help clarify where available information is sufficient for assessing and projecting population viability, and thereby reduce the risk that uncertainty is used as a reason to delay conservation action for imperiled populations.
 
@@
 More specifically, models that do not account for observation errors can be misleading [@schaub_integrated_2021].
-Several authors have pointed out that  misclassification of adults and missing calves in composition surveys can bias recruitment estimates from those surveys [@ellington_divergent_2020; @decesare_estimating_2012], but none have thus far examined sensitivity of Bayesian caribou population models to these sources of error [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019].
-We address this gap with a new bias correction term derived from a simple model of the recruitment survey observation process.
+Although several authors have pointed out that  misclassification of adults and missing calves in composition surveys can bias recruitment estimates from those surveys [@ellington_divergent_2020; @decesare_estimating_2012] none have thus far examined sensitivity of Bayesian caribou population models to these sources of error [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019].
+We address this gap with a bias correction term derived from a simple model of the recruitment survey observation process.
 
@@
 we used a modified version of Johnson et al's [-@johnson_science_2020]
-two-stage demographic model described by Dyson et al.
+demographic model described by Dyson et al.
 [-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
@@
 
-## Integration of local demographic data and national disturbance-demographic relationships in a Bayesian population model
+## Bayesian integration of local demographic data and national disturbance-demographic relationships
 
@@
 precisions for years with no observed mortality estimated using a
-separate Bayesian model. Observed estimated survival is a Gaussian
-distributed function of the true survival probability:
+separate Bayesian model. Observed estimated survival is a modelled as a Gaussian
+distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019]:
 \begin{equation}
-\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).
+\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
   (\#eq:hatSEst)
@@
 
-As in the simulations (eq \@ref(eq:barS)), we model survival probability as an adjusted function of buffered anthropogenic disturbance $A_t$ with a log link
-[@johnson_science_2020; @stewart_climate-informed_2023; @dyson_existing_2022].
-We add a Gaussian distributed random year effect to account for stochastic variation among years [@eacker_web-based_2019]:
+For consistency with [@johnson_science_2020], we model survival probability as an adjusted function of buffered anthropogenic disturbance $A_t$ with a log link
+[@stewart_climate-informed_2023; @dyson_existing_2022]. We add a Gaussian distributed random year effect to account for stochastic variation among years [@eacker_web-based_2019]:
 \begin{equation}
@@
 knitr::kable(tbl,caption="Prior means and standard errors of Gaussian distributed survival and reproduction model parameters.
-Standard errors are calculated by dividing Johnson et al's (2020) reported confidence intervals by 3.92.
-Multipliers on the standard errors (*) are calibrated so that the 95\\% prior prediction intervals for survival and recruitment from the Bayesian model match the range between the 2.5\\% and 97.5\\% quantiles of 1000 simulated survival
+Standard errors are calibrated so that the 95\\% prior prediction intervals for survival and recruitment from the Bayesian model match the range between the 2.5\\% and 97.5\\% quantiles of 1000 simulated survival
 and recruitment trajectories (Supplement C).")
@@
 
-As in the simulations, we assume a simple two-class demographic model, with a census in late
+As in the simulations, we assume a simple demographic model, with a census in late
 winter of surviving females and new recruits to the post-juvenile (aged
@@
 -->
+
+Models and analysis were implemented using the R language [@r_core_team_r_2023] and a variety of _tidyverse_ packages [@wickham_welcome_2019].
 We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the _R2jags_ R package [@su_r2jags_2021] to fit the Bayesian model.
@@
 We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the _mcmcplots_ R package [@curtis_mcmcplots_2018].
+Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>, and in the  _caribouMetrics_ R package available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.
 
@@
 To aid in interpretation of results and to highlight the advantages of selecting a metric that is directly relevant to the management objective of stable or increasing populations, we also show how integrating local monitoring data alters the distribution of differences between true population growth rates and Bayesian posterior mean estimated growth rates.
-
-## User interface and open, flexible, reuseable implementation
-
-Analysis was implemented using the R language [@r_core_team_r_2023] and a variety of _tidyverse} packages [@wickham_welcome_2019].
-To make the work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios, borrowing and building on code from an app developed by Eacker et al.
-[-@eacker_web-based_2019] where appropriate.
-Our app allows users to provide existing local data, or to make projections informed by simulated data as we have done in this analysis (similar to Figure
-\@ref(fig:examplePredictions)).
-The app allows advanced users to adjust many model parameters (supplemental Table A1).
-Default parameter values are set to allow novice users to get reasonable results quickly.
-Most of the code needed to run the app is in an R package, _caribouMetrics_, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g.
-@dyson_existing_2022; @stewart_climate-informed_2023].
-The _caribouMetrics_ package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.
-Find documentation and vignettes on the package website
-<https://landscitech.github.io/caribouMetrics/>.
-The Shiny user interface app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>.
-Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.
 
@@
 
+## Bayesian integration of local demographic data & national demographic-disturbance relationships
+
+The availability of boreal caribou demographic data varies widely across Canada due to large variation in habitat condition, population condition, available resources for monitoring, and remoteness.
+In a few regions enough data has been collected over time to characterize local demographic-disturbance relationships [@rudolph_demographic_2017;@hervieux_widespread_2013], but most populations have only been sparsely monitored for snapshot estimates of trend or abundance [@johnson_science_2020].
+As a result, in most regions it is necessary to make assumptions informed by observed variation in demographic rates with disturbance among populations and other relevant information about impacts of disturbance on caribou.
+However, high observed variation in demography where anthropogenic disturbance is low leads to high uncertainty in projections informed by national demographic-disturbance relationships alone [@stewart_climate-informed_2023;@dyson_existing_2022].
+In some cases, predictions of population growth rate informed by national demographic-disturbance relationships appear to disagree with estimates from local demographic data [@eccc_scientific_2011; @dyson_existing_2022], but without formal methods of accounting for uncertainty it can be difficult to distinguish true discrepancies from stochastic variation.
+
+We developed a simple Bayesian population model [building on @eacker_web-based_2019] to help address these challenges by acknowledging uncertainty associated with local demographic data and national demographic-disturbance relationships, and integrating these two sources of information to reduce uncertainty in projections of the impacts of changing landscape conditions.
+This Bayesian approach represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships.
+
+Our model assumes that increasing cumulative anthropogenic disturbance causes a log-linear decrease in caribou demographic rates (equations 4, 5, 10 and 13), reflecting a scientific consensus informed by a number of lines of evidence that cumulative anthropogenic disturbance has a negative impact on boreal caribou [@eccc_scientific_2011; @johnson_science_2020; @gouvernement_du_quebec_revue_2021].
+Local data can be integrated to estimate local demographic-disturbance parameters that differ from the national means, and in the absence of local data, appropriate priors lead to a range of outcomes consistent with the observed range of variation across the country [Figures 3 and 5 in @johnson_science_2020].
+Others [@moeller_integrating_2021; @mcnay_demographic_2022; @eacker_web-based_2019] have also used Bayesian models to estimate caribou demographic rates from local monitoring data, but our addition of prior information from national demographic-disturbance relationships is novel, and important for reducing the risk that lack of local data will be used as a reason to delay conservation action for highly disturbed and poorly monitored populations.
+Integrating demographic-disturbance relationships into a viability projection model clarifies that no additional information is needed to project outcomes where disturbance is high and actions to support recovery are not taken.
+
+There are a variety of possible actions intended to improve survival and reproduction of populations that are not otherwise self-sustaining [@mcnay_demographic_2022; @serrouya_predicting_2020].
+The simple models in this paper do not include these actions, and cannot be used to project outcomes for managed populations.
+Our intention is to reduce uncertainty that can be used to delay conservation action; there is nothing in our arguments, models or results that can or should be used to discourage much needed monitoring of the effects of conservation actions.
+Even when no additional additional monitoring is required for assessing self-sustainability, there are a variety of reasons to continue monitoring highly disturbed boreal caribou populations.
+Our point is that action is also required, and lack of monitoring should not be used as a reason to delay conservation measures for highly disturbed populations.
+
+Variation in anthropogenic disturbance across the Canadian boreal region is partially correlated with variation in climate and primary productivity because forestry occurs in higher productivity areas, and petroleum extraction activities are concentrated in Western Canada; concerns have been raised about confounding of demographic-disturbance relationships [@wilson_causal_2021;@neufeld_primary_2021; @neilson_direct_2022].
+One option for addressing these concerns would be to develop regional demographic-disturbance relationships for populations that are ecologically similar to one another (e.g.
+caribou inhabiting forests that are sufficiently productive for industrial forestry in Eastern Canada) [@eccc_scientific_2024], and use the priors from regional demographic-disturbance relationships to inform demographic projections within the region.
+However, while acknowledging and allowing for regional variation, we also need to guard against risks that uncertainty will be inflated and used as an excuse to delay action, a tactic that has been used in this and other decision contexts [@bissonnette_controverses_2020; @boan_climate_2018].
+In order to be informative and credible, regional demographic-disturbance analyses must include a sufficient number of observations spanning a range from high to low anthropogenic disturbance.
+Priors that lead to projections of outcomes for caribou that fall outside the observed range of variation across the country, characterized here (Figure 2) by 95% predictive intervals from a national analysis (Johnson et al.
+2020), are implausible.
+In other words, if we have seen no examples of self-sustaining caribou populations at a particularly high level of disturbance, it is unreasonable to project or expect this outcome.
+These premises constrain the plausible demographic-disturbance priors, and clarify that no additional information is needed to predict outcomes for caribou where anthropogenic disturbance is high (Figure 5) and no meaningful actions are taken to support recovery.
+
+## Allowing for composition survey errors
+
+Another novel aspect of this work is a term that allows for bias associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012].
+Lacking information about the true magnitude of this bias, we examined sensitivity to a range of assumed values, and found that adding unknown bias did increase uncertainty, but these increases were not large (supplement E).
+A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults and the probability of missing calves.
+Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment.
+Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out.
+We did not investigate sensitivity to the assumption that new recruits are 50% females, but we note that this is another possible source of error that would most likely lead to low recruitment estimates.
+
+Given lack of knowledge about true error rates, the most appropriate interpretation of our results is that better information on composition survey error rates is required in order to assess their importance.
+Fecal pellets are proving to be an interesting source of information on sex and age ratios in surveyed groups [@mcfarlane_genetic_2018;@jones_novel_2023]; for example, in analysis of 14 populations of boreal and southern mountain caribou, Jones et al.
+2023 found that sampled groups contained more females than males, and that mothers tend to stay with their daughters and sons, while fathers do not.
+These results are broadly consistent with our assumption that composition survey groups include mothers and younger animals of both sexes that can be difficult to classify visually.
+Documenting animals that are difficult to classify as “unknown” can help analysts minimize and account for classification bias [@ellington_divergent_2020], as can comparing classification results from multiple independent observers.
+The probability of missing calves is more difficult to quantify and account for using aerial survey methods alone, but likely increases with vegetation cover.
+
+Alternatives to aerial composition surveys for estimating recruitment include fecal pellet analysis [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022] and calf collaring [@ellington_divergent_2020].
+Each method has costs and benefits, and the reality is that monitoring everything everywhere all the time in great detail is neither feasible nor necessary.
+We hope that a bias correction term for the simple recruitment-mortality model [@hatter_moose_1991;@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020] that is used for many herd viability assessments will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is needed.
+
+## Limitations and Future Possibilities
+
+The main objective of Johnson et al's [-@johnson_science_2020;@eccc_scientific_2011] national analysis was to inform a national disturbance policy that applies to all ranges except SK1, and to inform a different policy for SK1 [@eccc_scientific_2011;@eccc_boreal_2019].
+A hierarchical analysis [@leasure_hierarchical_2019] would be better suited for estimating variation in demographic model parameters among populations, and distinguishing that from variation among years.
+However, data owners have not authorized the use of the data for that purpose.
+In order to proceed with the available published information we constrained our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower than higher than average remain lower or higher than average over time.
+The Bayesian model is not similarly constrained, and can accommodate evidence of a shift from unexpectedly low to unexpectedly low demographic rates over time, or vice versa.
+Switching from low to high each year is unlikely, but there is evidence of variation in the slope demographic-disturbance relationships among populations [@rudolph_demographic_2017] so it is important that the Bayesian model allows for a greater range of possibilities than our more constrained simulated example trajectories.
+Although our simulated trajectories do not include all plausible possibilities we believe they are sufficient to demonstrate how a Bayesian model can be used to integrate local data with national priors, and to show how monitoring requirements depend on landscape condition.
+
+Aspatial demographic models generally assume that observed animals are a random and representative sample of a fully mixed population in which survival and recruitment rates apply equally to all animals.
+However, metrics derived from fecal pellet analysis such as reproductive success [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022], age structure and pregnancy rates [@flasko_fecal_2017;@mcfarlane_identification_2022], and mother-offspring co-occurence [@jones_novel_2023] suggest high variation in fitness among individuals [@mcfarlane_genetic_2018] and spatial structure within populations [@jones_novel_2023;@mcfarlane_spatial_2021;@mcfarlane_identification_2022].
+Analyses of movement data also show spatial variation in risks and benefits for caribou and variation in habitat preferences and space use among individuals [e.g.
+@decesare_linking_2014;@avgar_spaceuse_2015].
+Although there is evidence of spatial structure and variation in demographic rates among individuals within populations, for most caribou populations there is not enough data available to estimate spatial variation in demographic model parameters.
+
+A number of additional complexities can be incorporated into aspatial Bayesian population models, including variation in demographic rates among individuals  [@armstrong_incorporating_2021], population size information from aerial or fecal dna surveys, and additional age classes [@moeller_integrating_2021; @mcnay_demographic_2022].
+However, in many contexts, population size and fecal dna survey information is not available, and a framework that includes these sources of information will be unnecessarily complicated.
+At present, our model is intended as a proof-of-concept to show the value of integrating national demographic-disturbance relationships with local data, accounting for composition survey errors, and assessing the value of local monitoring information.
+Rather than further complicate a proof-of-concept model applied to simulated data, our next step will be to collaborate with an interested decision maker to develop a tool informed by this analysis that meets their real local needs, recognizing that these needs will vary among monitoring programs [@nbckc_nbckc_2021;@nbckc_nbckc_2021].
+
+## Assessing the value of additional monitoring
+
+We applied the framework proposed by Dunham et al.
+(2023) to show that no additional local monitoring information is needed to predict whether boreal caribou populations in highly disturbed landscapes are self-sustaining.
+There is more variation in caribou demographic rates among study areas with low to moderate disturbance, and thus more need for local data to inform demographic projections.
+In these places, our analysis suggests a general pattern of diminishing returns from increasing monitoring effort, but more information on monitoring costs, objectives, and existing local data would be required to assess the adequacy and cost-effectiveness of monitoring scenarios in particular circumstances.
+Monitoring costs vary substantially with technologies used, the remoteness of herds, and the difficulty of collaring animals, which in turn depends on the number of currently collared animals, the population size, and vegetation across the landscape [@nbckc_boreal_2021].
+It is important to recognize that projecting population viability is only one possible objective for monitoring [@nbckc_boreal_2019], and that the impacts of monitoring on threatened animals should be considered [@nbckc_boreal_2019].
+Since monitoring costs, options and goals are context specific, we believe the most effective way forward will be to use this proof-of-concept model and analysis as a starting point for working with local decision makers to build decision support tools suited to their particular circumstances, needs and questions.
+
 # Acknowledgements

### 2024-05-24 analysis/paper/supplement.Rmd
Draft to coauthors.

@@
 
-# (APPENDIX) Supplementary Material for National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions {-}
+# (APPENDIX) Supplementary Material for Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions {-}
 
@@
 
+Prior parameter values should be set so that, given no monitoring data, the prior survival distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance.
+This condition is met when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (sSigmaMean & sSigmaSD, $\sigma^2_S$ in Table 2) (Figs \@ref(fig:PriorKSDists) and \@ref(fig:PriorAllFits)).
+Priors should also be vague enough so that, given intensive monitoring of populations where survival is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 60 collars per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local survival, rather than priors from the national model.
+Setting the mean and standard deviations of the prior distribution of the random effect of year on survival to moderately low values (sSigmaMean=0.034784, sSigmaSD=0.03, $\sigma^2_S$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDists), \@ref(fig:PriorAllFits) and \@ref(fig:PriorLowFits)).
+
+Similarly, prior parameter values should be set so that, given no monitoring data, the prior recruitment distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance.
+This condition is met when the multiplier on the prior standard deviation of the recruitment intercept (rIntSEMod: $\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (rSigmaMean & rSigmaSD, $\sigma^2_R$ in Table 2) (Figs \@ref(fig:PriorKSDistsRec) and \@ref(fig:PriorAllFitsRec)).
+Priors should also be vague enough so that, given intensive monitoring of populations where recruitment is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 9*60 cows per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local recruitment, rather than priors from the national model.
+Setting the mean and standard deviations of the prior distribution of the random effect of year on recruitment to moderately low values (rSigmaMean=0.23, rSigmaSD=0.22, $\sigma^2_R$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDistsRec), \@ref(fig:PriorAllFitsRec) and \@ref(fig:PriorHighFitsRec)).
+
 ```{r priorCalibration, cache=TRUE,echo=FALSE, message=FALSE,results='hide'}
@@
 
-Prior parameter values should be set so that, given no monitoring data, the prior survival distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance.
-This condition is met when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (sSigmaMean & sSigmaSD, $\sigma^2_S$ in Table 2) (Figs \@ref(fig:PriorKSDists) and \@ref(fig:PriorAllFits)).
-Priors should also be vague enough so that, given intensive monitoring of populations where survival is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 60 collars per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local survival, rather than priors from the national model.
-Setting the mean and standard deviations of the prior distribution of the random effect of year on survival to moderately low values (sSigmaMean=0.034784, sSigmaSD=0.03, $\sigma^2_S$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDists), \@ref(fig:PriorAllFits) and \@ref(fig:PriorLowFits)).
-
 ```{r PriorKSDistsRec, fig.cap="Similarity of the prior recruitment distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022].
@@
 ```
-
-Similarly, prior parameter values should be set so that, given no monitoring data, the prior recruitment distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance.
-This condition is met when the multiplier on the prior standard deviation of the recruitment intercept (rIntSEMod: $\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (rSigmaMean & rSigmaSD, $\sigma^2_R$ in Table 2) (Figs \@ref(fig:PriorKSDistsRec) and \@ref(fig:PriorAllFitsRec)).
-Priors should also be vague enough so that, given intensive monitoring of populations where recruitment is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 9*60 cows per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local recruitment, rather than priors from the national model.
-Setting the mean and standard deviations of the prior distribution of the random effect of year on recruitment to moderately low values (rSigmaMean=0.23, rSigmaSD=0.22, $\sigma^2_R$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDistsRec), \@ref(fig:PriorAllFitsRec) and \@ref(fig:PriorHighFitsRec)).
 

### 2024-05-31 analysis/paper/paper.Rmd
Adjusting priors

@@
 All regression coefficients are assumed to be Gaussian distributed, with
-priors derived from the expected values and 95% confidence intervals
-estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We
-calibrated the prior distributions so that the 95% prior prediction
+prior means from Johnson et al's (2020) regression model estimates (Table \@ref(tab:priors)). We
+calibrated the standard deviations of the priors so that the 95% prior predictive
 intervals for survival and recruitment from the Bayesian model match the
-range between the 2.5% and 97.5% quantiles of 1000 simulated survival
-and recruitment trajectories (Supplement C, Figure
+or exceed the 95% predictive intervals from Johnson et al's regression model (Supplement C, Figure
 \@ref(fig:examplePredictions)).
@@
 
-knitr::kable(tbl,caption="Prior means and standard errors of Gaussian distributed survival and reproduction model parameters.
-Standard errors are calibrated so that the 95\\% prior prediction intervals for survival and recruitment from the Bayesian model match the range between the 2.5\\% and 97.5\\% quantiles of 1000 simulated survival
-and recruitment trajectories (Supplement C).")
-#TO DO: indicate which of these are from Johnson, and which are calibrated.
+knitr::kable(tbl,caption="Prior means and standard deviations of Gaussian distributed survival and reproduction model parameters.
+Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the
+or exceed the 95% predictive intervals from Johnson et al's regression model (Supplement C).")
 ```

### 2024-05-31 analysis/paper/supplement.Rmd
Adjusting priors

@@
 
-Prior parameter values should be set so that, given no monitoring data, the prior survival distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance.
-This condition is met when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (sSigmaMean & sSigmaSD, $\sigma^2_S$ in Table 2) (Figs \@ref(fig:PriorKSDists) and \@ref(fig:PriorAllFits)).
-Priors should also be vague enough so that, given intensive monitoring of populations where survival is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 60 collars per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local survival, rather than priors from the national model.
-Setting the mean and standard deviations of the prior distribution of the random effect of year on survival to moderately low values (sSigmaMean=0.034784, sSigmaSD=0.03, $\sigma^2_S$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDists), \@ref(fig:PriorAllFits) and \@ref(fig:PriorLowFits)).
+We begin by setting random effects to zero, varying the prior standard deviations of the survival intercept (sIntSE is $\beta^S_0$ in Table 2) and recruitment intercept (rIntSE is $\beta^R_0$ in Table 2), and comparing the prior predictive intervals of the Bayesian model to the 95\% predictive intervals from Johnson et al.'s (2020) regression analysis in a case with no anthropogenic disturbance (top row of Fig \@ref(fig:InterceptsAndSlopes)).
+We set the prior standard deviation intercept parameters (sIntSE = 0.04, rIntSE = 0.25) so that the upper bounds of the Bayesian predictive intervals match the upper bounds of Johnson et al.'s predictive intervals.
 
-Similarly, prior parameter values should be set so that, given no monitoring data, the prior recruitment distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance.
-This condition is met when the multiplier on the prior standard deviation of the recruitment intercept (rIntSEMod: $\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (rSigmaMean & rSigmaSD, $\sigma^2_R$ in Table 2) (Figs \@ref(fig:PriorKSDistsRec) and \@ref(fig:PriorAllFitsRec)).
-Priors should also be vague enough so that, given intensive monitoring of populations where recruitment is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 9*60 cows per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local recruitment, rather than priors from the national model.
-Setting the mean and standard deviations of the prior distribution of the random effect of year on recruitment to moderately low values (rSigmaMean=0.23, rSigmaSD=0.22, $\sigma^2_R$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDistsRec), \@ref(fig:PriorAllFitsRec) and \@ref(fig:PriorHighFitsRec)).
+We then vary the prior standard deviations of the slopes of the relationships between anthropogenic disturbance and demographic rates (sAnthroSlopeSE is $\beta^S_R$ and rAnthroSlopeSE is $\beta^R_R$ in Table 2), and compare the prior predictive intervals of the Bayesian model to the 95\% predictive intervals from Johnson et al.'s (2020) regression analysis in a case with 90\% anthropogenic disturbance (bottom row of Fig \@ref(fig:InterceptsAndSlopes)).
+We set the prior standard deviation of the slopes (sAnthroSlopeSE = 0.0005, rAnthroSlopeSE = 0.006) so that the upper bound of the Bayesian predictive interval does not substantially exceed the upper bound of Johnson et al.'s predictive interval.
+
+The choice of priors for the random effects of year on survival and recruitment are less important because the random effects have no effect on expected population growth rate (Figs \@ref(fig:PriorAllFits) and \@ref(fig:PriorAllFitsR)).
+We set the hyperprior for the standard deviation of the survival random effect to vary uniformly between sSigmaMin = 0.00001 and sSigmaMax = 0.06 to exceed the variation among national model simulations by a moderate amount (Fig \@ref(fig:PriorAllFits)).
+We set the hyperprior for the standard deviation of the recruitment random effect to vary uniformly between rSigmaMin = 0.00001 and rSigmaMax = 0.2 to exceed the variation among national model simulations by a moderate amount (Fig \@ref(fig:PriorAllFitsR)).
+
+Given intensive monitoring (15 years of 60 collars per year) of populations where demographic rates are higher or lower than the national average (from 97.5% and 2.5% quantiles of the national model), and the slope of the anthropogenic disturbance relationship differs from the national model, these priors are vague enough to allow posteriors to reflect local demographic rates (Figs \@ref(fig:SurvBoth) and \@ref(fig:RecBoth)).
 
@@
 if(doPriorCalibration){
-
+  source(here::here("analysis/scripts/PriorsFromPIs.R"))
   source(here::here("analysis/scripts/calibratePriors.R"))
-}else{
-  KS<-read.csv(here::here("tabs/SurvKSAll.csv"))
-  KSRec<-read.csv(here::here("tabs/RecKSAll.csv"))
-
 }
-  #library(ggplot2);fig_widths=data.frame(two=7)
-  KS=subset(KS,scn=="All")
-  png(here::here("figs/PriorKSDists.png"),
-    height = 3, width = fig_widths["two"], units = "in",res=600)
-    base = ggplot(KS,aes(x=sIntSEMod,y=KSDistance,color=sSigmaMean))+
-    geom_point()+facet_wrap(~sSigmaSD,labeller="label_both")+theme_bw()+ylab("Kolmogorov-Smirnov Distance")+xlab("sIntSEMod: multiplier on SD of survival intercept")
-    print(base)
-  dev.off()
-
-  KSRec=subset(KSRec,scn=="All")
-  png(here::here("figs/PriorKSDistsRec.png"),
-    height = 3, width = fig_widths["two"], units = "in",res=600)
-    base = ggplot(KSRec,aes(x=rIntSEMod,y=KSDistance,color=rSigmaMean))+
-    geom_point()+facet_wrap(~rSigmaSD,labeller="label_both")+theme_bw()+ylab("Kolmogorov-Smirnov Distance")+xlab("rIntSEMod: multiplier on SD of recruitment intercept")
-    print(base)
-  dev.off()
 ```
 
-```{r PriorKSDists, fig.cap="Similarity of the prior survival distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022].
-The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (sSigmaMean & sSigmaSD, $\\sigma^2_S$ in Table 2).
-Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
-knitr::include_graphics(here::here("figs/PriorKSDists.png"))
-
+```{r InterceptsAndSlopes, fig.cap="Effect of variation in survival and recruitment prior standard deviations on the prior 95% predictive intervals and means of the Bayesian model (orange bands and lines) with no random effects (no interannual variation).
+Means and 95% predictive intervals from Johnson et al.'s 2020 regression model (black lines) summarize observed variation across the Canada." , cache=TRUE}
+knitr::include_graphics(here::here("figs/InterceptsAndSlopes.png"))
 ```
 
-```{r PriorAllFits, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring.
-In these examples, the standard deviation of the random effect prior (sSigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and anthropogenic disturbance is 0.
+```{r PriorAllFits, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('Bayesian') and simulated outcomes from the national model ('national'), assuming no local monitoring.
+In these examples anthropogenic disturbance is 0.
 Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}
 
-knitr::include_graphics(here::here("figs/SurvInterceptAllsSigmaSD0.03.png"))
+knitr::include_graphics(here::here("figs/SurvInterceptAllsIntSE0.04.png"))
 ```
 
-```{r PriorLowFits, fig.cap="Posterior predictive survival distributions from the Bayesian model ('local') assuming intensive monitoring (15 years of 60 collars per year) of populations from the 2.5 \\% quantile of the national model.
-In these examples, the standard deviation of the random effect prior (sSigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and  anthropogenic disturbance is 0.
-Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates.
-When the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is too low (<5) the insufficiently vague priors substantially influence the posterior distributions, yielding poor local parameter estimates.", cache=TRUE}
+```{r PriorAllFitsR, fig.cap="Differences between prior predictive recruitment distributions from the Bayesian model ('Bayesian') and simulated outcomes from the national model ('national'), assuming no local monitoring.
+In these examples anthropogenic disturbance is 0.
+Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}
 
-knitr::include_graphics(here::here("figs/SurvInterceptLowsSigmaSD0.03.png"))
+knitr::include_graphics(here::here("figs/RecInterceptAllrIntSE0.25.png"))
 ```
 
-```{r PriorAllFitsAnthro90, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring.
-In these examples, anthropogenic disturbance is 90\\%.
-When the multiplier on the prior standard deviation of the survival slope (sSlopeSEMod: $\\beta^S_a$ in Table 2) is 3, the posterior survival distribution is similar to the distribution of simulated outcomes from the national model.
-Bands show 95\\% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.", cache=TRUE}
-knitr::include_graphics(here::here("figs/SurvSlopeAll.png"))
+```{r SurvBoth, fig.cap="Posterior predicted survival from the Bayesian model fitted to 15 years of data, with 60 collars per year.
+Survival in these example populations is lower (sQuantile=0.025) or higher (sQuantile=0.975) than the national average, with no anthropogenic disturbance slope (sSlopeMod = 0), or an anthropogenic disturbance slope that is twice as steep as expected (sSlopeMod = 2).
+In these examples anthropogenic disturbance is 90%.
+Bands show 95% posterior predictive intervals, dots show true survival, and triangles show simulated observations.",  cache=TRUE}
+
+knitr::include_graphics(here::here("figs/SurvBoth.png"))
 ```
 
-```{r PriorHighFitsAnthro90, fig.cap="Posterior predictive survival distributions from the Bayesian model, assuming intensive monitoring (15 years of 60 collars per year) of populations from the 97.5 \\% quantile of the national model.
-In these examples, anthropogenic disturbance is 90\\%, and the effect of anthropogenic disturbance varies from none (sSlopeMod=0) through expected (sSlopeMod=1) to double the expected slope (sSlopeMod=2).
-All considered multipliers on the prior standard deviation of the survival slope (sSlopeSEMod: $\\beta^S_a$ in Table 2) are sufficiently vague to reflect local variation among populations.
-Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates.", cache=TRUE}
-knitr::include_graphics(here::here("figs/SurvSlopeHigh.png"))
-```
+```{r RecBoth, fig.cap="Posterior predicted recruitment from the Bayesian model fitted to 15 years of data, with 60 collars per year.
+Recruitment in these example populations is lower (rQuantile=0.025) or higher (rQuantile=0.975) than the national average, with no anthropogenic disturbance slope (rSlopeMod = 0), or an anthropogenic disturbance slope that is twice as steep as expected (rSlopeMod = 2).
+In these examples anthropogenic disturbance is 90%.
+Bands show 95% posterior predictive intervals, dots show recruitment, and triangles show simulated observations.",  cache=TRUE}
 
-```{r PriorKSDistsRec, fig.cap="Similarity of the prior recruitment distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022].
-The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (rIntSEMod: $\\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (rSigmaMean & rSigmaSD, $\\sigma^2_R$ in Table 2).
-Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
-knitr::include_graphics(here::here("figs/PriorKSDistsRec.png"))
-```
-
-```{r PriorAllFitsRec, fig.cap="Differences between prior predictive recruitment distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring.
-In these examples, the standard deviation of the random effect prior (rSigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and anthropogenic disturbance is 0.
-Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}
-knitr::include_graphics(here::here("figs/RecInterceptAllrSigmaSD0.22.png"))
-```
-
-```{r PriorHighFitsRec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model ('local') assuming intensive monitoring (9*60 cows per year for 15 years) of populations from the 97.5% quantile of the national model.
-In these examples, the standard deviation the random effect prior (rSigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and  anthropogenic disturbance is 0.
-Bands show 95% posterior predictive intervals, black dots show true recruitment rates from the simulated population, and open triangles show observed recruitment rates.", cache=TRUE}
-knitr::include_graphics(here::here("figs/RecInterceptHighrSigmaSD0.22.png"))
-```
-
-```{r PriorAllFitsAnthro90Rec, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring.
-In these examples, anthropogenic disturbance is 90\\%.
-When the multiplier on the prior standard deviation of the recruitment slope (rSlopeSEMod: $\\beta^R_a$ in Table 2) is 4, the posterior survival distribution is similar to the distribution of simulated outcomes from the national model.
-Bands show 95\\% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.", cache=TRUE}
-knitr::include_graphics(here::here("figs/RecSlopeAll.png"))
-```
-
-```{r PriorHighFitsAnthro90Rec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model, assuming intensive monitoring (9*60 cows per year for 15 years) of populations from the 97.5 \\% quantile of the national model.
-In these examples, anthropogenic disturbance is 90\\%, and the effect of anthropogenic disturbance varies from none (rSlopeMod=0) through expected (rSlopeMod=1) to double the expected slope (rSlopeMod=2).
-All considered multipliers on the prior standard deviation of the recruitment slope >1 (rSlopeSEMod: $\\beta^R_a$ in Table 2) are sufficiently vague to reflect local variation among populations.
-Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates.", cache=TRUE}
-knitr::include_graphics(here::here("figs/RecSlopeHigh.png"))
+knitr::include_graphics(here::here("figs/RecBoth.png"))
 ```

### 2024-05-31 R/CaribouDemographicFunctions.R

@@ -43,14 +43,14 @@ runScenario<-function(scns,quants=NULL,Anthro=NULL,survAnalysisMethod="KaplanMei
   return(scResults)
 }
 
-makeInterceptPlots <- function(scResults, addBit = "", facetVars = c("sIntSEMod","sSigmaMean"),
+makeInterceptPlots <- function(scResults, addBit = "", facetVars = NULL,
                                loopVars = NULL,
                                whichPlots = c("Adult female survival",
                                               "Population growth rate",
                                               "Recruitment",
                                               "Female population size"),
-                               survLow = 0.6, type = "png", useNational = T) {
-  # facetVars=c("sIntSEMod","sSigmaMean");loopVars="sSigmaSD";scResults=scResults
+                               survLow = 0.6, type = "png", useNational = T,lowLine=NULL,highLine=NULL) {
+  # facetVars=NULL;loopVars="sSigmaMax";scResults=scResults
   #useNational=T;type="png"; survLow=0.6
   if (!is.null(loopVars)) {
     loopSet <- unique(subset(scResults$rr.summary.all, select = loopVars))

### 2024-05-31 analysis/scripts/bayesianModelExamples.R

@@ -43,7 +43,7 @@ yr_scale1 <- scale_x_continuous(breaks = 2023:2072 %>% .[0:9*5+1],
                                 labels = 2009:2058 %>% .[0:9*5+1] )
 yr_scale2 <- scale_x_continuous(breaks = 2009:2058 %>% .[0:9*5+1],
                                 labels = 2009:2058 %>% .[0:9*5+1] )
-recPrior =  plotRes(priorResult, "Recruitment", lowBound=0, highBound = 0.75,
+recPrior =  plotRes(priorResult, "Recruitment", lowBound=0, highBound = 0.8,
                     legendPosition="none",breakInterval=breakInterval,
                     labFontSize=labFontSize) +
   yr_scale1 +
@@ -62,7 +62,7 @@ lambdaPrior =  plotRes(priorResult, "Population growth rate", lowBound=0.6,
 plot(lambdaPrior)
 
 posteriorResult = caribouMetrics:::runScnSet(scns[2,],eParsIn,simBig,getKSDists=F,printProgress=F)
-recPosterior =  plotRes(posteriorResult, "Recruitment", lowBound=0,highBound = 0.75,
+recPosterior =  plotRes(posteriorResult, "Recruitment", lowBound=0,highBound = 0.8,
                         legendPosition="none",breakInterval=breakInterval,
                         labFontSize=labFontSize)+
   yr_scale2 +
@@ -80,7 +80,7 @@ lambdaPosterior =  plotRes(posteriorResult, "Population growth rate", lowBound=0
   ylim(c(0, 1.2))
 plot(lambdaPosterior)
 
-leg <- plotRes(posteriorResult, "Recruitment", lowBound=0,highBound = 0.75,
+leg <- plotRes(posteriorResult, "Recruitment", lowBound=0,highBound = 0.8,
                legendPosition="left",breakInterval=breakInterval,labFontSize=labFontSize)
 leg <- ggpubr::get_legend(leg)
 
@@ -110,7 +110,7 @@ yr_scale1 <- scale_x_continuous(breaks = 2023:2072 %>% .[0:9*5+1],
                                 labels = 2009:2058 %>% .[0:9*5+1] )
 yr_scale2 <- scale_x_continuous(breaks = 2009:2058 %>% .[0:9*5+1],
                                 labels = 2009:2058 %>% .[0:9*5+1] )
-recPrior =  plotRes(priorResult, "Recruitment", lowBound=0, highBound = 0.75,
[…]

### 2024-05-31 analysis/scripts/calibratePriors.R

@@ -1,155 +1,66 @@
 #devtools::load_all(here::here())
-
+library(caribouMetrics)
 scnsApplyAll = data.frame(qMin=0,qMax=0,uMin=0,uMax=0,zMin=0,zMax=0,adjustR=F)
-scnsNoData = data.frame(obsYears=1,collarCount=0,cowMult=0,projYears=1)
+scnsNoData = data.frame(obsYears=1,collarCount=0,cowMult=0,projYears=4)
 scnsLow = data.frame(obsYears=15,collarCount=60,cowMult=9,projYears=1,rQuantile=0.025,sQuantile=0.025)
 scnsHigh = scnsLow;scnsHigh$rQuantile =0.975;scnsHigh$sQuantile=0.975
 ########################
-#Survival
-#vary sIntSEMod, sSigmaMean, and sSigmaSD in factorial array. Outcome of interest is KS distance
-sIntSEMod=seq(1,10,by=2);sSigmaMean=0.08696*seq(0.2,1,by=0.2);sSigmaSD=seq(0.01,0.09,by=0.02)
+#Survival random effects
+#vary sIntSE, sSigmaMin, and sSigmaMax in factorial array.
+sIntSE=getPriors()$l.Saf.Prior2;sSigmaMin=0.00001;sSigmaMax=seq(0.02,0.12,by=0.02)
 
 #Looking for low KS distance from full range of input sims when given almost no info
-scnsSInt=expand.grid(obsAnthroSlope = 0, projAnthroSlope = 0,sIntSEMod=sIntSEMod,
-                     sSigmaMean=sSigmaMean,sSigmaSD=sSigmaSD)
-scnsSInt$SigmaMean = scnsSInt$sSigmaMean
+scnsSInt=expand.grid(obsAnthroSlope = 0, projAnthroSlope = 0,sIntSE=sIntSE,
+                     sSigmaMin=sSigmaMin,sSigmaMax=sSigmaMax)
+scnsSInt$SigmaMin = scnsSInt$sSigmaMin
 scns=merge(scnsSInt,scnsApplyAll);scns=merge(scns,scnsNoData)
 scResults = runScenario(scns,survAnalysisMethod = "Exponential")
-KSAll = subset(scResults$ksDists,(Year==2023)&(Parameter=="Adult female survival"))
+KSAll = subset(scResults$ksDists,(Year==2023+scnsNoData$projYears)&(Parameter=="Adult female survival"))
[…]

### 2024-05-31 analysis/scripts/formatPriors.R

@@ -12,7 +12,7 @@ nameSet <- gsub(".Prior1","",names(rawPriors)[grepl("Prior1",names(rawPriors))],
 
 start=T
 for(n in nameSet){
-  crow = data.frame(name=n,Mean=rawPriors[[paste0(n,".Prior1")]],SE=rawPriors[[paste0(n,".Prior2")]])
+  crow = data.frame(name=n,Mean=rawPriors[[paste0(n,".Prior1")]],SD=rawPriors[[paste0(n,".Prior2")]])
 
   if(start){
     tbl = crow
@@ -25,10 +25,10 @@ for(n in nameSet){
 pars <- data.frame(name="l.R",Parameter="$\\beta^R_0$",Description="R intercept")
 pars<-rbind(pars,data.frame(name="beta.Rec.anthro",Parameter="$\\beta^R_a$",Description="R anthro slope"))
 pars<-rbind(pars,data.frame(name="beta.Rec.fire",Parameter="$\\beta^R_f$",Description="R fire slope"))
-pars<-rbind(pars,data.frame(name="sig.R",Parameter="$\\sigma^2_{R}$",Description="R random effect"))
+#pars<-rbind(pars,data.frame(name="sig.R",Parameter="$\\sigma^2_{R}$",Description="R random effect"))
 pars <- rbind(pars,data.frame(name="l.Saf",Parameter="$\\beta^S_0$",Description="S intercept"))
 pars<-rbind(pars,data.frame(name="beta.Saf",Parameter="$\\beta^S_a$",Description="S anthro slope"))
-pars<-rbind(pars,data.frame(name="sig.Saf",Parameter="$\\sigma^2_{S}$",Description="S random effect"))
+#pars<-rbind(pars,data.frame(name="sig.Saf",Parameter="$\\sigma^2_{S}$",Description="S random effect"))
 
 tbl<-merge(pars,tbl);tbl$name=NULL
 

### 2024-05-31 analysis/paper/paper.Rmd
Minor changes.

@@
 prior means from Johnson et al's (2020) regression model estimates (Table \@ref(tab:priors)). We
-calibrated the standard deviations of the priors so that the 95% prior predictive
+calibrated the standard deviations of the priors so that the the 95% prior predictive
 intervals for survival and recruitment from the Bayesian model match the
-or exceed the 95% predictive intervals from Johnson et al's regression model (Supplement C, Figure
+95% predictive intervals from Johnson et al's regression model (Supplement C, Figure
 \@ref(fig:examplePredictions)).
@@
 Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the
-or exceed the 95% predictive intervals from Johnson et al's regression model (Supplement C).")
+the 95% predictive intervals from Johnson et al's regression model (Supplement C).")
 ```

### 2024-06-17 analysis/paper/paper.Rmd
Updated results. Draft to MH.

@@
 To derive a composition survey bias correction term we assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$).
-The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is a multiplier that defines the apparent number of adult females as a function of the number of collared animals.
+The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of uncollared adult females to collared animals in the recruitment survey.
 The ratio of young bulls to uncollared adult females in the group is
@@
 ## Anthropogenic disturbance and monitoring scenarios
+
+For this investigation we assume collars are deployed in January, and
+last for up to 6 years. The target number of collared females $n$ varies
+among scenarios, as does the total duration of the monitoring program $d$ and the ratio of uncollared to collared adults in the recruitment survey $w$ 
+(Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be
+replaced each year ($o=1$), or there may be a 3 year gap between
+deployments ($o=4$).
+
+We focus on exploring a set of scenarios in which
+buffered anthropogenic disturbance, measured as the percentage of area within 500
+m of some type of anthropogenic disturbance [@johnson_science_2020; @eccc_scientific_2011],
+increases by 1% per year (Figure \@ref(fig:figDisturbance)).
+The duration of monitoring prior to the projection period varies among scenarios from 1 to 24 years.
+The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years.
+The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through medium-low (20\%) and medium-high (40\%) to high (60\%).
+These simple hypothetical scenarios span a large range to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change.
+Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.
 
@@
 ```
-
-For this investigation we assume collars are deployed in January, and
-last for up to 6 years. The target number of collared females $n$ varies
-among scenarios, as does the total duration of the monitoring program $d$ and the ratio of uncollared to collared adults in the recruitment survey $w$ 
-(Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be
-replaced each year ($o=1$), or there may be a 3 year gap between
-deployments ($o=4$).
-
-We focus on exploring a set of scenarios in which
-anthropogenic disturbance, measured as the percentage of area within 500
-m of disturbance [@johnson_science_2020; @eccc_scientific_2011],
-increases by 1% per year (Figure \@ref(fig:figDisturbance)).
-The duration of monitoring prior to the projection period varies among scenarios from 1 to 24 years.
-The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years.
-The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through med-low (20\%) and med-high (40\%) to high (60\%).
-These simple hypothetical scenarios span a large range to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change.
-Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.
 
@@
 These example population trajectories are informed by prior information about the distribution of possible parameter values and outcomes from the national analysis of demographic-disturbance relationships [@johnson_science_2020] so we call it the "national" model.
-Our method for selecting parameter values for each example trajectory yields a continuous set of alternative models (hypotheses) about the state of a particular caribou population, wherein there are more examples of parameter values that prior analysis suggests are more likely; these alternative scenarios informed by prior beliefs are required for analysis of the expected value of alternative sampling strategies [@runge_which_2011; @dunham_strategic_2023].
+Our method for selecting parameter values for each example trajectory yields a set of alternative models (hypotheses) about the state of a particular caribou population; these alternative scenarios informed by prior beliefs are used to assess the expected value of alternative sampling strategies [@runge_which_2011; @dunham_strategic_2023].
 We also use the simulated trajectories to confirm that our Bayesian model successfully integrates local data with priors from the national model.
@@
 \end{equation}
-where $w$ is a multiplier that defines the apparent number of adult
-females in the composition survey as a function of the number of
-collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
+where $w$ is the ratio of observed adult females to collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
 is the probability of misidentifying young bulls as adult females and
@@
 [@lacy_vortex_2017], and the number of juveniles recruiting to the
-post-juvenile class at the census is a binomially distributed function
+post-juvenile class at the census $(\dot{J}_{t})$ is a binomially distributed function
 of the number of surviving post-juvenile females and the adjusted
@@
 
-Following Dunham et al.
-[-@dunham_strategic_2023] we also calculate the expected value of sample information (EVSI) and the expected value of perfect information (EVPI) [@runge_which_2011;@yokota_value_2004], assigning a value of 1 to correct status assessments, and a value of 0 to incorrect assessments ($\nu_{\alpha,\psi}=1 \text{ when } \alpha=\psi, 0 \text{ otherwise}$).
-The expected value with perfect knowledge is $EV_{certainty}=1$.
-The difference between this and the expected score using knowledge of disturbance the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model gives EVPI, a measure of the maximum possible benefit that can be gained from local sampling:
+As suggested by Dunham et al.
+[-@dunham_strategic_2023] we also calculate the expected value of perfect information (EVPI) and the expected value of sample information (EVSI) [@runge_which_2011;@yokota_value_2004].
+We assign a value or utility of $\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$ matches the assigned a status $\alpha$, and value of $\nu_{\alpha,\psi}=0$ when the status assessment is incorrect.
+This is "the value ascribed by the decision-maker to the outcome, and is thus the measure of management performance" [@runge_which_2011].
+The expected value of current information is the average utility of decisions informed by knowledge of the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model:
 \begin{equation}
-EVPI_{\tau} = 1-max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
-  (\#eq:EVPI)
+EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
+  (\#eq:EVcurrent)
 \end{equation}
 where $P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior.
+The expected value under certainty (i.e.
+the expected value of a decision when the true state of the population is known) is $EV_{certainty}=1$.
+The difference between $EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum possible benefit that could theoretically be gained if the system were perfectly known.
+
 EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information: 
 \begin{equation}
-EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - (1-EVPI_{\tau}),
+EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - EV_{current},
   (\#eq:EVSI)
@@
 where $x$ is the sample information from a particular example trajectory.
-Simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) represent the distribution of possible values of $x$.
+In this study, we treat simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) as sample from the distribution of possible values of $x$, and assume each of these sample results is equally probable.
 $P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.
@@
 Uncertainty is also higher in our high anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because stochastic variation in population growth rates increase as populations become very small (as in top right panel of Fig \@ref(fig:examplePredictions)).
-These results confirm the Bayesian population model is working as it should.
 
-Our analysis becomes more interesting and informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of information (Figs \@ref(fig:figEVSI) and \@ref(fig:figEVPI)).
-When anthropogenic disturbance is high (right panels of Figs \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and \@ref(fig:figEVPI)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy.
+Our analysis becomes more informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of information (Figs \@ref(fig:figEVSI) and E.1).
+When anthropogenic disturbance is high (right panels of Figs \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy.
 Interestingly, when the projection period is long and anthropogenic disturbance is high, we see a negative EVSI (lower right panel of Fig \@ref(fig:figEVSI)), and we do not see an associated decrease in the probability of correctly assessing population status (lower right panel of Fig \@ref(fig:figStatusErrors)).
@@
 The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (supplement D).
-Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (supplement E).
+Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (supplement F).
 
-```{r examplePredictions, fig.cap="Prior (top row) and posterior (bottom row) predictions from the Bayesian model for an example population trajectory.
+```{r examplePredictions, fig.cap="Prior (a) and posterior (b) predictions from the Bayesian model for an example population trajectory.
 In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%.
@@
 
-```{r figEVPI, fig.cap="The expected value of perfect information (EVPI) varies among anthropogenic disturbance scenarios (see increasing, decreasing and stable scenarios in Figs \\@ref(fig:figDisturbance), D.1 and D.5 respectively) and the number of years projected.
-When disturbance is high EVPI is low, indicating that local sampling will not alter status assessments.
-Increasing the forecast horizon can either increase or decrease the EVPI as the anthropogenic disturbance trend makes outcomes more or less predictable.", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/EVPIall.png"))
-```
-
 # Discussion
@@
 
-Our model assumes that increasing cumulative anthropogenic disturbance causes a log-linear decrease in caribou demographic rates (equations 4, 5, 10 and 13), reflecting a scientific consensus informed by a number of lines of evidence that cumulative anthropogenic disturbance has a negative impact on boreal caribou [@eccc_scientific_2011; @johnson_science_2020; @gouvernement_du_quebec_revue_2021].
+Our model assumes that increasing cumulative anthropogenic disturbance causes a log-linear decrease in caribou demographic rates (equations 4, 5, 10 and 13), reflecting a scientific consensus informed by a number of lines of evidence that cumulative anthropogenic disturbance has a negative impact on boreal caribou [@eccc_scientific_2011; @johnson_science_2020; @gouvernement_du_quebec_revue_2021;@rudolph_demographic_2017;@sorensen_determining_2008;@hervieux_widespread_2013].
 Local data can be integrated to estimate local demographic-disturbance parameters that differ from the national means, and in the absence of local data, appropriate priors lead to a range of outcomes consistent with the observed range of variation across the country [Figures 3 and 5 in @johnson_science_2020].
@@
 
-Variation in anthropogenic disturbance across the Canadian boreal region is partially correlated with variation in climate and primary productivity because forestry occurs in higher productivity areas, and petroleum extraction activities are concentrated in Western Canada; concerns have been raised about confounding of demographic-disturbance relationships [@wilson_causal_2021;@neufeld_primary_2021; @neilson_direct_2022].
+Variation in anthropogenic disturbance across the Canadian boreal region is partially correlated with variation in climate and primary productivity [@neilson_direct_2022] because forestry occurs in higher productivity areas, and petroleum extraction activities are concentrated in Western Canada.
+Effects of disturbance on vegetation and foodweb dynamics vary with primary productivity [@neufeld_primary_2021], and some have expressed concern about confounding of demographic-disturbance relationships [@wilson_causal_2021].
 One option for addressing these concerns would be to develop regional demographic-disturbance relationships for populations that are ecologically similar to one another (e.g.

### 2024-06-17 analysis/paper/supplement.Rmd
Updated results. Draft to MH.

@@
 ```
-
 \newpage
@@
 
+# Expected value of perfect information
+
+```{r figEVPI, fig.cap="The expected value of perfect information (EVPI) varies among anthropogenic disturbance scenarios (see increasing, decreasing and stable scenarios in Figs 1, D.1 and D.5 respectively) and the number of years projected.
+When disturbance is high EVPI is low, indicating that local sampling will not alter status assessments.
+Increasing the forecast horizon can either increase or decrease the EVPI as the anthropogenic disturbance trend makes outcomes more or less predictable.", cache=TRUE}
+knitr::include_graphics(here::here("figs/forPaper/EVPIall.png"))
+```
+
+\newpage
+
 # Sensitivity to unkown bias in recruitment estimates from calf:cow surveys

### 2024-06-17 analysis/scripts/formatPriors.R

@@ -23,7 +23,7 @@ for(n in nameSet){
 }
 
 pars <- data.frame(name="l.R",Parameter="$\\beta^R_0$",Description="R intercept")
-pars<-rbind(pars,data.frame(name="beta.Rec.anthro",Parameter="$\\beta^R_a$",Description="R anthro slope"))
+pars<-rbind(pars,data.frame(name="beta.Rec.anthro",Parameter="$\\beta^R_a$",Description="R anthropogenic disturbance slope"))
 pars<-rbind(pars,data.frame(name="beta.Rec.fire",Parameter="$\\beta^R_f$",Description="R fire slope"))
 #pars<-rbind(pars,data.frame(name="sig.R",Parameter="$\\sigma^2_{R}$",Description="R random effect"))
 pars <- rbind(pars,data.frame(name="l.Saf",Parameter="$\\beta^S_0$",Description="S intercept"))

### 2024-06-17 analysis/scripts/viewSensitivityResults.R

@@ -13,7 +13,7 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 
 ########################
 #sensitivity
-setName = "s8"
+setName = "s8" #need to do both s7 and s8
 dir.create(paste0("figs/",setName),recursive=T)
 scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
 

### 2024-08-22 analysis/paper/paper.Rmd
ongoing

@@
 @hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021].
-Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou).
+Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013, @eacker_web-based_2019, and @lamb_effectiveness_2024 for examples of application to caribou).
 Recruitment is estimated from aerial surveys of the composition of groups of animals that include collared animals [@nbckc_nbckc_2022].
@@
 Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].
-Some efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019], following broader trends in the analysis of population dynamics [@schaub_integrated_2021].
+Some efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019;@lamb_effective_2024], following broader trends in the analysis of population dynamics [@schaub_integrated_2021].
 In Bayesian population models, all available demographic or population data (e.g.
@@
 However, there is very little monitoring data available for some caribou populations [@johnson_science_2020] and there is a risk that lack of local data will be used as a reason to delay conservation action [@boan_climate_2018].
-Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships, and we are concerned that models informed only by local demographic data could be used to inflate uncertainty about the status of poorly monitored populations.
+Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019;@lamb_effectiveness_2024] have not thus far included demographic-disturbance relationships, and we are concerned that models informed only by local demographic data could be used to inflate uncertainty about the status of poorly monitored populations.
 
@@
 
-Simulated local demographic and covariate data are provided as inputs to the Bayesian population model, including the observed state (dead or alive) of each collared animal in each month ($\hat{I}_{i,t,m}$), the observed number of adult females and calves in the recruitment surveys ($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per collared female in the composition survey $w$, and the measures of anthropogenic disturbance and fire ($A_t, F_t$).
-Following Eacker et al.
-[-@eacker_web-based_2019], recruitment rate $R_t$ is estimated from the observed number of calves $\hat{J}_t$ and adult female caribou $\hat{W}_t$ assuming a binomial distribution:
+Local demographic and covariate data provided as inputs to the Bayesian population model include the observed state (dead or alive) of each collared animal in each month ($\hat{I}_{i,t,m}$), the observed number of adult females and calves in the recruitment surveys ($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per collared female in the composition survey $w$, and the measures of anthropogenic disturbance and fire ($A_t, F_t$).
+Recruitment rate $R_t$ is estimated from the observed number of calves $\hat{J}_t$ and adult female caribou $\hat{W}_t$ assuming a Poisson distribution:
 \begin{equation}
-\hat{J}_t \sim \text{Binomial}(\hat{W}_t,R_t).
+\hat{J}_t \sim \text{Poisson}(\hat{W}_t,R_t).
   (\#eq:hatJEst)
@@
 
-Following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
+Following others [@eacker_web-based_2019;@lamb_effectiveness_2024], Kaplan-Meier
 estimates of observed adult female survival probability $\hat{S}_t$ and
@@
 separate Bayesian model. Observed estimated survival is a modelled as a Gaussian
-distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019]:
+distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019,@lamb_effectiveness_2024]:
 \begin{equation}
@@
 ensure R and S remain within reasonable bounds (not guaranteed given log
-link). Not yet documented here. \* note Eacker survival method gives
-biased results in years when \<12 months of survival data is available.
-\* Note I am dropping survival information for animals that are not
-alive at the beginning of a year to deal with censoring issues. There is
-probably some more elegant way to deal with this, but it won't make any
-difference as long as the collar deployment month is set to 1. \*
-consider & comment on alternative choices made in this example
-<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
-example in all particulars, change some aspects, or do comparisons? \*
+link).
+Not yet documented here.
+\* note Eacker survival method gives biased results if there is left censoring, or months without animals at risk.
+To address this we drop left censored observations and only estimate survival when at least some animals remain at risk by month 12.
+\*
+see alternative choices made in this example
+<https://poissonconsulting.shinyapps.io/bboushiny/>. See also Lamb et al, who uses two step KM estimation process for survival like Eacker. Stick with published examples. \*
 -->

### 2024-08-26 analysis/paper/paper.Rmd
minor text edits

@@
 Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$.
-Given these assumptions and definitions, the sex and bias-corrected recruitment rate $X$ can be written as function of the observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$ (see Supplement B for details).
+Given these assumptions and definitions, the sex and bias-corrected recruitment rate $X$ can be written as a function of the observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$ (see Supplement B for details).
 For simplicity, we combine the probabilities into a bias correction term $c$ that is integrated into the Bayesian population model and simulations (below):
@@
 as disturbance changes over time, so there is substantial persistent
-variation in recruitment and survival among example among populations
+variation in recruitment and survival among example populations
 (Figure \@ref(fig:exampleTrajectories)).
@@
 precisions for years with no observed mortality estimated using a
-separate Bayesian model. Observed estimated survival is a modelled as a Gaussian
+separate Bayesian model. Observed estimated survival is modelled as a Gaussian
 distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019,@lamb_effectiveness_2024]:
@@
 knitr::kable(tbl,caption="Prior means and standard deviations of Gaussian distributed survival and reproduction model parameters.
-Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the
-the 95% predictive intervals from Johnson et al's regression model (Supplement C).")
+Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the 95\\% predictive intervals from Johnson et al's regression model (Supplement C).")
 ```
@@
 [-@dunham_strategic_2023] we also calculate the expected value of perfect information (EVPI) and the expected value of sample information (EVSI) [@runge_which_2011;@yokota_value_2004].
-We assign a value or utility of $\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$ matches the assigned a status $\alpha$, and value of $\nu_{\alpha,\psi}=0$ when the status assessment is incorrect.
+We assign a value or utility of $\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$ matches the assigned status $\alpha$, and value of $\nu_{\alpha,\psi}=0$ when the status assessment is incorrect.
 This is "the value ascribed by the decision-maker to the outcome, and is thus the measure of management performance" [@runge_which_2011].
@@
 where $x$ is the sample information from a particular example trajectory.
-In this study, we treat simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) as sample from the distribution of possible values of $x$, and assume each of these sample results is equally probable.
+In this study, we treat simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) as samples from the distribution of possible values of $x$, and assume each of these sample results is equally probable.
 $P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.
@@
 At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Figs \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).
-The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (supplement D).
-Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (supplement F).
+The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (Supplement D).
+Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (Supplement F).
 
@@
 Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases.
-In these examples the composition cow multiplier $w$ is 6; varying $w$ does not alter qualitative conclusions (supplements D and E).", cache=TRUE}
+In these examples the composition cow multiplier $w$ is 6; varying $w$ does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
@@
 Our intention is to reduce uncertainty that can be used to delay conservation action; there is nothing in our arguments, models or results that can or should be used to discourage much needed monitoring of the effects of conservation actions.
-Even when no additional additional monitoring is required for assessing self-sustainability, there are a variety of reasons to continue monitoring highly disturbed boreal caribou populations.
+Even when no additional monitoring is required for assessing self-sustainability, there are a variety of reasons to continue monitoring highly disturbed boreal caribou populations.
 Our point is that action is also required, and lack of monitoring should not be used as a reason to delay conservation measures for highly disturbed populations.
@@
 Another novel aspect of this work is a term that allows for bias associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012].
-Lacking information about the true magnitude of this bias, we examined sensitivity to a range of assumed values, and found that adding unknown bias did increase uncertainty, but these increases were not large (supplement E).
+Lacking information about the true magnitude of this bias, we examined sensitivity to a range of assumed values, and found that adding unknown bias did increase uncertainty, but these increases were not large (Supplement E).
 A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults and the probability of missing calves.
@@
 However, data owners have not authorized the use of the data for that purpose.
-In order to proceed with the available published information we constrained our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower than higher than average remain lower or higher than average over time.
-The Bayesian model is not similarly constrained, and can accommodate evidence of a shift from unexpectedly low to unexpectedly low demographic rates over time, or vice versa.
-Switching from low to high each year is unlikely, but there is evidence of variation in the slope demographic-disturbance relationships among populations [@rudolph_demographic_2017] so it is important that the Bayesian model allows for a greater range of possibilities than our more constrained simulated example trajectories.
+In order to proceed with the available published information we constrained our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower than or higher than average remain lower or higher than average over time.
+The Bayesian model is not similarly constrained, and can accommodate evidence of a shift from unexpectedly low to unexpectedly high demographic rates over time, or vice versa.
+Switching from low to high each year is unlikely, but there is evidence of variation in the slope of demographic-disturbance relationships among populations [@rudolph_demographic_2017] so it is important that the Bayesian model allows for a greater range of possibilities than our more constrained simulated example trajectories.
 Although our simulated trajectories do not include all plausible possibilities we believe they are sufficient to demonstrate how a Bayesian model can be used to integrate local data with national priors, and to show how monitoring requirements depend on landscape condition.

### 2024-08-27 analysis/paper/paper.Rmd
fixing lamb ref, used visual mode, RStudio changed formatting

@@
 
-The number of species threatened with extinction is high and getting higher [@ipbes_global_2019].
-Reliable information on the status of populations and likely impacts of management are needed to inform conservation action but effective monitoring is challenging and costly [@legg_why_2006; @lindenmayer_effective_2018;@wintle_allocating_2010;@dunham_strategic_2023].
-Resources to support conservation and recovery are limited [@buxton_balancing_2022;@mccarthy_financial_2012;@gerber_conservation_2016;@wintle_spending_2019], delays increase risk [@martin_acting_2012], and some populations have been monitored to extinction [@lindenmayer_counting_2013].
-A recent analysis of action plans for Canadian species at risk found half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022].
-Analysis of the implications of monitoring outcomes for decisions can help clarify whether additional monitoring is necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012].
+The number of species threatened with extinction is high and getting
+higher [@ipbes_global_2019]. Reliable information on the status of
+populations and likely impacts of management are needed to inform
+conservation action but effective monitoring is challenging and costly
+[@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;
+@dunham_strategic_2023]. Resources to support conservation and recovery
+are limited [@buxton_balancing_2022; @mccarthy_financial_2012;
+@gerber_conservation_2016; @wintle_spending_2019], delays increase risk
+[@martin_acting_2012], and some populations have been monitored to
+extinction [@lindenmayer_counting_2013]. A recent analysis of action
+plans for Canadian species at risk found half of tasks are research and
+monitoring, and this proportion is highest for species with higher risk
+of extinction, highlighting a need to more carefully consider when,
+where, and why additional information is needed to inform conservation
+action [@buxton_balancing_2022]. Analysis of the implications of
+monitoring outcomes for decisions can help clarify whether additional
+monitoring is necessary to inform conservation action
+[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
+@gregory_structured_2012].
 
-Analysis methods that can help make better use of available monitoring data to inform decisions include Bayesian population modelling [@schaub_integrated_2021] and value of information analyses [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012].
-To be useful, these methods must be tailored to particular species and decision contexts.
-Here, we offer a simple Bayesian population model for boreal caribou (_Rangifer tarandus caribou_) that integrates information about the impacts of landscape change with available demographic data to reduce uncertainty in population viability projections for boreal caribou [building on @eacker_web-based_2019].
-Following the value of information analysis framework proposed by Dunham et al.
-(2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.
-Our approach is tailored to boreal caribou, but could be adapted for other species at risk found in multiple populations that are costly to monitor and sensitive to habitat disturbance.
+Analysis methods that can help make better use of available monitoring
+data to inform decisions include Bayesian population modelling
+[@schaub_integrated_2021] and value of information analyses
+[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
+@gregory_structured_2012]. To be useful, these methods must be tailored
+to particular species and decision contexts. Here, we offer a simple
+Bayesian population model for boreal caribou (*Rangifer tarandus
+caribou*) that integrates information about the impacts of landscape
+change with available demographic data to reduce uncertainty in
+population viability projections for boreal caribou [building on
+@eacker_web-based_2019]. Following the value of information analysis
+framework proposed by Dunham et al. (2023), we combine this Bayesian
+model with simulations of plausible population trajectories
+[@stewart_climate-informed_2023; @dyson_existing_2022] to show how the
+need for local monitoring varies with landscape condition, and to assess
+the ability of alternative monitoring scenarios to reduce the risk of
+errors in population viability projections. Our approach is tailored to
+boreal caribou, but could be adapted for other species at risk found in
+multiple populations that are costly to monitor and sensitive to habitat
+disturbance.
 
-Boreal caribou are widely and sparsely distributed through the Canadian boreal region across a gradient of anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020] that includes some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022;@ibisch_global_2016;@venter_sixteen_2016].
-Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020].
-The causes of these demographic-disturbance relationships include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011].
-Decreases in recruitment as disturbance increases are also observed over time in places where there is sufficient historic demographic data [@rudolph_demographic_2017;@hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships  [@johnson_science_2020].
-There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020;@neufeld_primary_2021; @fortin_forest_2017; @gagne_impact_2016], and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships [@stewart_climate-informed_2023;@dyson_existing_2022].
+Boreal caribou are widely and sparsely distributed through the Canadian
+boreal region across a gradient of anthropogenic disturbance
+[@eccc_scientific_2011; @johnson_science_2020] that includes some of the
+most intact and inaccessible places remaining on earth
+[@hirsh-pearson_canadas_2022; @ibisch_global_2016;
+@venter_sixteen_2016]. Across 58 boreal caribou study areas, both
+survival and recruitment decrease with the percentage of the study area
+that is disturbed, and no populations of caribou are considered
+self-sustaining in study areas with high anthropogenic disturbance
+[@eccc_scientific_2011; @johnson_science_2020]. The causes of these
+demographic-disturbance relationships include direct impacts on habitat
+availability and use, and indirect impacts on predators and alternate
+prey [@nbckc_boreal_2022; @eccc_scientific_2011]. Decreases in
+recruitment as disturbance increases are also observed over time in
+places where there is sufficient historic demographic data
+[@rudolph_demographic_2017; @hervieux_widespread_2013], but many places
+lack sufficient data to characterize local demographic-disturbance
+relationships [@johnson_science_2020]. There is also variation in
+demographic rates among study areas, particularly where anthropogenic
+disturbance is low [@johnson_science_2020; @neufeld_primary_2021;
+@fortin_forest_2017; @gagne_impact_2016], and this variation leads to
+uncertainty in projections derived from national disturbance-demographic
+relationships [@stewart_climate-informed_2023; @dyson_existing_2022].
 
-Interventions such as wolf culls, maternal penning, and restoration of linear features are intended to improve survival and reproduction of populations that are not otherwise self-sustaining [@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a need to assess and compare effects of boreal caribou conservation actions with careful monitoring.
-Our focus in this paper is assessing and projecting self-sustainability of populations in the absence of these interventions, to clarify when additional information is needed to reduce uncertainty about whether there is a need for action.
-Where actions are taken, outcomes should be monitored; the simple model we propose here does not include effects of management interventions, and cannot be used to project outcomes for managed populations.
+Interventions such as wolf culls, maternal penning, and restoration of
+linear features are intended to improve survival and reproduction of
+populations that are not otherwise self-sustaining
+[@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a
+need to assess and compare effects of boreal caribou conservation
+actions with careful monitoring. Our focus in this paper is assessing
+and projecting self-sustainability of populations in the absence of
+these interventions, to clarify when additional information is needed to
+reduce uncertainty about whether there is a need for action. Where
+actions are taken, outcomes should be monitored; the simple model we
+propose here does not include effects of management interventions, and
+cannot be used to project outcomes for managed populations.
 
-The most widely available source of boreal caribou demographic information is from cows fitted with telemetry collars [e.g.
-@courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@decesare_estimating_2012], though genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g.
-@hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021].
-Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013, @eacker_web-based_2019, and @lamb_effectiveness_2024 for examples of application to caribou).
-Recruitment is estimated from aerial surveys of the composition of groups of animals that include collared animals [@nbckc_nbckc_2022].
-Groups typically also contain uncollared animals, and a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020].
-Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa , and failing to detect calves [@ellington_divergent_2020; @decesare_estimating_2012].
+The most widely available source of boreal caribou demographic
+information is from cows fitted with telemetry collars [e.g.
+@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
+@decesare_estimating_2012], though genetic methods are playing an
+increasingly important role in some caribou monitoring programs [e.g.
+@hettinga_estimating_2012; @mcfarlane_genetic_2018;
+@moeller_integrating_2021; @flasko_fecal_2017; @jones_novel_2023;
+@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
+are typically captured via net-gun from aircraft in late winter, and
+affixed with with collars that send a mortality signal when no longer
+active [see @nbckc_nbckc_2021], thus allowing estimation of adult
+survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for
+original approach, and @mcloughlin_declines_2003,
+@hervieux_widespread_2013, @eacker_web-based_2019, and @lamb2024 for
+examples of application to caribou). Recruitment is estimated from
+aerial surveys of the composition of groups of animals that include
+collared animals [@nbckc_nbckc_2022]. Groups typically also contain
+uncollared animals, and a critical assumption in these surveys is that
+different demographic groups are equally detectable
+[@ellington_divergent_2020]. Errors that can bias composition survey
+results include misidentifying young bulls as cows or vice versa , and
+failing to detect calves [@ellington_divergent_2020;
+@decesare_estimating_2012].
 
-Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].
-Some efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019;@lamb_effective_2024], following broader trends in the analysis of population dynamics [@schaub_integrated_2021].
-In Bayesian population models, all available demographic or population data (e.g.
-abundance surveys, vital rate estimates, trend data) are combined into a single analysis that can permit novel parameter estimation, and improve precision, and account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021].
-These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021].
-However, there is very little monitoring data available for some caribou populations [@johnson_science_2020] and there is a risk that lack of local data will be used as a reason to delay conservation action [@boan_climate_2018].
-Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019;@lamb_effectiveness_2024] have not thus far included demographic-disturbance relationships, and we are concerned that models informed only by local demographic data could be used to inflate uncertainty about the status of poorly monitored populations.
+Survival and recruitment survey data can be combined in a simple
+recruitment-mortality population model [@hatter_moose_1991] that serves
+as the basis for many herd viability assessments
+[@decesare_estimating_2012; @serrouya_comparing_2017;
+@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].
+Some efforts have been made to improve caribou population viability
+projections and growth rate estimates using Bayesian methods
+[@moeller_integrating_2021; @mcnay_demographic_2022;
+@eacker_web-based_2019; @lamb2024], following broader trends in the
+analysis of population dynamics [@schaub_integrated_2021]. In Bayesian
+population models, all available demographic or population data (e.g.
+abundance surveys, vital rate estimates, trend data) are combined into a
+single analysis that can permit novel parameter estimation, and improve
+precision, and account for multiple sources of uncertainty
+[@schaub_integrated_2011; @eacker_assessing_2017;
+@schaub_integrated_2021]. These models are proving broadly useful for
+rapid detection of population declines, evaluation of management
+actions, and viability analysis of critically endangered species
+[@schaub_integrated_2021]. However, there is very little monitoring data
+available for some caribou populations [@johnson_science_2020] and there
+is a risk that lack of local data will be used as a reason to delay
+conservation action [@boan_climate_2018]. Bayesian caribou population
+models [@moeller_integrating_2021; @mcnay_demographic_2022;
+@eacker_web-based_2019; @lamb2024] have not thus far included
+demographic-disturbance relationships, and we are concerned that models
+informed only by local demographic data could be used to inflate
+uncertainty about the status of poorly monitored populations.
 
-We propose to reduce uncertainty in assessments and projections of local caribou population viability by integrating prior information from national analysis of demographic-disturbance relationships with available local demographic data in a simple Bayesian population model for boreal caribou [building on @eacker_web-based_2019].
-In our simple proof-of-concept model, information is shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020].
-A more rigorous statistical approach would be to integrate multiple populations into a hierarchical Bayesian analysis of variation within and among populations [@leasure_hierarchical_2019]; we have not done this because owners have not authorized the use of the data for this purpose, obtaining permission to use caribou demographic data is difficult, and most managers only have access to their own local data.
-Despite limitations, our simple Bayesian model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships.
-We show that integrating the most informative predictor of outcomes for boreal caribou across their range can substantially reduce uncertainty in local demographic projections and status assessments, improve viability projections on changing landscapes, and help clarify where, why and when additional data is required to inform decisions and project outcomes on changing landscapes.
-We hope that this approach can help clarify where available information is sufficient for assessing and projecting population viability, and thereby reduce the risk that uncertainty is used as a reason to delay conservation action for imperiled populations.
+We propose to reduce uncertainty in assessments and projections of local
+caribou population viability by integrating prior information from
+national analysis of demographic-disturbance relationships with
+available local demographic data in a simple Bayesian population model
+for boreal caribou [building on @eacker_web-based_2019]. In our simple
+proof-of-concept model, information is shared among populations using
+priors derived from a published analysis of variation in demographic
+rates among populations across Canada [@johnson_science_2020]. A more
+rigorous statistical approach would be to integrate multiple populations
+into a hierarchical Bayesian analysis of variation within and among
+populations [@leasure_hierarchical_2019]; we have not done this because
+owners have not authorized the use of the data for this purpose,
+obtaining permission to use caribou demographic data is difficult, and
+most managers only have access to their own local data. Despite
+limitations, our simple Bayesian model represents an important
+improvement over less formally integrated methods for estimating and
+projecting impacts of disturbance on population population viability
+[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
+accounts for uncertainties in local demographic data and
+demographic-disturbance relationships. We show that integrating the most
+informative predictor of outcomes for boreal caribou across their range
+can substantially reduce uncertainty in local demographic projections
+and status assessments, improve viability projections on changing
+landscapes, and help clarify where, why and when additional data is
+required to inform decisions and project outcomes on changing
+landscapes. We hope that this approach can help clarify where available
+information is sufficient for assessing and projecting population
+viability, and thereby reduce the risk that uncertainty is used as a
+reason to delay conservation action for imperiled populations.
 
-While we believe that using available information to reduce uncertainty is important, we also believe it is important to acknowledge remaining uncertainty.
-In general, parametric Bayesian population models include a number of simplifying assumptions, and misspecified models can yield biased or overconfident results [@schaub_integrated_2021].
-More specifically, models that do not account for observation errors can be misleading [@schaub_integrated_2021].
-Although several authors have pointed out that  misclassification of adults and missing calves in composition surveys can bias recruitment estimates from those surveys [@ellington_divergent_2020; @decesare_estimating_2012] none have thus far examined sensitivity of Bayesian caribou population models to these sources of error [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019].
-We address this gap with a bias correction term derived from a simple model of the recruitment survey observation process.
+While we believe that using available information to reduce uncertainty
+is important, we also believe it is important to acknowledge remaining
+uncertainty. In general, parametric Bayesian population models include a
+number of simplifying assumptions, and misspecified models can yield
+biased or overconfident results [@schaub_integrated_2021]. More
+specifically, models that do not account for observation errors can be
+misleading [@schaub_integrated_2021]. Although several authors have
+pointed out that misclassification of adults and missing calves in
+composition surveys can bias recruitment estimates from those surveys
+[@ellington_divergent_2020; @decesare_estimating_2012] none have thus
+far examined sensitivity of Bayesian caribou population models to these
+sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
+@eacker_web-based_2019]. We address this gap with a bias correction term
+derived from a simple model of the recruitment survey observation
+process.
 
@@
 
-We focus on analyzing survival of collared females (as in @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019), and the ratio of calves to cows in aerial surveys of groups that include collared females [@nbckc_nbckc_2022], because these remain the most widely available sources of demographic information for boreal caribou populations [@johnson_science_2020].
-Surveyed groups also contain uncollared animals so sample size is typically larger for recruitment than for survival estimates.
-Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa, and failing to detect calves [@ellington_divergent_2020; @decesare_estimating_2012].
+We focus on analyzing survival of collared females (as in
+@mcloughlin_declines_2003, @hervieux_widespread_2013 and
+@eacker_web-based_2019), and the ratio of calves to cows in aerial
+surveys of groups that include collared females [@nbckc_nbckc_2022],
+because these remain the most widely available sources of demographic
+information for boreal caribou populations [@johnson_science_2020].
+Surveyed groups also contain uncollared animals so sample size is
+typically larger for recruitment than for survival estimates. Errors
+that can bias composition survey results include misidentifying young
+bulls as cows or vice versa, and failing to detect calves
+[@ellington_divergent_2020; @decesare_estimating_2012].
 
-To derive a composition survey bias correction term we assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$).
-The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of uncollared adult females to collared animals in the recruitment survey.
+To derive a composition survey bias correction term we assume each group
+of animals in a calf:cow composition survey contains one or more
+collared adult females ($T$), and may also include: uncollared adult
+females misidentified as young bulls or unknown sex ($U$); correctly
+identified uncollared adult females ($V$); young bulls correctly
+identified as male or unknown sex ($O$); young bulls misidentified as
+uncollared adult females ($P$); observed calves ($J$); and unobserved
+calves ($K$). The apparent number of adult females in the group is
+$T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of
+uncollared adult females to collared animals in the recruitment survey.
 The ratio of young bulls to uncollared adult females in the group is
-$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$.
-Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$.
-Given these assumptions and definitions, the sex and bias-corrected recruitment rate $X$ can be written as a function of the observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$ (see Supplement B for details).
-For simplicity, we combine the probabilities into a bias correction term $c$ that is integrated into the Bayesian population model and simulations (below):
+$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying
+young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and
+$P=(O+P)u$. Given a probability $z$ of missing calves, we get
+$J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and
+bias-corrected recruitment rate $X$ can be written as a function of the
+observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
+bulls to adult females $q$, and the misidentification probabilities $u$
+and $z$ (see Supplement B for details). For simplicity, we combine the
+probabilities into a bias correction term $c$ that is integrated into
+the Bayesian population model and simulations (below):
 $$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$
@@
 last for up to 6 years. The target number of collared females $n$ varies
-among scenarios, as does the total duration of the monitoring program $d$ and the ratio of uncollared to collared adults in the recruitment survey $w$ 
-(Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be
-replaced each year ($o=1$), or there may be a 3 year gap between
-deployments ($o=4$).
+among scenarios, as does the total duration of the monitoring program
+$d$ and the ratio of uncollared to collared adults in the recruitment
+survey $w$ (Table \@ref(tab:monitoringScenarios)). Collars lost to
+mortality may be replaced each year ($o=1$), or there may be a 3 year
+gap between deployments ($o=4$).
 
-We focus on exploring a set of scenarios in which
-buffered anthropogenic disturbance, measured as the percentage of area within 500
-m of some type of anthropogenic disturbance [@johnson_science_2020; @eccc_scientific_2011],
-increases by 1% per year (Figure \@ref(fig:figDisturbance)).
-The duration of monitoring prior to the projection period varies among scenarios from 1 to 24 years.
-The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years.
-The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through medium-low (20\%) and medium-high (40\%) to high (60\%).
-These simple hypothetical scenarios span a large range to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change.
-Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.
+We focus on exploring a set of scenarios in which buffered anthropogenic
+disturbance, measured as the percentage of area within 500 m of some
+type of anthropogenic disturbance [@johnson_science_2020;
+@eccc_scientific_2011], increases by 1% per year (Figure
+\@ref(fig:figDisturbance)). The duration of monitoring prior to the
+projection period varies among scenarios from 1 to 24 years. The
+projection period begins in 2024, and we consider how uncertainty
+changes as the projection period increases from 5 to 20 years. The
+amount of anthropogenic disturbance at the beginning of the projection
+period varies from low (0%) through medium-low (20%) and medium-high
+(40%) to high (60%). These simple hypothetical scenarios span a large
+range to show how the need for local monitoring to reduce uncertainty
+depends on landscape state, not to represent the reality of landscape
+change. Scenarios in which anthropogenic disturbance remains the same or
+decreases over time are included in Supplement D.
 
@@
 
-To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics.
-These example population trajectories are informed by prior information about the distribution of possible parameter values and outcomes from the national analysis of demographic-disturbance relationships [@johnson_science_2020] so we call it the "national" model.
-Our method for selecting parameter values for each example trajectory yields a set of alternative models (hypotheses) about the state of a particular caribou population; these alternative scenarios informed by prior beliefs are used to assess the expected value of alternative sampling strategies [@runge_which_2011; @dunham_strategic_2023].
-We also use the simulated trajectories to confirm that our Bayesian model successfully integrates local data with priors from the national model.
-For reference, all variables and parameters are defined in supplemental Table A1.
+To investigate the effectiveness of various monitoring strategies for
+reducing uncertainty about population viability, and to show how this
+varies with landscape condition, we begin with a model for simulating
+example population dynamics and observations of those dynamics. These
+example population trajectories are informed by prior information about
+the distribution of possible parameter values and outcomes from the
+national analysis of demographic-disturbance relationships
+[@johnson_science_2020] so we call it the "national" model. Our method
+for selecting parameter values for each example trajectory yields a set
+of alternative models (hypotheses) about the state of a particular
+caribou population; these alternative scenarios informed by prior
+beliefs are used to assess the expected value of alternative sampling
+strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the
+simulated trajectories to confirm that our Bayesian model successfully
+integrates local data with priors from the national model. For
+reference, all variables and parameters are defined in supplemental
+Table A1.
 
@@
 
-To simulate demographic trajectories for example populations
-we used a modified version of Johnson et al's [-@johnson_science_2020]
-demographic model described by Dyson et al.
-[-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
-distributed with true survival probability $\dot{S}_t$:
-$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification
-biases (Supplement B) and (optionally) a delayed reproduction term to account for the reality that yearlings generally do not breed
-[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum potential recruitment of female calves per cow $\dot{X}_t$:
+To simulate demographic trajectories for example populations we used a
+modified version of Johnson et al's [-@johnson_science_2020] demographic
+model described by Dyson et al. [-@dyson_existing_2022]. Given a
+population of post-juvenile females at the beginning of year $t$,
+$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
+binomially distributed with true survival probability $\dot{S}_t$:
+$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow
+ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification biases
+(Supplement B) and (optionally) a delayed reproduction term to account
+for the reality that yearlings generally do not breed
+[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum
+potential recruitment of female calves per cow $\dot{X}_t$:
 \begin{equation}
@@
   (\#eq:dotX)
-\end{equation}
-where $w$ is the ratio of observed adult females to collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
-is the probability of misidentifying young bulls as adult females and
-vice versa, and $\dot{z}$ is the probability of missing a calf.
-We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6.
+\end{equation} where $w$ is the ratio of observed adult females to
+collared animals, $\dot{q}$ is the ratio of young bulls to adult
+females, $\dot{u}$ is the probability of misidentifying young bulls as
+adult females and vice versa, and $\dot{z}$ is the probability of
+missing a calf. We examine sensitivity to these unknown parameters by
+selecting values for each simulated example population from plausible
+ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and
+$\dot{q}$ varies uniformly between 0 and 0.6.
 
@@
 [@lacy_vortex_2017], and the number of juveniles recruiting to the
-post-juvenile class at the census $(\dot{J}_{t})$ is a binomially distributed function
-of the number of surviving post-juvenile females and the adjusted
-recruitment rate:
-\begin{equation}
+post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
+distributed function of the number of surviving post-juvenile females
+and the adjusted recruitment rate: \begin{equation}
   \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{N_0k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
@@
 
-Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et al. (2020) (Table A1), recruitment rate is lowest
-$(0.5\dot{X}_t)$ when $\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$
-at intermediate population sizes, and declines to $0.6\dot{X}_t$ as the
-population reaches carrying capacity of $k=100$ times the initial
-population size. The post-juvenile female population in the next year
-includes both survivors and new recruits:
-\begin{equation}
+Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et
+al. (2020) (Table A1), recruitment rate is lowest $(0.5\dot{X}_t)$ when
+$\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$ at intermediate
+population sizes, and declines to $0.6\dot{X}_t$ as the population
+reaches carrying capacity of $k=100$ times the initial population size.
+The post-juvenile female population in the next year includes both
+survivors and new recruits: \begin{equation}
 \dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t).
@@
 
+```{=tex}
 \begin{equation}
@@
 \end{equation}
-
+```
+```{=tex}
 \begin{equation}
@@
 \end{equation}
-
+```
 $\phi^R \sim \text{Normal}(19.862,2.229)$ and
@@
 as disturbance changes over time, so there is substantial persistent
-variation in recruitment and survival among example populations
-(Figure \@ref(fig:exampleTrajectories)).
-To ensure results are comparable across monitoring scenarios and that populations remain extant through the monitoring period, all simulations are started with $N_0=5000$ animals 24 years prior to the start of the projection period.
+variation in recruitment and survival among example populations (Figure
+\@ref(fig:exampleTrajectories)). To ensure results are comparable across
+monitoring scenarios and that populations remain extant through the
+monitoring period, all simulations are started with $N_0=5000$ animals
+24 years prior to the start of the projection period.
 
@@
 
-In our simulated example populations, mortality risk does not vary over time or
-among individuals, so the observed state (dead or alive) of a collared
-animal $i$ in month $m$ and year $t$ depends on the true survival rate
-$\dot{S}_t$ as follows:
+In our simulated example populations, mortality risk does not vary over
+time or among individuals, so the observed state (dead or alive) of a
+collared animal $i$ in month $m$ and year $t$ depends on the true
+survival rate $\dot{S}_t$ as follows: 
 \begin{equation}
@@
 \end{equation}
-The number of observed calves $\hat{J}_t$
-also depends on the unadjusted apparent recruitment for the population
-$\dot{R}_t$: 
+The number of observed calves $\hat{J}_t$ also depends on
+the unadjusted apparent recruitment for the population $\dot{R}_t$:
 \begin{equation}
@@
 
-Local demographic and covariate data provided as inputs to the Bayesian population model include the observed state (dead or alive) of each collared animal in each month ($\hat{I}_{i,t,m}$), the observed number of adult females and calves in the recruitment surveys ($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per collared female in the composition survey $w$, and the measures of anthropogenic disturbance and fire ($A_t, F_t$).
-Recruitment rate $R_t$ is estimated from the observed number of calves $\hat{J}_t$ and adult female caribou $\hat{W}_t$ assuming a Poisson distribution:
+Local demographic and covariate data provided as inputs to the Bayesian
+population model include the observed state (dead or alive) of each
+collared animal in each month ($\hat{I}_{i,t,m}$), the observed number
+of adult females and calves in the recruitment surveys
+($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per
+collared female in the composition survey $w$, and the measures of
+anthropogenic disturbance and fire ($A_t, F_t$). Recruitment rate $R_t$
+is estimated from the observed number of calves $\hat{J}_t$ and adult
+female caribou $\hat{W}_t$ assuming a Poisson distribution:
 \begin{equation}
@@
 
-Where appropriate, the Bayesian population model and national demographic model share the same structure, and we have selected notation to highlight the similarities.
-The difference between equations \@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to simulate observations.
-In the Bayesian model, in contrast, observations are provided, and used to estimate the unobserved recruitment rate $R_t$.
+Where appropriate, the Bayesian population model and national
+demographic model share the same structure, and we have selected
+notation to highlight the similarities. The difference between equations
+\@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an
+unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to
+simulate observations. In the Bayesian model, in contrast, observations
+are provided, and used to estimate the unobserved recruitment rate
+$R_t$.
 
@@
 disturbance and fire with a log link [@johnson_science_2020;
-@stewart_climate-informed_2023; @dyson_existing_2022]. In the Bayesian model we add a Gaussian
-distributed random year effect to account for stochastic variation among years [@eacker_web-based_2019]:
+@stewart_climate-informed_2023; @dyson_existing_2022]. In the Bayesian
+model we add a Gaussian distributed random year effect to account for
+stochastic variation among years [@eacker_web-based_2019]:
 \begin{equation}
@@
 
-In the simulation model (eq \@ref(eq:barR)), regression coefficient values ($\dot{\beta}^R_0,\dot{\beta}^R_a,\dot{\beta}^R_f$) are selected for each example population and used to calculate recruitment rate.
-In the Bayesian model (eq \@ref(eq:R)), unobserved regression coefficients ($\beta^R_0,\beta^R_a,\beta^R_f$) are estimated from local recruitment survey and disturbance observations.
+In the simulation model (eq \@ref(eq:barR)), regression coefficient
+values ($\dot{\beta}^R_0,\dot{\beta}^R_a,\dot{\beta}^R_f$) are selected
+for each example population and used to calculate recruitment rate. In
+the Bayesian model (eq \@ref(eq:R)), unobserved regression coefficients
+($\beta^R_0,\beta^R_a,\beta^R_f$) are estimated from local recruitment
+survey and disturbance observations.
 
-To account for bias and additional uncertainty caused by unknown misidentification biases (Supplement B) and delayed age of age of first reproduction
-[@eacker_web-based_2019; @decesare_estimating_2012] the estimated recruitment probability is adjusted to get expected female recruits per female:
-\begin{equation}
+To account for bias and additional uncertainty caused by unknown
+misidentification biases (Supplement B) and delayed age of age of first
+reproduction [@eacker_web-based_2019; @decesare_estimating_2012] the
+estimated recruitment probability is adjusted to get expected female
+recruits per female: \begin{equation}
 X_t=\frac{cR_t/2}{1+cR_t/2}; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
@@
 
-The composition survey bias correction term $c$ depends on the known apparent
-number of adult females per collared animal $w$, the unknown ratio of young
-bulls to adult females $q$, the unknown probability of misidentifying young
-bulls as adult females and vice versa $u$, and the unknown probability of
-missing a calf $z$ (Supplement B). Prior uncertainty about the value of the bias
-correction term $c$ is approximated with a Log-normal distribution
-(Supplement B). Given the apparent number of adult females per collared
-animal $w$ in each monitoring scenario (Table
-\@ref(tab:monitoringScenarios)), we calculate the mean and standard
-deviation of 10,000 samples of $\log{ c}$ assuming the ratio of young
-bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult
-misidentification probability $u$ varies uniformly between 0 and 0.2,
-and the probability of missing a calf $z$ varies uniformly between 0 and
-0.2. To maintain consistency with Johnson
-et al. [-@johnson_science_2020] we also provide an option to only adjust
-for sex ratio and misidentification biases $X_t=cR_t/2$.
+The composition survey bias correction term $c$ depends on the known
+apparent number of adult females per collared animal $w$, the unknown
+ratio of young bulls to adult females $q$, the unknown probability of
+misidentifying young bulls as adult females and vice versa $u$, and the
+unknown probability of missing a calf $z$ (Supplement B). Prior
+uncertainty about the value of the bias correction term $c$ is
+approximated with a Log-normal distribution (Supplement B). Given the
+apparent number of adult females per collared animal $w$ in each
+monitoring scenario (Table \@ref(tab:monitoringScenarios)), we calculate
+the mean and standard deviation of 10,000 samples of $\log{ c}$ assuming
+the ratio of young bulls to adult females $q$ varies uniformly between 0
+and 0.6, the adult misidentification probability $u$ varies uniformly
+between 0 and 0.2, and the probability of missing a calf $z$ varies
+uniformly between 0 and 0.2. To maintain consistency with Johnson et al.
+[-@johnson_science_2020] we also provide an option to only adjust for
+sex ratio and misidentification biases $X_t=cR_t/2$.
 
-Following others [@eacker_web-based_2019;@lamb_effectiveness_2024], Kaplan-Meier
+Following others [@eacker_web-based_2019; @lamb2024], Kaplan-Meier
 estimates of observed adult female survival probability $\hat{S}_t$ and
 associated precisions $\tau_t$ are estimated from known-fate radio
-collar data using the _survival_ R package [@therneau_modeling_2000], with
-precisions for years with no observed mortality estimated using a
-separate Bayesian model. Observed estimated survival is modelled as a Gaussian
-distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019,@lamb_effectiveness_2024]:
+collar data using the *survival* R package [@therneau_modeling_2000],
+with precisions for years with no observed mortality estimated using a
+separate Bayesian model. Observed estimated survival is modelled as a
+Gaussian distributed function of the true survival probability,
+truncated between 0 and 1, to allow sharing survival information across
+years with a random effect [@eacker_web-based_2019,@lamb2024]:
 \begin{equation}
@@
 
-For consistency with [@johnson_science_2020], we model survival probability as an adjusted function of buffered anthropogenic disturbance $A_t$ with a log link
-[@stewart_climate-informed_2023; @dyson_existing_2022]. We add a Gaussian distributed random year effect to account for stochastic variation among years [@eacker_web-based_2019]:
-\begin{equation}
+For consistency with [@johnson_science_2020], we model survival
+probability as an adjusted function of buffered anthropogenic
+disturbance $A_t$ with a log link [@stewart_climate-informed_2023;
+@dyson_existing_2022]. We add a Gaussian distributed random year effect
+to account for stochastic variation among years
+[@eacker_web-based_2019]: \begin{equation}
 S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).
@@
 individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed
-[@schaub_integrated_2021]:
-\begin{equation}
+[@schaub_integrated_2021]: \begin{equation}
 \hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).
@@
 All regression coefficients are assumed to be Gaussian distributed, with
-prior means from Johnson et al's (2020) regression model estimates (Table \@ref(tab:priors)). We
-calibrated the standard deviations of the priors so that the the 95% prior predictive
-intervals for survival and recruitment from the Bayesian model match the
-95% predictive intervals from Johnson et al's regression model (Supplement C, Figure
+prior means from Johnson et al's (2020) regression model estimates
+(Table \@ref(tab:priors)). We calibrated the standard deviations of the
+priors so that the the 95% prior predictive intervals for survival and
+recruitment from the Bayesian model match the 95% predictive intervals
+from Johnson et al's regression model (Supplement C, Figure
 \@ref(fig:examplePredictions)).
@@
 
-As in the simulations, we assume a simple demographic model, with a census in late
-winter of surviving females and new recruits to the post-juvenile (aged
-1+) class. The number of post-juvenile females that survive from year
-$t$ to the census $W_t$ is binomially distributed with survival
-probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The number of
-juveniles recruiting to the post-juvenile class at the census is a
-Poisson distributed function of the number of surviving post-juvenile
-females and the adjusted recruitment rate $X_t$:
+As in the simulations, we assume a simple demographic model, with a
+census in late winter of surviving females and new recruits to the
+post-juvenile (aged 1+) class. The number of post-juvenile females that
+survive from year $t$ to the census $W_t$ is binomially distributed with
+survival probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The
+number of juveniles recruiting to the post-juvenile class at the census
+is a Poisson distributed function of the number of surviving
+post-juvenile females and the adjusted recruitment rate $X_t$:
 $J_{t} \sim \text{Poisson}(W_tX_t)$. The post-juvenile female population
@@
 
+```{=html}
 <!-- Annotated equations for poster.
@@
 $$\underbrace{\hat{S_t}}_{\text{Observed survival (Kaplan-Meier estimate)}} \sim \text{Normal}(S_t,1/\tau_t).$$ -->
-
+```
+```{=html}
 <!--
@@
 -->
-
-Models and analysis were implemented using the R language [@r_core_team_r_2023] and a variety of _tidyverse_ packages [@wickham_welcome_2019].
-We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the _R2jags_ R package [@su_r2jags_2021] to fit the Bayesian model.
-We ran 15,000 iterations of 4 chain, discarded the first 10,000 interations, and used a thinning rate of 2.
-We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the _mcmcplots_ R package [@curtis_mcmcplots_2018].
-Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>, and in the  _caribouMetrics_ R package available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.
+```
+Models and analysis were implemented using the R language
+[@r_core_team_r_2023] and a variety of *tidyverse* packages
+[@wickham_welcome_2019]. We use the open source Gibbs sampling software
+JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
+to fit the Bayesian model. We ran 15,000 iterations of 4 chain,
+discarded the first 10,000 interations, and used a thinning rate of 2.
+We assessed convergence for a small random subset of example
+trajectories using a threshold for the Gelman-Rubin diagnostic
+[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
+posterior autocorrelation and traceplots produced by the *mcmcplots* R
+package [@curtis_mcmcplots_2018]. Code required to run analyses and
+reproduce figures in this paper is available at
+<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
+and in the *caribouMetrics* R package available on GitHub at
+<https://github.com/LandSciTech/caribouMetrics>.
 
@@
 relevant to decision making [@gregory_structured_2012]. For boreal
-caribou and other endangered species, a relevant outcome is the probability of correctly predicting the true population status (increasing or
-decreasing) [@dunham_strategic_2023]. To calculate this outcome for each disturbance and
-monitoring scenario (Fig \@ref(fig:figDisturbance)) we fit the
-Bayesian model to simulated observations from 500 example population trajectories (e.g. Fig
-\@ref(fig:exampleTrajectories)), and calculated the percentage of these 500 examples in
-which the Bayesian posterior mean population growth rate and the true
-population growth rate are both greater than 0.99, or both less than or equal to 0.99.
+caribou and other endangered species, a relevant outcome is the
+probability of correctly predicting the true population status
+(increasing or decreasing) [@dunham_strategic_2023]. To calculate this
+outcome for each disturbance and monitoring scenario (Fig
+\@ref(fig:figDisturbance)) we fit the Bayesian model to simulated
+observations from 500 example population trajectories (e.g. Fig
+\@ref(fig:exampleTrajectories)), and calculated the percentage of these
+500 examples in which the Bayesian posterior mean population growth rate
+and the true population growth rate are both greater than 0.99, or both
+less than or equal to 0.99.
 
-As suggested by Dunham et al.
-[-@dunham_strategic_2023] we also calculate the expected value of perfect information (EVPI) and the expected value of sample information (EVSI) [@runge_which_2011;@yokota_value_2004].
-We assign a value or utility of $\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$ matches the assigned status $\alpha$, and value of $\nu_{\alpha,\psi}=0$ when the status assessment is incorrect.
-This is "the value ascribed by the decision-maker to the outcome, and is thus the measure of management performance" [@runge_which_2011].
-The expected value of current information is the average utility of decisions informed by knowledge of the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model:
-\begin{equation}
+As suggested by Dunham et al. [-@dunham_strategic_2023] we also
+calculate the expected value of perfect information (EVPI) and the
+expected value of sample information (EVSI) [@runge_which_2011;
+@yokota_value_2004]. We assign a value or utility of
+$\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$
+matches the assigned status $\alpha$, and value of $\nu_{\alpha,\psi}=0$
+when the status assessment is incorrect. This is "the value ascribed by
+the decision-maker to the outcome, and is thus the measure of management
+performance" [@runge_which_2011]. The expected value of current
+information is the average utility of decisions informed by knowledge of
+the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and
+priors from the national model: \begin{equation}
 EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
   (\#eq:EVcurrent)
-\end{equation}
-where $P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior.
-The expected value under certainty (i.e.
-the expected value of a decision when the true state of the population is known) is $EV_{certainty}=1$.
-The difference between $EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum possible benefit that could theoretically be gained if the system were perfectly known.
+\end{equation} where $P(\psi|\tau)$ is the prior probability that the
+true state of the population is $\psi$, and each population is assigned
+a status $\alpha$ informed by the prior. The expected value under
+certainty (i.e. the expected value of a decision when the true state of
+the population is known) is $EV_{certainty}=1$. The difference between
+$EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum
+possible benefit that could theoretically be gained if the system were
+perfectly known.
 
-EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information: 
-\begin{equation}
+EVSI measures expected improvement from a particular sampling strategy
+as the difference between the expected value with and without additional
+sample information: \begin{equation}
 EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - EV_{current},
   (\#eq:EVSI)
-\end{equation}
-where $x$ is the sample information from a particular example trajectory.
-In this study, we treat simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) as samples from the distribution of possible values of $x$, and assume each of these sample results is equally probable.
-$P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.
+\end{equation} where $x$ is the sample information from a particular
+example trajectory. In this study, we treat simulated observations from
+500 example trajectories for each disturbance and monitoring scenario
+$\tau$ (Figure \@ref(fig:figDisturbance)) as samples from the
+distribution of possible values of $x$, and assume each of these sample
+results is equally probable. $P(\psi|x,\tau)$ is the posterior
+probability that the true state of the population is $\psi$, and each
+population is assigned a status $\alpha$ informed by the sample
+information.
 
-To aid in interpretation of results and to highlight the advantages of selecting a metric that is directly relevant to the management objective of stable or increasing populations, we also show how integrating local monitoring data alters the distribution of differences between true population growth rates and Bayesian posterior mean estimated growth rates.
+To aid in interpretation of results and to highlight the advantages of
+selecting a metric that is directly relevant to the management objective
+of stable or increasing populations, we also show how integrating local
+monitoring data alters the distribution of differences between true
+population growth rates and Bayesian posterior mean estimated growth
+rates.
 
@@
 
-Analysis of simulated example demographic trajectories confirms that our Bayesian model can effectively combine local demographic data with priors from national demographic-disturbance relationships to reduce uncertainty in boreal caribou demographic projections (Figs \@ref(fig:examplePredictions) and \@ref(fig:figDiffs)).
-As expected, uncertainty about the population growth rate decreases as sampling effort in each year (i.e.
-number of collars) or the number of years of monitoring (Fig \@ref(fig:figDiffs)) increase.
-Uncertainty is higher when the forecast horizon is long (compare 5 and 20 year projections in Fig \@ref(fig:figDiffs)).
-Uncertainty is also higher in our high anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because stochastic variation in population growth rates increase as populations become very small (as in top right panel of Fig \@ref(fig:examplePredictions)).
+Analysis of simulated example demographic trajectories confirms that our
+Bayesian model can effectively combine local demographic data with
+priors from national demographic-disturbance relationships to reduce
+uncertainty in boreal caribou demographic projections (Figs
+\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected,
+uncertainty about the population growth rate decreases as sampling
+effort in each year (i.e. number of collars) or the number of years of
+monitoring (Fig \@ref(fig:figDiffs)) increase. Uncertainty is higher
+when the forecast horizon is long (compare 5 and 20 year projections in
+Fig \@ref(fig:figDiffs)). Uncertainty is also higher in our high
+anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because
+stochastic variation in population growth rates increase as populations
+become very small (as in top right panel of Fig
+\@ref(fig:examplePredictions)).
 
-Our analysis becomes more informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of information (Figs \@ref(fig:figEVSI) and E.1).
-When anthropogenic disturbance is high (right panels of Figs \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy.
-Interestingly, when the projection period is long and anthropogenic disturbance is high, we see a negative EVSI (lower right panel of Fig \@ref(fig:figEVSI)), and we do not see an associated decrease in the probability of correctly assessing population status (lower right panel of Fig \@ref(fig:figStatusErrors)).
-This means that the long-term projections informed by local monitoring data are on average less precise but not less accurate, which can happen when there is very small prior probability that population is viable (orange band in the top right panel of Fig \@ref(fig:examplePredictions) falls below 1).
-At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Figs \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).
-The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (Supplement D).
-Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (Supplement F).
+Our analysis becomes more informative when we examine the probability of
+correctly assessing population status (Fig \@ref(fig:figStatusErrors))
+and the expected value of information (Figs \@ref(fig:figEVSI) and E.1).
+When anthropogenic disturbance is high (right panels of Figs
+\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
+of population status informed only by priors from the national
+demographic-disturbance relationships are accurate, and adding
+additional local monitoring data does not improve accuracy.
+Interestingly, when the projection period is long and anthropogenic
+disturbance is high, we see a negative EVSI (lower right panel of Fig
+\@ref(fig:figEVSI)), and we do not see an associated decrease in the
+probability of correctly assessing population status (lower right panel
+of Fig \@ref(fig:figStatusErrors)). This means that the long-term
+projections informed by local monitoring data are on average less
+precise but not less accurate, which can happen when there is very small
+prior probability that population is viable (orange band in the top
+right panel of Fig \@ref(fig:examplePredictions) falls below 1). At
+lower levels of disturbance, local monitoring can reduce uncertainty,
+but there are diminishing returns as the number of collars deployed or
+the number of years of monitoring increases (Figs
+\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the
+relationships between monitoring effort and the probability of correct
+status assessment depend on how anthropogenic disturbance is changing
+over time, but there are diminishing benefits of increased monitoring
+regardless of whether disturbance is increasing or decreasing
+(Supplement D). Adding unknown bias increases uncertainty, but these
+increases are not large for the set of parameters we investigate
+(Supplement F).
 
@@
 
-NOTE: Please provide comments and suggestions for the discussion here: <https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.
+NOTE: Please provide comments and suggestions for the discussion here:
+<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.
 
@@
 
-The availability of boreal caribou demographic data varies widely across Canada due to large variation in habitat condition, population condition, available resources for monitoring, and remoteness.
-In a few regions enough data has been collected over time to characterize local demographic-disturbance relationships [@rudolph_demographic_2017;@hervieux_widespread_2013], but most populations have only been sparsely monitored for snapshot estimates of trend or abundance [@johnson_science_2020].
-As a result, in most regions it is necessary to make assumptions informed by observed variation in demographic rates with disturbance among populations and other relevant information about impacts of disturbance on caribou.
-However, high observed variation in demography where anthropogenic disturbance is low leads to high uncertainty in projections informed by national demographic-disturbance relationships alone [@stewart_climate-informed_2023;@dyson_existing_2022].
-In some cases, predictions of population growth rate informed by national demographic-disturbance relationships appear to disagree with estimates from local demographic data [@eccc_scientific_2011; @dyson_existing_2022], but without formal methods of accounting for uncertainty it can be difficult to distinguish true discrepancies from stochastic variation.
+The availability of boreal caribou demographic data varies widely across
+Canada due to large variation in habitat condition, population
+condition, available resources for monitoring, and remoteness. In a few
+regions enough data has been collected over time to characterize local
+demographic-disturbance relationships [@rudolph_demographic_2017;
+@hervieux_widespread_2013], but most populations have only been sparsely
+monitored for snapshot estimates of trend or abundance
+[@johnson_science_2020]. As a result, in most regions it is necessary to
+make assumptions informed by observed variation in demographic rates
+with disturbance among populations and other relevant information about
+impacts of disturbance on caribou. However, high observed variation in
+demography where anthropogenic disturbance is low leads to high
+uncertainty in projections informed by national demographic-disturbance
+relationships alone [@stewart_climate-informed_2023;
+@dyson_existing_2022]. In some cases, predictions of population growth
+rate informed by national demographic-disturbance relationships appear
+to disagree with estimates from local demographic data
+[@eccc_scientific_2011; @dyson_existing_2022], but without formal
+methods of accounting for uncertainty it can be difficult to distinguish
+true discrepancies from stochastic variation.
 
-We developed a simple Bayesian population model [building on @eacker_web-based_2019] to help address these challenges by acknowledging uncertainty associated with local demographic data and national demographic-disturbance relationships, and integrating these two sources of information to reduce uncertainty in projections of the impacts of changing landscape conditions.
-This Bayesian approach represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships.
+We developed a simple Bayesian population model [building on
+@eacker_web-based_2019] to help address these challenges by
+acknowledging uncertainty associated with local demographic data and
+national demographic-disturbance relationships, and integrating these
+two sources of information to reduce uncertainty in projections of the
+impacts of changing landscape conditions. This Bayesian approach
+represents an important improvement over less formally integrated
+methods for estimating and projecting impacts of disturbance on
+population population viability [@eccc_scientific_2011;
+@johnson_science_2020] because it reconciles and accounts for
+uncertainties in local demographic data and demographic-disturbance
+relationships.
 
-Our model assumes that increasing cumulative anthropogenic disturbance causes a log-linear decrease in caribou demographic rates (equations 4, 5, 10 and 13), reflecting a scientific consensus informed by a number of lines of evidence that cumulative anthropogenic disturbance has a negative impact on boreal caribou [@eccc_scientific_2011; @johnson_science_2020; @gouvernement_du_quebec_revue_2021;@rudolph_demographic_2017;@sorensen_determining_2008;@hervieux_widespread_2013].
-Local data can be integrated to estimate local demographic-disturbance parameters that differ from the national means, and in the absence of local data, appropriate priors lead to a range of outcomes consistent with the observed range of variation across the country [Figures 3 and 5 in @johnson_science_2020].
-Others [@moeller_integrating_2021; @mcnay_demographic_2022; @eacker_web-based_2019] have also used Bayesian models to estimate caribou demographic rates from local monitoring data, but our addition of prior information from national demographic-disturbance relationships is novel, and important for reducing the risk that lack of local data will be used as a reason to delay conservation action for highly disturbed and poorly monitored populations.
-Integrating demographic-disturbance relationships into a viability projection model clarifies that no additional information is needed to project outcomes where disturbance is high and actions to support recovery are not taken.
+Our model assumes that increasing cumulative anthropogenic disturbance
+causes a log-linear decrease in caribou demographic rates (equations 4,
+5, 10 and 13), reflecting a scientific consensus informed by a number of
+lines of evidence that cumulative anthropogenic disturbance has a
+negative impact on boreal caribou [@eccc_scientific_2011;
+@johnson_science_2020; @gouvernement_du_quebec_revue_2021;
+@rudolph_demographic_2017; @sorensen_determining_2008;
+@hervieux_widespread_2013]. Local data can be integrated to estimate
+local demographic-disturbance parameters that differ from the national
+means, and in the absence of local data, appropriate priors lead to a
+range of outcomes consistent with the observed range of variation across
+the country [Figures 3 and 5 in @johnson_science_2020]. Others
+[@moeller_integrating_2021; @mcnay_demographic_2022;
+@eacker_web-based_2019] have also used Bayesian models to estimate
+caribou demographic rates from local monitoring data, but our addition
+of prior information from national demographic-disturbance relationships
+is novel, and important for reducing the risk that lack of local data
+will be used as a reason to delay conservation action for highly
+disturbed and poorly monitored populations. Integrating
+demographic-disturbance relationships into a viability projection model
+clarifies that no additional information is needed to project outcomes
+where disturbance is high and actions to support recovery are not taken.
 
-There are a variety of possible actions intended to improve survival and reproduction of populations that are not otherwise self-sustaining [@mcnay_demographic_2022; @serrouya_predicting_2020].
-The simple models in this paper do not include these actions, and cannot be used to project outcomes for managed populations.
-Our intention is to reduce uncertainty that can be used to delay conservation action; there is nothing in our arguments, models or results that can or should be used to discourage much needed monitoring of the effects of conservation actions.
-Even when no additional monitoring is required for assessing self-sustainability, there are a variety of reasons to continue monitoring highly disturbed boreal caribou populations.
-Our point is that action is also required, and lack of monitoring should not be used as a reason to delay conservation measures for highly disturbed populations.
+There are a variety of possible actions intended to improve survival and
+reproduction of populations that are not otherwise self-sustaining
+[@mcnay_demographic_2022; @serrouya_predicting_2020]. The simple models
+in this paper do not include these actions, and cannot be used to
+project outcomes for managed populations. Our intention is to reduce
+uncertainty that can be used to delay conservation action; there is
+nothing in our arguments, models or results that can or should be used
+to discourage much needed monitoring of the effects of conservation
+actions. Even when no additional monitoring is required for assessing
+self-sustainability, there are a variety of reasons to continue
+monitoring highly disturbed boreal caribou populations. Our point is
+that action is also required, and lack of monitoring should not be used
+as a reason to delay conservation measures for highly disturbed
+populations.
 
-Variation in anthropogenic disturbance across the Canadian boreal region is partially correlated with variation in climate and primary productivity [@neilson_direct_2022] because forestry occurs in higher productivity areas, and petroleum extraction activities are concentrated in Western Canada.
-Effects of disturbance on vegetation and foodweb dynamics vary with primary productivity [@neufeld_primary_2021], and some have expressed concern about confounding of demographic-disturbance relationships [@wilson_causal_2021].
-One option for addressing these concerns would be to develop regional demographic-disturbance relationships for populations that are ecologically similar to one another (e.g.
-caribou inhabiting forests that are sufficiently productive for industrial forestry in Eastern Canada) [@eccc_scientific_2024], and use the priors from regional demographic-disturbance relationships to inform demographic projections within the region.
-However, while acknowledging and allowing for regional variation, we also need to guard against risks that uncertainty will be inflated and used as an excuse to delay action, a tactic that has been used in this and other decision contexts [@bissonnette_controverses_2020; @boan_climate_2018].
-In order to be informative and credible, regional demographic-disturbance analyses must include a sufficient number of observations spanning a range from high to low anthropogenic disturbance.
-Priors that lead to projections of outcomes for caribou that fall outside the observed range of variation across the country, characterized here (Figure 2) by 95% predictive intervals from a national analysis (Johnson et al.
-2020), are implausible.
-In other words, if we have seen no examples of self-sustaining caribou populations at a particularly high level of disturbance, it is unreasonable to project or expect this outcome.
-These premises constrain the plausible demographic-disturbance priors, and clarify that no additional information is needed to predict outcomes for caribou where anthropogenic disturbance is high (Figure 5) and no meaningful actions are taken to support recovery.
+Variation in anthropogenic disturbance across the Canadian boreal region
+is partially correlated with variation in climate and primary
+productivity [@neilson_direct_2022] because forestry occurs in higher
+productivity areas, and petroleum extraction activities are concentrated
+in Western Canada. Effects of disturbance on vegetation and foodweb
+dynamics vary with primary productivity [@neufeld_primary_2021], and
+some have expressed concern about confounding of demographic-disturbance
+relationships [@wilson_causal_2021]. One option for addressing these
+concerns would be to develop regional demographic-disturbance
+relationships for populations that are ecologically similar to one
+another (e.g. caribou inhabiting forests that are sufficiently
+productive for industrial forestry in Eastern Canada)
+[@eccc_scientific_2024], and use the priors from regional
+demographic-disturbance relationships to inform demographic projections
+within the region. However, while acknowledging and allowing for
+regional variation, we also need to guard against risks that uncertainty
+will be inflated and used as an excuse to delay action, a tactic that
+has been used in this and other decision contexts
+[@bissonnette_controverses_2020; @boan_climate_2018]. In order to be
+informative and credible, regional demographic-disturbance analyses must
+include a sufficient number of observations spanning a range from high
+to low anthropogenic disturbance. Priors that lead to projections of
+outcomes for caribou that fall outside the observed range of variation
+across the country, characterized here (Figure 2) by 95% predictive
+intervals from a national analysis (Johnson et al. 2020), are
+implausible. In other words, if we have seen no examples of
+self-sustaining caribou populations at a particularly high level of
+disturbance, it is unreasonable to project or expect this outcome. These
+premises constrain the plausible demographic-disturbance priors, and
+clarify that no additional information is needed to predict outcomes for
+caribou where anthropogenic disturbance is high (Figure 5) and no
+meaningful actions are taken to support recovery.
 
@@
 
-Another novel aspect of this work is a term that allows for bias associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012].
-Lacking information about the true magnitude of this bias, we examined sensitivity to a range of assumed values, and found that adding unknown bias did increase uncertainty, but these increases were not large (Supplement E).
-A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults and the probability of missing calves.
-Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment.
-Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out.
-We did not investigate sensitivity to the assumption that new recruits are 50% females, but we note that this is another possible source of error that would most likely lead to low recruitment estimates.
+Another novel aspect of this work is a term that allows for bias
+associated with misclassification of adults and missing calves in
+composition surveys [@ellington_divergent_2020;
+@decesare_estimating_2012]. Lacking information about the true magnitude
+of this bias, we examined sensitivity to a range of assumed values, and
+found that adding unknown bias did increase uncertainty, but these
+increases were not large (Supplement E). A key reason for this result is
+that we assumed the same range of values for the probability of
+misclassifying adults and the probability of missing calves. Presuming
+that surveyed groups contain more cows than bulls, and that there is an
+equal probability of misclassifying cows and young bulls and vice versa,
+then misclassification errors cause high estimated recruitment. Missing
+calves causes low estimated recruitment, and together these errors tend
+to cancel one another out. We did not investigate sensitivity to the
+assumption that new recruits are 50% females, but we note that this is
+another possible source of error that would most likely lead to low
+recruitment estimates.
 
-Given lack of knowledge about true error rates, the most appropriate interpretation of our results is that better information on composition survey error rates is required in order to assess their importance.
-Fecal pellets are proving to be an interesting source of information on sex and age ratios in surveyed groups [@mcfarlane_genetic_2018;@jones_novel_2023]; for example, in analysis of 14 populations of boreal and southern mountain caribou, Jones et al.
-2023 found that sampled groups contained more females than males, and that mothers tend to stay with their daughters and sons, while fathers do not.
-These results are broadly consistent with our assumption that composition survey groups include mothers and younger animals of both sexes that can be difficult to classify visually.
-Documenting animals that are difficult to classify as “unknown” can help analysts minimize and account for classification bias [@ellington_divergent_2020], as can comparing classification results from multiple independent observers.
-The probability of missing calves is more difficult to quantify and account for using aerial survey methods alone, but likely increases with vegetation cover.
+Given lack of knowledge about true error rates, the most appropriate
+interpretation of our results is that better information on composition
+survey error rates is required in order to assess their importance.
+Fecal pellets are proving to be an interesting source of information on
+sex and age ratios in surveyed groups [@mcfarlane_genetic_2018;
+@jones_novel_2023]; for example, in analysis of 14 populations of boreal
+and southern mountain caribou, Jones et al. 2023 found that sampled
+groups contained more females than males, and that mothers tend to stay
+with their daughters and sons, while fathers do not. These results are
+broadly consistent with our assumption that composition survey groups
+include mothers and younger animals of both sexes that can be difficult
+to classify visually. Documenting animals that are difficult to classify
+as “unknown” can help analysts minimize and account for classification
+bias [@ellington_divergent_2020], as can comparing classification
+results from multiple independent observers. The probability of missing
+calves is more difficult to quantify and account for using aerial survey
+methods alone, but likely increases with vegetation cover.
 
-Alternatives to aerial composition surveys for estimating recruitment include fecal pellet analysis [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022] and calf collaring [@ellington_divergent_2020].
-Each method has costs and benefits, and the reality is that monitoring everything everywhere all the time in great detail is neither feasible nor necessary.
-We hope that a bias correction term for the simple recruitment-mortality model [@hatter_moose_1991;@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020] that is used for many herd viability assessments will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is needed.
+Alternatives to aerial composition surveys for estimating recruitment
+include fecal pellet analysis [@mcfarlane_genetic_2018;
+@mcfarlane_spatial_2021; @mcfarlane_identification_2022] and calf
+collaring [@ellington_divergent_2020]. Each method has costs and
+benefits, and the reality is that monitoring everything everywhere all
+the time in great detail is neither feasible nor necessary. We hope that
+a bias correction term for the simple recruitment-mortality model
+[@hatter_moose_1991; @decesare_estimating_2012;
+@serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011;
+@johnson_science_2020] that is used for many herd viability assessments
+will enable and encourage practitioners to examine the sensitivity of
+their conclusions to plausible composition survey error rates, and help
+clarify thinking about where, when, and why additional information is
+needed.
 
@@
 
-The main objective of Johnson et al's [-@johnson_science_2020;@eccc_scientific_2011] national analysis was to inform a national disturbance policy that applies to all ranges except SK1, and to inform a different policy for SK1 [@eccc_scientific_2011;@eccc_boreal_2019].
-A hierarchical analysis [@leasure_hierarchical_2019] would be better suited for estimating variation in demographic model parameters among populations, and distinguishing that from variation among years.
-However, data owners have not authorized the use of the data for that purpose.
-In order to proceed with the available published information we constrained our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower than or higher than average remain lower or higher than average over time.
-The Bayesian model is not similarly constrained, and can accommodate evidence of a shift from unexpectedly low to unexpectedly high demographic rates over time, or vice versa.
-Switching from low to high each year is unlikely, but there is evidence of variation in the slope of demographic-disturbance relationships among populations [@rudolph_demographic_2017] so it is important that the Bayesian model allows for a greater range of possibilities than our more constrained simulated example trajectories.
-Although our simulated trajectories do not include all plausible possibilities we believe they are sufficient to demonstrate how a Bayesian model can be used to integrate local data with national priors, and to show how monitoring requirements depend on landscape condition.
+The main objective of Johnson et al's [-@johnson_science_2020;
+@eccc_scientific_2011] national analysis was to inform a national
+disturbance policy that applies to all ranges except SK1, and to inform
+a different policy for SK1 [@eccc_scientific_2011; @eccc_boreal_2019]. A
+hierarchical analysis [@leasure_hierarchical_2019] would be better
+suited for estimating variation in demographic model parameters among
+populations, and distinguishing that from variation among years.
+However, data owners have not authorized the use of the data for that
+purpose. In order to proceed with the available published information we
+constrained our example trajectory simulations by assuming that
+populations remain in their quantile, so populations that are lower than
+or higher than average remain lower or higher than average over time.
+The Bayesian model is not similarly constrained, and can accommodate
+evidence of a shift from unexpectedly low to unexpectedly high
+demographic rates over time, or vice versa. Switching from low to high
+each year is unlikely, but there is evidence of variation in the slope
+of demographic-disturbance relationships among populations
+[@rudolph_demographic_2017] so it is important that the Bayesian model
+allows for a greater range of possibilities than our more constrained
+simulated example trajectories. Although our simulated trajectories do
+not include all plausible possibilities we believe they are sufficient
+to demonstrate how a Bayesian model can be used to integrate local data
+with national priors, and to show how monitoring requirements depend on
+landscape condition.
 
-Aspatial demographic models generally assume that observed animals are a random and representative sample of a fully mixed population in which survival and recruitment rates apply equally to all animals.
-However, metrics derived from fecal pellet analysis such as reproductive success [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022], age structure and pregnancy rates [@flasko_fecal_2017;@mcfarlane_identification_2022], and mother-offspring co-occurence [@jones_novel_2023] suggest high variation in fitness among individuals [@mcfarlane_genetic_2018] and spatial structure within populations [@jones_novel_2023;@mcfarlane_spatial_2021;@mcfarlane_identification_2022].
-Analyses of movement data also show spatial variation in risks and benefits for caribou and variation in habitat preferences and space use among individuals [e.g.
-@decesare_linking_2014;@avgar_spaceuse_2015].
-Although there is evidence of spatial structure and variation in demographic rates among individuals within populations, for most caribou populations there is not enough data available to estimate spatial variation in demographic model parameters.
+Aspatial demographic models generally assume that observed animals are a
+random and representative sample of a fully mixed population in which
+survival and recruitment rates apply equally to all animals. However,
+metrics derived from fecal pellet analysis such as reproductive success
+[@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;
+@mcfarlane_identification_2022], age structure and pregnancy rates
+[@flasko_fecal_2017; @mcfarlane_identification_2022], and
+mother-offspring co-occurence [@jones_novel_2023] suggest high variation
+in fitness among individuals [@mcfarlane_genetic_2018] and spatial
+structure within populations [@jones_novel_2023;
+@mcfarlane_spatial_2021; @mcfarlane_identification_2022]. Analyses of
+movement data also show spatial variation in risks and benefits for
+caribou and variation in habitat preferences and space use among
+individuals [e.g. @decesare_linking_2014; @avgar_spaceuse_2015].
+Although there is evidence of spatial structure and variation in
+demographic rates among individuals within populations, for most caribou
+populations there is not enough data available to estimate spatial
+variation in demographic model parameters.
 
-A number of additional complexities can be incorporated into aspatial Bayesian population models, including variation in demographic rates among individuals  [@armstrong_incorporating_2021], population size information from aerial or fecal dna surveys, and additional age classes [@moeller_integrating_2021; @mcnay_demographic_2022].
-However, in many contexts, population size and fecal dna survey information is not available, and a framework that includes these sources of information will be unnecessarily complicated.
-At present, our model is intended as a proof-of-concept to show the value of integrating national demographic-disturbance relationships with local data, accounting for composition survey errors, and assessing the value of local monitoring information.
-Rather than further complicate a proof-of-concept model applied to simulated data, our next step will be to collaborate with an interested decision maker to develop a tool informed by this analysis that meets their real local needs, recognizing that these needs will vary among monitoring programs [@nbckc_nbckc_2021;@nbckc_nbckc_2021].
+A number of additional complexities can be incorporated into aspatial
+Bayesian population models, including variation in demographic rates
+among individuals [@armstrong_incorporating_2021], population size
+information from aerial or fecal dna surveys, and additional age classes
+[@moeller_integrating_2021; @mcnay_demographic_2022]. However, in many
+contexts, population size and fecal dna survey information is not
+available, and a framework that includes these sources of information
+will be unnecessarily complicated. At present, our model is intended as
+a proof-of-concept to show the value of integrating national
+demographic-disturbance relationships with local data, accounting for
+composition survey errors, and assessing the value of local monitoring
+information. Rather than further complicate a proof-of-concept model
+applied to simulated data, our next step will be to collaborate with an
+interested decision maker to develop a tool informed by this analysis
+that meets their real local needs, recognizing that these needs will
+vary among monitoring programs [@nbckc_nbckc_2021; @nbckc_nbckc_2021].
 
@@
 
-We applied the framework proposed by Dunham et al.
-(2023) to show that no additional local monitoring information is needed to predict whether boreal caribou populations in highly disturbed landscapes are self-sustaining.
-There is more variation in caribou demographic rates among study areas with low to moderate disturbance, and thus more need for local data to inform demographic projections.
-In these places, our analysis suggests a general pattern of diminishing returns from increasing monitoring effort, but more information on monitoring costs, objectives, and existing local data would be required to assess the adequacy and cost-effectiveness of monitoring scenarios in particular circumstances.
-Monitoring costs vary substantially with technologies used, the remoteness of herds, and the difficulty of collaring animals, which in turn depends on the number of currently collared animals, the population size, and vegetation across the landscape [@nbckc_boreal_2021].
-It is important to recognize that projecting population viability is only one possible objective for monitoring [@nbckc_boreal_2019], and that the impacts of monitoring on threatened animals should be considered [@nbckc_boreal_2019].
-Since monitoring costs, options and goals are context specific, we believe the most effective way forward will be to use this proof-of-concept model and analysis as a starting point for working with local decision makers to build decision support tools suited to their particular circumstances, needs and questions.
+We applied the framework proposed by Dunham et al. (2023) to show that
+no additional local monitoring information is needed to predict whether
+boreal caribou populations in highly disturbed landscapes are
+self-sustaining. There is more variation in caribou demographic rates
+among study areas with low to moderate disturbance, and thus more need
+for local data to inform demographic projections. In these places, our
+analysis suggests a general pattern of diminishing returns from
+increasing monitoring effort, but more information on monitoring costs,
+objectives, and existing local data would be required to assess the
+adequacy and cost-effectiveness of monitoring scenarios in particular
+circumstances. Monitoring costs vary substantially with technologies
+used, the remoteness of herds, and the difficulty of collaring animals,
+which in turn depends on the number of currently collared animals, the
+population size, and vegetation across the landscape
+[@nbckc_boreal_2021]. It is important to recognize that projecting
+population viability is only one possible objective for monitoring
+[@nbckc_boreal_2019], and that the impacts of monitoring on threatened
+animals should be considered [@nbckc_boreal_2019]. Since monitoring
+costs, options and goals are context specific, we believe the most
+effective way forward will be to use this proof-of-concept model and
+analysis as a starting point for working with local decision makers to
+build decision support tools suited to their particular circumstances,
+needs and questions.
 

### 2024-08-27 analysis/paper/paper.Rmd
update affiliation address

@@
 institute:
-  - ECCC: Environment and Climate Change Canada 
+  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
 date: "`r format(Sys.time(), '%d %B, %Y')`"

### 2024-08-27 analysis/scripts/viewSensitivityResults.R
capitalize axis lab

@@ -196,7 +196,7 @@ for(i in 1:length(pages)){
         height = 4, width = 7.48, units = "in",res=600)
     base=ggplot(subset(probs,pageLabB==pp),aes(x=obsYears,y=Mean,col=CorrectStatus))+geom_point(shape="-",size=3,alpha=0.5)+
       facet_grid(YearsOfProjection~AnthroScn,labeller="label_both")+
-      theme_bw()+xlab("years of monitoring")+ylab("Estimated mean population growth rate")
+      theme_bw()+xlab("Years of monitoring")+ylab("Estimated mean population growth rate")
     print(base)
     dev.off()
 
@@ -204,7 +204,7 @@ for(i in 1:length(pages)){
         height = 4, width = 7.48, units = "in",res=600)
     base=ggplot(subset(probs,pageLabB==pp),aes(x=obsYears,y=trueMean,col=CorrectStatus))+geom_point(shape="-",size=3,alpha=0.5)+
       facet_grid(YearsOfProjection~AnthroScn,labeller="label_both")+
-      theme_bw()+xlab("years of monitoring")+ylab("True population growth rate")
+      theme_bw()+xlab("Years of monitoring")+ylab("True population growth rate")
     print(base)
     dev.off()
 
@@ -225,7 +225,7 @@ for(i in 1:length(pages)){
     base=ggplot(subset(probs,(pageLab==pp)&(RenewalInterval==1)),aes(x=as.factor(obsYears),y=LambdaDiff,col=NumCollars,fill=NumCollars,group=grp))+
       geom_violin(alpha=0.5)+ylim(-0.15,0.15)+
       facet_grid(YearsOfProjection~AnthroScn,labeller="label_both")+labs(color="Number of\n Collars",fill="Number of\n Collars")+
-      theme_bw()+xlab("years of monitoring")+ylab("Difference between true growth rate and posterior mean")+
+      theme_bw()+xlab("Years of monitoring")+ylab("Difference between true growth rate and posterior mean")+
       scale_color_discrete(type=(pal4))+scale_fill_discrete(type=(pal4))
     print(base)
     dev.off()
@@ -308,7 +308,7 @@ for(i in 1:length(pages)){
         height = 4, width = 7.48, units = "in",res=600)
     base=ggplot(subset(probsSum,pageLabC==pp),aes(x=obsYears,y=1-propWrong,col=NumCollars,linetype=RenewalInterval,group=grp))+geom_line()+
[…]

### 2024-08-27 analysis/paper/paper.Rmd
fixed equations

@@
 potential recruitment of female calves per cow $\dot{X}_t$:
+
 \begin{equation}
@@
   (\#eq:dotX)
-\end{equation} where $w$ is the ratio of observed adult females to
+\end{equation}
+
+where $w$ is the ratio of observed adult females to
 collared animals, $\dot{q}$ is the ratio of young bulls to adult
@@
 distributed function of the number of surviving post-juvenile females
-and the adjusted recruitment rate: \begin{equation}
+and the adjusted recruitment rate:
+
+\begin{equation}
   \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{N_0k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
@@
 The post-juvenile female population in the next year includes both
-survivors and new recruits: \begin{equation}
+survivors and new recruits:
+
+\begin{equation}
 \dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t).
@@
 
-```{=tex}
 \begin{equation}
@@
 \end{equation}
-```
-```{=tex}
+
 \begin{equation}
@@
 \end{equation}
-```
+
 $\phi^R \sim \text{Normal}(19.862,2.229)$ and
@@
 estimated recruitment probability is adjusted to get expected female
-recruits per female: \begin{equation}
+recruits per female: 
+\begin{equation}
 X_t=\frac{cR_t/2}{1+cR_t/2}; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
@@
 to account for stochastic variation among years
-[@eacker_web-based_2019]: \begin{equation}
+[@eacker_web-based_2019]: 
+\begin{equation}
 S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).
@@
 individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed
-[@schaub_integrated_2021]: \begin{equation}
+[@schaub_integrated_2021]: 
+\begin{equation}
 \hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).
@@
 the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and
-priors from the national model: \begin{equation}
+priors from the national model:
+\begin{equation}
 EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
   (\#eq:EVcurrent)
-\end{equation} where $P(\psi|\tau)$ is the prior probability that the
+\end{equation}
+where $P(\psi|\tau)$ is the prior probability that the
 true state of the population is $\psi$, and each population is assigned
@@
 as the difference between the expected value with and without additional
-sample information: \begin{equation}
+sample information: 
+\begin{equation}
 EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - EV_{current},
   (\#eq:EVSI)
-\end{equation} where $x$ is the sample information from a particular
+\end{equation}
+where $x$ is the sample information from a particular
 example trajectory. In this study, we treat simulated observations from

### 2024-08-27 analysis/paper/paper.Rmd

@@
 information is the average utility of decisions informed by knowledge of
-the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and
+the disturbance scenario $\tau$ (Figure \@ref(fig:figDisturbance)) and
 priors from the national model:
@@
 priors from national demographic-disturbance relationships to reduce
-uncertainty in boreal caribou demographic projections (Figs
+uncertainty in boreal caribou demographic projections (Figures
 \@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected,
@@
 effort in each year (i.e. number of collars) or the number of years of
-monitoring (Fig \@ref(fig:figDiffs)) increase. Uncertainty is higher
+monitoring (Figure \@ref(fig:figDiffs)) increase. Uncertainty is higher
 when the forecast horizon is long (compare 5 and 20 year projections in
-Fig \@ref(fig:figDiffs)). Uncertainty is also higher in our high
-anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because
+Figure \@ref(fig:figDiffs)). Uncertainty is also higher in our high
+anthropogenic disturbance scenarios (Figure \@ref(fig:figDiffs)) because
 stochastic variation in population growth rates increase as populations
-become very small (as in top right panel of Fig
+become very small (as in top right panel of Figure
 \@ref(fig:examplePredictions)).
@@
 Our analysis becomes more informative when we examine the probability of
-correctly assessing population status (Fig \@ref(fig:figStatusErrors))
-and the expected value of information (Figs \@ref(fig:figEVSI) and E.1).
-When anthropogenic disturbance is high (right panels of Figs
+correctly assessing population status (Figure \@ref(fig:figStatusErrors))
+and the expected value of information (Figures \@ref(fig:figEVSI) and E.1).
+When anthropogenic disturbance is high (right panels of Figures
 \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
@@
 Interestingly, when the projection period is long and anthropogenic
-disturbance is high, we see a negative EVSI (lower right panel of Fig
+disturbance is high, we see a negative EVSI (lower right panel of Figure
 \@ref(fig:figEVSI)), and we do not see an associated decrease in the
 probability of correctly assessing population status (lower right panel
-of Fig \@ref(fig:figStatusErrors)). This means that the long-term
+of Figure \@ref(fig:figStatusErrors)). This means that the long-term
 projections informed by local monitoring data are on average less
@@
 prior probability that population is viable (orange band in the top
-right panel of Fig \@ref(fig:examplePredictions) falls below 1). At
+right panel of Figure \@ref(fig:examplePredictions) falls below 1). At
 lower levels of disturbance, local monitoring can reduce uncertainty,
 but there are diminishing returns as the number of collars deployed or
-the number of years of monitoring increases (Figs
+the number of years of monitoring increases (Figures
 \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the

### 2024-08-27 analysis/paper/paper.Rmd
add author contributions, number sections

@@
 
+# Author contributions
+**Josie Hughes:** Conceptualization, Formal analysis, Methodology, Investigation,
+Supervision, Software, Writing – original draft, Writing – review and editing, 
+Visualization
+**Sarah Endicott:** Resources, Software, Writing – review and editing
+**Anna M Calvert:** Conceptualization, Writing – original draft, Writing – review and editing
+**Cheryl A Johnson:** Conceptualization, Writing – review and editing
+
 <!-- The following line inserts a page break  -->

### 2024-08-30 analysis/paper/paper.Rmd
Switch to beta distributed interannual variation, and other changes.

@@
 runScenarios = F #To do: set up to run full analysis from within this document.
-doPriorCalibration=F #Note the prior calibration is time consuming.
 
@@
 figSet = c("diffs","distScns","EVSI","power","EVPIall")
+
+if(runScenarios){
+    source(here::here("analysis/scripts/0_nationalModelTrajectories.R"))
+    source(here::here("analysis/scripts/0_bayesianModelExamples.R"))
+    #create scenarios with sensitivity1.R and run in cloud
+    #run viewSensitivityResults.R for S8 and S7
+    #run EVSIPlots.R
+}
 
@@
 }
-
 ```
@@
 @novomestky_package_2016]:
-$\dot{R}_t \sim \text{TruncatedBeta}(\bar{R}_t,\nu_R,l_R,h_R); \dot{S}_t \sim \text{TruncatedBeta}(\bar{S}_t,\nu_S,l_S,h_S)$.
-Coefficients of variation among years $(\nu_R,\nu_S)$ and
+$\dot{R}_t \sim \text{TruncatedBeta}(\bar{R}_t,v_R,l_R,h_R); \dot{S}_t \sim \text{TruncatedBeta}(\bar{S}_t,v_S,l_S,h_S)$.
+Coefficients of variation among years $(v_R,v_S)$ and
 maximum/minimum values $l_R,h_R,l_S,h_S$ for recruitment and survival
@@
 ```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction.
+To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variability ($v_S,v_R$) by 75\\% in these examples.
 Orange dots show means of 500 example trajectories.
@@
 is estimated from the observed number of calves $\hat{J}_t$ and adult
-female caribou $\hat{W}_t$ assuming a Poisson distribution:
-\begin{equation}
-\hat{J}_t \sim \text{Poisson}(\hat{W}_t,R_t).
+female caribou $\hat{W}_t$ assuming a Binomial distribution:
+\begin{equation}
+\hat{J}_t \sim \text{Binomial}(\hat{W}_t,R_t).
   (\#eq:hatJEst)
@@
 In both models recruitment probability is a function of anthropogenic
-disturbance and fire with a log link [@johnson_science_2020;
-@stewart_climate-informed_2023; @dyson_existing_2022]. In the Bayesian
-model we add a Gaussian distributed random year effect to account for
-stochastic variation among years [@eacker_web-based_2019]:
-\begin{equation}
-\log(R_t)=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t+\epsilon^R_t; \epsilon^R_t \sim \text{Normal}(0,\sigma^2_{R}).
+disturbance and fire with a log link and Beta distributed interannual variation [@johnson_science_2020;
+@stewart_climate-informed_2023; @dyson_existing_2022]. The coefficient of variation $\nu_R$ is constant, while the mean $\tilde{R_t}$ and standard deviation $\sigma_{R,t}$ vary with disturbance:
+\begin{equation}
+log(\tilde{R_t})=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t
+  (\#eq:Rtilde)
+\end{equation}
+
+\begin{equation}
+R_t \sim \text{Beta}(\tilde{R_t},\sigma_{R,t});
+\sigma_{R,t} = min(\nu_R \tilde{R_t},\sqrt{\tilde{R_t} (1-\tilde{R_t})})
 (\#eq:R)
@@
 
-For consistency with [@johnson_science_2020], we model survival
+For consistency with Johnson et al. [-@johnson_science_2020], we model survival
 probability as an adjusted function of buffered anthropogenic
-disturbance $A_t$ with a log link [@stewart_climate-informed_2023;
-@dyson_existing_2022]. We add a Gaussian distributed random year effect
-to account for stochastic variation among years
-[@eacker_web-based_2019]: 
-\begin{equation}
-S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).
+disturbance $A_t$ with a log link and Beta distributed interannual variation with coefficient of variation $\nu_S$, mean $\tilde{S_t}$ and standard deviation $\sigma_{S,t}$ [@stewart_climate-informed_2023;
+@dyson_existing_2022]: 
+\begin{equation}
+\tilde{S_t}=(46 e^{\beta^S_0+\beta^S_a A_t}-0.5)/45
+  (\#eq:Stilde)
+\end{equation}
+
+\begin{equation}
+S_t \sim \text{Beta}(\tilde{S_t},\sigma_{S,t});
+\sigma_{S,t} = min(\nu_S \tilde{S_t},\sqrt{\tilde{S_t} (1-\tilde{S_t})})
 (\#eq:S)
@@
 wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
-individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed
-[@schaub_integrated_2021]: 
+individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed: 
 \begin{equation}
@@
 @yokota_value_2004]. We assign a value or utility of
-$\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$
-matches the assigned status $\alpha$, and value of $\nu_{\alpha,\psi}=0$
+$\eta_{\alpha,\psi}=1$ when the true state of the population $\psi$
+matches the assigned status $\alpha$, and value of $\eta_{\alpha,\psi}=0$
 when the status assessment is incorrect. This is "the value ascribed by
@@
 \begin{equation}
-EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
+EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\eta_{\alpha,\psi}],
   (\#eq:EVcurrent)
@@
 \begin{equation}
-EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - EV_{current},
+EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\eta_{\alpha,\psi}]] - EV_{current},
   (\#eq:EVSI)

### 2024-08-30 analysis/paper/supplement.Rmd
ongoing
Switch to beta distributed interannual variation, and other changes.

@@
 |$\beta^R_f$||Fire slope for recruitment.|
-|$\sigma^2_{R}$||Variance in random effect of year on recruitment.|
+|$\nu_{R}$||Coefficient of variation in recruitment among years.|
 |$\beta^S_0$||Survival intercept.|
 |$\beta^S_a$||Anthropogenic disturbance slope for survival.|
-|$\sigma^2_{S}$||Variance in random effect of year on survival.|
+|$\nu_{S}$||Coefficient of variation in survival among years.|
 |**True state**|
@@
 |$r_{max}=1.3$||Maximum population growth rate.|
-|$\nu_R=0.46$||Coefficient of variation in $\dot{R}_t$ among years.|
-|$\nu_S=0.08696$||Coefficient of variation in $\dot{S}_t$ among years.|
+|$v_R=0.46$||Coefficient of variation in $\dot{R}_t$ among years.|
+|$v_S=0.08696$||Coefficient of variation in $\dot{S}_t$ among years.|
 |$l_R=0$||Minimum value for $\dot{R}_t$.|
-|$l_S=0$||Minimum value for $\dot{S}_t$.|
-|$h_R=0$||Maximum value for $\dot{R}_t$.|
-|$h_S=0$||Maximum value for $\dot{S}_t$.|
+|$l_S=0.61$||Minimum value for $\dot{S}_t$.|
+|$h_R=0.82$||Maximum value for $\dot{R}_t$.|
+|$h_S=1$||Maximum value for $\dot{S}_t$.|
 |**Observation model parameters**|
@@
 |$\alpha=\text{viable or not}$|binary|Assigned population status.|
-|$\nu_{\alpha,\psi}=1 \text{ when } \alpha=\psi, 0 \text{ otherwise}$|binary|Value of classification decision.|
+|$\eta_{\alpha,\psi}=1 \text{ when } \alpha=\psi, 0 \text{ otherwise}$|binary|Value of classification decision.|
 |$x$||Simulated observations from an example population.|
@@
 
-The choice of priors for the random effects of year on survival and recruitment are less important because the random effects have no effect on expected population growth rate (Figs \@ref(fig:PriorAllFits) and \@ref(fig:PriorAllFitsR)).
-We set the hyperprior for the standard deviation of the survival random effect to vary uniformly between sSigmaMin = 0.00001 and sSigmaMax = 0.06 to exceed the variation among national model simulations by a moderate amount (Fig \@ref(fig:PriorAllFits)).
-We set the hyperprior for the standard deviation of the recruitment random effect to vary uniformly between rSigmaMin = 0.00001 and rSigmaMax = 0.2 to exceed the variation among national model simulations by a moderate amount (Fig \@ref(fig:PriorAllFitsR)).
+The choice of priors for the random effects of year on survival and recruitment are less important because the random effects have little impact on expected population growth rate (Figs \@ref(fig:PriorAllFits) and \@ref(fig:PriorAllFitsR)).
+We set the prior for the coefficient of variation of survival among years $\nu_S$ to vary uniformly between sNuMin = 0.001 and sNuMax = 0.13 to match the variation among national model simulations (Fig \@ref(fig:PriorAllFits)).
+We set the prior for the coefficient of variation of recruitment of among years $\nu_S$ to vary uniformly between rNuMin = 0.001 and rNuMax = 0.7 to match the variation among national model simulations (Fig \@ref(fig:PriorAllFitsR)).
 
@@
 if(doPriorCalibration){
-  source(here::here("analysis/scripts/PriorsFromPIs.R"))
-  source(here::here("analysis/scripts/calibratePriors.R"))
+  source(here::here("analysis/scripts/0_PriorsFromPIs.R"))
+  source(here::here("analysis/scripts/0_calibratePriors.R"))
 }

### 2024-08-30 R/CaribouDemographicFunctions.R

@@ -50,7 +50,7 @@ makeInterceptPlots <- function(scResults, addBit = "", facetVars = NULL,
                                               "Recruitment",
                                               "Female population size"),
                                survLow = 0.6, type = "png", useNational = T,lowLine=NULL,highLine=NULL) {
-  # facetVars=NULL;loopVars="sSigmaMax";scResults=scResults
+  # facetVars=NULL;loopVars="sNuMax";scResults=scResults
   #useNational=T;type="png"; survLow=0.6
   if (!is.null(loopVars)) {
     loopSet <- unique(subset(scResults$rr.summary.all, select = loopVars))

### 2024-08-30 analysis/scripts/0_PriorsFromPIs.R

@@ -1,0 +1,119 @@
+#devtools::load_all(here::here())
+
+# Get some priors from PIs of Johnson et al regression
+library(caribouMetrics)
+library(tidyverse)
+library(ggpubr)
+theme_set(theme_bw())
+library(RColorBrewer)
+
+dia_shp <- 23
+
+err_col <- "grey50"
+
+baseDir <- "."
+
+#numbers from Johnson et al for validation
+johnsonCompare <- read.csv(paste0(baseDir,"/data/Johnson et al. figures5_6.csv"))
+
+getAnswers <-function(scns,par,johnsonCompare,vName){
+  scResults = runScenario(scns,survAnalysisMethod = "Exponential")
+  rr = subset(scResults$rr.summary.all,Parameter == par)
+  rr$ylow = rr[["Lower 95% CRI"]]
+  rr$yhigh = rr[["Upper 95% CRI"]]
+  d = unique(rr$iAnthro)
+  if(par=="Recruitment"){
+    rr$lowLine =johnsonCompare$Rlow_95_pre[johnsonCompare$anthro==d]/100
+    rr$highLine = johnsonCompare$Rhigh_95_pred[johnsonCompare$anthro==d]/100
+    rr$meanLine = johnsonCompare$Rresp[johnsonCompare$anthro==d]/100
+  }else{
+    rr$lowLine =johnsonCompare$Slow_95_pred[johnsonCompare$anthro==d]
+    rr$highLine = johnsonCompare$Shigh_95_pred[johnsonCompare$anthro==d]
+    rr$meanLine = johnsonCompare$Sresp[johnsonCompare$anthro==d]
+  }
+  rr$vv = rr[[vName]]
+  pp1 = ggplot(rr,aes(x=vv,y=Mean,ymin=ylow,ymax=yhigh))+
+    geom_line(col="#EF8A62")+geom_ribbon(fill="#EF8A62",alpha=0.2)+
+    geom_hline(yintercept=unique(rr$lowLine),colour="black")+
+    geom_hline(yintercept=unique(rr$highLine),colour="black")+
+    geom_hline(yintercept=unique(rr$meanLine),colour="black",linetype="dashed")+
+    labs(title=paste0(d,"% anthropogenic disturbance"),y=par,x=vName)
+  #plot(pp1)
+  return(list(data=rr,plot=pp1))
+}
+
+rrAll=list()
[…]

### 2024-08-30 analysis/scripts/0_bayesianModelExamples.R

@@ -1,0 +1,198 @@
+# Prior and posterior predictions from Bayesian model
+library(caribouMetrics)
+library(tidyverse)
+library(ggpubr)
+theme_set(theme_bw())
+library(RColorBrewer)
+
+baseDir <- "."
+
+monitoringScns = data.frame(obsYears=c(1,16),collarCount=c(0,30),cowMult=c(6),collarInterval=c(1),
+                            assessmentYrs=c(3))
+stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
+stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
+
+stateScns$sQuantile=0.8
+stateScns$rQuantile = 0.8
+scns=merge(monitoringScns,stateScns)
+
+scns$iAnthro = 0
+scns$tA = scns$iAnthro+(scns$obsYears)*scns$obsAnthroSlope
+scns$projYears = 50-scns$obsYears
+scns$N0 = 5000
+scns$adjustR = TRUE
+#scns$assessmentYrs=3
+
+####################
+eParsIn = list()
+eParsIn$cowCounts <- data.frame(Year = 1981:2023,
+                                Count = 100,
+                                Class = "cow")
+eParsIn$freqStartsByYear <- data.frame(Year = 1981:2023,
+                                       numStarts = 30)
+eParsIn$collarOnTime=1
+eParsIn$collarOffTime=12
+eParsIn$collarNumYears=1
+labFontSize = 10; breakInterval=5
+
+###############################
+#With bias adjustment
+simBig<-getSimsNational(replicates=3000,forceUpdate=T)
+
+priorResult = caribouMetrics:::runScnSet(scns[1,],eParsIn,simBig,getKSDists=F,printProgress=F)
+priorResult$obs.all=NULL
+yr_scale1 <- scale_x_continuous(breaks = 2023:2072 %>% .[0:9*5+1],
+                                labels = 2009:2058 %>% .[0:9*5+1] )
+yr_scale2 <- scale_x_continuous(breaks = 2009:2058 %>% .[0:9*5+1],
+                                labels = 2009:2058 %>% .[0:9*5+1] )
+recPrior =  plotRes(priorResult, "Recruitment", lowBound=0, highBound = 0.85,
+                    legendPosition="none",breakInterval=breakInterval,
+                    labFontSize=labFontSize) +
+  yr_scale1 +
+  labs(tag = "a")
+plot(recPrior)
[…]

### 2024-08-30 analysis/scripts/0_calibratePriors.R

@@ -1,0 +1,69 @@
+#devtools::load_all(here::here())
+library(caribouMetrics)
+scnsApplyAll = data.frame(qMin=0,qMax=0,uMin=0,uMax=0,zMin=0,zMax=0,adjustR=F)
+scnsNoData = data.frame(obsYears=1,collarCount=0,cowMult=0,projYears=4)
+scnsLow = data.frame(obsYears=15,collarCount=60,cowMult=9,projYears=1,rQuantile=0.025,sQuantile=0.025)
+scnsHigh = scnsLow;scnsHigh$rQuantile =0.975;scnsHigh$sQuantile=0.975
+########################
+#Survival random effects
+#vary sIntSE, sNuMin, and sNuMax in factorial array.
+sIntSE=getPriors()$l.Saf.Prior2;sNuMin=0.01;sNuMax=seq(0.07,0.15,by=0.01)
+
+#Looking for low KS distance from full range of input sims when given almost no info
+scnsSInt=expand.grid(obsAnthroSlope = 0, projAnthroSlope = 5,sIntSE=sIntSE,
+                     sNuMin=sNuMin,sNuMax=sNuMax)
+scnsSInt$NuMin = scnsSInt$sNuMin
+scns=merge(scnsSInt,scnsApplyAll);scns=merge(scns,scnsNoData)
+scns$projYears=20
+scResults = runScenario(scns,survAnalysisMethod = "Exponential")
+KSAll = subset(scResults$ksDists,(Year==2023+scnsNoData$projYears)&(Parameter=="Adult female survival"))
+addBit = paste0("InterceptAll")
+scResults$obs.all = NULL
+makeInterceptPlots(scResults,addBit=addBit,facetVars=c("sNuMax"),loopVars="sIntSE",whichPlots=c("Adult female survival"))
+
+########################
+#Recruitment random effects
+#vary rIntSE, rNuMin, and rNuMax in factorial array.
+#devtools::load_all(here::here())
+rIntSE=getPriors()$l.R.Prior2;rNuMin=0.01;rNuMax=seq(0.1,0.9,by=0.1)
+
+#Looking for low KS distance from full range of input sims when given almost no info
+scnsRInt=expand.grid(obsAnthroSlope = 0, projAnthroSlope = 5,rIntSE=rIntSE,
+                     rNuMin=rNuMin,rNuMax=rNuMax)
[…]

### 2024-08-30 analysis/scripts/0_nationalModelTrajectories.R

@@ -1,0 +1,256 @@
+# Example trajectories from national model
+library(caribouMetrics)
+library(tidyverse)
+library(ggpubr)
+theme_set(theme_bw())
+library(RColorBrewer)
+
+dia_shp <- 23
+
+err_col <- "grey50"
+
+baseDir <- "."
+
+#numbers from Johnson et al for validation
+johnsonCompare <- read.csv(paste0(baseDir,"/data/Johnson et al. figures5_6.csv"))
+
+# get outputs from demographics
+covTableSim <- data.frame(Anthro = c(0, 10, 20, 30, 40, 50,
+                                     60, 70, 80, 90),
+                          Fire = 0,
+                          fire_excl_anthro = 0)
+covTableSim$polygon <- paste0("Missisa_", covTableSim$Anthro + 1)
+covTableSim$area <- "FarNorth"
+covTableSim$Total_dist <- covTableSim$Anthro + covTableSim$fire_excl_anthro
+
+popGrowthPars <- demographicCoefficients(500,
+  modelVersion = "Johnson",
+  survivalModelNumber = "M1",
+  recruitmentModelNumber = "M4"
+)
+
+popGrowthParsSmall <- demographicCoefficients(35,
+  modelVersion = "Johnson",
+  survivalModelNumber = "M1",
+  recruitmentModelNumber = "M4"
+)
+
+rateSamples <- demographicRates(
+  covTable = covTableSim,
+  popGrowthPars = popGrowthParsSmall,
+  ignorePrecision = F, returnSample = T, useQuantiles = T
+)
+
+rateSamplesLarge <- demographicRates(
+  covTable = covTableSim,
+  popGrowthPars = popGrowthPars,
+  ignorePrecision = F, returnSample = T, useQuantiles = T
+)
+
+rateSummaries <- demographicRates(
+  covTable = covTableSim, popGrowthPars = popGrowthPars,
+  ignorePrecision = F, returnSample = F, useQuantiles = F
+)
+
+#for each sample, select bias correction term from distributions
+
+str(popGrowthPars)
+
+cr = data.frame(replicate=unique(rateSamples$replicate))
+nr= nrow(cr)
[…]

### 2024-08-30 analysis/scripts/1_sensitivity1.R

@@ -1,0 +1,71 @@
+#app_path <- "C:/Users/HughesJo/Documents/gitprojects/BayesianCaribouDemographicProjection"
+#devtools::load_all(here::here())
+
+########################
+#sensitivity
+setName = "s7"
+monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
+                             cowMult=c(3,6,9),collarInterval=c(1,4),
+                             assessmentYrs=c(3))
+#TO DO - in next iteration, remove multiple years of one collar
+monitoringScns = subset(monitoringScns, !((obsYears>1)&(collarCount==0))&!((collarCount==1)&(cowMult>3))&!((collarCount==0)&(collarInterval>1)))
+stateScns = data.frame(tA=c(0,20,40,60,0,20,40,60,60,40,20),
+                       obsAnthroSlope=c(0,1,1,1,0,0,0,0,0,0,0),
+                       projAnthroSlope=c(1,1,1,1,0,0,0,0,-1,-1,-1))
+stateScns = merge(stateScns,data.frame(rep=seq(1:500)))
+
+stateScns$sQuantile=runif(nrow(stateScns),min=0.01,max=0.99)
+stateScns$rQuantile = runif(nrow(stateScns),min=0.01,max=0.99)
+#monitoringScns=rbind(monitoringScns,minimalScn)
+scns=merge(monitoringScns,stateScns)
+
+scns$preYears = max(scns$obsYears)-scns$obsYears
+
+scns$iAnthro = scns$tA-(scns$obsYears+scns$preYears-1)*scns$obsAnthroSlope
+scns$projYears = 20
+unique(scns$iAnthro)
+scns$repBatch = ceiling(scns$rep/50)
+table(scns$repBatch)
+scns$N0 = 5000
+scns$adjustR = TRUE
+
+scns$pageLab = paste0("cmult",scns$cowMult,"ay",scns$assessmentYrs,"aSf",scns$projAnthroSlope,"repBatch",scns$repBatch)
+scns$pageId = as.numeric(as.factor(scns$pageLab))
+
+write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
+pages=unique(scns$pageLab)
+
+setName="s9"
+scns = subset(scns,(rep==1)&(collarCount==30)
[…]

### 2024-08-30 analysis/scripts/2_viewSensitivityResults.R

@@ -1,0 +1,351 @@
+#devtools::load_all(here::here())
+library(tidyverse)
+library(ggplot2)
+library(caribouMetrics)
+library(RColorBrewer)
+
+pal2 = brewer.pal(7,"RdBu")[c(2,6)]
+pal4 = brewer.pal(5,"RdBu")[c(1,2,4,5)]
+
+pal4b = brewer.pal(5,"Purples")[c(2,3,4,5)]
+
+scn_defaults <- eval(formals(getScenarioDefaults))
+
+########################
+#sensitivity
+setName = "s8" #need to do both s7 and s8
+dir.create(paste0("figs/",setName),recursive=T)
+scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
+
+
+table(scns$pageId)
+
+nrow(scns)
+pagesa=unique(scns$pageId)
+
+pages = pagesa
+for (p in pagesa){
+  #p=pagesa[1]
+  if(!file.exists(paste0("results/",setName,"/rTest",p,".Rds"))){pages=pages[pages!=p]}
+}
+
+batchStrip<-function(l,batches=c(10,seq(1:9))){
+  for(b in batches){
+    l=gsub(paste0("repBatch",b),"",l,fixed=T)
+  }
+  return(l)
+}
+
+pages=sort(pages)
+
+combine=T
+addEV = F
+addProbs=F
+for(i in 1:length(pages)){
+  #combine=F;i=50
+  cpageId=pages[i]
+
+  if(i==length(pages)){
+    nextP = "TrepBatch1"
+  }else{
+    nextP = unique(scns$pageLab[scns$pageId==pages[i+1]])
+  }
+
+  p = unique(scns$pageLab[scns$pageId==cpageId])
+
+  print(paste(i,p))
+
+  if(combine&(cpageId>1)){
+    scNew = readRDS(paste0("results/",setName,"/rTest",cpageId,".Rds"))
+    for(n in names(scNew)){
+      if(n=="errorLog"){
+        scResults[[n]]=c(scResults[[n]],scNew[[n]])
+      }else{
+        scResults[[n]]=rbind(scResults[[n]],scNew[[n]])
+      }
+    }
+  }else{
+    scResults = readRDS(paste0("results/",setName,"/rTest",cpageId,".Rds"))
+  }
+
[…]

### 2024-08-30 analysis/scripts/3_EVSIPlots.R

@@ -1,0 +1,95 @@
+########################
+#EVSI: see Dunham et al. 2023 and references therein.
+#Run after viewSensitivityResults.R
+
+library(tidyverse)
+library(ggplot2)
+library(caribouMetrics)
+library(RColorBrewer)
+
+pal2 = brewer.pal(7,"RdBu")[c(2,6)]
+pal4 = brewer.pal(5,"RdBu")[c(1,2,4,5)]
+
+pal4b = brewer.pal(5,"Purples")[c(2,3,4,5)]
+
+scn_defaults <- eval(formals(getScenarioDefaults))
+
+
+batchStrip<-function(l,batches=c(10,seq(1:9))){
+  for(b in batches){
+    l=gsub(paste0("repBatch",b),"",l,fixed=T)
+  }
+  return(l)
+}
+
+setName = "s8"
+probsSum = read.csv(here::here(paste0("tabs/EVsample",setName,".csv")),stringsAsFactors = F)
+
+probsSum$pageLabC = batchStrip(probsSum$pageLab)
+pagesC=unique(probsSum$pageLabC)
+
+unique(probsSum$AnthroScn)
+probsSum$AnthroScn = factor(probsSum$AnthroScn, levels = c("low","low-med","med-high","high"))
+
+EVuncertainty = subset(probsSum,collarCount==0)
+#Resume here - calculate expectations only from priors
+#i.e. obsYears == 0 results
+str(EVuncertainty)
+EVuncertainty=subset(EVuncertainty,select=c(EVsample,EVvar,YearsOfProjection,AnthroScn,pageLab,pageLabC))
+names(EVuncertainty)[names(EVuncertainty)=="EVsample"]="EVuncertainty"
+hist(EVuncertainty$EVvar)#expect low values here. Method used to calculate EVuncertainty assumes priors probs do not differ among replicates.
+
+EVuncertainty$EVPI = 1-EVuncertainty$EVuncertainty
+
+#EVPI across more scenarios
+
+str(EVuncertainty)
+
+EVuncertainty$cmult=substr(EVuncertainty$pageLabC,6,6)
+
+EVuncertainty$trend=substr(EVuncertainty$pageLabC,13,13)
+EVuncertainty$trend[EVuncertainty$trend=="-"]= "decreasing"
+EVuncertainty$trend[EVuncertainty$trend=="0"]= "stable"
+EVuncertainty$trend[EVuncertainty$trend=="1"]= "increasing"
[…]

### 2024-08-30 analysis/scripts/X_BiasModelSensitivity.R

@@ -1,0 +1,57 @@
+library(ggplot2)
+library(tidyverse)
+library(caribouMetrics)
+#Notes
+#q: assume more females than males in calf:cow group, so q<1
+
+p=expand.grid(delay=c(T,F),q=c(0,0.3,0.6),w=c(1,3,6),
+              u=c(0,0.1,0.2),z=seq(0,0.2,length.out=3),R=seq(0.15,0.6,length.out=10))
+
+getX<-function(R,delay,q,w,u,z){
+  X = 0.5*R*w*(q*u+1-u)/((q*u+w-u)*(1-z))
+  X=X*(1-delay)+delay*X/(1+X)
+  return(X)
+}
+
+p$X = getX(p$R,p$delay,p$q,p$w,p$u,p$z)
+
+p$grp = paste(p$u,p$z)
+names(p)
+p$u=as.factor(p$u)
+p$z=as.factor(p$z)
+
+base = ggplot(subset(p,!delay),aes(x=R,y=2*X/R,col=z,linetype=u,group=grp))+
+  geom_hline(yintercept=1)+geom_line()+facet_grid(q~w,labeller=label_both)+theme_bw()
+plot(base)
+
+base2 = ggplot(subset(p,delay),aes(x=R,y=2*X/R,col=z,linetype=u,group=grp))+
+  geom_hline(yintercept=1)+geom_line()+facet_grid(q~w,labeller=label_both)+theme_bw()
+plot(base2)
+
+#############
+#approximate distribution for Bayesian model - lognormal
+cr=expand.grid(w = c(2,3,4,5,6,7,8,9),rep=seq(1,10000))
+nr=nrow(cr)
+cr$c = compositionBiasCorrection(q=runif(nr,0,0.6),w=cr$w,u=runif(nr,0,0.2),z=runif(nr,0,0.2))
+
+cs = compositionBiasCorrection(q=runif(nr,0,0.6),w=cr$w,u=runif(nr,0,0.2),z=runif(nr,0,0.2),approx=T)
+
+cr = merge(cr,cs)
+cr$ca= exp(rnorm(nrow(cr),mean=cr$mu,sd=cr$sig2^0.5))
+
+base = ggplot(cr,aes(x=c))+geom_histogram(aes(y=..density..))+
+  facet_wrap(~w,nrow=3,labeller=label_both)+theme_bw()+
[…]

### 2024-08-30 analysis/scripts/X_EVPIFromNationalSims.R

@@ -1,0 +1,41 @@
+
+#Calculation of EVPI (expected value of perfect information) with prior probabilities from national model
+#See Dunham et al. 2023 and references therein.
+library(caribouMetrics)
+library(tidyverse)
+
+#Step 1: get prior belief that true state is s p0_sd
+simsBig = getSimsNational(replicates=5000)
+lambdas = subset(simsBig$samples,Parameter=="Population growth rate")
+lambdas$s1 = (lambdas$Value>0.99)
+
+p0 = lambdas %>%
+  group_by(Anthro) %>%
+  summarise(p0_sd = mean(s1))
+p0$s = 1
+
+p0_s0 = p0; p0_s0$s = 0; p0_s0$p0_sd=1-p0_s0$p0_sd
+p0= rbind(p0,p0_s0)
+
+#step 2: value matrix v_as
+v = expand.grid(s = c(0,1),a=c(0,1))
+v$v_as = as.numeric(v$s==v$a)
+
+#step 3: merge
+vals = merge(v,p0)
+
+#step 4: calculate
+#Note expected value of certainty in this case is 1
+vals$innerProd = vals$p0_sd*vals$v_as
+
+Es = vals %>%
+  group_by(Anthro,a) %>%
+  summarise(Es = sum(innerProd))
+
+EVPI = Es %>%
+  group_by(Anthro) %>%
+  summarise(EVPI = 1-max(Es),EVuncertainty=max(Es))
+
+write.csv(EVPI,paste0("tabs/EVPI.csv"),row.names=F)
+
+plot(EVPI~Anthro,data=EVPI)

### 2024-08-30 analysis/scripts/bayesianModelExamples.R

@@ -1,196 +1,0 @@
-# Prior and posterior predictions from Bayesian model
-library(caribouMetrics)
-library(tidyverse)
-library(ggpubr)
-theme_set(theme_bw())
-library(RColorBrewer)
-
-baseDir <- "."
-
-monitoringScns = data.frame(obsYears=c(1,16),collarCount=c(0,30),cowMult=c(6),collarInterval=c(1),assessmentYrs=c(3))
-stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
-stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
-
-stateScns$sQuantile=0.8
-stateScns$rQuantile = 0.8
-scns=merge(monitoringScns,stateScns)
-
-scns$iAnthro = 0
-scns$tA = scns$iAnthro+(scns$obsYears)*scns$obsAnthroSlope
-scns$projYears = 50-scns$obsYears
-scns$N0 = 5000
-scns$adjustR = TRUE
-
-####################
-eParsIn = list()
-eParsIn$cowCounts <- data.frame(Year = 1981:2023,
-                                Count = 100,
-                                Class = "cow")
-eParsIn$freqStartsByYear <- data.frame(Year = 1981:2023,
-                                       numStarts = 30)
-eParsIn$collarOnTime=1
-eParsIn$collarOffTime=12
-eParsIn$collarNumYears=1
-labFontSize = 10; breakInterval=5
-
-###############################
-#With bias adjustment
-simBig<-getSimsNational(replicates=3000,forceUpdate=T)
-
-priorResult = caribouMetrics:::runScnSet(scns[1,],eParsIn,simBig,getKSDists=F,printProgress=F)
-priorResult$obs.all=NULL
-yr_scale1 <- scale_x_continuous(breaks = 2023:2072 %>% .[0:9*5+1],
-                                labels = 2009:2058 %>% .[0:9*5+1] )
-yr_scale2 <- scale_x_continuous(breaks = 2009:2058 %>% .[0:9*5+1],
-                                labels = 2009:2058 %>% .[0:9*5+1] )
-recPrior =  plotRes(priorResult, "Recruitment", lowBound=0, highBound = 0.8,
-                    legendPosition="none",breakInterval=breakInterval,
-                    labFontSize=labFontSize) +
-  yr_scale1 +
-  labs(tag = "a")
-plot(recPrior)
-survPrior =  plotRes(priorResult, "Adult female survival", lowBound=0.6,
[…]

### 2024-08-30 analysis/scripts/calibratePriors.R

@@ -1,66 +1,0 @@
-#devtools::load_all(here::here())
-library(caribouMetrics)
-scnsApplyAll = data.frame(qMin=0,qMax=0,uMin=0,uMax=0,zMin=0,zMax=0,adjustR=F)
-scnsNoData = data.frame(obsYears=1,collarCount=0,cowMult=0,projYears=4)
-scnsLow = data.frame(obsYears=15,collarCount=60,cowMult=9,projYears=1,rQuantile=0.025,sQuantile=0.025)
-scnsHigh = scnsLow;scnsHigh$rQuantile =0.975;scnsHigh$sQuantile=0.975
-########################
-#Survival random effects
-#vary sIntSE, sSigmaMin, and sSigmaMax in factorial array.
-sIntSE=getPriors()$l.Saf.Prior2;sSigmaMin=0.00001;sSigmaMax=seq(0.02,0.12,by=0.02)
-
-#Looking for low KS distance from full range of input sims when given almost no info
-scnsSInt=expand.grid(obsAnthroSlope = 0, projAnthroSlope = 0,sIntSE=sIntSE,
-                     sSigmaMin=sSigmaMin,sSigmaMax=sSigmaMax)
-scnsSInt$SigmaMin = scnsSInt$sSigmaMin
-scns=merge(scnsSInt,scnsApplyAll);scns=merge(scns,scnsNoData)
-scResults = runScenario(scns,survAnalysisMethod = "Exponential")
-KSAll = subset(scResults$ksDists,(Year==2023+scnsNoData$projYears)&(Parameter=="Adult female survival"))
-addBit = paste0("InterceptAll")
-scResults$obs.all = NULL
-makeInterceptPlots(scResults,addBit=addBit,facetVars=c("sSigmaMax"),loopVars="sIntSE",whichPlots=c("Adult female survival"))
-
-########################
-#Recruitment random effects
-#vary rIntSE, rSigmaMin, and rSigmaMax in factorial array.
-rIntSE=getPriors()$l.R.Prior2;rSigmaMin=0.00001;rSigmaMax=seq(0.05,0.3,by=0.05)
-
-#Looking for low KS distance from full range of input sims when given almost no info
-scnsRInt=expand.grid(obsAnthroSlope = 0, projAnthroSlope = 0,rIntSE=rIntSE,
[…]

### 2024-08-30 analysis/scripts/nationalModelTrajectories.R

@@ -1,252 +1,0 @@
-# Example trajectories from national model
-library(caribouMetrics)
-library(tidyverse)
-library(ggpubr)
-theme_set(theme_bw())
-library(RColorBrewer)
-
-dia_shp <- 23
-
-err_col <- "grey50"
-
-baseDir <- "."
-
-#numbers from Johnson et al for validation
-johnsonCompare <- read.csv(paste0(baseDir,"/data/Johnson et al. figures5_6.csv"))
-
-# get outputs from demographics
-covTableSim <- data.frame(Anthro = c(0, 10, 20, 30, 40, 50,
-                                     60, 70, 80, 90),
-                          Fire = 0,
-                          fire_excl_anthro = 0)
-covTableSim$polygon <- paste0("Missisa_", covTableSim$Anthro + 1)
-covTableSim$area <- "FarNorth"
-covTableSim$Total_dist <- covTableSim$Anthro + covTableSim$fire_excl_anthro
-
-popGrowthPars <- demographicCoefficients(500,
-  modelVersion = "Johnson",
-  survivalModelNumber = "M1",
-  recruitmentModelNumber = "M4"
-)
-
-popGrowthParsSmall <- demographicCoefficients(35,
-  modelVersion = "Johnson",
-  survivalModelNumber = "M1",
-  recruitmentModelNumber = "M4"
-)
-
-rateSamples <- demographicRates(
-  covTable = covTableSim,
-  popGrowthPars = popGrowthParsSmall,
-  ignorePrecision = F, returnSample = T, useQuantiles = T
-)
-
-rateSamplesLarge <- demographicRates(
-  covTable = covTableSim,
-  popGrowthPars = popGrowthPars,
-  ignorePrecision = F, returnSample = T, useQuantiles = T
-)
-
-rateSummaries <- demographicRates(
-  covTable = covTableSim, popGrowthPars = popGrowthPars,
-  ignorePrecision = F, returnSample = F, useQuantiles = F
-)
-
-#for each sample, select bias correction term from distributions
-
-str(popGrowthPars)
-
-cr = data.frame(replicate=unique(rateSamples$replicate))
-nr= nrow(cr)
[…]

### 2024-08-30 analysis/scripts/sensitivity1.R

@@ -1,71 +1,0 @@
-#app_path <- "C:/Users/HughesJo/Documents/gitprojects/BayesianCaribouDemographicProjection"
-#devtools::load_all(here::here())
-
-########################
-#sensitivity
-setName = "s7"
-monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
-                             cowMult=c(3,6,9),collarInterval=c(1,4),
-                             assessmentYrs=c(3))
-#TO DO - in next iteration, remove multiple years of one collar
-monitoringScns = subset(monitoringScns, !((obsYears>1)&(collarCount==0))&!((collarCount==1)&(cowMult>3))&!((collarCount==0)&(collarInterval>1)))
-stateScns = data.frame(tA=c(0,20,40,60,0,20,40,60,60,40,20),
-                       obsAnthroSlope=c(0,1,1,1,0,0,0,0,0,0,0),
-                       projAnthroSlope=c(1,1,1,1,0,0,0,0,-1,-1,-1))
-stateScns = merge(stateScns,data.frame(rep=seq(1:500)))
-
-stateScns$sQuantile=runif(nrow(stateScns),min=0.01,max=0.99)
-stateScns$rQuantile = runif(nrow(stateScns),min=0.01,max=0.99)
-#monitoringScns=rbind(monitoringScns,minimalScn)
-scns=merge(monitoringScns,stateScns)
-
-scns$preYears = max(scns$obsYears)-scns$obsYears
-
-scns$iAnthro = scns$tA-(scns$obsYears+scns$preYears-1)*scns$obsAnthroSlope
-scns$projYears = 20
-unique(scns$iAnthro)
-scns$repBatch = ceiling(scns$rep/50)
-table(scns$repBatch)
-scns$N0 = 5000
-scns$adjustR = TRUE
-
-scns$pageLab = paste0("cmult",scns$cowMult,"ay",scns$assessmentYrs,"aSf",scns$projAnthroSlope,"repBatch",scns$repBatch)
-scns$pageId = as.numeric(as.factor(scns$pageLab))
-
-write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
-pages=unique(scns$pageLab)
-
-setName="s9"
-scns = subset(scns,(rep==1)&(collarCount==30)
[…]

### 2024-08-30 analysis/scripts/viewSensitivityResults.R

@@ -1,351 +1,0 @@
-#devtools::load_all(here::here())
-library(tidyverse)
-library(ggplot2)
-library(caribouMetrics)
-library(RColorBrewer)
-
-pal2 = brewer.pal(7,"RdBu")[c(2,6)]
-pal4 = brewer.pal(5,"RdBu")[c(1,2,4,5)]
-
-pal4b = brewer.pal(5,"Purples")[c(2,3,4,5)]
-
-scn_defaults <- eval(formals(getScenarioDefaults))
-
-########################
-#sensitivity
-setName = "s8" #need to do both s7 and s8
-dir.create(paste0("figs/",setName),recursive=T)
-scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
-
-
-table(scns$pageId)
-
-nrow(scns)
-pagesa=unique(scns$pageId)
-
-pages = pagesa
-for (p in pagesa){
-  #p=pagesa[1]
-  if(!file.exists(paste0("results/",setName,"/rTest",p,".Rds"))){pages=pages[pages!=p]}
-}
-
-batchStrip<-function(l,batches=c(10,seq(1:9))){
-  for(b in batches){
-    l=gsub(paste0("repBatch",b),"",l,fixed=T)
-  }
-  return(l)
-}
-
-pages=sort(pages)
-
-combine=T
-addEV = F
-addProbs=F
-for(i in 1:length(pages)){
-  #combine=F;i=50
-  cpageId=pages[i]
-
-  if(i==length(pages)){
-    nextP = "TrepBatch1"
-  }else{
-    nextP = unique(scns$pageLab[scns$pageId==pages[i+1]])
-  }
-
-  p = unique(scns$pageLab[scns$pageId==cpageId])
-
-  print(paste(i,p))
-
-  if(combine&(cpageId>1)){
-    scNew = readRDS(paste0("results/",setName,"/rTest",cpageId,".Rds"))
-    for(n in names(scNew)){
-      if(n=="errorLog"){
-        scResults[[n]]=c(scResults[[n]],scNew[[n]])
-      }else{
-        scResults[[n]]=rbind(scResults[[n]],scNew[[n]])
-      }
-    }
-  }else{
-    scResults = readRDS(paste0("results/",setName,"/rTest",cpageId,".Rds"))
-  }
-
[…]

### 2024-09-13 analysis/paper/paper.Rmd
New simulation set & partial update of results.

@@
 \begin{equation}
-  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{N_0k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
+  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
   (\#eq:dotJ)
@@
 population sizes, and declines to $0.6\dot{X}_t$ as the population
-reaches carrying capacity of $k=100$ times the initial population size.
+reaches carrying capacity of $k=50000$.
 The post-juvenile female population in the next year includes both

### 2024-09-13 analysis/paper/supplement.Rmd
New simulation set & partial update of results.

@@
 |$p_k=0.6$||Recruitment multiplier at carrying capacity.|
-|$k=100$||Carrying capacity multiplier.|
+|$k=50000$||Carrying capacity.|
 |$b=4$||Density dependence shape parameter.|
@@
 
-The choice of priors for the random effects of year on survival and recruitment are less important because the random effects have little impact on expected population growth rate (Figs \@ref(fig:PriorAllFits) and \@ref(fig:PriorAllFitsR)).
+The choice of priors for interannual variation in survival and recruitment are less important because these have little impact on expected population growth rate (Figs \@ref(fig:PriorAllFits) and \@ref(fig:PriorAllFitsR)).
 We set the prior for the coefficient of variation of survival among years $\nu_S$ to vary uniformly between sNuMin = 0.001 and sNuMax = 0.13 to match the variation among national model simulations (Fig \@ref(fig:PriorAllFits)).

### 2024-09-13 analysis/scripts/1_sensitivity1.R

@@ -3,6 +3,42 @@
 
 ########################
 #sensitivity
+setName = "s10"
+monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
+                             cowMult=c(6),collarInterval=c(1),
+                             assessmentYrs=c(3))
+#TO DO - in next iteration, remove multiple years of one collar
+monitoringScns = subset(monitoringScns, !((obsYears>1)&(collarCount==0))&!((collarCount==1)&(cowMult>3))&!((collarCount==0)&(collarInterval>1)))
+stateScns = data.frame(tA=c(0,20,40,60),
+                       obsAnthroSlope=c(0,1,1,1),
+                       projAnthroSlope=c(1,1,1,1)
+                       )
+stateScns = merge(stateScns,data.frame(rep=seq(1:500)))
+stateScns = merge(stateScns,data.frame(interannualVar=c("list(R_CV=0.46,S_CV=0.087)","list(R_CV=0.23,S_CV=0.0435)")))
+stateScns$sQuantile=runif(nrow(stateScns),min=0.01,max=0.99)
+stateScns$rQuantile = runif(nrow(stateScns),min=0.01,max=0.99)
+#monitoringScns=rbind(monitoringScns,minimalScn)
+scns=merge(monitoringScns,stateScns)
+
+scns$preYears = max(scns$obsYears)-scns$obsYears
+
+scns$iAnthro = scns$tA-(scns$obsYears+scns$preYears-1)*scns$obsAnthroSlope
+scns$projYears = 20
+unique(scns$iAnthro)
+scns$repBatch = ceiling(scns$rep/50)
+table(scns$repBatch)
+scns$N0 = 5000
+scns$adjustR = TRUE
+
+scns$pageLab = paste0("cmult",scns$cowMult,"ay",scns$assessmentYrs,"aSf",scns$projAnthroSlope,"repBatch",scns$repBatch)
+scns$pageId = as.numeric(as.factor(scns$pageLab))
+nrow(scns)
+
+write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
+pages=unique(scns$pageLab)
+
+
+
 setName = "s7"
 monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
                              cowMult=c(3,6,9),collarInterval=c(1,4),

### 2024-09-13 analysis/scripts/sensitivityMinimalTest.R

@@ -9,10 +9,10 @@ cpageId <- args[1]
 cDir = getwd()
 
 library(caribouMetrics)
-
-# cpageId <- 31;n_reps <- 1;cDir = "C:/Users/HughesJo/Documents/gitprojects/Caribou-Demographic-Projection-Paper"
-
 setName = args[2]
+
+# setName="s10"; cpageId <- 1;n_reps <- 1;cDir = "C:/Users/HughesJo/Documents/gitprojects/Caribou-Demographic-Projection-Paper"
+
 
 #######################
 dir.create(paste0(cDir,"/figs/",setName),recursive=T)
@@ -46,10 +46,12 @@ message("batch ", cpageId, " started")
 
 str(scns)
 
+str(subset(scns,interannualVar=="list(R_CV=0.23,S_CV=0.0435)"))
 if(n_reps=="all"){
   scResults = caribouMetrics:::runScnSet(scns,eParsIn,simBig,getKSDists=F,printProgress=F)
 }else{
-  scResults = caribouMetrics:::runScnSet(scns[1:n_reps,],eParsIn,simBig,getKSDists=F,printProgress=F)
+  #scResults = caribouMetrics:::runScnSet(scns[1:n_reps,],eParsIn,simBig,getKSDists=F,printProgress=F)
+  scResults = runScnSet(scns[3801:3801,],eParsIn,simBig,getKSDists=F,printProgress=F)
 }
 
 saveRDS(scResults,paste0("results/",setName,"/rTest",cpageId,n_reps,cpageId,".Rds"))

### 2024-09-23 analysis/paper/paper.Rmd
increase assessmentYrs to reduce impact of interannual variability

@@
 caribou*) that integrates information about the impacts of landscape 
-change with available demographic data to reduce uncertainty in
-population viability projections for boreal caribou [building on
-@eacker_web-based_2019]. Following the value of information analysis
+change [@johnson_science_2020] with available demographic data [building on @eacker_web-based_2019] to reduce uncertainty in
+population viability projections. Following the value of information analysis
 framework proposed by Dunham et al. (2023), we combine this Bayesian
@@
 original approach, and @mcloughlin_declines_2003,
-@hervieux_widespread_2013, @eacker_web-based_2019, and @lamb2024 for
+@hervieux_widespread_2013, @eacker_web-based_2019, @lamb_effectiveness_2024 and @dalgarno_bboutools_2024 for
 examples of application to caribou). Recruitment is estimated from
 aerial surveys of the composition of groups of animals that include
-collared animals [@nbckc_nbckc_2022]. Groups typically also contain
+collared animals [@nbckc_nbckc_2022, @eacker_web-based_2019, @dalgarno_bboutools_2024]. Groups typically also contain
 uncollared animals, and a critical assumption in these surveys is that
@@
 [@decesare_estimating_2012; @serrouya_comparing_2017;
-@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].
-Some efforts have been made to improve caribou population viability
+@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].
+Various efforts have been made to improve caribou population viability
 projections and growth rate estimates using Bayesian methods
 [@moeller_integrating_2021; @mcnay_demographic_2022;
-@eacker_web-based_2019; @lamb2024], following broader trends in the
+@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024], following broader trends in the
 analysis of population dynamics [@schaub_integrated_2021]. In Bayesian
-population models, all available demographic or population data (e.g.
+population models, available demographic or population data (e.g.
 abundance surveys, vital rate estimates, trend data) are combined into a
@@
 models [@moeller_integrating_2021; @mcnay_demographic_2022;
-@eacker_web-based_2019; @lamb2024] have not thus far included
-demographic-disturbance relationships, and we are concerned that models
+@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not thus far included demographic-disturbance relationships, and we are concerned that models
 informed only by local demographic data could be used to inflate
@@
 available local demographic data in a simple Bayesian population model
-for boreal caribou [building on @eacker_web-based_2019]. In our simple
-proof-of-concept model, information is shared among populations using
-priors derived from a published analysis of variation in demographic
-rates among populations across Canada [@johnson_science_2020]. A more
+for boreal caribou. In our simple proof-of-concept model, information is 
+shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020]. A more
 rigorous statistical approach would be to integrate multiple populations
@@
 sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
-@eacker_web-based_2019]. We address this gap with a bias correction term
-derived from a simple model of the recruitment survey observation
+@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. We address this gap with a bias correction term derived from a simple model of the recruitment survey observation
 process.
@@
 
-resultScenario = "S7"
+resultScenario = "S10"
 resultSet ="cmult6ay3aSf1"
@@
 
-We focus on analyzing survival of collared females (as in
-@mcloughlin_declines_2003, @hervieux_widespread_2013 and
-@eacker_web-based_2019), and the ratio of calves to cows in aerial
+We focus on analyzing survival of collared females and the ratio of calves to cows in aerial
 surveys of groups that include collared females [@nbckc_nbckc_2022],
@@
 strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the
-simulated trajectories to confirm that our Bayesian model successfully
+simulated trajectories to confirm that the Bayesian model successfully
 integrates local data with priors from the national model. For
@@
 ```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction.
-To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variability ($v_S,v_R$) by 75\\% in these examples.
+To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variation by 75\\% in these examples ($v_R=0.115,v_S=0.02175$).
 Orange dots show means of 500 example trajectories.
@@
 
-Following others [@eacker_web-based_2019; @lamb2024], Kaplan-Meier
+Following others [@eacker_web-based_2019; @lamb_effectiveness_2024], Kaplan-Meier
 estimates of observed adult female survival probability $\hat{S}_t$ and
@@
 Gaussian distributed function of the true survival probability,
-truncated between 0 and 1, to allow sharing survival information across
-years with a random effect [@eacker_web-based_2019,@lamb2024]:
+truncated between 0 and 1 [@eacker_web-based_2019,@lamb_effectiveness_2024]:
 \begin{equation}
@@
 500 examples in which the Bayesian posterior mean population growth rate
-and the true population growth rate are both greater than 0.99, or both
-less than or equal to 0.99.
+and the true population growth rate over the full projection period are both greater than 0.99, or both less than or equal to 0.99.
 
@@
 when the forecast horizon is long (compare 5 and 20 year projections in
-Figure \@ref(fig:figDiffs)). Uncertainty is also higher in our high
+Figure \@ref(fig:figDiffs)), and when interannual variation is high (Figure
+\@ref(fig:examplePredictions)). Uncertainty is also higher in our high
 anthropogenic disturbance scenarios (Figure \@ref(fig:figDiffs)) because
@@
 lower levels of disturbance, local monitoring can reduce uncertainty,
-but there are diminishing returns as the number of collars deployed or
+but in some circumstances there are diminishing returns as the number of collars deployed or
 the number of years of monitoring increases (Figures
-\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the
+\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). High interannual variation makes reducing uncertainty more difficult (Figures
+\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
 relationships between monitoring effort and the probability of correct
@@
 
-```{r examplePredictions, fig.cap="Prior (a) and posterior (b) predictions from the Bayesian model for an example population trajectory.
+```{r examplePredictions, fig.cap="Prior (a) and posterior predictions from the Bayesian model for population trajectories with high (b: $v_R=0.46,v_S=0.087$) and lower (c: $v_R=0.23,v_S=0.0435$) interannual variation.
 In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%.
@@
 The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading).
-Local population data (open triangles) reduces uncertainty (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99).
+Local population data (open triangles) reduces uncertainty more when interannual variation is lower (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99).
 When projected populations become very small, stochastic variation in projected population growth rate increases (as in the top right panel).", cache=TRUE}
@@
 Differences are larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations.
-In these examples the composition survey cow multiplier $w$ is 6, and collars are renewed each year.", cache=TRUE}
+In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
@@
 
-```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
+```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
 When disturbance is high the probability of correct status assessment is high even without local monitoring.
 Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases.
-In these examples the composition cow multiplier $w$ is 6; varying $w$ does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
+In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
 knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
@@
 
-```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort.
+```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort.
 When disturbance is high EVSI is low, indicating that local sampling does not alter status assessments.
-Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases.
+Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but in some circumstances there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases.
 See Figure \\@ref(fig:figStatusErrors) for additional scenario details.", cache=TRUE}
@@
 [@moeller_integrating_2021; @mcnay_demographic_2022;
-@eacker_web-based_2019] have also used Bayesian models to estimate
+@eacker_web-based_2019; @dalgarno_bboutools_2024] have also used Bayesian models to estimate
 caribou demographic rates from local monitoring data, but our addition
@@
 self-sustainability, there are a variety of reasons to continue
-monitoring highly disturbed boreal caribou populations. Our point is
+monitoring highly disturbed boreal caribou populations. The point we wish to emphasize is
 that action is also required, and lack of monitoring should not be used
@@
 However, data owners have not authorized the use of the data for that
-purpose. In order to proceed with the available published information we
+purpose. To develop a proof-of-concept model with available published information we
 constrained our example trajectory simulations by assuming that
@@
 landscape condition.
+
+For this analysis we used Kaplan-Meier estimates of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ [as in @eacker_web-based_2019; @lamb_effectiveness_2024], and assumed Beta distributed interannual variability [as in @johnson_science_2020], but we acknowledge there are other possible approaches [e.g.
+@dalgarno_bboutools_2024].
+We believe the proof-of-concept model presented in this paper is adequate for making general points about the integration of national
+demographic-disturbance relationships with local data, and accounting for
+composition survey errors.
+However, in future we intend to use the more fully developed and documented 'bboutools' R package [@dalgarno_bboutools_2024] as a starting point for work with local decision makers to develop tools they can use inform local monitoring decisions.
 
@@
 for local data to inform demographic projections. In these places, our
-analysis suggests a general pattern of diminishing returns from
+analysis suggests that there may be diminishing returns from
 increasing monitoring effort, but more information on monitoring costs,

### 2024-09-23 analysis/scripts/0_bayesianModelExamples.R

@@ -11,16 +11,19 @@ monitoringScns = data.frame(obsYears=c(1,16),collarCount=c(0,30),cowMult=c(6),co
                             assessmentYrs=c(3))
 stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
 stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
+stateScns = merge(stateScns,data.frame(interannualVar=c("list(R_CV=0.46,S_CV=0.087)","list(R_CV=0.23,S_CV=0.0435)")))
 
 stateScns$sQuantile=0.8
 stateScns$rQuantile = 0.8
 scns=merge(monitoringScns,stateScns)
 
+scns = subset(scns,!(grepl("0.23",interannualVar,fixed=T)&(collarCount==0)))
 scns$iAnthro = 0
 scns$tA = scns$iAnthro+(scns$obsYears)*scns$obsAnthroSlope
 scns$projYears = 50-scns$obsYears
 scns$N0 = 5000
 scns$adjustR = TRUE
+
 #scns$assessmentYrs=3
 
 ####################
@@ -82,18 +85,39 @@ lambdaPosterior =  plotRes(posteriorResult, "Population growth rate", lowBound=0
   ylim(c(0, 1.2))
 plot(lambdaPosterior)
 
+posteriorResultB = caribouMetrics:::runScnSet(scns[3,],eParsIn,simBig,getKSDists=F,printProgress=F)
+recPosteriorB =  plotRes(posteriorResultB, "Recruitment", lowBound=0,highBound = 0.85,
+                        legendPosition="none",breakInterval=breakInterval,
+                        labFontSize=labFontSize)+
+  yr_scale2 +
+  labs(tag = "c")
+plot(recPosteriorB)
+survPosteriorB =  plotRes(posteriorResultB, "Adult female survival", lowBound=0.6,
+                         legendPosition="none",breakInterval=breakInterval,
+                         labFontSize=labFontSize)+
+  yr_scale2
+plot(survPosteriorB)
+lambdaPosteriorB =  plotRes(posteriorResultB, "Population growth rate", lowBound=0.5,
+                           legendPosition="none",breakInterval=breakInterval,
+                           labFontSize=labFontSize)+
+  yr_scale2 +
+  ylim(c(0, 1.2))
+plot(lambdaPosteriorB)
+
+
 leg <- plotRes(posteriorResult, "Recruitment", lowBound=0,highBound = 0.85,
[…]

### 2024-09-23 analysis/scripts/1_sensitivity1.R

@@ -6,7 +6,7 @@
 setName = "s10"
 monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
                              cowMult=c(6),collarInterval=c(1),
-                             assessmentYrs=c(3))
+                             assessmentYrs=c(5))
 #TO DO - in next iteration, remove multiple years of one collar
 monitoringScns = subset(monitoringScns, !((obsYears>1)&(collarCount==0))&!((collarCount==1)&(cowMult>3))&!((collarCount==0)&(collarInterval>1)))
 stateScns = data.frame(tA=c(0,20,40,60),
@@ -33,11 +33,12 @@ scns$adjustR = TRUE
 scns$pageLab = paste0("cmult",scns$cowMult,"ay",scns$assessmentYrs,"aSf",scns$projAnthroSlope,"repBatch",scns$repBatch)
 scns$pageId = as.numeric(as.factor(scns$pageLab))
 nrow(scns)
+length(unique(scns$pageLab))
 
 write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
 pages=unique(scns$pageLab)
 
-
+nrow(scns)
 
 setName = "s7"
 monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
@@ -67,6 +68,8 @@ scns$adjustR = TRUE
 
 scns$pageLab = paste0("cmult",scns$cowMult,"ay",scns$assessmentYrs,"aSf",scns$projAnthroSlope,"repBatch",scns$repBatch)
 scns$pageId = as.numeric(as.factor(scns$pageLab))
+nrow(scns)
+length(unique(scns$pageLab))
 
 write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
 pages=unique(scns$pageLab)

### 2024-09-23 analysis/scripts/2_viewSensitivityResults.R

@@ -13,7 +13,7 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 
 ########################
 #sensitivity
-setName = "s8" #need to do both s7 and s8
+setName = "s10" #need to do both s7 and s8
 dir.create(paste0("figs/",setName),recursive=T)
 scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
 
@@ -24,6 +24,27 @@ nrow(scns)
 pagesa=unique(scns$pageId)
 
 pages = pagesa
+
+str(scns)
+
+setLTYVar <-function(scns){
+  if(is.element("interannualVar",names(scns))){
+    scns$ltyVariable = "low"
+    scns$ltyVariable[grepl("0.46",scns$interannualVar,fixed=T)]="high"
+  }else{
+    scns$ltyVariable=scns$collarInterval
+  }
+  return(scns)
+}
+
+scns = setLTYVar(scns)
+if(is.element("interannualVar",names(scns))){
+  ltyLabel = "Interannual\nVariation"
+}else{
+  ltyLabel = "Collar\nRenewal\nInterval"
+}
+
+
 for (p in pagesa){
   #p=pagesa[1]
   if(!file.exists(paste0("results/",setName,"/rTest",p,".Rds"))){pages=pages[pages!=p]}
@@ -41,8 +62,9 @@ pages=sort(pages)
 combine=T
 addEV = F
 addProbs=F
+
 for(i in 1:length(pages)){
-  #combine=F;i=50
+  #combine=F;i=10
   cpageId=pages[i]
 
   if(i==length(pages)){
@@ -72,12 +94,15 @@ for(i in 1:length(pages)){
     combine=F
   }else{combine=T;next}
 
+  scResults$rr.summary.all = setLTYVar(scResults$rr.summary.all)
 
+  scResults$obs.all = setLTYVar(scResults$obs.all)
   #figure out how to count out errors.
   head(scResults$rr.summary.all)
   unique(scResults$rr.summary.all$collarInterval)
   #show examples projections
-  exResults = subset(scResults$rr.summary.all,(collarCount==30)&(collarInterval==1)&(Parameter=="Population growth rate"))
+  exResults = subset(scResults$rr.summary.all,(collarCount==30)&(ltyVariable==scns$ltyVariable[1])&(Parameter=="Population growth rate"))
+
   exResults$startYear = exResults$startYear+exResults$preYears
[…]

### 2024-09-23 analysis/scripts/3_EVSIPlots.R

@@ -14,7 +14,6 @@ pal4b = brewer.pal(5,"Purples")[c(2,3,4,5)]
 
 scn_defaults <- eval(formals(getScenarioDefaults))
 
-
 batchStrip<-function(l,batches=c(10,seq(1:9))){
   for(b in batches){
     l=gsub(paste0("repBatch",b),"",l,fixed=T)
@@ -22,7 +21,7 @@ batchStrip<-function(l,batches=c(10,seq(1:9))){
   return(l)
 }
 
-setName = "s8"
+setName = "s10"
 probsSum = read.csv(here::here(paste0("tabs/EVsample",setName,".csv")),stringsAsFactors = F)
 
 probsSum$pageLabC = batchStrip(probsSum$pageLab)
@@ -31,11 +30,18 @@ pagesC=unique(probsSum$pageLabC)
 unique(probsSum$AnthroScn)
 probsSum$AnthroScn = factor(probsSum$AnthroScn, levels = c("low","low-med","med-high","high"))
 
+
+if(is.element("interannualVar",names(probsSum))){
+  ltyLabel = "Interannual\nVariation"
+}else{
+  ltyLabel = "Collar\nRenewal\nInterval"
+}
+
+
 EVuncertainty = subset(probsSum,collarCount==0)
-#Resume here - calculate expectations only from priors
 #i.e. obsYears == 0 results
 str(EVuncertainty)
-EVuncertainty=subset(EVuncertainty,select=c(EVsample,EVvar,YearsOfProjection,AnthroScn,pageLab,pageLabC))
+EVuncertainty=subset(EVuncertainty,select=c(EVsample,EVvar,YearsOfProjection,AnthroScn,ltyVariable,pageLab,pageLabC))
 names(EVuncertainty)[names(EVuncertainty)=="EVsample"]="EVuncertainty"
 hist(EVuncertainty$EVvar)#expect low values here. Method used to calculate EVuncertainty assumes priors probs do not differ among replicates.
 
@@ -52,7 +58,6 @@ EVuncertainty$trend[EVuncertainty$trend=="-"]= "decreasing"
 EVuncertainty$trend[EVuncertainty$trend=="0"]= "stable"
 EVuncertainty$trend[EVuncertainty$trend=="1"]= "increasing"
 EVuncertainty$YearsOfProjection =as.factor(EVuncertainty$YearsOfProjection)
-
 
 png(here::here(paste0("figs/",setName,"/EVPIall.png")),
[…]

### 2024-09-23 analysis/scripts/sensitivityMinimalTest.R

@@ -54,6 +54,7 @@ if(n_reps=="all"){
   scResults = runScnSet(scns[3801:3801,],eParsIn,simBig,getKSDists=F,printProgress=F)
 }
 
+unique(scResults$rr.summary.all$Parameter)
 saveRDS(scResults,paste0("results/",setName,"/rTest",cpageId,n_reps,cpageId,".Rds"))
 
 message("batch ", cpageId, " complete")

### 2024-10-03 analysis/paper/paper.Rmd
Various - modifying methods and results...

@@
 for boreal caribou. In our simple proof-of-concept model, information is 
-shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020]. A more
-rigorous statistical approach would be to integrate multiple populations
-into a hierarchical Bayesian analysis of variation within and among
-populations [@leasure_hierarchical_2019]; we have not done this because
-owners have not authorized the use of the data for this purpose,
-obtaining permission to use caribou demographic data is difficult, and
-most managers only have access to their own local data. Despite
-limitations, our simple Bayesian model represents an important
-improvement over less formally integrated methods for estimating and
+shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020].
+The model represents an important improvement over less formally integrated methods for estimating and
 projecting impacts of disturbance on population population viability
@@
 resultScenario = "S10"
-resultSet ="cmult6ay3aSf1"
+resultSet ="cmult6ay5aSf1"
 figSet = c("diffs","distScns","EVSI","power","EVPIall")
@@
 In all scenarios, the projection period begins in 2024, and the monitoring period occurs prior to the projection period.", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))
+knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay5aSf1.png"))
 
@@
 $\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow
-ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification biases
-(Supplement B) and (optionally) a delayed reproduction term to account
-for the reality that yearlings generally do not breed
-[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum
+ratio $\dot{R}_t$ is adjusted for sex ratio and misidentification biases
+(Supplement B) to get maximum
 potential recruitment of female calves per cow $\dot{X}_t$:
@@
 \begin{equation}
-  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
+  \dot{X}_t=\dot{c}\dot{R}_t/2;\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
   (\#eq:dotX)
@@
 
-```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction.
+```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships.
 To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variation by 75\\% in these examples ($v_R=0.115,v_S=0.02175$).
 Orange dots show means of 500 example trajectories.
-For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}
-
-knitr::include_graphics(here::here("analysis/paper/figs/DemographicRatesDelay.png"))
+For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models.", cache=TRUE}
+
+knitr::include_graphics(here::here("analysis/paper/figs/DemographicRates.png"))
 
@@
 
-To account for bias and additional uncertainty caused by unknown
-misidentification biases (Supplement B) and delayed age of age of first
-reproduction [@eacker_web-based_2019; @decesare_estimating_2012] the
+To account for sex ratio of calves and additional uncertainty caused by unknown
+misidentification errors (Supplement B) the
 estimated recruitment probability is adjusted to get expected female
@@
 \begin{equation}
-X_t=\frac{cR_t/2}{1+cR_t/2}; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
+X_t=cR_t/2; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
   (\#eq:X)
@@
 between 0 and 0.2, and the probability of missing a calf $z$ varies
-uniformly between 0 and 0.2. To maintain consistency with Johnson et al.
-[-@johnson_science_2020] we also provide an option to only adjust for
-sex ratio and misidentification biases $X_t=cR_t/2$.
+uniformly between 0 and 0.2.
 
@@
 number of juveniles recruiting to the post-juvenile class at the census
-is a Poisson distributed function of the number of surviving
+is a binomially distributed function of the number of surviving
 post-juvenile females and the adjusted recruitment rate $X_t$:
-$J_{t} \sim \text{Poisson}(W_tX_t)$. The post-juvenile female population
+$J_{t} \sim \text{Binomial}(W_t,X_t)$. The post-juvenile female population
 in the next year includes both survivors and new recruits:
@@
 
-```{=html}
-<!-- Annotated equations for poster.
-$$J_{t} \sim \text{Poisson}(W_tX_t)$$
-$$\underbrace{N_{t+1}}_{\text{Adult females in year t+1}}=\underbrace{W_t}_{\text{Surviving adults}}+\underbrace{J_t}_{\text{New recruits to adult age class}}$$
-$$W_{t} \sim \text{Binomial}(N_t,\underbrace{S_t}_{\text{Adult survival probability}})$$
-$$S_t=\underbrace{\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})}_{\text{Adjustment as in in Johnson et al.
-2020}}=\beta^S_0+ \underbrace{\beta^S_aA_t}_{\text{Anthropogenic disturbance effect}}+\underbrace{\epsilon^S_t}_{\text{Random effect of year}}.$$
-$$J_{t} \sim \text{Poisson}(W_t\underbrace{R_t/2}_{\text{Recruitment rate}})$$
-$$\log(R_t)=\beta^R_0+\underbrace{\beta^R_a A_t}_{\text{Anthropogenic disturbance effect}}+ \underbrace{\beta^R_fF_t}_{\text{Fire effect}}+\underbrace{\epsilon^R_t}_{\text{Random effect of year}}.$$
-$$\underbrace{\hat{C}_t}_{\text{Observed calves}} \sim \text{Binomial}(\underbrace{\hat{W}_t}_{\text{Observed cows}},R_t)$$
-$$\underbrace{\hat{S_t}}_{\text{Observed survival (Kaplan-Meier estimate)}} \sim \text{Normal}(S_t,1/\tau_t).$$ -->
-```
 ```{=html}
@@
 
-```{r examplePredictions, fig.cap="Prior (a) and posterior predictions from the Bayesian model for population trajectories with high (b: $v_R=0.46,v_S=0.087$) and lower (c: $v_R=0.23,v_S=0.0435$) interannual variation.
+```{r examplePredictions, fig.cap="Prior (a) and posterior predictions from the Bayesian model for population trajectories with low (b: $v_R=0.23,v_S=0.0435$) and high (c: $v_R=0.46,v_S=0.087$) interannual variation.
 In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%.
@@
 In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay3aSf1.png"))
+knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay5aSf1.png"))
 ```
@@
 In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
+knitr::include_graphics(here::here("figs/forPaper/powercmult6ay5aSf1.png"))
 ```
@@
 
-knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay3aSf1.png"))
+knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay5aSf1.png"))
 ```
@@
 populations, and distinguishing that from variation among years.
-However, data owners have not authorized the use of the data for that
-purpose. To develop a proof-of-concept model with available published information we
-constrained our example trajectory simulations by assuming that
-populations remain in their quantile, so populations that are lower than
-or higher than average remain lower or higher than average over time.
+We have not done this because data owners have not authorized the use of the data for that purpose, and obtaining permission to use caribou demographic data is challenging.
+To move ahead with available published information we constrained our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower than or higher than average remain lower or higher than average over time.
 The Bayesian model is not similarly constrained, and can accommodate
@@
 allows for a greater range of possibilities than our more constrained
-simulated example trajectories. Although our simulated trajectories do
+simulated example trajectories. Although the simulated trajectories do
 not include all plausible possibilities we believe they are sufficient
 to demonstrate how a Bayesian model can be used to integrate local data
-with national priors, and to show how monitoring requirements depend on
+with national priors, and to show that monitoring requirements depend on
 landscape condition.
@@
 @dalgarno_bboutools_2024].
-We believe the proof-of-concept model presented in this paper is adequate for making general points about the integration of national
+The proof-of-concept model presented in this paper is adequate for making general points about the integration of national
 demographic-disturbance relationships with local data, and accounting for
 composition survey errors.
-However, in future we intend to use the more fully developed and documented 'bboutools' R package [@dalgarno_bboutools_2024] as a starting point for work with local decision makers to develop tools they can use inform local monitoring decisions.
+However, the implementation is a research prototype, not a fully developed analysis tool.
+The more fully developed and documented *bboutools* R package [@dalgarno_bboutools_2024] is a better starting point for development of tools to enable local decision makers to inform decisions with their  existing data.
+A possible way forward would be to extend that package to include demographic-disturbance relationships and composition survey errors.
 
@@
 caribou and variation in habitat preferences and space use among
-individuals [e.g. @decesare_linking_2014; @avgar_spaceuse_2015].
-Although there is evidence of spatial structure and variation in
-demographic rates among individuals within populations, for most caribou
-populations there is not enough data available to estimate spatial
-variation in demographic model parameters.
-
-A number of additional complexities can be incorporated into aspatial
-Bayesian population models, including variation in demographic rates
+individuals [e.g. @decesare_linking_2014; @avgar_spaceuse_2015]. It is possible to integrate complexities such as variation in demographic rates
 among individuals [@armstrong_incorporating_2021], population size
-information from aerial or fecal dna surveys, and additional age classes
-[@moeller_integrating_2021; @mcnay_demographic_2022]. However, in many
-contexts, population size and fecal dna survey information is not
-available, and a framework that includes these sources of information
-will be unnecessarily complicated. At present, our model is intended as
-a proof-of-concept to show the value of integrating national
-demographic-disturbance relationships with local data, accounting for
+information from aerial or fecal dna surveys, and variation in demographic rates among age classes [@moeller_integrating_2021; @mcnay_demographic_2022] into Bayesian population models.
+However, in many contexts there is not enough to data to estimate more complex model parameters.
+
+At present, our analysis is intended as a proof-of-concept to show the value of integrating national demographic-disturbance relationships with local data, accounting for
 composition survey errors, and assessing the value of local monitoring
-information. Rather than further complicate a proof-of-concept model
-applied to simulated data, our next step will be to collaborate with an
-interested decision maker to develop a tool informed by this analysis
-that meets their real local needs, recognizing that these needs will
-vary among monitoring programs [@nbckc_nbckc_2021; @nbckc_nbckc_2021].
-
-## Assessing the value of additional monitoring
-
-We applied the framework proposed by Dunham et al. (2023) to show that
+information. We applied the framework proposed by Dunham et al. (2023) to show that
 no additional local monitoring information is needed to predict whether
@@
 costs, options and goals are context specific, we believe the most
-effective way forward will be to use this proof-of-concept model and
-analysis as a starting point for working with local decision makers to
-build decision support tools suited to their particular circumstances,
-needs and questions.
+effective way forward will be to use this proof-of-concept analysis, existing local data, and the more fully developed *bboutools* R package [@dalgarno_bboutools_2024] as a starting point for working with interested decision makers to develop monitoring decision support tools that meet real local needs, recognizing that those needs will vary [@nbckc_nbckc_2021;@nbckc_nbckc_2022].
 

### 2024-10-03 analysis/paper/supplement.Rmd
Various - modifying methods and results...

@@
 |$\bar{\lambda}_{y,t}$|rate|Average population growth over $y$ years.|
-|$X_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|
+|$X_t$|rate|Recruitment adjusted for sex ratio and misidentification bias.|
 |$W_t$|animals|Surviving adult females before recruitment.|
@@
 |$\dot{N}_t$|animals|Adult female population size.|
-|$\dot{X}_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|
+|$\dot{X}_t$|rate|Recruitment adjusted for sex ratio and misidentification bias.|
 |$\dot{W}_t$|animals|Surviving adult females before recruitment.|
@@
 $$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$
-
-If we also adjust for delayed age at first reproduction (DeCesare et al. 2012; Eacker et al. 2019), the adjusted recruitment rate becomes $$X=\frac{cR/2}{1+cR/2}.$$
 

### 2024-10-03 R/CaribouDemographicFunctions.R

@@ -34,9 +34,9 @@ runScenario<-function(scns,quants=NULL,Anthro=NULL,survAnalysisMethod="KaplanMei
   #Get full set of sims for comparison
 
   if(is.null(quants)){
-    simBig<-getSimsNational(adjustR=unique(scns$adjustR))#If called with default parameters, use saved object to speed things up.
+    simBig<-getSimsNational()#If called with default parameters, use saved object to speed things up.
   }else{
-    simBig<-getSimsNational(useQuantiles = quants,Anthro=Anthro,adjustR=unique(scns$adjustR))#If called with default parameters, use saved object to speed things up.
+    simBig<-getSimsNational(useQuantiles = quants,Anthro=Anthro)#If called with default parameters, use saved object to speed things up.
   }
 
   scResults = caribouMetrics:::runScnSet(scns,eParsIn,simBig,survAnalysisMethod,getKSDists=getKSDists,printProgress=T)

### 2024-10-03 analysis/scripts/0_PriorsFromPIs.R

@@ -16,6 +16,8 @@ baseDir <- "."
 #numbers from Johnson et al for validation
 johnsonCompare <- read.csv(paste0(baseDir,"/data/Johnson et al. figures5_6.csv"))
 
+
+
 getAnswers <-function(scns,par,johnsonCompare,vName){
   scResults = runScenario(scns,survAnalysisMethod = "Exponential")
   rr = subset(scResults$rr.summary.all,Parameter == par)
@@ -44,7 +46,7 @@ getAnswers <-function(scns,par,johnsonCompare,vName){
 
 rrAll=list()
 #Johnson analysis does not include interannual variation. So omit that to set the survival and reproduction priors.
-scnsApplyAll = data.frame(qMin=0,qMax=0,uMin=0,uMax=0,zMin=0,zMax=0,adjustR=F,iFire=4.27)
+scnsApplyAll = data.frame(qMin=0,qMax=0,uMin=0,uMax=0,zMin=0,zMax=0,iFire=4.27)
 scnsNoData = data.frame(obsYears=1,collarCount=0,cowMult=0,projYears=1)
 scnsLow = data.frame(obsYears=15,collarCount=60,cowMult=9,projYears=1,rQuantile=0.025,sQuantile=0.025)
 scnsHigh = scnsLow;scnsHigh$rQuantile =0.975;scnsHigh$sQuantile=0.975

### 2024-10-03 analysis/scripts/0_bayesianModelExamples.R

@@ -8,7 +8,7 @@ library(RColorBrewer)
 baseDir <- "."
 
 monitoringScns = data.frame(obsYears=c(1,16),collarCount=c(0,30),cowMult=c(6),collarInterval=c(1),
-                            assessmentYrs=c(3))
+                            assessmentYrs=c(1))
 stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
 stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
 stateScns = merge(stateScns,data.frame(interannualVar=c("list(R_CV=0.46,S_CV=0.087)","list(R_CV=0.23,S_CV=0.0435)")))
@@ -21,8 +21,7 @@ scns = subset(scns,!(grepl("0.23",interannualVar,fixed=T)&(collarCount==0)))
 scns$iAnthro = 0
 scns$tA = scns$iAnthro+(scns$obsYears)*scns$obsAnthroSlope
 scns$projYears = 50-scns$obsYears
-scns$N0 = 5000
-scns$adjustR = TRUE
+scns$N0 = 2000
 
 #scns$assessmentYrs=3
 
@@ -63,7 +62,7 @@ lambdaPrior =  plotRes(priorResult, "Population growth rate", lowBound=0.4,
                        legendPosition="none",breakInterval=breakInterval,
                        labFontSize=labFontSize)+
   yr_scale1 +
-  ylim(c(0, 1.2))
+  ylim(c(0, 1.3))
 plot(lambdaPrior)
 
 posteriorResult = caribouMetrics:::runScnSet(scns[2,],eParsIn,simBig,getKSDists=F,printProgress=F)
@@ -71,7 +70,7 @@ recPosterior =  plotRes(posteriorResult, "Recruitment", lowBound=0,highBound = 0
                         legendPosition="none",breakInterval=breakInterval,
                         labFontSize=labFontSize)+
   yr_scale2 +
-  labs(tag = "b")
+  labs(tag = "c")
 plot(recPosterior)
 survPosterior =  plotRes(posteriorResult, "Adult female survival", lowBound=0.6,
                          legendPosition="none",breakInterval=breakInterval,
@@ -82,7 +81,7 @@ lambdaPosterior =  plotRes(posteriorResult, "Population growth rate", lowBound=0
                            legendPosition="none",breakInterval=breakInterval,
                            labFontSize=labFontSize)+
[…]

### 2024-10-03 analysis/scripts/0_calibratePriors.R

@@ -1,6 +1,6 @@
 #devtools::load_all(here::here())
 library(caribouMetrics)
-scnsApplyAll = data.frame(qMin=0,qMax=0,uMin=0,uMax=0,zMin=0,zMax=0,adjustR=F)
+scnsApplyAll = data.frame(qMin=0,qMax=0,uMin=0,uMax=0,zMin=0,zMax=0)
 scnsNoData = data.frame(obsYears=1,collarCount=0,cowMult=0,projYears=4)
 scnsLow = data.frame(obsYears=15,collarCount=60,cowMult=9,projYears=1,rQuantile=0.025,sQuantile=0.025)
 scnsHigh = scnsLow;scnsHigh$rQuantile =0.975;scnsHigh$sQuantile=0.975

### 2024-10-03 analysis/scripts/0_nationalModelTrajectories.R

@@ -120,26 +120,26 @@ pars2 <- cbind(pars, caribouPopGrowth(pars$N0,
 ))
 
 oo <- pars2 %>%
-  select(Anthro, lambda, fullGrp, rrp) %>%
+  select(Anthro, lambdaTrue, fullGrp, rrp) %>%
   group_by(fullGrp, Anthro) %>%
-  summarise(lambda = median(lambda))
+  summarise(lambdaTrue = median(lambdaTrue))
 
 ooT <- pars1 %>%
-  select(Anthro, lambda, fullGrp, rrp) %>%
+  select(Anthro, lambdaTrue, fullGrp, rrp) %>%
   group_by(fullGrp, Anthro) %>%
-  summarise(lambda = median(lambda))
+  summarise(lambdaTrue = median(lambdaTrue))
 
-oo$lambdaH <- oo$lambda
-oo$lambdaL <- oo$lambda
+oo$lambdaH <- oo$lambdaTrue
+oo$lambdaL <- oo$lambdaTrue
 
 plot_lambda <- ggplot(oo,
-                      aes(x = Anthro, y = lambda, ymin = lambdaH, ymax = lambdaL)) +
+                      aes(x = Anthro, y = lambdaTrue, ymin = lambdaH, ymax = lambdaL)) +
   geom_line(size = 0.5,
-            aes(x = Anthro, y = lambda, group = fullGrp, color = fullGrp),
+            aes(x = Anthro, y = lambdaTrue, group = fullGrp, color = fullGrp),
             alpha = 1) + scale_color_manual(values=pal)+
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
   xlab("Anthropogenic disturbance (%)") +
-  ylab(expression("Population Growth Rate " * lambda)) +
+  ylab(expression("Population Growth Rate " * dot(lambda)[t])) +
   theme(legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
 
 str(pars1)
@@ -160,6 +160,22 @@ plot_recruitment3 <- ggplot(data = rateSummaries,
   theme(legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
 plot(plot_recruitment3)
 
+plot_recruitment1 <- ggplot(data = rateSummaries,
+                            aes(x = Anthro, y = R_bar,
+                                ymin = R_PIlow, ymax = R_PIhigh)) +
[…]

### 2024-10-03 analysis/scripts/1_sensitivity1.R

@@ -6,7 +6,7 @@
 setName = "s10"
 monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
                              cowMult=c(6),collarInterval=c(1),
-                             assessmentYrs=c(5))
+                             assessmentYrs=c(1))
 #TO DO - in next iteration, remove multiple years of one collar
 monitoringScns = subset(monitoringScns, !((obsYears>1)&(collarCount==0))&!((collarCount==1)&(cowMult>3))&!((collarCount==0)&(collarInterval>1)))
 stateScns = data.frame(tA=c(0,20,40,60),
@@ -27,8 +27,7 @@ scns$projYears = 20
 unique(scns$iAnthro)
 scns$repBatch = ceiling(scns$rep/50)
 table(scns$repBatch)
-scns$N0 = 5000
-scns$adjustR = TRUE
+scns$N0 = 2000
 
 scns$pageLab = paste0("cmult",scns$cowMult,"ay",scns$assessmentYrs,"aSf",scns$projAnthroSlope,"repBatch",scns$repBatch)
 scns$pageId = as.numeric(as.factor(scns$pageLab))
@@ -38,12 +37,19 @@ length(unique(scns$pageLab))
 write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
 pages=unique(scns$pageLab)
 
-nrow(scns)
+###################
+#Without bias
+setName = "s12"
+scns$qMax = 0;scns$uMax=0;scns$zMax=0
+write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
+pages=unique(scns$pageLab)
 
-setName = "s7"
+###################
+#Different anthropogenic disturbance scenarios, collar intervals. With bias and default interannual variability.
+setName = "s11"
 monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
                              cowMult=c(3,6,9),collarInterval=c(1,4),
-                             assessmentYrs=c(3))
+                             assessmentYrs=c(1))
 #TO DO - in next iteration, remove multiple years of one collar
[…]

### 2024-10-03 analysis/scripts/2_viewSensitivityResults.R

@@ -61,7 +61,7 @@ pages=sort(pages)
 
 combine=T
 addEV = F
-addProbs=F
+addProbs = F
 
 for(i in 1:length(pages)){
   #combine=F;i=10
@@ -191,24 +191,24 @@ for(i in 1:length(pages)){
   sizeTrue$parameter=NULL;sizeTrue$type=NULL;sizeTrue$Mean=NULL
   setdiff(names(statusTrue),names(probs))
 
-  sizeProj = subset(scResults$rr.summary.all,(Parameter=="Female population size"),select=c(label,Year,Mean))
-  sizeProj$projSize = sizeProj$Mean; sizeProj$Mean=NULL
+  #sizeProj = subset(scResults$rr.summary.all,(Parameter=="Female population size"),select=c(label,Year,Mean))
+  #sizeProj$projSize = sizeProj$Mean; sizeProj$Mean=NULL
 
   probs = merge(probs,statusTrue)
   probs = merge(probs,sizeTrue)
-  probs = merge(probs,sizeProj)
-  probs$trueChange = probs$trueSize/probs$N0
-  probs$predChange = probs$projSize/probs$N0
+  #probs = merge(probs,sizeProj)
+  probs$trueChange = probs$trueMean#probs$trueSize/probs$N0
+  probs$predChange = probs$Mean#probs$projSize/probs$N0
 
   probs$viableTrue = (probs$trueChange>0.99)&(probs$trueSize>10)
 
-  probs$viablePred = (probs$predChange>0.99)&(probs$projSize>10)
+  probs$viablePred = (probs$predChange>0.99)#&(probs$projSize>10)
 
   probs$wrong = probs$viableTrue != probs$viablePred
 
   probs$CorrectStatus[probs$wrong]="no"
   probs$CorrectStatus[!probs$wrong]="yes"
-  probs$LambdaDiff = probs$trueMean-probs$Mean
+  probs$LambdaDiff = probs$trueChange-probs$predChange
   #probs$LambdaDiff[(probs$projSize<=10)]=NA
 
   probs$pageLabB = paste0(batchStrip(probs$pageLab),"st",probs$collarCount,"ri",probs$ltyVariable)

### 2024-10-03 analysis/scripts/sensitivityMinimalTest.R

@@ -8,10 +8,10 @@ cpageId <- args[1]
 
 cDir = getwd()
 
-library(caribouMetrics)
 setName = args[2]
 
 # setName="s10"; cpageId <- 1;n_reps <- 1;cDir = "C:/Users/HughesJo/Documents/gitprojects/Caribou-Demographic-Projection-Paper"
+library(caribouMetrics)
 
 
 #######################

### 2024-10-09 analysis/paper/paper.Rmd
Ongoing - manuscript and method tweaking.

@@
 resultScenario = "S10"
-resultSet ="cmult6ay5aSf1"
+resultSet ="cmult6ay1aSf1"
 figSet = c("diffs","distScns","EVSI","power","EVPIall")
@@
 In all scenarios, the projection period begins in 2024, and the monitoring period occurs prior to the projection period.", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay5aSf1.png"))
+knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay1aSf1.png"))
 
@@
 varies with landscape condition, we begin with a model for simulating
-example population dynamics and observations of those dynamics. These
-example population trajectories are informed by prior information about
-the distribution of possible parameter values and outcomes from the
-national analysis of demographic-disturbance relationships
-[@johnson_science_2020] so we call it the "national" model. Our method
-for selecting parameter values for each example trajectory yields a set
-of alternative models (hypotheses) about the state of a particular
-caribou population; these alternative scenarios informed by prior
+example population dynamics and observations of those dynamics.
+Demographic parameters for the simulated trajectories are sampled from the regression models that summarize national demographic-disturbance relationships
+[@johnson_science_2020] so we call this the "national" model.
+Together, a set of example trajectories represent alternative hypotheses about the true state of a particular caribou population; these alternative scenarios informed by prior
 beliefs are used to assess the expected value of alternative sampling
@@
 simulated trajectories to confirm that the Bayesian model successfully
-integrates local data with priors from the national model. For
+integrates local data with priors from the national model, and that in the absence of local data Bayesian model outcomes match expectations from the national model. For
 reference, all variables and parameters are defined in supplemental
@@
 population sizes, and declines to $0.6\dot{X}_t$ as the population
-reaches carrying capacity of $k=50000$.
+reaches carrying capacity of $k=10000$.
 The post-juvenile female population in the next year includes both
@@
 24 years prior to the start of the projection period.
+Realized population growth rate is $\dot{\lambda}_t=\dot{N}_t/\dot{N}_{t-1}$ Expected population growth rate (without interannual variation, demographic stochasticity, or effects of population density) is $\bar{\lambda}_t=\bar{S}_t(1+\dot{c}\bar{R}_t/2)$.
 
@@
 
-To assess sensitivity of results to the survival estimation method and
-handle cases in which there is little or no survival data, we also
-implemented an alternative parametric exponential survival model,
+To handle cases in which there is little or no survival data, we use an alternative parametric exponential survival model,
 wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
@@
 priors so that the the 95% prior predictive intervals for survival and
-recruitment from the Bayesian model match the 95% predictive intervals
+recruitment from the Bayesian model match the 95% prediction intervals
 from Johnson et al's regression model (Supplement C, Figure
@@
 knitr::kable(tbl,caption="Prior means and standard deviations of Gaussian distributed survival and reproduction model parameters.
-Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the 95\\% predictive intervals from Johnson et al's regression model (Supplement C).")
+Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the 95\\% prediction intervals from Johnson et al's regression model (Supplement C).")
 ```
@@
 census in late winter of surviving females and new recruits to the
-post-juvenile (aged 1+) class. The number of post-juvenile females that
-survive from year $t$ to the census $W_t$ is binomially distributed with
-survival probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The
-number of juveniles recruiting to the post-juvenile class at the census
-is a binomially distributed function of the number of surviving
-post-juvenile females and the adjusted recruitment rate $X_t$:
-$J_{t} \sim \text{Binomial}(W_t,X_t)$. The post-juvenile female population
-in the next year includes both survivors and new recruits:
-$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
-when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
-stable estimates of population status, we report average population
-growth over a period of $y=3$ years, calculated as
-$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
-post-juvenile population size $N_0$ is 5000 animals. For reference, all
-variables and parameters are defined in supplemental Table A1.
-
-```{=html}
-<!--
-Gritty details for the record: \* note distribution of random effect parameters are truncated to
-ensure R and S remain within reasonable bounds (not guaranteed given log
-link).
-Not yet documented here.
-\* note Eacker survival method gives biased results if there is left censoring, or months without animals at risk.
-To address this we drop left censored observations and only estimate survival when at least some animals remain at risk by month 12.
-\*
-see alternative choices made in this example
-<https://poissonconsulting.shinyapps.io/bboushiny/>. See also Lamb et al, who uses two step KM estimation process for survival like Eacker. Stick with published examples. \*
--->
-```
+post-juvenile (aged 1+) class.
+Expected population growth rate (without interannual variation or demographic stochasticity) is $\tilde{\lambda}_t=\tilde{S}_t(1+c\tilde{R}_t/2)$.
+For reference, all variables and parameters are defined in supplemental Table A1.
+
 Models and analysis were implemented using the R language
@@
 \@ref(fig:exampleTrajectories)), and calculated the percentage of these
-500 examples in which the Bayesian posterior mean population growth rate
-and the true population growth rate over the full projection period are both greater than 0.99, or both less than or equal to 0.99.
+500 examples in which the Bayesian posterior expected population growth rate
+and the true expected population growth rate are both greater than 0.99, or both less than or equal to 0.99.
 
@@
 monitoring data alters the distribution of differences between true
-population growth rates and Bayesian posterior mean estimated growth
+expected population growth rates and Bayesian posterior expected growth
 rates.
@@
 of Figure \@ref(fig:figStatusErrors)). This means that the long-term
-projections informed by local monitoring data are on average less
-precise but not less accurate, which can happen when there is very small
-prior probability that population is viable (orange band in the top
-right panel of Figure \@ref(fig:examplePredictions) falls below 1). At
+projections informed by local monitoring data are less
+precise but not less accurate, which can happen if the prior viability predictions are more certain than the posterior predictions (e.g 
+left side of right panels in Figure \@ref(fig:examplePredictions)). At
 lower levels of disturbance, local monitoring can reduce uncertainty,
-but in some circumstances there are diminishing returns as the number of collars deployed or
+there are diminishing returns as the number of collars deployed or
 the number of years of monitoring increases (Figures
-\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). High interannual variation makes reducing uncertainty more difficult (Figures
+\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). Reducing uncertainty is more difficult when interannual variation is high (Figures
 \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
@@
 The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading).
-Local population data (open triangles) reduces uncertainty more when interannual variation is lower (bottom row), but when disturbance is very high local data is not required to predict viability (population growth <= 0.99).
-When projected populations become very small, stochastic variation in projected population growth rate increases (as in the top right panel).", cache=TRUE}
+Local population data (open triangles) reduces uncertainty about demographic rates, but local data is not required to predict viability (population growth <= 0.99) on highly disturbed landscapes.", cache=TRUE}
 
@@
 
-```{r figDiffs, fig.cap="The distribution of differences between true population growth rates and Bayesian posterior mean growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
+```{r figDiffs, fig.cap="The distribution of differences between true expected population growth rates and Bayesian posterior expected growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
 Differences decrease as the number of years of monitoring and the number of collars increase.
-Differences are larger when anthropogenic disturbance is high because stochastic variation in population growth rate is higher for small populations.
 In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay5aSf1.png"))
+knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay1aSf1.png"))
 ```
@@
 When disturbance is high the probability of correct status assessment is high even without local monitoring.
-Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases.
+Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as the number of years or collars increases.
 In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/powercmult6ay5aSf1.png"))
+knitr::include_graphics(here::here("figs/forPaper/powercmult6ay1aSf1.png"))
 ```
@@
 ```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort.
-When disturbance is high EVSI is low, indicating that local sampling does not alter status assessments.
-Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but in some circumstances there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases.
+When disturbance is high EVSI is low, indicating that local sampling does not change predicted status.Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring (number of years) as the number of years increases.
+Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction.
 See Figure \\@ref(fig:figStatusErrors) for additional scenario details.", cache=TRUE}
 
-knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay5aSf1.png"))
+knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay1aSf1.png"))
 ```
@@
 outcomes for caribou that fall outside the observed range of variation
-across the country, characterized here (Figure 2) by 95% predictive
+across the country, characterized here (Figure 2) by 95% prediction
 intervals from a national analysis (Johnson et al. 2020), are
@@
 calves causes low estimated recruitment, and together these errors tend
-to cancel one another out. We did not investigate sensitivity to the
-assumption that new recruits are 50% females, but we note that this is
-another possible source of error that would most likely lead to low
-recruitment estimates.
+to cancel one another out.
 
@@
 to classify visually. Documenting animals that are difficult to classify
-as “unknown” can help analysts minimize and account for classification
-bias [@ellington_divergent_2020], as can comparing classification
+as “unknown” can help analysts account for classification uncertainties [@ellington_divergent_2020;@dalgarno_bboutools_2024], as can comparing classification
 results from multiple independent observers. The probability of missing
@@
 the time in great detail is neither feasible nor necessary. We hope that
-a bias correction term for the simple recruitment-mortality model
-[@hatter_moose_1991; @decesare_estimating_2012;
+a bias correction term for the simple recruitment-
+mortality model that is used for many herd viability assessments [@hatter_moose_1991; @decesare_estimating_2012;
 @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011;
-@johnson_science_2020] that is used for many herd viability assessments
-will enable and encourage practitioners to examine the sensitivity of
-their conclusions to plausible composition survey error rates, and help
-clarify thinking about where, when, and why additional information is
+@johnson_science_2020; @dalgarno_bboutools_2024] will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is
 needed.

### 2024-10-09 analysis/scripts/0_bayesianModelExamples.R

@@ -13,8 +13,8 @@ stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
 stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
 stateScns = merge(stateScns,data.frame(interannualVar=c("list(R_CV=0.46,S_CV=0.087)","list(R_CV=0.23,S_CV=0.0435)")))
 
-stateScns$sQuantile=0.8
-stateScns$rQuantile = 0.8
+stateScns$sQuantile=0.9
+stateScns$rQuantile = 0.9
 scns=merge(monitoringScns,stateScns)
 
 scns = subset(scns,!(grepl("0.23",interannualVar,fixed=T)&(collarCount==0)))

### 2024-10-09 analysis/scripts/2_viewSensitivityResults.R

@@ -200,7 +200,7 @@ for(i in 1:length(pages)){
   probs$trueChange = probs$trueMean#probs$trueSize/probs$N0
   probs$predChange = probs$Mean#probs$projSize/probs$N0
 
-  probs$viableTrue = (probs$trueChange>0.99)&(probs$trueSize>10)
+  probs$viableTrue = (probs$trueChange>0.99)#&(probs$trueSize>10)
 
   probs$viablePred = (probs$predChange>0.99)#&(probs$projSize>10)
 

### 2024-10-09 analysis/paper/paper.Rmd
Draft MS sent to co-authors

@@
 We propose a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and reduce the risk that lack of local data will be used as a reason to delay conservation action.
-The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process.
+The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a term derived from a simple model of the recruitment survey observation process.
 We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.
 Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships.
-At lower disturbance levels where uncertainty is higher, deploying fewer collars for more years may be better than intensive short-term monitoring for improving the reliability of status projections.
+At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years may be better than intensive short-term monitoring for improving the reliability of status projections.
 keywords: |
@@
 active [see @nbckc_nbckc_2021], thus allowing estimation of adult
-survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for
-original approach, and @mcloughlin_declines_2003,
-@hervieux_widespread_2013, @eacker_web-based_2019, @lamb_effectiveness_2024 and @dalgarno_bboutools_2024 for
-examples of application to caribou). Recruitment is estimated from
+survival [e.g. @mcloughlin_declines_2003;
+@hervieux_widespread_2013; @eacker_web-based_2019, @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. Recruitment is estimated from
 aerial surveys of the composition of groups of animals that include
-collared animals [@nbckc_nbckc_2022, @eacker_web-based_2019, @dalgarno_bboutools_2024]. Groups typically also contain
-uncollared animals, and a critical assumption in these surveys is that
+collared animals [@nbckc_nbckc_2022, @eacker_web-based_2019, @dalgarno_bboutools_2024]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that
 different demographic groups are equally detectable
@@
 Survival and recruitment survey data can be combined in a simple
-recruitment-mortality population model [@hatter_moose_1991] that serves
+recruitment-mortality population model [@hatter_moose_1991;@hatter_revisiting_2020] that serves
 as the basis for many herd viability assessments
@@
 [@moeller_integrating_2021; @mcnay_demographic_2022;
-@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024], following broader trends in the
-analysis of population dynamics [@schaub_integrated_2021]. In Bayesian
-population models, available demographic or population data (e.g.
-abundance surveys, vital rate estimates, trend data) are combined into a
-single analysis that can permit novel parameter estimation, and improve
-precision, and account for multiple sources of uncertainty
+@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024], following broader trends in the analysis of population dynamics [@schaub_integrated_2021].
+An advantage of Bayesian demographic models is that they account for multiple sources of uncertainty
 [@schaub_integrated_2011; @eacker_assessing_2017;
-@schaub_integrated_2021]. These models are proving broadly useful for
-rapid detection of population declines, evaluation of management
-actions, and viability analysis of critically endangered species
-[@schaub_integrated_2021]. However, there is very little monitoring data
+@schaub_integrated_2021]. However, there is very little monitoring data
 available for some caribou populations [@johnson_science_2020] and there
@@
 models [@moeller_integrating_2021; @mcnay_demographic_2022;
-@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not thus far included demographic-disturbance relationships, and we are concerned that models
-informed only by local demographic data could be used to inflate
-uncertainty about the status of poorly monitored populations.
-
-We propose to reduce uncertainty in assessments and projections of local
+@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not thus far included demographic-disturbance relationships, and we are concerned that uncertainty in models
+informed only by local demographic data could be used to delay conservation action for poorly monitored populations. We propose to reduce uncertainty in assessments and projections of local
 caribou population viability by integrating prior information from
-national analysis of demographic-disturbance relationships with
+national analysis of demographic-disturbance relationships [@johnson_science_2020] with
 available local demographic data in a simple Bayesian population model
-for boreal caribou. In our simple proof-of-concept model, information is 
-shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020].
-The model represents an important improvement over less formally integrated methods for estimating and
-projecting impacts of disturbance on population population viability
+for boreal caribou.
+
+While we believe that using available information to reduce uncertainty
+is important, we also believe it is important to acknowledge remaining
+uncertainty. In general, parametric Bayesian population models include a
+number of simplifying assumptions, and misspecified models can yield
+biased or overconfident results [@schaub_integrated_2021]. More
+specifically, models that do not account for observation errors can be
+misleading [@schaub_integrated_2021]. Although several authors have
+pointed out that misclassification of adults and missing calves in
+composition surveys can bias recruitment estimates from those surveys
+[@ellington_divergent_2020; @decesare_estimating_2012], none have thus
+far examined sensitivity of Bayesian caribou population models to these
+sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
+@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. We address this gap with a bias term derived from a simple model of the recruitment survey observation
+process.
+
+Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability
 [@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
@@
 
-While we believe that using available information to reduce uncertainty
-is important, we also believe it is important to acknowledge remaining
-uncertainty. In general, parametric Bayesian population models include a
-number of simplifying assumptions, and misspecified models can yield
-biased or overconfident results [@schaub_integrated_2021]. More
-specifically, models that do not account for observation errors can be
-misleading [@schaub_integrated_2021]. Although several authors have
-pointed out that misclassification of adults and missing calves in
-composition surveys can bias recruitment estimates from those surveys
-[@ellington_divergent_2020; @decesare_estimating_2012] none have thus
-far examined sensitivity of Bayesian caribou population models to these
-sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
-@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. We address this gap with a bias correction term derived from a simple model of the recruitment survey observation
-process.
-
 # Methods
@@
 
-## Caribou population monitoring data and composition survey bias correction
+## Caribou population monitoring data and composition survey errors
 
@@
 
-To derive a composition survey bias correction term we assume each group
+To derive a composition survey bias term we assume each group
 of animals in a calf:cow composition survey contains one or more
@@
 and $z$ (see Supplement B for details). For simplicity, we combine the
-probabilities into a bias correction term $c$ that is integrated into
+probabilities into a bias term $c$ that is integrated into
 the Bayesian population model and simulations (below):
@@
 
-For this investigation we assume collars are deployed in January, and
+In simulated monitoring scenarios, we assume collars are deployed in January, and
 last for up to 6 years. The target number of collared females $n$ varies
@@
 simulated trajectories to confirm that the Bayesian model successfully
-integrates local data with priors from the national model, and that in the absence of local data Bayesian model outcomes match expectations from the national model. For
+integrates local data, and that prior predictions meet expectations from the national model. For
 reference, all variables and parameters are defined in supplemental
 Table A1.
-
-### A model for simulating example population dynamics informed by national disturbance-demographic relationships
 
@@
 binomially distributed with true survival probability $\dot{S}_t$:
-$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow
-ratio $\dot{R}_t$ is adjusted for sex ratio and misidentification biases
-(Supplement B) to get maximum
+$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$.
+
+Realized recruitment rate varies with population density
+[@lacy_vortex_2017], and the number of juveniles recruiting to the
+post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
+distributed function of the number of surviving post-juvenile females ($\dot{W}_t$)
+and the maximum
+potential recruitment of female calves per cow ($\dot{X}_t$):
+
+\begin{equation}
+  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}),
+  (\#eq:dotJ)
+\end{equation}
+
+Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et
+al. (2020) (Table A1), recruitment rate is lowest $(0.5\dot{X}_t)$ when
+$\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$ at intermediate
+population sizes, and declines to $0.6\dot{X}_t$ as the population
+reaches carrying capacity of $k=10000$.
+The post-juvenile female population in the next year includes both
+survivors and new recruits:
+
+\begin{equation}
+\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t).
+  (\#eq:dotN)
+\end{equation}
+
+The observable calf:cow
+ratio $\dot{R}_t$ is adjusted for sex ratio and composition survey errors
+(Supplement B) to get the maximum
 potential recruitment of female calves per cow $\dot{X}_t$:
@@
 $\dot{q}$ varies uniformly between 0 and 0.6.
-
-Realized recruitment rate varies with population density
-[@lacy_vortex_2017], and the number of juveniles recruiting to the
-post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
-distributed function of the number of surviving post-juvenile females
-and the adjusted recruitment rate:
-
-\begin{equation}
-  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
-  (\#eq:dotJ)
-\end{equation}
-
-Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et
-al. (2020) (Table A1), recruitment rate is lowest $(0.5\dot{X}_t)$ when
-$\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$ at intermediate
-population sizes, and declines to $0.6\dot{X}_t$ as the population
-reaches carrying capacity of $k=10000$.
-The post-juvenile female population in the next year includes both
-survivors and new recruits:
-
-\begin{equation}
-\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t).
-  (\#eq:dotN)
-\end{equation}
 
@@
 
-### Simulation of observed survival and recruitment
-
 In our simulated example populations, mortality risk does not vary over
@@
 
-Where appropriate, the Bayesian population model and national
-demographic model share the same structure, and we have selected
-notation to highlight the similarities. The difference between equations
+Where appropriate, the Bayesian and national population models share the same structure, and we have selected notation to highlight the similarities. The difference between equations
 \@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an
@@
 
-To account for sex ratio of calves and additional uncertainty caused by unknown
-misidentification errors (Supplement B) the
-estimated recruitment probability is adjusted to get expected female
+To account for sex ratio of calves and additional uncertainty caused by composition survey errors errors (Supplement B) estimated recruitment is adjusted to get expected female
 recruits per female: 
@@
 
-The composition survey bias correction term $c$ depends on the known
+The composition survey bias term $c$ depends on the known
 apparent number of adult females per collared animal $w$, the unknown
@@
 unknown probability of missing a calf $z$ (Supplement B). Prior
-uncertainty about the value of the bias correction term $c$ is
+uncertainty about the value of the bias term $c$ is
 approximated with a Log-normal distribution (Supplement B). Given the
@@
 monitoring scenario (Table \@ref(tab:monitoringScenarios)), we calculate
-the mean and standard deviation of 10,000 samples of $\log{ c}$ assuming
+the mean and standard deviation of 10,000 samples of $\log{c}$ assuming
 the ratio of young bulls to adult females $q$ varies uniformly between 0
@@
 Following others [@eacker_web-based_2019; @lamb_effectiveness_2024], Kaplan-Meier
-estimates of observed adult female survival probability $\hat{S}_t$ and
-associated precisions $\tau_t$ are estimated from known-fate radio
+estimates [@kaplan_nonparametric_1958] of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ are estimated from known-fate radio
 collar data using the *survival* R package [@therneau_modeling_2000],
@@
 Gaussian distributed function of the true survival probability,
-truncated between 0 and 1 [@eacker_web-based_2019,@lamb_effectiveness_2024]:
+truncated between 0 and 1:
 \begin{equation}
@@
 
+Again assuming a simple demographic model with a
+census in late winter of surviving females and new recruits to the
+post-juvenile class [@hatter_moose_1991; @hatter_revisiting_2020], the expected population growth rate (without interannual variation or demographic stochasticity) is $\tilde{\lambda}_t=\tilde{S}_t(1+c\tilde{R}_t/2)$.
+
 All regression coefficients are assumed to be Gaussian distributed, with
-prior means from Johnson et al's (2020) regression model estimates
+prior means from Johnson et al's (2020) regression models
 (Table \@ref(tab:priors)). We calibrated the standard deviations of the
@@
 
-As in the simulations, we assume a simple demographic model, with a
-census in late winter of surviving females and new recruits to the
-post-juvenile (aged 1+) class.
-Expected population growth rate (without interannual variation or demographic stochasticity) is $\tilde{\lambda}_t=\tilde{S}_t(1+c\tilde{R}_t/2)$.
-For reference, all variables and parameters are defined in supplemental Table A1.
-
-Models and analysis were implemented using the R language
+For reference, all variables and parameters are defined in supplemental Table A1. Models and analysis were implemented using the R language
 [@r_core_team_r_2023] and a variety of *tidyverse* packages
@@
 
-As suggested by Dunham et al. [-@dunham_strategic_2023] we also
+Following Dunham et al. [-@dunham_strategic_2023] we also
 calculate the expected value of perfect information (EVPI) and the
@@
 disturbance is high, we see a negative EVSI (lower right panel of Figure
-\@ref(fig:figEVSI)), and we do not see an associated decrease in the
+\@ref(fig:figEVSI)), but we do not see an associated decrease in the
 probability of correctly assessing population status (lower right panel
@@
 left side of right panels in Figure \@ref(fig:examplePredictions)). At
-lower levels of disturbance, local monitoring can reduce uncertainty,
-there are diminishing returns as the number of collars deployed or
-the number of years of monitoring increases (Figures
+lower levels of disturbance, local monitoring can reduce uncertainty, but
+there are diminishing returns as the number of years of monitoring increases (Figures
 \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). Reducing uncertainty is more difficult when interannual variation is high (Figures
@@
 regardless of whether disturbance is increasing or decreasing
-(Supplement D). Adding unknown bias increases uncertainty, but these
-increases are not large for the set of parameters we investigate
+(Supplement D). Acknowledging the possibility of composition survey errors increases uncertainty, but these increases are not large for the set of parameters we investigate
 (Supplement F).
@@
 The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading).
-Local population data (open triangles) reduces uncertainty about demographic rates, but local data is not required to predict viability (population growth <= 0.99) on highly disturbed landscapes.", cache=TRUE}
+Local population data (open triangles) reduces uncertainty about demographic rates, but local data is not required to predict viability (population growth <= 0.99) of highly disturbed populations.", cache=TRUE}
 
@@
 When disturbance is high the probability of correct status assessment is high even without local monitoring.
-Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as the number of years or collars increases.
+Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as the number of years increases.
 In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
@@
 ```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort.
-When disturbance is high EVSI is low, indicating that local sampling does not change predicted status.Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring (number of years) as the number of years increases.
+When disturbance is high EVSI is low, indicating that local sampling does not change predicted status.Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases.
 Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction.
@@
 
-We developed a simple Bayesian population model [building on
-@eacker_web-based_2019] to help address these challenges by
+We propose a simple Bayesian population model that helps address these challenges by
 acknowledging uncertainty associated with local demographic data and
@@
 nothing in our arguments, models or results that can or should be used
-to discourage much needed monitoring of the effects of conservation
+to discourage monitoring of the effectiveness of conservation
 actions. Even when no additional monitoring is required for assessing
@@
 
-Another novel aspect of this work is a term that allows for bias
-associated with misclassification of adults and missing calves in
+Another novel aspect of this work is a term that acknowledges possible bias
+caused by misclassification of adults and missing calves in
 composition surveys [@ellington_divergent_2020;
 @decesare_estimating_2012]. Lacking information about the true magnitude
-of this bias, we examined sensitivity to a range of assumed values, and
-found that adding unknown bias did increase uncertainty, but these
+of these errors, we examined sensitivity to a range of assumed values, and
+found that acknowledging bias did increase uncertainty, but these
 increases were not large (Supplement E). A key reason for this result is
@@
 to cancel one another out.
-
-Given lack of knowledge about true error rates, the most appropriate
-interpretation of our results is that better information on composition
-survey error rates is required in order to assess their importance.
-Fecal pellets are proving to be an interesting source of information on
-sex and age ratios in surveyed groups [@mcfarlane_genetic_2018;
-@jones_novel_2023]; for example, in analysis of 14 populations of boreal
-and southern mountain caribou, Jones et al. 2023 found that sampled
+Given lack of knowledge about true composition survey error rates, the most appropriate interpretation of our results is that better information is required in order to assess the importance of these errors.
+
+Alternatives to aerial composition surveys for estimating recruitment
+include fecal pellet analysis [@mcfarlane_genetic_2018;
+@mcfarlane_spatial_2021; @mcfarlane_identification_2022] and calf
+collaring [@ellington_divergent_2020]. In an analysis of fecal pellets from 14 populations of boreal and southern mountain caribou, Jones et al. [-@jones_novel_2023] found that sampled
 groups contained more females than males, and that mothers tend to stay
@@
 results from multiple independent observers. The probability of missing
-calves is more difficult to quantify and account for using aerial survey
-methods alone, but likely increases with vegetation cover.
-
-Alternatives to aerial composition surveys for estimating recruitment
-include fecal pellet analysis [@mcfarlane_genetic_2018;
-@mcfarlane_spatial_2021; @mcfarlane_identification_2022] and calf
-collaring [@ellington_divergent_2020]. Each method has costs and
-benefits, and the reality is that monitoring everything everywhere all
-the time in great detail is neither feasible nor necessary. We hope that
-a bias correction term for the simple recruitment-
-mortality model that is used for many herd viability assessments [@hatter_moose_1991; @decesare_estimating_2012;
+calves likely increases with vegetation cover, and is more difficult to quantify.
+
+Each survey method has costs and benefits, and the reality is that monitoring everything everywhere all the time in great detail is neither feasible nor necessary. We hope that
+adding a bias term to the simple recruitment-mortality model that is used for many herd viability assessments [@hatter_moose_1991;@hatter_revisiting_2020;@decesare_estimating_2012;
 @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011;
@@
 populations, and distinguishing that from variation among years.
-We have not done this because data owners have not authorized the use of the data for that purpose, and obtaining permission to use caribou demographic data is challenging.
+We have not reanalyzed the national data using different methods because the data is owned by many jurisdictions, and obtaining permission from all of these jurisdictions would be difficult and time-consuming.
 To move ahead with available published information we constrained our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower than or higher than average remain lower or higher than average over time.
-The Bayesian model is not similarly constrained, and can accommodate
+Note that the Bayesian model is not similarly constrained, and can accommodate
 evidence of a shift from unexpectedly low to unexpectedly high
@@
 composition survey errors.
-However, the implementation is a research prototype, not a fully developed analysis tool.
-The more fully developed and documented *bboutools* R package [@dalgarno_bboutools_2024] is a better starting point for development of tools to enable local decision makers to inform decisions with their  existing data.
-A possible way forward would be to extend that package to include demographic-disturbance relationships and composition survey errors.
+However, it is a research prototype, not a fully developed analysis tool.
+The more fully developed and documented *bboutools* R package [@dalgarno_bboutools_2024] is likely a better starting point for development of tools to enable local decision makers to inform decisions with their  existing data.
+A possible next step is to to extend that package to include demographic-disturbance relationships and composition survey errors.
 
@@
 information from aerial or fecal dna surveys, and variation in demographic rates among age classes [@moeller_integrating_2021; @mcnay_demographic_2022] into Bayesian population models.
-However, in many contexts there is not enough to data to estimate more complex model parameters.
-
-At present, our analysis is intended as a proof-of-concept to show the value of integrating national demographic-disturbance relationships with local data, accounting for
+However, in many contexts there is not enough to data to estimate more complex model parameters, and simple models remain relevant.
+
+Our proof-of-concept analysis shows the value of integrating national demographic-disturbance relationships with local data, accounting for
 composition survey errors, and assessing the value of local monitoring

### 2024-10-09 analysis/paper/supplement.Rmd
Ongoing - manuscript and method tweaking.
Draft MS sent to co-authors

@@
 |$S_t$|probability|Adult female survival.|
-|$N_t$|animals|Adult female population size.|
-|$\bar{\lambda}_{y,t}$|rate|Average population growth over $y$ years.|
-|$X_t$|rate|Recruitment adjusted for sex ratio and misidentification bias.|
-|$W_t$|animals|Surviving adult females before recruitment.|
-|$J_t$|animals|New recruits to adult age class.|
-|$\lambda_t$|rate|Annual population growth.|
+|$X_t$|rate|Recruitment adjusted for sex ratio and composition survey errors.|
 |**Estimated Parameters**|
@@
 |$\nu_{S}$||Coefficient of variation in survival among years.|
+|$\tilde{lambda}_{t}$|rate|Expected annual population growth.|
 |**True state**|
@@
 |$\bar{S}_t$|probability|Expected adult female survival.|
+|$\dot{\lambda}_t$|rate|Annual population growth.|
+|$\bar{lambda}_{t}$|rate|Expected annual population growth.|
 |$\dot{\beta}^R_0$||Recruitment intercept.|
@@
 |$\dot{\beta}^S_a$||Anthropogenic disturbance slope for survival.|
+|$\phi^R \sim \text{Normal}(19.862,2.229)$||Precision of recruitment-disturbance relationship.|
+|$\phi^S \sim \text{Normal}(63.733,8.311)$||Precision of survival-disturbance relationship.|
 |$p_0=1$||Maximum recruitment multiplier.|
 |$p_k=0.6$||Recruitment multiplier at carrying capacity.|
-|$k=50000$||Carrying capacity.|
+|$k=10000$||Carrying capacity.|
 |$b=4$||Density dependence shape parameter.|
@@
 |$o=1 \text{ or } 4$|years|Years between collar deployments.|
-|$y=3$|years|Assessment period for population growth rate.|
 |$w=3,6,\text{ or } 9$|ratio|The apparent number of adult females per collared animal in composition survey.|

### 2024-10-09 analysis/scripts/0_bayesianModelExamples.R

@@ -225,11 +225,12 @@ priorResult = caribouMetrics:::runScnSet(scnsN,eParsIn,simBig,getKSDists=F,print
 
 priorResult$obs.all=NULL; priorResult$sim.all=NULL
 
+
 yr_scale1 <- scale_x_continuous(breaks = 2023:2072 %>% .[0:9*5+1],
                                 labels = 2009:2058 %>% .[0:9*5+1] )
 yr_scale2 <- scale_x_continuous(breaks = 2009:2058 %>% .[0:9*5+1],
                                 labels = 2009:2058 %>% .[0:9*5+1] )
-lambdaPrior1 =  plotRes(priorResult, "Population growth rate", lowBound=0.5,
+lambdaPrior1 =  plotRes(priorResult, "Population growth rate", lowBound=0,
                        legendPosition="none",breakInterval=breakInterval,
                        labFontSize=labFontSize,facetVars=c("uMax","zMax"))+
   yr_scale1 +

### 2024-10-15 cloud/setup_azure.sh
cloud run s11 and s12

@@ -19,7 +19,12 @@ sastoken=`az storage container generate-sas --account-name ecdcwls --expiry $end
 # I actually created the SASURL manually bc it wouldn't let me do more than 7 days through CLI
 sasurl=https://ecdcwls.blob.core.windows.net/sendicott/?$sastoken
 
-setName="s10"
+setName="s12"
+
+# 30 batches have 5550 runs and 60 have 7400 it would be smart to have them on
+# the same nodes so the node can close when finished
+# 30: 4 nodes with 8 slots
+# 60: 8 nodes with 8 slots
 
 nBatches=10
 
@@ -33,13 +38,25 @@ rm -r cloud/pool_json
 # Updates sasurl and setName in files, might need to change pool slots and target nodes
 Rscript --vanilla "cloud/make_batch_scripts.R" $setName $sasurl
 
+# Check what container is empty
+az storage blob list -c sendicott --account-name ecdcwls --sas-token $sastoken \
+--query "[].{name:name}" --output yaml
+
 az storage copy -d $sasurl -s cloud/task_scripts --recursive
 
 #### Create pool, job, tasks ##########################
 az batch pool create --json-file cloud/pool_json/caribou_add_pool1.json
 az batch job create --pool-id $poolName --id $jobName
 
-for ((i=1;i<=nBatches;i++))
+# all batches
+# for ((i=1;i<=nBatches;i++))
+
+# this approach only sort of worked because some tasks had to be restarted
+# Only batches with 5550 rows
+# for i in 1 2 3 4 5 6 7 8 9 10 31 32 33 34 35 36 37 38 39 40 61 62 63 64 65 66 67 68 69 70
+
+# Only batches with 7400 rows
+for i in 11 12 13 14 15 16 17 18 19 20 21 22 23 24 25 26 27 28 29 30 41 42 43 44 45 46 47 48 49 50 51 52 53 54 55 56 57 58 59 60 71 72 73 74 75 76 77 78 79 80 81 82 83 84 85 86 87 88 89 90
 do
 	echo "setting up task" $i
 	az batch task create --json-file cloud/task_jsons/caribouDemo$i.json --job-id $jobName
@@ -82,7 +99,7 @@ az batch task list --job-id $jobName --query "{tasks: [?state == 'completed'].[i
 # Summary of task counts by state
 az batch job task-counts show --job-id $jobName
 
-# added taskslotspernode
+# added taskslotspernode # doesn't seem to recognize needs more nodes when slots full
[…]

### 2024-10-15 analysis/paper/paper.Rmd
revisions in progress

@@
 Demographic projections derived from local population-specific data are  uncertain for populations with limited monitoring.
-We propose a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and reduce the risk that lack of local data will be used as a reason to delay conservation action.
+We propose a simple Bayesian population model that integrates prior information from a national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and to reduce the risk that lack of local data will be used as a reason to delay conservation action.
 The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a term derived from a simple model of the recruitment survey observation process.
@@
 populations and likely impacts of management are needed to inform
-conservation action but effective monitoring is challenging and costly
+conservation action, but effective monitoring is challenging and costly
 [@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;
@@
 are limited [@buxton_balancing_2022; @mccarthy_financial_2012;
-@gerber_conservation_2016; @wintle_spending_2019], delays increase risk
+@gerber_conservation_2016; @wintle_spending_2019], delays in action increase risk
 [@martin_acting_2012], and some populations have been monitored to
 extinction [@lindenmayer_counting_2013]. A recent analysis of action
-plans for Canadian species at risk found half of tasks are research and
+plans for Canadian species at risk found that half of tasks are research and
 monitoring, and this proportion is highest for species with higher risk
@@
 action [@buxton_balancing_2022]. Analysis of the implications of
-monitoring outcomes for decisions can help clarify whether additional
+monitoring outcomes for decisions can help to clarify whether additional
 monitoring is necessary to inform conservation action
@@
 
-Analysis methods that can help make better use of available monitoring
+Analysis methods that can make better use of available monitoring
 data to inform decisions include Bayesian population modelling
@@
 [@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
-@gregory_structured_2012]. To be useful, these methods must be tailored
+@gregory_structured_2012]. But to be useful, these methods must be tailored
 to particular species and decision contexts. Here, we offer a simple
@@
 model with simulations of plausible population trajectories
-[@stewart_climate-informed_2023; @dyson_existing_2022] to show how the
-need for local monitoring varies with landscape condition, and to assess
+[@stewart_climate-informed_2023; @dyson_existing_2022] to (i) show how the
+value of local monitoring varies with landscape condition, and (ii) assess
 the ability of alternative monitoring scenarios to reduce the risk of
@@
 survival and recruitment decrease with the percentage of the study area
-that is disturbed, and no populations of caribou are considered
-self-sustaining in study areas with high anthropogenic disturbance
-[@eccc_scientific_2011; @johnson_science_2020]. The causes of these
+that is disturbed, and no populations of caribou inhabiting areas with high anthropogenic disturbance are considered self-sustaining [@eccc_scientific_2011; @johnson_science_2020].
+The mechanisms behind these
 demographic-disturbance relationships include direct impacts on habitat
@@
 recruitment as disturbance increases are also observed over time in
-places where there is sufficient historic demographic data
+places where there are historic demographic data
 [@rudolph_demographic_2017; @hervieux_widespread_2013], but many places
@@
 [@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a
-need to assess and compare effects of boreal caribou conservation
-actions with careful monitoring. Our focus in this paper is assessing
+need to evaluate and compare effects of boreal caribou conservation
+actions with careful monitoring. Our focus in this paper is on assessing
 and projecting self-sustainability of populations in the absence of
@@
 reduce uncertainty about whether there is a need for action. Where
-actions are taken, outcomes should be monitored; the simple model we
+actions are taken, outcomes should be monitored; the model we
 propose here does not include effects of management interventions, and
@@
 can substantially reduce uncertainty in local demographic projections
-and status assessments, improve viability projections on changing
-landscapes, and help clarify where, why and when additional data is
+and status assessments, can improve viability projections on changing
+landscapes, and can help clarify where, why and when additional data is
 required to inform decisions and project outcomes on changing
-landscapes. We hope that this approach can help clarify where available
+landscapes. We hope that this approach will help to identify where available
 information is sufficient for assessing and projecting population
@@
 the Bayesian population model and simulations (below):
-$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$
+
+\begin{equation}
+  c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2.
+  (\#eq:c)
+\end{equation}
 
@@
 In simulated monitoring scenarios, we assume collars are deployed in January, and
-last for up to 6 years. The target number of collared females $n$ varies
+continue to function for up to 6 years. The target number of collared females $n$ varies
 among scenarios, as does the total duration of the monitoring program
@@
 variation in recruitment and survival among example populations (Figure
-\@ref(fig:exampleTrajectories)). To ensure results are comparable across
+\@ref(fig:exampleTrajectories)). To ensure that results are comparable across
 monitoring scenarios and that populations remain extant through the
@@
 ```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships.
-To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variation by 75\\% in these examples ($v_R=0.115,v_S=0.02175$).
+To highlight how variation among trajectories persists as disturbance changes, we have reduced interannual variation by 75\\% in these examples ($v_R=0.115,v_S=0.02175$).
 Orange dots show means of 500 example trajectories.
@@
 
-For consistency with Johnson et al. [-@johnson_science_2020], we model survival
+As done by Johnson et al. [-@johnson_science_2020], we model survival
 probability as an adjusted function of buffered anthropogenic
@@
 JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
-to fit the Bayesian model. We ran 15,000 iterations of 4 chain,
+to fit the Bayesian model. We ran 15,000 iterations of 4 chains,
 discarded the first 10,000 interations, and used a thinning rate of 2.
@@
 Bayesian model can effectively combine local demographic data with
-priors from national demographic-disturbance relationships to reduce
+priors from a national demographic-disturbance relationships to reduce
 uncertainty in boreal caribou demographic projections (Figures
@@
 effort in each year (i.e. number of collars) or the number of years of
-monitoring (Figure \@ref(fig:figDiffs)) increase. Uncertainty is higher
-when the forecast horizon is long (compare 5 and 20 year projections in
-Figure \@ref(fig:figDiffs)), and when interannual variation is high (Figure
-\@ref(fig:examplePredictions)). Uncertainty is also higher in our high
-anthropogenic disturbance scenarios (Figure \@ref(fig:figDiffs)) because
-stochastic variation in population growth rates increase as populations
-become very small (as in top right panel of Figure
-\@ref(fig:examplePredictions)).
+monitoring (Figure \@ref(fig:figDiffs)) increase. Reducing uncertainty is more challenging when interannual variation is high (Figure
+\@ref(fig:examplePredictions)). Increasing the forecast horizon [@petchey_ecological_2015] does not increase uncertainty about expected population growth (compare 5 and 20 year projections in
+Figure \@ref(fig:figDiffs)) in these scenarios.
 
@@
 condition, available resources for monitoring, and remoteness. In a few
-regions enough data has been collected over time to characterize local
+regions enough data have been collected over time to characterize local
 demographic-disturbance relationships [@rudolph_demographic_2017;
@@
 
-We propose a simple Bayesian population model that helps address these challenges by
+We have proposed a simple Bayesian population model that helps address these challenges by
 acknowledging uncertainty associated with local demographic data and
@@
 @johnson_science_2020] because it reconciles and accounts for
-uncertainties in local demographic data and demographic-disturbance
+uncertainties in local demographic data and in demographic-disturbance
 relationships.
@@
 causes a log-linear decrease in caribou demographic rates (equations 4,
-5, 10 and 13), reflecting a scientific consensus informed by a number of
-lines of evidence that cumulative anthropogenic disturbance has a
+5, 10 and 13), reflecting a scientific consensus that cumulative anthropogenic disturbance has a
 negative impact on boreal caribou [@eccc_scientific_2011;
@@
 productive for industrial forestry in Eastern Canada)
-[@eccc_scientific_2024], and use the priors from regional
+[@eccc_scientific_2024], and to use the priors from regional
 demographic-disturbance relationships to inform demographic projections
@@
 of these errors, we examined sensitivity to a range of assumed values, and
-found that acknowledging bias did increase uncertainty, but these
+found that while acknowledging bias did increase uncertainty, these
 increases were not large (Supplement E). A key reason for this result is
@@
 that surveyed groups contain more cows than bulls, and that there is an
-equal probability of misclassifying cows and young bulls and vice versa,
+equal probability of misclassifying cows as young bulls and vice versa,
 then misclassification errors cause high estimated recruitment. Missing
@@
 populations, and distinguishing that from variation among years.
-We have not reanalyzed the national data using different methods because the data is owned by many jurisdictions, and obtaining permission from all of these jurisdictions would be difficult and time-consuming.
+We have not reanalyzed the national data using different methods because the data are owned by many jurisdictions, and obtaining permission from all of these jurisdictions would be difficult and time-consuming.
 To move ahead with available published information we constrained our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower than or higher than average remain lower or higher than average over time.
@@
 of demographic-disturbance relationships among populations
-[@rudolph_demographic_2017] so it is important that the Bayesian model
+[@rudolph_demographic_2017], so it is important that the Bayesian model
 allows for a greater range of possibilities than our more constrained
 simulated example trajectories. Although the simulated trajectories do
-not include all plausible possibilities we believe they are sufficient
+not include all plausible possibilities, we believe they are sufficient
 to demonstrate how a Bayesian model can be used to integrate local data
@@
 The more fully developed and documented *bboutools* R package [@dalgarno_bboutools_2024] is likely a better starting point for development of tools to enable local decision makers to inform decisions with their  existing data.
-A possible next step is to to extend that package to include demographic-disturbance relationships and composition survey errors.
+A possible next step is to extend that package to include demographic-disturbance relationships and composition survey errors.
 

### 2024-10-15 analysis/paper/paper.Rmd
submission prep in progress

@@
 
-NOTE: please provide comments and suggestions for the introduction here:
-<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit?usp=sharing&ouid=111541730342945675365&rtpof=true&sd=true>.
-
 The number of species threatened with extinction is high and getting
@@
 is a risk that lack of local data will be used as a reason to delay
-conservation action [@boan_climate_2018]. Bayesian caribou population
+conservation action [@bissonnette_controverses_2020; @boan_climate_2018]. Bayesian caribou population
 models [@moeller_integrating_2021; @mcnay_demographic_2022;
@@
 viability, and thereby reduce the risk that uncertainty is used as a
-reason to delay conservation action for imperiled populations.
+reason to delay conservation action for imperiled populations [@bissonnette_controverses_2020; @boan_climate_2018].
 
@@
 binomially distributed with true survival probability $\dot{S}_t$:
-$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$.
-
-Realized recruitment rate varies with population density
-[@lacy_vortex_2017], and the number of juveniles recruiting to the
+$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Realized recruitment rate varies with population density [@lacy_vortex_2017], and the number of juveniles recruiting to the
 post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
@@
 
-NOTE: Please provide comments and suggestions for the discussion here:
-<https://docs.google.com/document/d/1RD7bJ6yhtBJULjJFRJfHzObQscteyObM/edit>.
-
 ## Bayesian integration of local demographic data & national demographic-disturbance relationships
@@
 methods for estimating and projecting impacts of disturbance on
-population population viability [@eccc_scientific_2011;
+population caribou population viability [@eccc_scientific_2011;
 @johnson_science_2020] because it reconciles and accounts for
@@
 will be used as a reason to delay conservation action for highly
-disturbed and poorly monitored populations. Integrating
+disturbed and poorly monitored populations [@bissonnette_controverses_2020; @boan_climate_2018]. Integrating
 demographic-disturbance relationships into a viability projection model

### 2024-10-15 analysis/paper/paper.Rmd
bib changes

--- 2024-10-15 analysis/paper/paper.Rmd
@@
 bibliography: CaribouDemographicModel.bib
-csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
+csl: "../templates/ecological-informatics.csl" # Insert path for the bib-style
 abstract: |

### 2024-10-16 analysis/paper/paper.Rmd
add supp mat captions

@@
 
+# Supplementary material
+Supplement A: Definitions of variables and parameters in the Bayesian population model
+Supplement B: Model of bias in recruitment estimates from calf:cow surveys
+Supplement C: Calibration of priors
+Supplement D: Alternative disturbance and monitoring scenarios
+Supplement E: Expected value of perfect information
+Supplement F: Sensitivity to unknown bias in recruitment estimates from calf:cow surveys
+
 # Acknowledgements

### 2024-10-16 analysis/paper/paper.Rmd
improve table 1 formatting

@@
 
-knitr::kable(tbl,caption="Monitoring scenario parameters.")
+tbl <- tbl %>% separate_wider_delim(name, names = c("symbol", "name"), delim = ", ") %>% 
+  select(-name, symbol, units, description, values)
+
+knitr::kable(tbl,caption="Monitoring scenario parameters and values used in simulations.",
+             col.names = str_to_title(colnames(tbl)))
 ```

### 2024-10-16 analysis/paper/paper.Rmd
fixing things

--- 2024-10-15 analysis/paper/paper.Rmd
@@
 bibliography: CaribouDemographicModel.bib
-csl: "../templates/ecological-informatics.csl" # Insert path for the bib-style
+csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
 abstract: |
@@
 survival [e.g. @mcloughlin_declines_2003;
-@hervieux_widespread_2013; @eacker_web-based_2019, @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. Recruitment is estimated from
+@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. Recruitment is estimated from
 aerial surveys of the composition of groups of animals that include
-collared animals [@nbckc_nbckc_2022, @eacker_web-based_2019, @dalgarno_bboutools_2024]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that
+collared animals [@nbckc_nbckc_2022; @eacker_web-based_2019; @dalgarno_bboutools_2024]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that
 different demographic groups are equally detectable

### 2024-10-16 analysis/paper/paper.Rmd
Arrange figures and captions at end

--- 2024-10-16 analysis/paper/paper.Rmd
@@
 
-```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios.
-Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%).
-These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, and not intended to reflect real landscape change.
-In all scenarios, the projection period begins in 2024, and the monitoring period occurs prior to the projection period.", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay1aSf1.png"))
-
-```
-
 ```{=html}
@@
 Realized population growth rate is $\dot{\lambda}_t=\dot{N}_t/\dot{N}_{t-1}$ Expected population growth rate (without interannual variation, demographic stochasticity, or effects of population density) is $\bar{\lambda}_t=\bar{S}_t(1+\dot{c}\bar{R}_t/2)$.
-
-```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships.
-To highlight how variation among trajectories persists as disturbance changes, we have reduced interannual variation by 75\\% in these examples ($v_R=0.115,v_S=0.02175$).
-Orange dots show means of 500 example trajectories.
-For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models.", cache=TRUE}
-
-knitr::include_graphics(here::here("analysis/paper/figs/DemographicRates.png"))
-
-```
 
@@
 (Supplement F).
-
-```{r examplePredictions, fig.cap="Prior (a) and posterior predictions from the Bayesian model for population trajectories with low (b: $v_R=0.23,v_S=0.0435$) and high (c: $v_R=0.46,v_S=0.087$) interannual variation.
-In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%.
-The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys.
-The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading).
-Local population data (open triangles) reduces uncertainty about demographic rates, but local data is not required to predict viability (population growth <= 0.99) of highly disturbed populations.", cache=TRUE}
-
-knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamples.png"))
-```
-
-```{r figDiffs, fig.cap="The distribution of differences between true expected population growth rates and Bayesian posterior expected growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
-Differences decrease as the number of years of monitoring and the number of collars increase.
-In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay1aSf1.png"))
-```
-
-```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios).
-When disturbance is high the probability of correct status assessment is high even without local monitoring.
-Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as the number of years increases.
-In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/powercmult6ay1aSf1.png"))
-```
-
-```{r figEVSI, fig.cap="The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort.
-When disturbance is high EVSI is low, indicating that local sampling does not change predicted status.Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases.
-Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction.
-See Figure \\@ref(fig:figStatusErrors) for additional scenario details.", cache=TRUE}
-
-knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay1aSf1.png"))
-```
 
@@
 
+# Figures
+
+```{r figDisturbance, fig.cap=".", cache=TRUE}
+knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay1aSf1.png"))
+
+```
+
+```{r exampleTrajectories, fig.cap=".", cache=TRUE}
+knitr::include_graphics(here::here("analysis/paper/figs/DemographicRates.png"))
+
+```
+
+```{r examplePredictions, fig.cap=".", cache=TRUE}
+
+knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamples.png"))
+```
+
+```{r figDiffs, fig.cap=".", cache=TRUE}
+knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay1aSf1.png"))
+```
+
+```{r figStatusErrors, fig.cap=".", cache=TRUE}
+knitr::include_graphics(here::here("figs/forPaper/powercmult6ay1aSf1.png"))
+```
+
+```{r figEVSI, fig.cap=".", cache=TRUE}
+
+knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay1aSf1.png"))
+```
+
+## Figure Captions
+Figure 1: Anthropogenic disturbance and monitoring scenarios.
+Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%).
+These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, and not intended to reflect real landscape change.
+In all scenarios, the projection period begins in 2024, and the monitoring period occurs prior to the projection period.
+
+Figure 2: Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships.
+To highlight how variation among trajectories persists as disturbance changes, we have reduced interannual variation by 75% in these examples ($v_R=0.115,v_S=0.02175$).
+Orange dots show means of 500 example trajectories.
+For reference, black lines and and shaded bands show means and 95% predictive intervals from Johnson et al's (2020) regression models.
+
+Figure 3: Prior (a) and posterior predictions from the Bayesian model for population trajectories with low (b: $v_R=0.23,v_S=0.0435$) and high (c: $v_R=0.46,v_S=0.087$) interannual variation.
+In this example scenario anthropogenic disturbance increases by 2% per year from 0 to 100%.
+The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys.
+The prior means and 95% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5% and 97.5% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading).
+Local population data (open triangles) reduces uncertainty about demographic rates, but local data is not required to predict viability (population growth <= 0.99) of highly disturbed populations.
+
+Figure 4: The distribution of differences between true expected population growth rates and Bayesian posterior expected growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
+Differences decrease as the number of years of monitoring and the number of collars increase.
+In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).
+
+Figure 5: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
+When disturbance is high the probability of correct status assessment is high even without local monitoring.
+Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as the number of years increases.
+In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).
+
+Figure 6: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort.
+When disturbance is high EVSI is low, indicating that local sampling does not change predicted status.Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases.
+Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction.
+See Figure \@ref(fig:figStatusErrors) for additional scenario details.
+
 ### Colophon

### 2024-10-16 analysis/paper/paper.Rmd

@@
 Figure 6: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort.
-When disturbance is high EVSI is low, indicating that local sampling does not change predicted status.Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases.
+When disturbance is high EVSI is low, indicating that local sampling does not change predicted status.
+Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases.
 Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction.

### 2024-10-16 analysis/scripts/0_bayesianModelExamples.R
save figs to pdf, change width, make caps in labels consistent sentence case

@@ -117,9 +117,12 @@ plts <- ggpubr::ggarrange(recPrior, survPrior, lambdaPrior,
 ggpubr::ggarrange(plts, leg, ncol = 2, widths = c(6,1),heights=0.75)+bgcolor("white")
 
 ggsave(paste0(baseDir,"/analysis/paper/figs/bayesianModelExamples.png"),
-       width = 12*0.8, height = 4.6*2, units = "in",
+       width = 9.6*0.779, height = 9.2*0.779, units = "in",
        dpi = 1200)
 
+ggsave(paste0(baseDir,"/analysis/paper/figs_submit/bayesianModelExamples.pdf"),
+       width = 9.6*0.779, height = 9.2*0.779, units = "in",
+       dpi = 1200)
 ###############################
 #Without bias adjustment
 

### 2024-10-16 analysis/scripts/0_nationalModelTrajectories.R

@@ -139,7 +139,7 @@ plot_lambda <- ggplot(oo,
             alpha = 1) + scale_color_manual(values=pal)+
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
   xlab("Anthropogenic disturbance (%)") +
-  ylab(expression("Population Growth Rate " * dot(lambda)[t])) +
+  ylab(expression("Population growth rate " * dot(lambda)[t])) +
   theme(legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
 
 str(pars1)
@@ -156,7 +156,7 @@ plot_recruitment3 <- ggplot(data = rateSummaries,
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
   scale_y_continuous(limits = c(0, 0.5), breaks = c(0, 0.1, 0.2, 0.3, 0.4,0.5)) +
   xlab("Anthropogenic disturbance (%)") +
-  ylab(expression("Adjusted Recruitment " * dot(X)[t])) +
+  ylab(expression("Adjusted recruitment " * dot(X)[t])) +
   theme(legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
 plot(plot_recruitment3)
 
@@ -184,8 +184,8 @@ base1 <- ggplot(data = rateSummaries,
   scale_color_manual(values=pal)+
   geom_line(data=johnsonCompare, colour = "black", size = 1.5) +
   geom_line(colour = "#ef8a62", size = 2, linetype = "dotted") +
-  xlab("Anthropogenic Disturbance (%)") +
-  ylab(expression("Adult Female Survival " * dot(S)[t])) +
+  xlab("Anthropogenic disturbance (%)") +
+  ylab(expression("Adult female survival " * dot(S)[t])) +
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
   scale_y_continuous(limits = c(0.5, 1)) +
   theme(legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
@@ -195,7 +195,9 @@ plot(base1)
 ggpubr::ggarrange(plot_recruitment3,base1,plot_lambda,labels = "",
                   ncol = 3, vjust = 1)
 
[…]

### 2024-10-16 analysis/scripts/2_viewSensitivityResults.R

@@ -39,9 +39,9 @@ setLTYVar <-function(scns){
 
 scns = setLTYVar(scns)
 if(is.element("interannualVar",names(scns))){
-  ltyLabel = "Interannual\nVariation"
+  ltyLabel = "Interannual\nvariation"
 }else{
-  ltyLabel = "Collar\nRenewal\nInterval"
+  ltyLabel = "Collar\nrenewal\ninterval"
 }
 
 
@@ -168,12 +168,18 @@ for(i in 1:length(pages)){
   timelineLabs = unique(subset(distScns,(Year==startYear)|(Year>=2023),select=c(Year,Timeline,Anthro,DisturbanceScn,grp)))
   timelineLabs = timelineLabs[order(timelineLabs$Year),]
 
-  png(here::here(paste0("figs/",setName,"/distScns",batchStrip(p),".png")),
-      height = 4, width = 5.51, units = "in",res=600)
   base=ggplot(distScns,aes(x=Year,y=Anthro,col=DisturbanceScn,group=grp))+geom_line()+geom_point(data=timelineLabs)+
     theme_bw()+theme(axis.text.x = element_text(angle=90,vjust=0.5,hjust=1,size=8)) + labs(color="Anthropogenic\nDisturbance\nScenario")+
     scale_x_continuous(name="Timeline", breaks=timelineLabs$Year, labels=timelineLabs$Timeline)+ylab("Anthropogenic Disturbance")+
     scale_color_discrete(type=(pal4b))
+
+  png(here::here(paste0("figs/",setName,"/distScns",batchStrip(p),".png")),
+      height = 4, width = 5.51, units = "in",res=600)
+  print(base)
+  dev.off()
+
+  pdf(here::here(paste0("figs/",setName,"/distScns",batchStrip(p),".pdf")),
+      height = 4, width = 5.51)
   print(base)
   dev.off()
 
@@ -247,13 +253,19 @@ for(i in 1:length(pages)){
   probs$CollarYrs = as.numeric(as.character(probs$NumCollars))*probs$obsYears
   for(pp in pagesCa){
     #pp=pagesCa[1]
-    png(here::here(paste0("figs/",setName,"/diffs",pp,".png")),
-        height = 4, width = 7.48, units = "in",res=600)
[…]

### 2024-10-16 analysis/scripts/3_EVSIPlots.R

@@ -32,9 +32,9 @@ probsSum$AnthroScn = factor(probsSum$AnthroScn, levels = c("low","low-med","med-
 
 
 if(is.element("interannualVar",names(probsSum))){
-  ltyLabel = "Interannual\nVariation"
+  ltyLabel = "Interannual\nvariation"
 }else{
-  ltyLabel = "Collar\nRenewal\nInterval"
+  ltyLabel = "Collar\nrenewal\ninterval"
 }
 
 
@@ -88,12 +88,18 @@ probsSum$CollarYrs = as.numeric(as.character(probsSum$NumCollars))*probsSum$obsY
 EVuncertainty$YearsOfProjection=as.factor(EVuncertainty$YearsOfProjection)
 for(pp in pagesC){
   #pp=pagesC[1]
+  base=ggplot(subset(probsSum,pageLabC==pp),aes(x=obsYears,y=EVSI,col=NumCollars,linetype=RenewalInterval,group=grp))+geom_line()+
+    facet_grid(YearsOfProjection~AnthroScn,labeller="label_both")+labs(color="Number of\n collars", linetype=ltyLabel)+
+    theme_bw()+xlab("Years of monitoring")+ylab("Expected value of sample information EVSI")+
+    scale_color_discrete(type=(pal4))
+
   png(here::here(paste0("figs/",setName,"/EVSI",pp,".png")),
       height = 3.7, width = 7.48, units = "in",res=600)
-  base=ggplot(subset(probsSum,pageLabC==pp),aes(x=obsYears,y=EVSI,col=NumCollars,linetype=RenewalInterval,group=grp))+geom_line()+
-    facet_grid(YearsOfProjection~AnthroScn,labeller="label_both")+labs(color="Number of\n Collars", linetype=ltyLabel)+
-    theme_bw()+xlab("years of monitoring")+ylab("Expected Value of Sample Information EVSI")+
-    scale_color_discrete(type=(pal4))
+  print(base)
+  dev.off()
+
+  pdf(here::here(paste0("figs/",setName,"/EVSI",pp,".pdf")),
+      height = 3.7, width = 7.48)
   print(base)
   dev.off()
 

### 2024-10-17 analysis/paper/paper.Rmd
save pdfs in figs_submit folder and name according to figure number

@@
   file.copy(currentFile, newFile,overwrite=T)
+
+  currentFile = here::here(paste0("figs/",resultScenario,"/",f,addBit,".pdf"))
+  newFile = here::here(paste0("analysis/paper/figs_submit/",f,addBit,".pdf"))
+  file.copy(currentFile, newFile,overwrite=T)
 }
@@
 knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay1aSf1.png"))
-
+file.rename(here::here("analysis/paper/figs_submit/distScnscmult6ay1aSf1.pdf"), 
+            here::here("analysis/paper/figs_submit/Figure1_distScnscmult6ay1aSf1.pdf"))
 ```
@@
 knitr::include_graphics(here::here("analysis/paper/figs/DemographicRates.png"))
-
+file.rename(here::here("analysis/paper/figs_submit/DemographicRates.pdf"), 
+            here::here("analysis/paper/figs_submit/Figure2_DemographicRates.pdf"))
 ```
@@
 ```{r examplePredictions, fig.cap=".", cache=TRUE}
-
 knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamples.png"))
+file.rename(here::here("analysis/paper/figs_submit/bayesianModelExamples.pdf"), 
+            here::here("analysis/paper/figs_submit/Figure3_bayesianModelExamples.pdf"))
 ```
@@
 knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay1aSf1.png"))
+file.rename(here::here("analysis/paper/figs_submit/diffscmult6ay1aSf1.pdf"), 
+            here::here("analysis/paper/figs_submit/Figure4_diffscmult6ay1aSf1.pdf"))
 ```
@@
 knitr::include_graphics(here::here("figs/forPaper/powercmult6ay1aSf1.png"))
+file.rename(here::here("analysis/paper/figs_submit/powercmult6ay1aSf1.pdf"), 
+            here::here("analysis/paper/figs_submit/Figure5_powercmult6ay1aSf1.pdf"))
 ```
@@
 ```{r figEVSI, fig.cap=".", cache=TRUE}
-
 knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay1aSf1.png"))
+file.rename(here::here("analysis/paper/figs_submit/EVSIcmult6ay1aSf1.pdf"), 
+            here::here("analysis/paper/figs_submit/Figure6_EVSIcmult6ay1aSf1.pdf"))
 ```

### 2024-10-17 analysis/paper/paper.Rmd
prepping for submission

--- 2024-10-16 analysis/paper/paper.Rmd
@@
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  woodland caribou (*Rangifer tarandus caribou*); anthropogenic disturbance scenarios; population viability; value of information; monitoring decisions; species at risk; prior information
 highlights: |
@@
 [@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
-@gregory_structured_2012]. But to be useful, these methods must be tailored
+@gregory_structured_2012].
+For sparsely monitored populations, incorporating prior information from elsewhere can help reduce uncertainty [@drummond_combining_2018; @parlato_incorporating_2021; @tufto_bayesian_2000].
+To be useful, these methods must be tailored
 to particular species and decision contexts. Here, we offer a simple
 Bayesian population model for boreal caribou (*Rangifer tarandus
-caribou*) that integrates information about the impacts of landscape 
+caribou*) that integrates prior information about the impacts of landscape 
 change [@johnson_science_2020] with available demographic data [building on @eacker_web-based_2019] to reduce uncertainty in
@@
 @decesare_estimating_2012], though genetic methods are playing an
-increasingly important role in some caribou monitoring programs [e.g.
+increasingly important role in some caribou monitoring programs [e.g.@flasko_fecal_2017;
 @hettinga_estimating_2012; @mcfarlane_genetic_2018;
-@moeller_integrating_2021; @flasko_fecal_2017; @jones_novel_2023;
+@moeller_integrating_2021; @jones_novel_2023;
 @mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
@@
 active [see @nbckc_nbckc_2021], thus allowing estimation of adult
-survival [e.g. @mcloughlin_declines_2003;
-@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. Recruitment is estimated from
+survival [e.g. @dalgarno_bboutools_2024; @mcloughlin_declines_2003;
+@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024]. Recruitment is estimated from
 aerial surveys of the composition of groups of animals that include
@@
 $J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and
-bias-corrected recruitment rate $X$ can be written as a function of the
+bias-adjusted recruitment rate $X$ can be written as a function of the
 observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
@@
 distributed function of the number of surviving post-juvenile females ($\dot{W}_t$)
-and the maximum
-potential recruitment of female calves per cow ($\dot{X}_t$):
+and the maximum potential recruitment of female calves per cow ($\dot{X}_t$):
 
@@
 \begin{equation}
-\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t).
+\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t),
   (\#eq:dotN)
 \end{equation}
+
+where the maximum potential population growth rate is $r_{max}=1.3$.
 
@@
 caribou and variation in habitat preferences and space use among
-individuals [e.g. @decesare_linking_2014; @avgar_spaceuse_2015]. It is possible to integrate complexities such as variation in demographic rates
+individuals [e.g. @avgar_spaceuse_2015; @decesare_linking_2014]. It is possible to integrate complexities such as variation in demographic rates
 among individuals [@armstrong_incorporating_2021], population size
@@
 # Acknowledgements
+This work was informed by a variety of discussions with more people than we can name, but here are a few.
+Members of the Monitoring Working Group of the National Boreal Caribou Knowledge Consortium and the Scientific Assessment of Federal and Provincial Frameworks for the Conservation of Boreal Caribou in Ontario helped us understand context.
+Tal Avgar and Micheline Manseau encouraged us to think about composition survey errors.
+Comments from Joseph Bennett and Daniel Eacker on early versions of this work motivated a variety of improvements.
+Note that our acknowledgement of these contributions does not imply endorsement; we also acknowledge disagreement on some points, and that all remaining errors are our own.
 
 # Author contributions
-**Josie Hughes:** Conceptualization, Formal analysis, Methodology, Investigation,
-Supervision, Software, Writing – original draft, Writing – review and editing, 
-Visualization
+**Josie Hughes:** Conceptualization, Methodology, Formal analysis, Investigation,
+Supervision, Software, Visualization, Writing – original draft, Writing – review and editing
 **Sarah Endicott:** Resources, Software, Writing – review and editing
-**Anna M Calvert:** Conceptualization, Writing – original draft, Writing – review and editing
-**Cheryl A Johnson:** Conceptualization, Writing – review and editing
+**Anna M Calvert:** Conceptualization, Methodology, Writing – original draft, Writing – review and editing
+**Cheryl A Johnson:** Conceptualization, Methodology, Writing – review and editing
 

### 2024-10-17 analysis/paper/supplement.Rmd
add supp mat captions
prepping for submission

@@
 bibliography: CaribouDemographicModel.bib
-csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
+csl: "../templates/ecological-informatics.csl" # Insert path for the bib-style
 
@@
 
-resultScenario = "S7"
-noBiasResultScenario = "S8"
+resultScenario = "S11"
+noBiasResultScenario = "S12"
 
@@
 |$i$|ID|Individual animal identifier.|
-|**Observed**|
+|**Simulated observations of example trajectories**|
 |$A_t$|%|Buffered anthropogenic disturbance.|
@@
 |$\hat{I}_{i,t,m}$|0 or 1|Indicator of whether animal $i$ is dead or alive.|
-|**Estimated state**|
+|**Parameters estimated in the Bayesian model**|
 |$R_t$|rate|Recruitment (calves per cow).|
@@
 |$X_t$|rate|Recruitment adjusted for sex ratio and composition survey errors.|
-|**Estimated Parameters**|
 |$\beta^R_0$||Recruitment intercept.|
@@
 |$\nu_{S}$||Coefficient of variation in survival among years.|
+|$\tilde{R}_{t}$|rate|Expected recruitment (calves per cow).|
+|$\tilde{S}_{t}$|rate|Expected adult female survival.|
 |$\tilde{lambda}_{t}$|rate|Expected annual population growth.|
-|**True state**|
+|**True state in simulated example trajectories**|
 |$\dot{R}_t$|rate|Recruitment (calves per cow).|
@@
 |$w=3,6,\text{ or } 9$|ratio|The apparent number of adult females per collared animal in composition survey.|
-|$\dot{q}=\text{uniform}(0,0.6)$|ratio|True ratio of young bulls to adult females in composition survey.|
-|$\dot{u}=\text{uniform}(0,0.2)$|probability|True probability of misidentifying young bulls as adult females and vice versa in composition survey.|
-|$\dot{z}=\text{uniform}(0,0.2)$|probability|True probability of missing calves in composition survey.|
-|$\dot{c}$||True bias correction term in simulation model (Supplement B).|
-|$q=\text{uniform}(0,0.6)$|ratio|Unknown ratio of young bulls to adult females in composition survey.|
-|$u=\text{uniform}(0,0.2)$|probability|Unknown probability of misidentifying young bulls as adult females and vice versa in composition survey.|
-|$z=\text{uniform}(0,0.2)$|probability|Unknown probability of missing calves in composition survey.|
-|$c$||Unknown bias correction term in Bayesian model (Supplement B).|
+|$\dot{q}=\text{uniform}(0,0.6)$|ratio|True ratio of young bulls to adult females in simulated example composition survey.|
+|$\dot{u}=\text{uniform}(0,0.2)$|probability|True probability of misidentifying young bulls as adult females and vice versa in simulated composition survey.|
+|$\dot{z}=\text{uniform}(0,0.2)$|probability|True probability of missing calves in simulated composition survey.|
+|$\dot{c}$||True bias in simulated example data (Supplement B).|
+|$q=\text{uniform}(0,0.6)$|ratio|Unknown ratio of young bulls to adult females in composition survey (used to set bias prior in Bayesian model) .|
+|$u=\text{uniform}(0,0.2)$|probability|Unknown probability of misidentifying young bulls as adult females and vice versa in composition survey (used to set bias prior in Bayesian model).|
+|$z=\text{uniform}(0,0.2)$|probability|Unknown probability of missing calves in composition survey (used to set bias prior Bayesian model).|
+|$c$||Unknown bias in Bayesian model (Supplement B).|
 |**Expected value of perfect and sample information**|
@@
 
-# Calibration of Priors
+# Calibration of priors
 
@@
 ```{r figDisturbanceAnthroStatic, fig.cap="Static anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult6ay3aSf0.png")))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult6ay1aSf0.png")))
 
@@
 
-```{r powercmult3ay3aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios).
+```{r powercmult3ay1aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios).
 In these examples there are 3 cows per collared cow in recruitment surveys.
@@
 ", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult3ay3aSf0.png")))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult3ay1aSf0.png")))
 ```
 
-```{r powercmult6ay3aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios).
+```{r powercmult6ay1aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios).
 In these examples there are 6 cows per collared cow in recruitment surveys.
@@
 ", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult6ay3aSf0.png")))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult6ay1aSf0.png")))
 ```
 
-```{r powercmult9ay3aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios).
+```{r powercmult9ay1aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios).
 In these examples there are 9 cows per collared cow in recruitment surveys.
@@
 ", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult9ay3aSf0.png")))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult9ay1aSf0.png")))
 ```
@@
 ```{r figDisturbanceAnthroDeclining, fig.cap="Declining anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult6ay3aSf-1.png")))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult6ay1aSf-1.png")))
 ```
 
-```{r powercmult3ay3aSfm1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroDeclining) for declining anthropogenic disturbance scenarios).
+```{r powercmult3ay1aSfm1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroDeclining) for declining anthropogenic disturbance scenarios).
 In these examples there are 3 cows per collared cow in recruitment surveys.
@@
 ", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult3ay3aSf-1.png")))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult3ay1aSf-1.png")))
 ```
 
-```{r powercmult6ay3aSfm1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroDeclining) for declining anthropogenic disturbance scenarios).
+```{r powercmult6ay1aSfm1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroDeclining) for declining anthropogenic disturbance scenarios).
 In these examples there are 6 cows per collared cow in recruitment surveys.
@@
 ", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult6ay3aSf-1.png")))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult6ay1aSf-1.png")))
 ```
 
-```{r powercmult9ay3aSfm1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroDeclining) for declining anthropogenic disturbance scenarios).
+```{r powercmult9ay1aSfm1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroDeclining) for declining anthropogenic disturbance scenarios).
 In these examples there are 9 cows per collared cow in recruitment surveys.
@@
 ", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult9ay3aSf-1.png")))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult9ay1aSf-1.png")))
 ```
@@
 Increasing the forecast horizon can either increase or decrease the EVPI as the anthropogenic disturbance trend makes outcomes more or less predictable.", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/EVPIall.png"))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/EVPIall.png")))
 ```
@@
 
-# Sensitivity to unkown bias in recruitment estimates from calf:cow surveys
+# Sensitivity to unknown bias in recruitment estimates from calf:cow surveys
 
-Adding unknown unknown bias does increase uncertainty, but these increases are not large (compare Figs \@ref(fig:powercmult3ay3aSf1)/\@ref(fig:powercmult3ay3aSf1NoBias) and \@ref(fig:powercmult9ay3aSf1)/\@ref(fig:powercmult9ay3aSf1NoBias)).
+Adding unknown unknown bias does increase uncertainty, but these increases are not large (compare Figs 4 and \@ref(fig:powercmult6ay1aSf1NoBias)).
 A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults, and the probability of missing calves (supplement B).
@@
 
-```{r powercmult3ay3aSf1, fig.cap="The probability of a correct status assessment with unknown bias in recruitment estimates from calf:cow surveys.
-In these examples there are 3 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
+```{r powercmult6ay1aSf1NoBias, fig.cap="The probability of a correct status assessment with unknown bias in recruitment estimates from calf:cow surveys.
+In these examples there are 6 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
 See Figure 4 for additional details.
 ", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult3ay3aSf1.png")))
-```
-
-```{r powercmult3ay3aSf1NoBias, fig.cap="The probability of a correct status assessment without bias in recruitment estimates from calf:cow surveys.
-In these examples there are 3 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
-See Figure 4 for additional details.
-", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powercmult3ay3aSf1.png")))
-```
-
-```{r powercmult9ay3aSf1, fig.cap="The probability of a correct status assessment with unknown bias in recruitment estimates from calf:cow surveys.
-In these examples there are 9 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
-See Figure 4 for additional details.
-", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult9ay3aSf1.png")))
-```
-
-```{r powercmult9ay3aSf1NoBias, fig.cap="The probability of a correct status assessment without bias in recruitment estimates from calf:cow surveys.
-In these examples there are 9 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
-See Figure 4 for additional details.
-", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powercmult9ay3aSf1.png")))
+knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powercmult6ay1aSf1.png")))
 ```
@@
 # which R packages and versions?
-#if ("devtools" %in% installed.packages()) devtools::session_info()
+if ("devtools" %in% installed.packages()) devtools::session_info()
 ```
@@
 # what commit is this file at? 
-#if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
 ```

### 2024-10-17 analysis/scripts/2_viewSensitivityResults.R

@@ -13,7 +13,7 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 
 ########################
 #sensitivity
-setName = "s11" #need to do both s7 and s8
+setName = "s11" #need to do s10, s11, and s12
 dir.create(paste0("figs/",setName),recursive=T)
 scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
 
@@ -28,7 +28,7 @@ pages = pagesa
 str(scns)
 
 setLTYVar <-function(scns){
-  if(is.element("interannualVar",names(scns))){
+  if(is.element("interannualVar",names(scns))&&(length(unique(scns$interannualVar))>1)){
     scns$ltyVariable = "low"
     scns$ltyVariable[grepl("0.46",scns$interannualVar,fixed=T)]="high"
   }else{
@@ -118,6 +118,9 @@ for(i in 1:length(pages)){
   exResults$Anthro2023 = pmax(0,exResults$tA)#pmax(0,exResults$tA-2) #correcting for error - change back in round 4
 
 
+  unique(scResults$obs.all$collarCount)
+  unique(scResults$obs.all$ltyVariable)
+  unique(scResults$obs.all$parameter)
   obs = subset(scResults$obs.all,(collarCount==30)&(ltyVariable==scns$ltyVariable[1])&(parameter=="Population growth rate"))
   obs$startYear = obs$startYear+obs$preYears
   obs = merge(obs,unique(subset(exResults,select=c(tA,obsYears,rQuantile,sQuantile,quantile,grp,Anthro2023))))

### 2024-10-17 analysis/scripts/3_EVSIPlots.R

@@ -21,7 +21,7 @@ batchStrip<-function(l,batches=c(10,seq(1:9))){
   return(l)
 }
 
-setName = "s10"
+setName = "s11"
 probsSum = read.csv(here::here(paste0("tabs/EVsample",setName,".csv")),stringsAsFactors = F)
 
 probsSum$pageLabC = batchStrip(probsSum$pageLab)
@@ -59,8 +59,10 @@ EVuncertainty$trend[EVuncertainty$trend=="0"]= "stable"
 EVuncertainty$trend[EVuncertainty$trend=="1"]= "increasing"
 EVuncertainty$YearsOfProjection =as.factor(EVuncertainty$YearsOfProjection)
 
+subset(EVuncertainty, (AnthroScn=="high")&(trend=="decreasing"))
+
 png(here::here(paste0("figs/",setName,"/EVPIall.png")),
-    height = 3.7, width = 7.48, units = "in",res=600)
+    height = 3.8, width = 7.48, units = "in",res=600)
 base=ggplot(EVuncertainty,aes(x=YearsOfProjection,y=EVPI,col=cmult,group=pageLabC))+geom_point()+
   facet_grid(trend~AnthroScn,labeller="label_both")+labs(color="cow multiplier")+
   theme_bw()+xlab("years of projection")+ylab("Expected Value of Perfect Information EVPI")+

### 2024-10-17 analysis/scripts/formatScenarios.R

@@ -1,7 +1,7 @@
 
 library(caribouMetrics)
 
-rawScns<-read.csv(here::here("tabs/s6.csv"))
+rawScns<-read.csv(here::here("tabs/s10.csv"))
 nrow(rawScns)
 rawScns = subset(rawScns,collarCount!=1)
 

### 2024-10-17 analysis/paper/paper.Rmd
more submission prep

@@
 highlights: |
-  These are the highlights. 
+
+  * Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed.
+
+  * We show how Bayesian integration of prior information from a national analysis of demographic-disturbance relationships can reduce uncertainty about the viability of sparsely monitored populations.
+
+  * Integrating prior information about demographic-disturbance relationships clarifies that no additional information is needed to project population viability where disturbance is high and there are no actions to support recovery.
+
+  * At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years is likely better than intensive short-term monitoring for reducing uncertainty about population viability.
+
+  * We introduce a method for acknowledging potential errors in recruitment surveys to help clarify where and when these these errors are likely to alter conclusions about population viability.
+
 editor_options: 
@@
 @decesare_estimating_2012], though genetic methods are playing an
-increasingly important role in some caribou monitoring programs [e.g.@flasko_fecal_2017;
+increasingly important role in some caribou monitoring programs [e.g. @flasko_fecal_2017;
 @hettinga_estimating_2012; @mcfarlane_genetic_2018;

### 2024-10-17 analysis/scripts/0_nationalModelTrajectories.R

@@ -138,9 +138,9 @@ plot_lambda <- ggplot(oo,
             aes(x = Anthro, y = lambdaTrue, group = fullGrp, color = fullGrp),
             alpha = 1) + scale_color_manual(values=pal)+
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
-  xlab("Anthropogenic disturbance (%)") +
+  xlab("Anthropogenic disturbance") +
   ylab(expression("Population growth rate " * dot(lambda)[t])) +
-  theme(legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
+  theme(axis.title=element_text(size=rel(0.95)),legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
 
 str(pars1)
 
@@ -155,9 +155,9 @@ plot_recruitment3 <- ggplot(data = rateSummaries,
   geom_line(colour = "#ef8a62", size = 2, linetype = "dotted") +
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
   scale_y_continuous(limits = c(0, 0.5), breaks = c(0, 0.1, 0.2, 0.3, 0.4,0.5)) +
-  xlab("Anthropogenic disturbance (%)") +
+  xlab("Anthropogenic disturbance") +
   ylab(expression("Adjusted recruitment " * dot(X)[t])) +
-  theme(legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
+  theme(axis.title=element_text(size=rel(0.95)),legend.position = "none", plot.margin = margin(l = 0.6, unit = "cm"))
 plot(plot_recruitment3)
 
 plot_recruitment1 <- ggplot(data = rateSummaries,
@@ -171,9 +171,9 @@ plot_recruitment1 <- ggplot(data = rateSummaries,
   geom_line(colour = "#ef8a62", size = 2, linetype = "dotted") +
   scale_x_continuous(limits = c(-1, 90), breaks = c(0, 20, 40, 60, 80)) +
   scale_y_continuous(limits = c(0, 0.7), breaks = c(0, 0.1, 0.2, 0.3, 0.4,0.5,0.6,0.7)) +
-  xlab("Anthropogenic disturbance (%)") +
+  xlab("Anthropogenic disturbance") +
   ylab(expression("Recruitment " * dot(R)[t])) +
[…]

### 2024-10-17 analysis/scripts/2_viewSensitivityResults.R
more submission prep

@@ -13,7 +13,7 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 
 ########################
 #sensitivity
-setName = "s11" #need to do s10, s11, and s12
+setName = "s10" #need to do s10, s11, and s12
 dir.create(paste0("figs/",setName),recursive=T)
 scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
 

### 2024-10-18 analysis/scripts/3_EVSIPlots.R
more prep for submission

@@ -21,7 +21,7 @@ batchStrip<-function(l,batches=c(10,seq(1:9))){
   return(l)
 }
 
-setName = "s11"
+setName = "s10"
 probsSum = read.csv(here::here(paste0("tabs/EVsample",setName,".csv")),stringsAsFactors = F)
 
 probsSum$pageLabC = batchStrip(probsSum$pageLab)

### 2025-01-23 analysis/paper/paper.Rmd
add data statement
MS revisions in progress

@@
 are typically captured via net-gun from aircraft in late winter, and
-affixed with with collars that send a mortality signal when no longer
+affixed with collars that send a mortality signal when no longer
 active [see @nbckc_nbckc_2021], thus allowing estimation of adult
@@
 
-Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability
+Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population viability
 [@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
@@
 
+## Overview of methods and notation
+
+To investigate how the accuracy and precision of predictions vary with both landscape condition and monitoring effort we analyze simulated data using a Bayesian model with informative priors and a composition survey bias term (section 2.1).
+Steps in our analysis procedure include defining anthopogenic disturbance and monitoring scenarios (section 2.2), simulating population dynamics and monitoring (section 2.3) to produce example demographic trajectories for each scenario, and analyzing example demographic trajectories with a Bayesian model that enables integration of (simulated) local demographic data with prior knowledge of demographic-disturbance relationships (section 2.4).
+Outcomes of interest include expected population growth rate, the probability of correctly assessing population status, and the expected value of local information (section 2.5).
+We refer to the model used to simulate example trajectories as the "national" model because parameters are from a national analysis of demographic-disturbance relationships [@johnson_science_2020].
+We distinguish this national model from the Bayesian model that is used to analyze simulated example demographic trajectories.
+Where appropriate the national model and the Bayesian model share the same structure, and we have selected notation to highlight the similarities.
+Bayesian model estimates are distinguished from observed values, estimated expected values, true values, and true expected values using $\hat{\text{ }}$ $\tilde{\text{ }}$ $\dot{\text{ }}$ and $\bar{\text{ }}$ symbols respectively; for example, $S_t$ is the Bayesian estimate of survival in year $t$, $\hat{S}_t$ is observed survival, $\tilde{S}_t$ is the Bayesian estimate of expected survival (without interannual variation), $\dot{S}_t$ is true survival, and $\bar{S_t}$ is true expected survival (without interannual variation).
+For reference, all variables and parameters are defined in supplemental Table A1.
+
 ```{r setup2}
@@
 
-We focus on analyzing survival of collared females and the ratio of calves to cows in aerial
+We focus on analyzing survival of radio-collared females and the ratio of calves to cows in aerial
 surveys of groups that include collared females [@nbckc_nbckc_2022],
@@
 information for boreal caribou populations [@johnson_science_2020].
-Surveyed groups also contain uncollared animals so sample size is
-typically larger for recruitment than for survival estimates. Errors
-that can bias composition survey results include misidentifying young
+GPS radio collars report animal locations at regular intervals, and cease moving when animals die.
+Aerial recruitment surveys are usually done in March or April, and radio-collared females are often used to locate groups of animals.
+Surveyors use visual cues to distinguish calves that have survived their first winter (~ 9 months old) from adult cows and bulls.
+Surveyed groups also contain uncollared adults so sample size is typically larger for recruitment than for survival estimates.
+Errors that can bias composition survey results include misidentifying young
 bulls as cows or vice versa, and failing to detect calves
@@
 $T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of
-uncollared adult females to collared animals in the recruitment survey.
+adult females to collared animals in the recruitment survey.
 The ratio of young bulls to uncollared adult females in the group is
@@
 observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
-bulls to adult females $q$, and the misidentification probabilities $u$
+bulls to uncollared adult females $q$, and the misidentification probabilities $u$
 and $z$ (see Supplement B for details). For simplicity, we combine the
@@
 among scenarios, as does the total duration of the monitoring program
-$d$ and the ratio of uncollared to collared adults in the recruitment
+$d$ and the ratio of observed adult females to
+collared animals in the recruitment
 survey $w$ (Table \@ref(tab:monitoringScenarios)). Collars lost to
@@
 
-To investigate the effectiveness of various monitoring strategies for
-reducing uncertainty about population viability, and to show how this
-varies with landscape condition, we begin with a model for simulating
-example population dynamics and observations of those dynamics.
-Demographic parameters for the simulated trajectories are sampled from the regression models that summarize national demographic-disturbance relationships
-[@johnson_science_2020] so we call this the "national" model.
-Together, a set of example trajectories represent alternative hypotheses about the true state of a particular caribou population; these alternative scenarios informed by prior
-beliefs are used to assess the expected value of alternative sampling
-strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the
-simulated trajectories to confirm that the Bayesian model successfully
-integrates local data, and that prior predictions meet expectations from the national model. For
-reference, all variables and parameters are defined in supplemental
-Table A1.
-
-To simulate demographic trajectories for example populations we used a
-modified version of Johnson et al's [-@johnson_science_2020] demographic
-model described by Dyson et al. [-@dyson_existing_2022]. Given a
-population of post-juvenile females at the beginning of year $t$,
+To investigate the effectiveness of various monitoring strategies, and to show how this
+varies with landscape condition, we begin by simulating example demographic trajectories and observations of those trajectories.
+To simulated example trajectories we use a modified version of Johnson et al's [-@johnson_science_2020] demographic model described by Dyson et al.
+[-@dyson_existing_2022].
+For brevity, we refer to this as the "national" model.
+For clarity, we include a full description of the model in the following paragraphs.
+Together, a set of example trajectories from the national model represent alternative hypotheses about the true state of a particular caribou population.
+To assess the accuracy of population status assessments and the expected value of alternative monitoring strategies [@runge_which_2011; @dunham_strategic_2023] we analyze each trajectory of simulated observations using a Bayesian model (section 2.4).
+We also use the sets of simulated trajectories to confirm that prior predictions from our Bayesian model are consistent with knowledge of the observed distribution of outcomes for caribou across the country [@johnson_science_2020].
+
+Given a population of post-juvenile females at the beginning of year $t$,
 $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
@@
 $\dot{q}$ varies uniformly between 0 and 0.6.
+The distinction between the true composition bias term $\dot{c}$ and the term $c$ we include in the Bayesian model (eq \@ref(eq:c)) is that the (simulated) truth is known, and the Bayesian model bias is uncertain.
 
@@
 variation in recruitment and survival among example populations (Figure
-\@ref(fig:exampleTrajectories)). To ensure that results are comparable across
-monitoring scenarios and that populations remain extant through the
-monitoring period, all simulations are started with $N_0=5000$ animals
-24 years prior to the start of the projection period.
-Realized population growth rate is $\dot{\lambda}_t=\dot{N}_t/\dot{N}_{t-1}$ Expected population growth rate (without interannual variation, demographic stochasticity, or effects of population density) is $\bar{\lambda}_t=\bar{S}_t(1+\dot{c}\bar{R}_t/2)$.
+\@ref(fig:exampleTrajectories)).
 
@@
 \end{equation}
+
+To ensure that results are comparable across monitoring scenarios and that populations remain extant through the
+monitoring period, all simulations are started with $N_0=5000$ animals
+24 years prior to the start of the projection period.
+Realized population growth rate is $\dot{\lambda}_t=\dot{N}_t/\dot{N}_{t-1}$ Expected population growth rate (without interannual variation, demographic stochasticity, or effects of population density) is $\bar{\lambda}_t=\bar{S}_t(1+\dot{c}\bar{R}_t/2)$.
 
@@
 
-To account for sex ratio of calves and additional uncertainty caused by composition survey errors errors (Supplement B) estimated recruitment is adjusted to get expected female
-recruits per female: 
-\begin{equation}
-X_t=cR_t/2; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
-  (\#eq:X)
-\end{equation}
-
-The composition survey bias term $c$ depends on the known
+To account for sex ratio of calves and additional uncertainty caused by composition survey errors (Supplement B) estimated recruitment is adjusted to get expected female
+recruits per female $X_t=cR_t/2$. The composition survey bias term $c$ (eq \@ref(eq:c)) depends on the known
 apparent number of adult females per collared animal $w$, the unknown
@@
 \@ref(fig:examplePredictions)).
+To show the value of informative priors, we also include a weakly informative scenario with prior means of the intercept parameters from Dalgarno et al.
+[-@dalgarno_bboutools_2024] ($\beta^R_0=-0.3689, \beta^S_0=-0.2126776$), prior means of the slope parameters set to zero, and standard deviations for all parameters (Table \@ref(tab:priors)) multiplied by ten.
 
@@
 outcome for each disturbance and monitoring scenario (Fig
-\@ref(fig:figDisturbance)) we fit the Bayesian model to simulated
-observations from 500 example population trajectories (e.g. Fig
-\@ref(fig:exampleTrajectories)), and calculated the percentage of these
+\@ref(fig:figDisturbance)) we simulate observations from 500 example trajectories (e.g. Fig
+\@ref(fig:exampleTrajectories)), fit the Bayesian model to each of these, and calculated the percentage of these
 500 examples in which the Bayesian posterior expected population growth rate
@@
 
-EVSI measures expected improvement from a particular sampling strategy
+EVSI measures expected improvement from a particular monitoring strategy
 as the difference between the expected value with and without additional
@@
 uncertainty in boreal caribou demographic projections (Figures
-\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected,
+\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)).
+When local data is limited (e.g.
+4 years, 15 collars per year) the model with informative priors yields predictions that are consistent with knowledge of observed outcomes across the country, and the model with weakly informative priors does not (Figure
+\@ref(fig:examplePredictionsLowEffort)). As expected,
 uncertainty about the population growth rate decreases as sampling
 effort in each year (i.e. number of collars) or the number of years of
-monitoring (Figure \@ref(fig:figDiffs)) increase. Reducing uncertainty is more challenging when interannual variation is high (Figure
-\@ref(fig:examplePredictions)). Increasing the forecast horizon [@petchey_ecological_2015] does not increase uncertainty about expected population growth (compare 5 and 20 year projections in
+monitoring (Figure \@ref(fig:figDiffs)) increase.
+Interannual variation makes reducing uncertainty challenging, and insufficiently informative priors can yield predictions that are both imprecise and inaccurate even in a case where there is 16 years of local monitoring data (Figure \@ref(fig:examplePredictions)).
+Increasing the forecast horizon [@petchey_ecological_2015] does not increase uncertainty about expected population growth (compare 5 and 20 year projections in
 Figure \@ref(fig:figDiffs)) in these scenarios.
 
-Our analysis becomes more informative when we examine the probability of
+Examining the probability of
 correctly assessing population status (Figure \@ref(fig:figStatusErrors))
-and the expected value of information (Figures \@ref(fig:figEVSI) and E.1).
+and the expected value of information (Figures \@ref(fig:figEVSI) and E.1) helps clarify how the need for monitoring varies with landscape condition.
 When anthropogenic disturbance is high (right panels of Figures
@@
 
+# Data statement
+Code required to create the data, run analyses and reproduce figures in this paper is available at https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper, and in the caribouMetrics R package available on GitHub at https://github.com/LandSciTech/caribouMetrics/tree/EI-paper-submission.
+
 # Acknowledgements
@@
 
+```{r examplePredictionsLowEffort, fig.cap=".", cache=TRUE}
+knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamplesLowEffort.png"))
+file.rename(here::here("analysis/paper/figs_submit/bayesianModelExamplesLowEffort.pdf"), 
+            here::here("analysis/paper/figs_submit/Figure3_bayesianModelExamplesLowEffort.pdf"))
+```
+
 ```{r figDiffs, fig.cap=".", cache=TRUE}
@@
 
-Figure 3: Prior (a) and posterior predictions from the Bayesian model for population trajectories with low (b: $v_R=0.23,v_S=0.0435$) and high (c: $v_R=0.46,v_S=0.087$) interannual variation.
+Figure 3: Prior (a) and posterior predictions from the Bayesian model for population trajectories with informative (a,b) and weakly informative (c) priors given lower interannual variation ($v_R=0.23,v_S=0.435$) and intensive monitoring.
 In this example scenario anthropogenic disturbance increases by 2% per year from 0 to 100%.
 The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys.
-The prior means and 95% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5% and 97.5% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading).
-Local population data (open triangles) reduces uncertainty about demographic rates, but local data is not required to predict viability (population growth <= 0.99) of highly disturbed populations.
-
-Figure 4: The distribution of differences between true expected population growth rates and Bayesian posterior expected growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
+The prior means and 95% predictive intervals (a, orange lines and shading) are similar to the means and ranges between the 2.5% and 97.5% quantiles of 3000 simulated survival and recruitment trajectories from the national model (a, blue lines and shading).
+Local population data (open triangles) reduces uncertainty about demographic rates (b), though that is not needed to predict viability (population growth <= 0.99) of highly disturbed populations (a).
+With weakly informative priors the model does not provide accurate or precise predictions of future viability after 16 years of monitoring in a case where anthropogenic disturbance is increasing to 90% (c).
+
+Figure 4: Posterior predictions from the Bayesian model for population trajectories with informative (a) and weakly informative (b) priors given higher interannual variation ($v_R=0.46,v_S=0.087$) and limited monitoring.
+The population is monitored for 4 years with 15 collars per year, and 3 cows per collared cow in the composition surveys.
+See Figure 3 for disturbance trends and other details.
+The limited local data (open triangles) reduces uncertainty (orange shading) about the observed current state of the population, but is not sufficient to reduce future uncertainty.
+Given limited data the model with informative priors (a) yields predictions (orange bands) that are consistent with knowledge of observed outcomes across the country (blue shading), and the model with weakly informative priors (b) does not.
+
+Figure 5: The distribution of differences between true expected population growth rates and Bayesian posterior expected growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
 Differences decrease as the number of years of monitoring and the number of collars increase.
@@
 
-Figure 5: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
+Figure 6: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
 When disturbance is high the probability of correct status assessment is high even without local monitoring.
@@
 
-Figure 6: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort.
-When disturbance is high EVSI is low, indicating that local sampling does not change predicted status.
+Figure 7: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort.
+When disturbance is high EVSI is low, indicating that local monitoring does not change predicted status.
 Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases.

### 2025-01-23 analysis/scripts/0_bayesianModelExamplesA.R

@@ -11,14 +11,16 @@ monitoringScns = data.frame(obsYears=c(1,16),collarCount=c(0,30),cowMult=c(6),co
                             assessmentYrs=c(1))
 stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
 stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
-stateScns = merge(stateScns,data.frame(interannualVar=c("list(R_CV=0.46,S_CV=0.087)")))
+#stateScns = merge(stateScns,data.frame(interannualVar=c("list(R_CV=0.46,S_CV=0.087)")))
+stateScns = merge(stateScns,data.frame(interannualVar=c("list(R_CV=0.23,S_CV=0.0435)")))
 stateScns = merge(stateScns,data.frame(survivalModelNumber = c("M1","M0"), recruitmentModelNumber=c("M4","M0"),
                                        rAnthroSlopeSE = c(0.006,0.06),rFireSlopeSE = c(0.002,0.02),
                                        sAnthroSlopeSE = c(0.0005,0.005),sIntSE = c(0.04,0.4),
-                                       rIntSE = c(0.25,2.5)))
-stateScns$sQuantile=0.9
-stateScns$rQuantile = 0.9
+                                       rIntSE = c(0.25,2.5),sQuantile=c(0.9,0.9),rQuantile=c(0.9,0.9)))
+
 scns=merge(monitoringScns,stateScns)
+#scns$obsYears[scns$survivalModelNumber=="M0"]=4
+#scns$obsYears[scns$survivalModelNumber=="M0"]=4
 
 scns = subset(scns,!(grepl("2.5",rIntSE,fixed=T)&(collarCount==0)))
 scns$iAnthro = 0
@@ -259,3 +261,86 @@ ggsave(paste0(baseDir,"/analysis/paper/figs/bayesianModelBiasSensitivity.png"),
        width = 12*0.8, height = 3.6*2, units = "in",
        dpi = 1200)
 
+###################################
+#Low effort scenario
+monitoringScns = data.frame(obsYears=c(1,4),collarCount=c(0,15),cowMult=c(3),collarInterval=c(1),
+                            assessmentYrs=c(1))
+stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
+stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
+stateScns = merge(stateScns,data.frame(interannualVar=c("list(R_CV=0.46,S_CV=0.087)")))
[…]

### 2025-01-23 analysis/scripts/1_sensitivity1.R

@@ -25,13 +25,14 @@ scns$preYears = max(scns$obsYears)-scns$obsYears
 scns$iAnthro = scns$tA-(scns$obsYears+scns$preYears-1)*scns$obsAnthroSlope
 scns$projYears = 20
 unique(scns$iAnthro)
-scns$repBatch = ceiling(scns$rep/50)
+scns$repBatch = ceiling(scns$rep/10)
 table(scns$repBatch)
 scns$N0 = 5000
 
 scns$pageLab = paste0("cmult",scns$cowMult,"ay",scns$assessmentYrs,"aSf",scns$projAnthroSlope,"repBatch",scns$repBatch)
 scns$pageId = as.numeric(as.factor(scns$pageLab))
 nrow(scns)
+
 length(unique(scns$pageLab))
 
 write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
@@ -39,14 +40,14 @@ pages=unique(scns$pageLab)
 
 ###################
 #Without bias
-setName = "s14"
+setName = "s15" #14 is the same, just bigger batches.
 scns$qMax = 0;scns$uMax=0;scns$zMax=0
 write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
 pages=unique(scns$pageLab)
 
 ###################
 #Different anthropogenic disturbance scenarios, collar intervals. With bias and default interannual variability.
-setName = "s11"
+setName = "s15" #formerly s11
 monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
                              cowMult=c(3,6,9),collarInterval=c(1,4),
                              assessmentYrs=c(1))
@@ -67,7 +68,7 @@ scns$preYears = max(scns$obsYears)-scns$obsYears
 scns$iAnthro = scns$tA-(scns$obsYears+scns$preYears-1)*scns$obsAnthroSlope
 scns$projYears = 20
 unique(scns$iAnthro)
-scns$repBatch = ceiling(scns$rep/50)
+scns$repBatch = ceiling(scns$rep/10)
 table(scns$repBatch)
 scns$N0 = 5000
 
@@ -79,10 +80,10 @@ length(unique(scns$pageLab))
 write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
 pages=unique(scns$pageLab)
 
-setName="s9"
-scns = subset(scns,(rep==1)&(collarCount==30)
[…]

### 2025-01-23 analysis/scripts/2_viewSensitivityResults.R

@@ -3,6 +3,7 @@ library(tidyverse)
 library(ggplot2)
 library(caribouMetrics)
 library(RColorBrewer)
+library(scales)
 
 pal2 = brewer.pal(7,"RdBu")[c(2,6)]
 pal4 = brewer.pal(5,"RdBu")[c(1,2,4,5)]
@@ -13,7 +14,7 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 
 ########################
 #sensitivity
-setName = "s12" #need to do s10, s11, and s12
+setName = "s10" #need to do s10, s11, and s12
 dir.create(paste0("figs/",setName),recursive=T)
 scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
 
@@ -138,7 +139,6 @@ for(i in 1:length(pages)){
   print(base)
   dev.off()
 
-
   probs = subset(scResults$rr.summary.all,(Parameter=="Population growth rate"))
   probs$startYear = probs$startYear+probs$preYears
   probs$projectionTime = probs$Year-2023
@@ -188,6 +188,22 @@ for(i in 1:length(pages)){
 
   probs = subset(probs,is.element(projectionTime,c(5,20)))
 
+  unique(scResults$obs.all$parameter)
+
+  #get true survey bias
+  cTrue = subset(scResults$obs.all,(type=="true")&is.element(parameter,c("Recruitment"))&
+                   (Year>(startYear+obsYears-1)))
+  cTrue$R = cTrue$Mean
+  cTrue$parameter=NULL;cTrue$type=NULL;cTrue$Mean=NULL
+  ccTrue = subset(scResults$obs.all,(type=="true")&is.element(parameter,c("Adjusted recruitment"))&
+                   (Year>(startYear+obsYears-1)))
+  ccTrue$X = ccTrue$Mean
+  ccTrue$parameter=NULL;ccTrue$type=NULL;ccTrue$Mean=NULL
+  cTrue = merge(cTrue, ccTrue)
+  ccTrue = NULL
+  cTrue$c = cTrue$X*2/cTrue$R
+  cTrue$startYear=cTrue$startYear+cTrue$preYears
+
   #plot probability of making an error about true population state
   statusTrue = subset(scResults$obs.all,(type=="true")&(parameter=="Population growth rate")&(Year>(startYear+obsYears-1)))
   statusTrue$startYear=statusTrue$startYear+statusTrue$preYears
[…]

### 2025-01-23 analysis/scripts/formatPriors.R

@@ -32,3 +32,7 @@ pars<-rbind(pars,data.frame(name="beta.Saf",Parameter="$\\beta^S_a$",Description
 
 tbl<-merge(pars,tbl);tbl$name=NULL
 
+
+
+
+

### 2025-01-23 cloud/make_batch_scripts.R

@@ -5,8 +5,10 @@ if(0){
   keyring::key_set("Azure_subnetId")
   # retrieves the secret
   keyring::key_get("Azure_SASURL")
+  keyring::key_get("Azure_subnetId")
 }
 cargs <- commandArgs(trailingOnly = TRUE)
+#cargs=c("s14","https://ecdcwls.blob.core.windows.net/jhughes/?se=2025-01-24T17%3A50Z&sp=racwdli&sv=2022-11-02&sr=c&skoid=536c15ec-3c37-4b86-ae47-a3fefedad9c4&sktid=740c5fd3-6e8b-4176-9cc9-454dbe4e62c4&skt=2025-01-17T17%3A50%3A13Z&ske=2025-01-24T17%3A50%3A00Z&sks=b&skv=2022-11-02&sig=py82KrESgpDj/yy2aIgki%2BhfZfTKgkTYF%2BUq0muzXn4%3D")
 setName = cargs[1]
 allScns = read.csv(paste0("tabs/",setName,".csv"))
 scn_nums <- unique(allScns$pageId)
@@ -71,7 +73,7 @@ purrr::walk(scn_nums, make_task)
 outfile <-  file(paste0("cloud/pool_json/caribou_add_pool1.json"), "wb")
 readLines("cloud/caribou_add_pool.json") |>
   stringr::str_replace_all("<subnetId>", keyring::key_get("Azure_subnetId")) |>
-  stringr::str_replace_all("<id>", paste0("sendicott_caribouDemo_", setName)) |>
+  stringr::str_replace_all("<id>", paste0("jhughes_caribouDemo_", setName)) |>
   stringr::str_replace_all("<n_nodes>", as.character(nNodes)) |>
   stringr::str_replace_all("<n_slots>", as.character(nSlotsPerNode)) |>
   stringr::str_replace_all("<vmSize>", as.character(vmSizeUse)) |>

### 2025-01-23 cloud/setup_azure.sh

@@ -2,7 +2,7 @@
 #############################################
 
 #### Parameters to change
-setName="s10"
+setName="s13"
 
 
 #####Login etc##################
@@ -19,13 +19,15 @@ allocState:allocationState}" \
 
 #### Move files to container ##############
 end=`date -u -d "7 days" '+%Y-%m-%dT%H:%MZ'`
-sastoken=`az storage container generate-sas --account-name ecdcwls --expiry $end --name sendicott --permissions racwdli -o tsv --auth-mode login --as-user`
+sastoken=`az storage container generate-sas --account-name ecdcwls --expiry $end --name jhughes --permissions racwdli -o tsv --auth-mode login --as-user`
 
+subnetid=$(az network vnet subnet list --resource-group EcDc-WLS-rg --vnet-name EcDc-WLS-vnet \
+--query "[?name=='EcDc-WLS_compute-cluster-snet'].id" --output tsv)
 #
-sasurl=https://ecdcwls.blob.core.windows.net/sendicott/?$sastoken
+sasurl=https://ecdcwls.blob.core.windows.net/jhughes/?$sastoken
 
-jobName="sendicott_job_"$setName
-poolName="sendicott_caribouDemo_"$setName
+jobName="jhughes_job_"$setName
+poolName="jhughes_caribouDemo_"$setName
 # delete old versions of scripts
 rm -r cloud/task_scripts
 rm -r cloud/task_jsons
@@ -36,7 +38,7 @@ rm -r cloud/pool_json
 nBatches=$(Rscript --vanilla "cloud/make_batch_scripts.R" $setName $sasurl)
 
 # Check that container is empty
-az storage blob list -c sendicott --account-name ecdcwls --sas-token $sastoken \
+az storage blob list -c jhughes --account-name ecdcwls --sas-token $sastoken \
 --query "[].{name:name}" --output yaml
 
 # Upload scripts to use in tasks
@@ -94,7 +96,7 @@ az batch task show --job-id $jobName \
 --query "{state: state, executionInfo: executionInfo}" --output yaml
 
 # download output file for a task
-taskNum=10
+taskNum=12
 
 az batch task file download --task-id caribou-demog_sens_batch$taskNum \
 --job-id $jobName --file-path "wd/nohup_"$taskNum".out" \
@@ -114,15 +116,15 @@ az batch task list --job-id $jobName --query "{tasks: [?state == 'completed'].[i
 #### Download results ##########################
 
 # Check what results have been added to the storage container
[…]

### 2025-01-23 analysis/paper/supplement.Rmd
fix lambdas
MS revisions in progress

@@
 resultScenario = "S11"
-noBiasResultScenario = "S12"
+noBiasResultScenario = "S14"
 
@@
 |$\tilde{S}_{t}$|rate|Expected adult female survival.|
-|$\tilde{lambda}_{t}$|rate|Expected annual population growth.|
+|$\tilde{\lambda}_{t}$|rate|Expected annual population growth.|
 |**True state in simulated example trajectories**|
@@
 |$\dot{\lambda}_t$|rate|Annual population growth.|
-|$\bar{lambda}_{t}$|rate|Expected annual population growth.|
+|$\bar{\lambda}_{t}$|rate|Expected annual population growth.|
 |$\dot{\beta}^R_0$||Recruitment intercept.|
@@
 We begin by setting random effects to zero, varying the prior standard deviations of the survival intercept (sIntSE is $\beta^S_0$ in Table 2) and recruitment intercept (rIntSE is $\beta^R_0$ in Table 2), and comparing the prior predictive intervals of the Bayesian model to the 95\% predictive intervals from Johnson et al.'s (2020) regression analysis in a case with no anthropogenic disturbance (top row of Fig \@ref(fig:InterceptsAndSlopes)).
-We set the prior standard deviation intercept parameters (sIntSE = 0.04, rIntSE = 0.25) so that the upper bounds of the Bayesian predictive intervals match the upper bounds of Johnson et al.'s predictive intervals.
+We set the prior standard deviation intercept parameters (sIntSE = 0.06, rIntSE = 0.35) so that the lower bounds of the Bayesian predictive intervals match the upper bounds of Johnson et al.'s predictive intervals.
 
@@
 ```{r figEVPI, fig.cap="The expected value of perfect information (EVPI) varies among anthropogenic disturbance scenarios (see increasing, decreasing and stable scenarios in Figs 1, D.1 and D.5 respectively) and the number of years projected.
-When disturbance is high EVPI is low, indicating that local sampling will not alter status assessments.
+When disturbance is high EVPI is low, indicating that local monitoring will not alter status assessments.
 Increasing the forecast horizon can either increase or decrease the EVPI as the anthropogenic disturbance trend makes outcomes more or less predictable.", cache=TRUE}

### 2025-01-23 analysis/scripts/0_PriorsFromPIs.R

@@ -66,7 +66,7 @@ plot(rrAll[[vName]]$plot)
 #set sIntSE = 0.04. Set to match high bound.
 
 #Now vary sAnthroSlopeSE to match Johnson predictive interval when anthropogenic disturbance is high.
-sIntSE=0.04; sNuMin=0;sNuMax=0
+sIntSE=0.06; sNuMin=0;sNuMax=0
 sAnthroSlopeSE=seq(0.0001,0.001,by=0.0001)
 
 scnsSSlope=expand.grid(obsAnthroSlope = 0, projAnthroSlope = 0,iAnthro=90,
@@ -94,7 +94,7 @@ plot(rrAll[[vName]]$plot)
 #set rIntSE = 0.25. Set to match upper bound of Johnson.
 
 #Now vary rAnthroSlopeSE to match Johnson predictive interval when anthropogenic disturbance is high.
-rIntSE=0.25;rNuMin=0;rNuMax=0;
+rIntSE=0.35;rNuMin=0;rNuMax=0;
 rAnthroSlopeSE=seq(0.001,0.01,by=0.001)
 
 scnsRSlope=expand.grid(obsAnthroSlope = 0, projAnthroSlope = 0,iAnthro=90,

### 2025-01-23 analysis/scripts/0_bayesianModelExamples.R

@@ -1,255 +1,0 @@
-# Prior and posterior predictions from Bayesian model
-library(caribouMetrics)
-library(tidyverse)
-library(ggpubr)
-theme_set(theme_bw())
-library(RColorBrewer)
-
-baseDir <- "."
-
-monitoringScns = data.frame(obsYears=c(1,16),collarCount=c(0,30),cowMult=c(6),collarInterval=c(1),
-                            assessmentYrs=c(1))
-stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
-stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
-stateScns = merge(stateScns,data.frame(interannualVar=c("list(R_CV=0.46,S_CV=0.087)","list(R_CV=0.23,S_CV=0.0435)")))
-
-stateScns$sQuantile=0.9
-stateScns$rQuantile = 0.9
-scns=merge(monitoringScns,stateScns)
-
-scns = subset(scns,!(grepl("0.23",interannualVar,fixed=T)&(collarCount==0)))
-scns$iAnthro = 0
-scns$tA = scns$iAnthro+(scns$obsYears)*scns$obsAnthroSlope
-scns$projYears = 50-scns$obsYears
-scns$N0 = 2000
-
-#scns$assessmentYrs=3
-
-####################
-eParsIn = list()
-eParsIn$cowCounts <- data.frame(Year = 1981:2023,
-                                Count = 100,
-                                Class = "cow")
-eParsIn$freqStartsByYear <- data.frame(Year = 1981:2023,
-                                       numStarts = 30)
-eParsIn$collarOnTime=1
-eParsIn$collarOffTime=12
-eParsIn$collarNumYears=1
-labFontSize = 10; breakInterval=5
-
-###############################
-#With bias adjustment
-simBig<-getSimsNational(replicates=3000,forceUpdate=T)
-
-priorResult = caribouMetrics:::runScnSet(scns[1,],eParsIn,simBig,getKSDists=F,printProgress=F)
-priorResult$obs.all=NULL
-yr_scale1 <- scale_x_continuous(breaks = 2023:2072 %>% .[0:9*5+1],
-                                labels = 2009:2058 %>% .[0:9*5+1] )
-yr_scale2 <- scale_x_continuous(breaks = 2009:2058 %>% .[0:9*5+1],
-                                labels = 2009:2058 %>% .[0:9*5+1] )
[…]

### 2025-01-23 analysis/scripts/0_bayesianModelExamplesA.R

@@ -15,12 +15,10 @@ stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
 stateScns = merge(stateScns,data.frame(interannualVar=c("list(R_CV=0.23,S_CV=0.0435)")))
 stateScns = merge(stateScns,data.frame(survivalModelNumber = c("M1","M0"), recruitmentModelNumber=c("M4","M0"),
                                        rAnthroSlopeSE = c(0.006,0.06),rFireSlopeSE = c(0.002,0.02),
-                                       sAnthroSlopeSE = c(0.0005,0.005),sIntSE = c(0.04,0.4),
-                                       rIntSE = c(0.25,2.5),sQuantile=c(0.9,0.9),rQuantile=c(0.9,0.9)))
+                                       sAnthroSlopeSE = c(0.0005,0.005),sIntSE = c(0.06,0.6),
+                                       rIntSE = c(0.35,2.5),sQuantile=c(0.9,0.9),rQuantile=c(0.9,0.9)))
 
 scns=merge(monitoringScns,stateScns)
-#scns$obsYears[scns$survivalModelNumber=="M0"]=4
-#scns$obsYears[scns$survivalModelNumber=="M0"]=4
 
 scns = subset(scns,!(grepl("2.5",rIntSE,fixed=T)&(collarCount==0)))
 scns$iAnthro = 0

### 2025-01-23 analysis/scripts/0_bayesianModelExamplesB.R

@@ -1,175 +1,0 @@
-# Prior and posterior predictions from Bayesian model
-library(caribouMetrics)
-library(tidyverse)
-library(ggpubr)
-theme_set(theme_bw())
-library(RColorBrewer)
-
-baseDir <- "."
-
-monitoringScns = data.frame(obsYears=c(24),collarCount=c(60),cowMult=c(6),collarInterval=c(1),
-                            assessmentYrs=c(1))
-stateScns = data.frame(obsAnthroSlope=c(1),projAnthroSlope=c(1))
-stateScns = merge(stateScns,data.frame(rep=seq(1:1)))
-stateScns = merge(stateScns,data.frame(interannualVar=c(NA,"list(R_CV=0.46,S_CV=0.087)","list(R_CV=0.23,S_CV=0.0435)")))
-stateScns = merge(stateScns,data.frame(sQuantile = c(0.1,0.5,0.9),rQuantile=c(0.1,0.5,0.9)))
-stateScns$qMax = 0;stateScns$uMax=0;stateScns$zMax=0
-scns=merge(monitoringScns,stateScns)
-
-scns$iAnthro = 0
-scns$tA = scns$iAnthro+(scns$obsYears)*scns$obsAnthroSlope
-scns$projYears = 50-scns$obsYears
-scns$N0 = 2000
-
-scns
-#scns$assessmentYrs=3
-
-####################
-eParsIn = list()
-eParsIn$cowCounts <- data.frame(Year = 1981:2023,
-                                Count = 100,
-                                Class = "cow")
-eParsIn$freqStartsByYear <- data.frame(Year = 1981:2023,
-                                       numStarts = 30)
-eParsIn$collarOnTime=1
-eParsIn$collarOffTime=12
-eParsIn$collarNumYears=1
-labFontSize = 10; breakInterval=5
-
-###############################
-#With bias adjustment
-simBig<-getSimsNational(replicates=3000,forceUpdate=T)
-
-results = caribouMetrics:::runScnSet(scns,eParsIn,simBig,getKSDists=F,printProgress=F)
-
-localPlotFn<-function(modTables,parameter,lowBound,highBound,facetVars,legendPosition,breakInterval){
-  allRes <- modTables$rr.summary.all
-  obs <- modTables$obs.all
-  simRange <- NULL#modTables$sim.all
-
-  pal2 = c("#EF8A62","#67A9CF")#brewer.pal(7,"RdBu")[c(2,6)]
-
[…]

### 2025-01-23 analysis/scripts/1_sensitivity1.R

@@ -40,7 +40,7 @@ pages=unique(scns$pageLab)
 
 ###################
 #Without bias
-setName = "s15" #14 is the same, just bigger batches.
+setName = "s16" #14 is the same, just bigger batches.
 scns$qMax = 0;scns$uMax=0;scns$zMax=0
 write.csv(scns,paste0("tabs/",setName,".csv"),row.names=F)
 pages=unique(scns$pageLab)

### 2025-01-23 analysis/scripts/2_viewSensitivityResults.R

@@ -14,11 +14,12 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 
 ########################
 #sensitivity
-setName = "s10" #need to do s10, s11, and s12
+setName = "s14" #need to do s13, s14, and s15
 dir.create(paste0("figs/",setName),recursive=T)
 scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
 
-
+numBatches= max(scns$repBatch)
+str(scns)
 table(scns$pageId)
 
 nrow(scns)
@@ -30,7 +31,8 @@ str(scns)
 
 setLTYVar <-function(scns){
   if(is.element("interannualVar",names(scns))&&(length(unique(scns$interannualVar))>1)){
-    scns$ltyVariable = "low"
+    scns$ltyVariable = "none"
+    scns$ltyVariable[grepl("0.23",scns$interannualVar,fixed=T)]="low"
     scns$ltyVariable[grepl("0.46",scns$interannualVar,fixed=T)]="high"
   }else{
     scns$ltyVariable=scns$collarInterval
@@ -51,7 +53,8 @@ for (p in pagesa){
   if(!file.exists(paste0("results/",setName,"/rTest",p,".Rds"))){pages=pages[pages!=p]}
 }
 
-batchStrip<-function(l,batches=c(10,seq(1:9))){
+
+batchStrip<-function(l,batches=c(seq(10,numBatches),seq(1:9))){
   for(b in batches){
     l=gsub(paste0("repBatch",b),"",l,fixed=T)
   }
@@ -65,7 +68,7 @@ addEV = F
 addProbs = F
 
 for(i in 1:length(pages)){
-  #combine=F;i=10
+  #combine=F;i=12
   cpageId=pages[i]
 
   if(i==length(pages)){
@@ -91,7 +94,7 @@ for(i in 1:length(pages)){
     scResults = readRDS(paste0("results/",setName,"/rTest",cpageId,".Rds"))
   }
 
-  if((as.numeric(strsplit(p,"repBatch")[[1]][2])!=10)&(as.numeric(strsplit(nextP,"repBatch")[[1]][2])<=as.numeric(strsplit(p,"repBatch")[[1]][2]))){
+  if((as.numeric(strsplit(p,"repBatch")[[1]][2])!=numBatches)&(as.numeric(strsplit(nextP,"repBatch")[[1]][2])<=as.numeric(strsplit(p,"repBatch")[[1]][2]))){
[…]

### 2025-01-23 cloud/setup_azure.sh

@@ -2,8 +2,7 @@
 #############################################
 
 #### Parameters to change
-setName="s13"
-
+setName="s16"
 
 #####Login etc##################
 # Accept the default tenant by pressing enter when prompted

### 2025-01-24 analysis/paper/paper.Rmd
Revisions in progress.

@@
 resultSet ="cmult6ay1aSf1"
-figSet = c("diffs","distScns","EVSI","power","EVPIall")
+figSet = c("diffslow","distScns","EVSI","power","EVPIall")
 
@@
 
-Examining the probability of
-correctly assessing population status (Figure \@ref(fig:figStatusErrors))
+Examining the probability of correctly assessing population status (Figure \@ref(fig:figStatusErrors))
 and the expected value of information (Figures \@ref(fig:figEVSI) and E.1) helps clarify how the need for monitoring varies with landscape condition.
@@
 projections informed by local monitoring data are less
-precise but not less accurate, which can happen if the prior viability predictions are more certain than the posterior predictions (e.g 
-left side of right panels in Figure \@ref(fig:examplePredictions)). At
-lower levels of disturbance, local monitoring can reduce uncertainty, but
+precise but not less accurate, which can happen if the prior viability predictions are more certain than the posterior predictions (e.g left side of right panels in Figure \@ref(fig:examplePredictions)).
+At
+lower levels of disturbance, local monitoring can increase accuracy, but
 there are diminishing returns as the number of years of monitoring increases (Figures
-\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). Reducing uncertainty is more difficult when interannual variation is high (Figures
-\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
+\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).
+Increasing accuracy is more difficult when interannual variation is high (Figures \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
 relationships between monitoring effort and the probability of correct
@@
 (Supplement F).
+
+Averaging over the full set of example trajectories for each scenario (Figures \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)) obscures variation among them that is important for monitoring decisions.
+For example, where disturbance is low but increasing, in the near-term monitoring is most valuable for identifying populations with lower than expected population growth rates (row 1 of Figure \@(fig:figStatusErrorsLow)), but refining knowledge of the current population state also helps identify populations that are likely to remain viable as disturbance increases (row 3 of Figure \@(fig:figStatusErrorsLow)).
+Status assessment accuracy is much higher among examples where the 95% credible interval for expected population growth rate does not include one (xx %) than among examples where it does (xx %), and this information could be used to guide monitoring.
+A more nuanced option is to recognize that the posterior probability that the population is viable ($P(\psi|x,\tau)$ in eq \@ref(eq:EVSI)) is an indicator of error probability (Figure \@ref(fig:figProbViability)).
+One could inform a decision about whether more information is needed for a particular population by selecting an error tolerance (e.g.
+the probability of correct status assessment should be >0.98) and using that threshold to distinguish cases where the posterior probability of viability is sufficiently certain (i.e.
+near enough to 0 or 1) from cases where more information is needed (Figure \@ref(fig:figProbViability)).
 
@@
 ```{r figDiffs, fig.cap=".", cache=TRUE}
-knitr::include_graphics(here::here("figs/forPaper/diffscmult6ay1aSf1.png"))
-file.rename(here::here("analysis/paper/figs_submit/diffscmult6ay1aSf1.pdf"), 
-            here::here("analysis/paper/figs_submit/Figure4_diffscmult6ay1aSf1.pdf"))
+knitr::include_graphics(here::here("figs/forPaper/diffslowcmult6ay1aSf1.png"))
+file.rename(here::here("analysis/paper/figs_submit/diffslowcmult6ay1aSf1.pdf"), 
+            here::here("analysis/paper/figs_submit/Figure4_diffslowcmult6ay1aSf1.pdf"))
 ```

### 2025-01-24 analysis/scripts/2_viewSensitivityResults.R

@@ -16,13 +16,16 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 #sensitivity
 setName = "s14" #need to do s13, s14, and s15
 dir.create(paste0("figs/",setName),recursive=T)
+dir.create(paste0("tabs/",setName),recursive=T)
+
 scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
+
 
 numBatches= max(scns$repBatch)
 str(scns)
 table(scns$pageId)
 
-nrow(scns)
+1942/(nrow(scns)/60)
 pagesa=unique(scns$pageId)
 
 pages = pagesa
@@ -121,7 +124,6 @@ for(i in 1:length(pages)){
   exResults$grp = paste(exResults$type,exResults$quantile, exResults$tA,exResults$obsYears)
   exResults$Anthro2023 = pmax(0,exResults$tA)#pmax(0,exResults$tA-2) #correcting for error - change back in round 4
 
-
   #unique(scResults$obs.all$collarCount)
   #unique(scResults$obs.all$ltyVariable)
   pars = unique(scResults$obs.all$parameter)
@@ -147,7 +149,7 @@ for(i in 1:length(pages)){
   names(probs)= gsub("_Population growth rate","",names(probs),fixed=T)
   probs$startYear = probs$startYear+probs$preYears
   probs$projectionTime = probs$Year-2023
-  probs$Anthro2023 = pmax(0,probs$tA)#pmax(0,probs$tA-2) #correcting for error - change back in round 4
+  probs$Anthro2023 = probs$tA
   probs$YearsOfProjection = probs$projectionTime
 
   #See disturbance scenarios
@@ -195,21 +197,15 @@ for(i in 1:length(pages)){
   #make wide obs table
   obsWide <- subset(scResults$obs.all,(type=="true"))%>%
     pivot_wider(names_from = parameter, values_from = Mean)
-    str(obsWide)
   obsWide$startYear=obsWide$startYear+obsWide$preYears
-
-  check = subset(obsWide,is.na(interannualVar))
-  check$expLambda = check[["Adult female survival"]]*(1+check[["Adjusted recruitment"]])
-  range(check$expLambda-check[["Population growth rate"]])
-  check=NULL
-
   obsWide$c = obsWide[["Adjusted recruitment"]]*2/obsWide[["Recruitment"]]
 
   #now combine and make small
   names(obsWide)[names(obsWide)=="Population growth rate"]= "trueMean"
[…]

### 2025-01-24 cloud/setup_azure.sh

@@ -15,7 +15,6 @@ state:state, enableAutoScale:enableAutoScale,
 allocState:allocationState}" \
 --output table
 
-
 #### Move files to container ##############
 end=`date -u -d "7 days" '+%Y-%m-%dT%H:%MZ'`
 sastoken=`az storage container generate-sas --account-name ecdcwls --expiry $end --name jhughes --permissions racwdli -o tsv --auth-mode login --as-user`

### 2025-01-27 analysis/paper/paper.Rmd
MS revisions in progress

@@
 
-resultScenario = "S10"
+resultScenario = "S14"
 resultSet ="cmult6ay1aSf1"
-figSet = c("diffslow","distScns","EVSI","power","EVPIall")
+figSet = c("diffslow","distScns","EVSI","power","EVPIall","powerLam","probViability")
 
@@
 [-@dyson_existing_2022].
-For brevity, we refer to this as the "national" model.
-For clarity, we include a full description of the model in the following paragraphs.
+For brevity, we refer to this as the "national" model, and for clarity we include a full description in the following paragraphs.
 Together, a set of example trajectories from the national model represent alternative hypotheses about the true state of a particular caribou population.
@@
 In our simulated example populations, mortality risk does not vary over
-time or among individuals, so the observed state (dead or alive) of a
-collared animal $i$ in month $m$ and year $t$ depends on the true
-survival rate $\dot{S}_t$ as follows: 
-\begin{equation}
-\hat{I}_{i,t,m} \sim \text{Bernoulli}(\dot{S}^{1/12}_t\hat{I}_{i,t,m-1}).
+among seasons or individuals, so the number of collared cows at the time of the composition survey $\hat{T}_t$ depends on the true survival rate $\dot{S}_t$ and the number of collared cows at the beginning of the year $\hat{I}_t$ as follows:
+\begin{equation}
+\hat{T}_t \sim \text{Binomial}(\hat{I}_t,\dot{S}_t).
   (\#eq:hatI)
@@
 Local demographic and covariate data provided as inputs to the Bayesian
-population model include the observed state (dead or alive) of each
-collared animal in each month ($\hat{I}_{i,t,m}$), the observed number
+population model include the number of collared cows at the start of the year ($\hat{I}_{t}$), the number of collared animals that survive the year ($\hat{T}_{t}$), the observed number
 of adult females and calves in the recruitment surveys
@@
 
-Following others [@eacker_web-based_2019; @lamb_effectiveness_2024], Kaplan-Meier
-estimates [@kaplan_nonparametric_1958] of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ are estimated from known-fate radio
-collar data using the *survival* R package [@therneau_modeling_2000],
-with precisions for years with no observed mortality estimated using a
-separate Bayesian model. Observed estimated survival is modelled as a
-Gaussian distributed function of the true survival probability,
-truncated between 0 and 1:
-\begin{equation}
-\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
+Survival $S_t$ is estimated from the observed number of collared cows at the beginning of each year $\hat{I}_t$ and the number of those that survive the year $\hat{T}_t$ assuming a Binomial distribution:
+\begin{equation}
+\hat{T}_t \sim \text{Binomial}(\hat{I}_t,S_t).
   (\#eq:hatSEst)
 \end{equation}
+
+Note that for analysis of real known-fate radio collar data data one should use a survival analysis method that accounts for variation throughout the year in both survival rate and the number of collared animals [e.g.
+@dalgarno_bboutools_2024; @eacker_web-based_2019; @lamb_effectiveness_2024].
+However, our implementation of a two-step non-parametric survival analysis method [following @eacker_web-based_2019] yielded slightly biased survival estimates.
 
@@
 (\#eq:S)
-\end{equation}
-
-To handle cases in which there is little or no survival data, we use an alternative parametric exponential survival model,
-wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
-individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed: 
-\begin{equation}
-\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).
-  (\#eq:hatIEst)
 \end{equation}
@@
 and in the *caribouMetrics* R package available on GitHub at
-<https://github.com/LandSciTech/caribouMetrics>.
+<https://github.com/LandSciTech/caribouMetrics/tree/EI-paper-submission>.
 
@@
 file.rename(here::here("analysis/paper/figs_submit/bayesianModelExamplesLowEffort.pdf"), 
-            here::here("analysis/paper/figs_submit/Figure3_bayesianModelExamplesLowEffort.pdf"))
+            here::here("analysis/paper/figs_submit/Figure4_bayesianModelExamplesLowEffort.pdf"))
 ```
@@
 file.rename(here::here("analysis/paper/figs_submit/diffslowcmult6ay1aSf1.pdf"), 
-            here::here("analysis/paper/figs_submit/Figure4_diffslowcmult6ay1aSf1.pdf"))
+            here::here("analysis/paper/figs_submit/Figure5_diffslowcmult6ay1aSf1.pdf"))
 ```
@@
 file.rename(here::here("analysis/paper/figs_submit/powercmult6ay1aSf1.pdf"), 
-            here::here("analysis/paper/figs_submit/Figure5_powercmult6ay1aSf1.pdf"))
+            here::here("analysis/paper/figs_submit/Figure6_powercmult6ay1aSf1.pdf"))
 ```
@@
 file.rename(here::here("analysis/paper/figs_submit/EVSIcmult6ay1aSf1.pdf"), 
-            here::here("analysis/paper/figs_submit/Figure6_EVSIcmult6ay1aSf1.pdf"))
+            here::here("analysis/paper/figs_submit/Figure7_EVSIcmult6ay1aSf1.pdf"))
+```
+
+```{r figStatusErrorsLow, fig.cap=".", cache=TRUE}
+knitr::include_graphics(here::here("figs/forPaper/powerLamcmult6ay1aSf1.png"))
+file.rename(here::here("analysis/paper/figs_submit/powerLamcmult6ay1aSf1.pdf"), 
+            here::here("analysis/paper/figs_submit/Figure8_powerLamcmult6ay1aSf1.pdf"))
+```
+
+```{r figProbViability, fig.cap=".", cache=TRUE}
+knitr::include_graphics(here::here("figs/forPaper/probViabilitycmult6ay1aSf1.png"))
+file.rename(here::here("analysis/paper/figs_submit/powerLamcmult6ay1aSf1.pdf"), 
+            here::here("analysis/paper/figs_submit/Figure9_probViabilitycmult6ay1aSf1.pdf"))
 ```
@@
 
-Figure 4: Posterior predictions from the Bayesian model for population trajectories with informative (a) and weakly informative (b) priors given higher interannual variation ($v_R=0.46,v_S=0.087$) and limited monitoring.
-The population is monitored for 4 years with 15 collars per year, and 3 cows per collared cow in the composition surveys.
+Figure 4: Posterior predictions from the Bayesian model for population trajectories with informative (a) and weakly informative (b) priors given higher interannual variation ($v_R=0.46,v_S=0.087$) and limited monitoring (4 years, 15 collars per year, 3 cows per collared cow in composition surveys).
 See Figure 3 for disturbance trends and other details.
@@
 
-Figure 6: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
+Figure 6: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
 When disturbance is high the probability of correct status assessment is high even without local monitoring.
-Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as the number of years increases.
+Where disturbance is low or moderate, increasing the number of years of monitoring increases accuracy, but there are diminishing returns on additional monitoring as the number of years increases.
 In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).
 
-Figure 7: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort.
+Figure 7: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort.
 When disturbance is high EVSI is low, indicating that local monitoring does not change predicted status.
@@
 
+Figure 8: In the low anthropogenic disturbance scenario (Figure \@ref(fig:figDisturbance)) probability of a correct status assessment varies with the true growth rate of the population (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
+When anthropogenic disturbance is low and expected to increase (Figure \@ref(fig:figDisturbance)) increasing the number of years of monitoring helps to distinguish populations with current demographic rates substantially lower than 1 (top left), and to identify populations with higher than expected growth rates that fare better as disturbance increases (lower right).
+Increasing the number of collars more effectively increases near-term accuracy when interannual variation is low (top left).
+There is little advantage to increasing the number of collars if the goal is long-term projection (bottom row).
+Increasing the accuracy of status assessment for populations with growth rates close to 1 is more difficult (middle columns).
+In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year.
+
+Figure 9: The posterior probability that the population is viable ($P(\psi|x,\tau)$ in eq \@ref(eq:EVSI)) predicts the probability of a correct status assessment (estimated from the 114,000 example cases in Figures 5 to 7 with a generalized additive model).
+Status assessment accuracy increases as the probability of viability nears 0 or 1.
+Cases in which there is no need for additional monitoring if one is willing to accept a 2% chance of error fall above the dashed line.
+
 ### Colophon

### 2025-01-27 analysis/paper/supplement.Rmd
MS revisions in progress

@@
 |$\hat{W}_t$|animals|Cows observed in calf:cow survey.|
-|$\hat{T}_t$|animals|Number of collared cows.|
-|$\hat{S}_t$|probability|Kaplan-Meier estimate of observed adult female survival.|
-|$\tau_t$||Precision of Kaplan-Meier survival estimate.|
-|$\hat{I}_{i,t,m}$|0 or 1|Indicator of whether animal $i$ is dead or alive.|
+|$\hat{T}_t$|animals|Number of collared cows at end of year.|
+|$\hat{I}_{t}$|animals|Number of collared cows at beginning of year.|
 |**Parameters estimated in the Bayesian model**|
@@
 
-Adding unknown unknown bias does increase uncertainty, but these increases are not large (compare Figs 4 and \@ref(fig:powercmult6ay1aSf1NoBias)).
+Adding unknown unknown bias does increase uncertainty, but these increases are not large (compare Figs 6 and \@ref(fig:figStatusErrorsNoBias)).
 A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults, and the probability of missing calves (supplement B).
@@
 
-```{r powercmult6ay1aSf1NoBias, fig.cap="The probability of a correct status assessment with unknown bias in recruitment estimates from calf:cow surveys.
+```{r figDiffsNoBias, fig.cap="The distribution of differences between true expected population growth rates and Bayesian posterior expected growth rates without bias in recruitment estimates from calf:cow surveys.
+In these examples there is no interannual variation, 6 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
+See Figure 6 for additional details.", cache=TRUE}
+knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/diffsnonecmult6ay1aSf1.png")))
+```
+
+```{r figStatusErrorsNoBias, fig.cap="The probability of a correct status assessment without bias in recruitment estimates from calf:cow surveys.
 In these examples there are 6 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
-See Figure 4 for additional details.
+See Figure 6 for additional details.
 ", cache=TRUE}
 knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powercmult6ay1aSf1.png")))
+```
+
+```{r figStatusErrorsLowNoBias, fig.cap=".", cache=TRUE}
+knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powerLamcmult6ay1aSf1.png")))
+```
+
+```{r figProbViabilityNoBias, fig.cap=".", cache=TRUE}
+knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/probViabilitycmult6ay1aSf1.png")))
 ```

### 2025-01-27 analysis/scripts/1_sensitivity1.R

@@ -50,7 +50,7 @@ pages=unique(scns$pageLab)
 #Different anthropogenic disturbance scenarios, collar intervals. With bias and default interannual variability.
 setName = "s15" #formerly s11
 monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60),
-                             cowMult=c(3,6,9),collarInterval=c(1,4),
+                             cowMult=c(3,9),collarInterval=c(1,4),
                              assessmentYrs=c(1))
 #TO DO - in next iteration, remove multiple years of one collar
 monitoringScns = subset(monitoringScns, !((obsYears>1)&(collarCount==0))&!((collarCount==1)&(cowMult>3))&!((collarCount==0)&(collarInterval>1)))
@@ -69,7 +69,7 @@ scns$preYears = max(scns$obsYears)-scns$obsYears
 scns$iAnthro = scns$tA-(scns$obsYears+scns$preYears-1)*scns$obsAnthroSlope
 scns$projYears = 20
 unique(scns$iAnthro)
-scns$repBatch = ceiling(scns$rep/4)
+scns$repBatch = ceiling(scns$rep/12)
 table(scns$repBatch)
 scns$N0 = 5000
 

### 2025-01-27 cloud/make_batch_scripts.R

@@ -8,7 +8,7 @@ if(0){
   keyring::key_get("Azure_subnetId")
 }
 cargs <- commandArgs(trailingOnly = TRUE)
-#cargs=c("s14","https://ecdcwls.blob.core.windows.net/jhughes/?se=2025-01-24T17%3A50Z&sp=racwdli&sv=2022-11-02&sr=c&skoid=536c15ec-3c37-4b86-ae47-a3fefedad9c4&sktid=740c5fd3-6e8b-4176-9cc9-454dbe4e62c4&skt=2025-01-17T17%3A50%3A13Z&ske=2025-01-24T17%3A50%3A00Z&sks=b&skv=2022-11-02&sig=py82KrESgpDj/yy2aIgki%2BhfZfTKgkTYF%2BUq0muzXn4%3D")
+#cargs=c("s15","https://ecdcwls.blob.core.windows.net/jhughes/?se=2025-01-24T17%3A50Z&sp=racwdli&sv=2022-11-02&sr=c&skoid=536c15ec-3c37-4b86-ae47-a3fefedad9c4&sktid=740c5fd3-6e8b-4176-9cc9-454dbe4e62c4&skt=2025-01-17T17%3A50%3A13Z&ske=2025-01-24T17%3A50%3A00Z&sks=b&skv=2022-11-02&sig=py82KrESgpDj/yy2aIgki%2BhfZfTKgkTYF%2BUq0muzXn4%3D")
 setName = cargs[1]
 allScns = read.csv(paste0("tabs/",setName,".csv"))
 scn_nums <- unique(allScns$pageId)
@@ -22,7 +22,7 @@ sasurl <- cargs[2]
 vmSizes <- c(2,4,8,16,32)
 
 # assuming 12 is the max nodes of 27 for all LERS that we want to use at a time
-nSlotsPerNodeOpts <- nBatches/1:12
+nSlotsPerNodeOpts <- nBatches/1:16
 
 wchnSlotsPerNode <- sapply(nSlotsPerNodeOpts,
                            function(x) min((vmSizes - x)[(vmSizes - x) >= 0]))|>

### 2025-02-21 analysis/paper/paper.Rmd
preparing EI submission

@@
 ---
-title: "Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions"
+title: "Integration of national demographic-disturbance relationships and local data can improve caribou population viability projections and inform monitoring decisions"
 author:
@@
   Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed.
-There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but where disturbance is high we have no evidence of self-sustaining populations across the country.
+There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but no populations inhabiting areas with high anthropogenic disturbance are considered self-sustaining.
 Demographic projections derived from local population-specific data are uncertain for populations with limited monitoring.
-We propose a simple Bayesian population model that integrates prior information from a national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and to reduce the risk that lack of local data will be used as a reason to delay conservation action.
-The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a term derived from a simple model of the recruitment survey observation process.
+We propose a simple Bayesian population model that integrates prior information from a national analysis of demographic-disturbance relationships with available local demographic data to improve population viability projections, and to reduce the risk that a lack of local data will be used as a reason to delay conservation action.
+The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves, through a term derived from a simple model of the recruitment survey observation process.
 We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.
 Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships.
-At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years may be better than intensive short-term monitoring for improving the reliability of status projections.
+At lower disturbance levels where initial uncertainty is high, each additional year of monitoring provides less new information.
+The estimated probability of viability indicates whether more information is needed to improve accuracy of population viability projections.
 keywords: |
@@
 
-  * We show how Bayesian integration of prior information from a national analysis of demographic-disturbance relationships can reduce uncertainty about the viability of sparsely monitored populations.
+  * We show how Bayesian integration of local data and prior information from a national analysis of demographic-disturbance relationships can reduce uncertainty about the viability of sparsely monitored populations.
 
@@
 
-  * At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years is likely better than intensive short-term monitoring for reducing uncertainty about population viability.
-
-  * We introduce a method for acknowledging potential errors in recruitment surveys to help clarify where and when these these errors are likely to alter conclusions about population viability.
+  * At lower disturbance levels where uncertainty is higher there are diminishing returns on additional monitoring, and the estimated probability of viability indicates whether more information is needed.
+
+  * We introduce a method for acknowledging potential errors in recruitment surveys to help clarify where and when these errors are likely to alter conclusions about population viability.
 
@@
 
-The number of species threatened with extinction is high and getting
-higher [@ipbes_global_2019]. Reliable information on the status of
-populations and likely impacts of management are needed to inform
-conservation action, but effective monitoring is challenging and costly
-[@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;
-@dunham_strategic_2023]. Resources to support conservation and recovery
-are limited [@buxton_balancing_2022; @mccarthy_financial_2012;
-@gerber_conservation_2016; @wintle_spending_2019], delays in action increase risk
-[@martin_acting_2012], and some populations have been monitored to
-extinction [@lindenmayer_counting_2013]. A recent analysis of action
-plans for Canadian species at risk found that half of tasks are research and
-monitoring, and this proportion is highest for species with higher risk
-of extinction, highlighting a need to more carefully consider when,
-where, and why additional information is needed to inform conservation
-action [@buxton_balancing_2022]. Analysis of the implications of
-monitoring outcomes for decisions can help to clarify whether additional
-monitoring is necessary to inform conservation action
-[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
-@gregory_structured_2012].
-
-Analysis methods that can make better use of available monitoring
-data to inform decisions include Bayesian population modelling
-[@schaub_integrated_2021] and value of information analyses
-[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
-@gregory_structured_2012].
+The number of species threatened with extinction is high and getting higher [@ipbes_global_2019].
+Reliable information on the status of populations and likely impacts of management are needed to inform conservation action, but effective monitoring is challenging and costly [@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;@dunham_strategic_2023].
+Resources to support conservation and recovery are limited [@buxton_balancing_2022; @mccarthy_financial_2012;@gerber_conservation_2016; @wintle_spending_2019], delays in action increase risk [@martin_acting_2012], and some populations have been monitored to extinction [@lindenmayer_counting_2013].
+A recent analysis of action plans for Canadian species at risk found that half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022].
+Analysis of the implications of monitoring outcomes can help to clarify whether additional monitoring is necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;@gregory_structured_2012].
+
+Boreal caribou are widely and sparsely distributed through the Canadian boreal region across a gradient of anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020] that includes some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022; @ibisch_global_2016;
+@venter_sixteen_2016].
+Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the area that is disturbed, and no populations inhabiting areas with high anthropogenic disturbance are considered viable in the absence of interventions such as predator control or maternal penning (i.e.
+self-sustaining) [@eccc_scientific_2011; @johnson_science_2020].
+Identifying populations that are declining is important for clarifying where and when conservation action is needed.
+Disturbance affects caribou demography both through direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022; @eccc_scientific_2011].
+Decreasing recruitment with increased disturbance has also been observed over time where historic demographic data exist [@rudolph_demographic_2017; @hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships [@johnson_science_2020].
+There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020; @neufeld_primary_2021;
+@fortin_forest_2017; @gagne_impact_2016], which leads to uncertainty in projections that rely only on national-scale relationships [@stewart_climate-informed_2023; @dyson_existing_2022].
+
+Analysis methods that can make better use of available monitoring data to inform conservation decisions include Bayesian population modelling [@schaub_integrated_2021] and value of information analyses [@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;@gregory_structured_2012].
 For sparsely monitored populations, incorporating prior information from elsewhere can help reduce uncertainty [@drummond_combining_2018; @parlato_incorporating_2021; @tufto_bayesian_2000].
-To be useful, these methods must be tailored
-to particular species and decision contexts. Here, we offer a simple
-Bayesian population model for boreal caribou (*Rangifer tarandus
-caribou*) that integrates prior information about the impacts of landscape 
-change [@johnson_science_2020] with available demographic data [building on @eacker_web-based_2019] to reduce uncertainty in
-population viability projections. Following the value of information analysis
-framework proposed by Dunham et al. (2023), we combine this Bayesian
-model with simulations of plausible population trajectories
-[@stewart_climate-informed_2023; @dyson_existing_2022] to (i) show how the
-value of local monitoring varies with landscape condition, and (ii) assess
-the ability of alternative monitoring scenarios to reduce the risk of
-errors in population viability projections. Our approach is tailored to
-boreal caribou, but could be adapted for other species at risk found in
-multiple populations that are costly to monitor and sensitive to habitat
-disturbance.
-
-Boreal caribou are widely and sparsely distributed through the Canadian
-boreal region across a gradient of anthropogenic disturbance
-[@eccc_scientific_2011; @johnson_science_2020] that includes some of the
-most intact and inaccessible places remaining on earth
-[@hirsh-pearson_canadas_2022; @ibisch_global_2016;
-@venter_sixteen_2016]. Across 58 boreal caribou study areas, both
-survival and recruitment decrease with the percentage of the study area
-that is disturbed, and no populations of caribou inhabiting areas with high anthropogenic disturbance are considered self-sustaining [@eccc_scientific_2011; @johnson_science_2020].
-The mechanisms behind these
-demographic-disturbance relationships include direct impacts on habitat
-availability and use, and indirect impacts on predators and alternate
-prey [@nbckc_boreal_2022; @eccc_scientific_2011]. Decreases in
-recruitment as disturbance increases are also observed over time in
-places where there are historic demographic data
-[@rudolph_demographic_2017; @hervieux_widespread_2013], but many places
-lack sufficient data to characterize local demographic-disturbance
-relationships [@johnson_science_2020]. There is also variation in
-demographic rates among study areas, particularly where anthropogenic
-disturbance is low [@johnson_science_2020; @neufeld_primary_2021;
-@fortin_forest_2017; @gagne_impact_2016], and this variation leads to
-uncertainty in projections derived from national disturbance-demographic
-relationships [@stewart_climate-informed_2023; @dyson_existing_2022].
-
-Interventions such as wolf culls, maternal penning, and restoration of
-linear features are intended to improve survival and reproduction of
-populations that are not otherwise self-sustaining
-[@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a
-need to evaluate and compare effects of boreal caribou conservation
-actions with careful monitoring. Our focus in this paper is on assessing
-and projecting self-sustainability of populations in the absence of
-these interventions, to clarify when additional information is needed to
-reduce uncertainty about whether there is a need for action. Where
-actions are taken, outcomes should be monitored; the model we
-propose here does not include effects of management interventions, and
-cannot be used to project outcomes for managed populations.
+Various efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021; @mcnay_demographic_2022;@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024].
+An advantage of Bayesian demographic models is that they account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021].
+However, Bayesian caribou population models developed to-date [@moeller_integrating_2021; @mcnay_demographic_2022;@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not included prior information about the observed distribution of demographic rates across the country that could reduce uncertainty about the status of poorly monitored populations.
+
+Here, we offer a simple Bayesian population model for boreal caribou (*Rangifer tarandus
+caribou*) that integrates prior information about the impacts of landscape change [i.e.
+demographic-disturbance relationships in @johnson_science_2020] with available demographic data to reduce uncertainty in population viability projections.
+Note that for the purposes of this analysis we consider a population to be viable if the population growth rate is > 0.99.
+We also offer a novel method of acknowledging and accounting for errors in composition surveys.
+Following a value information analysis framework for conservation status assessments [@dunham_strategic_2023], we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023; @dyson_existing_2022] to (i) show how the value of local monitoring varies with landscape condition, and (ii) assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections.
+
+Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population viability [@eccc_scientific_2011; @johnson_science_2020] because it accounts for uncertainties in both local data and demographic-disturbance relationships.
+We show that integrating the most informative predictor of outcomes for boreal caribou across their range can substantially reduce uncertainty in local demographic projections and improve the accuracy of population viability projections on changing landscapes.
+This approach can help clarify where available information is – and is not – sufficient for assessing and projecting population viability, and thereby reduce the risk that uncertainty is used as a reason to delay conservation action for imperiled populations [@bissonnette_controverses_2020; @boan_climate_2018].
+Our approach is tailored to boreal caribou, but could be adapted for other species at risk found in multiple populations that are costly to monitor and sensitive to habitat disturbance.
+
+# Methods
+
+## Overview of methods and notation
+
+To investigate how the accuracy and precision of predictions vary with both landscape condition and monitoring effort, we analyze simulated data using a Bayesian model with informative priors (section 2.4) and a composition survey bias term (section 2.1).
+Steps in our analysis procedure include defining anthopogenic disturbance and monitoring scenarios (section 2.2), simulating population dynamics and monitoring (section 2.3) to produce example demographic trajectories for each scenario, and analyzing example demographic trajectories with a Bayesian model that enables integration of (simulated) local demographic data with prior knowledge of demographic-disturbance relationships (section 2.4).
+Outcomes of interest include expected population growth rate, the probability of correctly assessing population status, the posterior probability of population viability, and the expected value of local information (section 2.5).
+We refer to the model used to simulate example trajectories as the "national" model because parameters are from a national analysis of demographic-disturbance relationships [@johnson_science_2020].
+We distinguish this national model from the Bayesian model that is used to analyze simulated example demographic trajectories.
+Where appropriate the national model and the Bayesian model share the same structure, and we have selected notation to highlight the similarities.
+Bayesian model estimates are distinguished from observed values, estimated expected values, true values, and true expected values using $\hat{\text{ }}$ $\tilde{\text{ }}$ $\dot{\text{ }}$ and $\bar{\text{ }}$ symbols respectively; for example, $S_t$ is the Bayesian estimate of survival in year $t$, $\hat{S}_t$ is observed survival, $\tilde{S}_t$ is the Bayesian estimate of expected survival (without stochastic interannual variation), $\dot{S}_t$ is true survival, and $\bar{S_t}$ is true expected survival (without stochastic interannual variation).
+For reference, all variables and parameters are defined in supplemental Table A1.
+
+```{r setup2}
+library(ggplot2)
+library(dplyr)
+library(stringr)
+library(tidyr)
+library(purrr)
+library(ggpubr)
+library(MetBrewer)
+library(here)
+library(tmap)
+library(sf)
+
+devtools::load_all(here::here())
+
+#set paths
+data_path_raw <- "analysis/data/raw_data/"
+data_path_drvd <- "analysis/data/derived_data"
+
+pal_nm <- "Redon"
+
+fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)
+
+runScenarios = F #To do: set up to run full analysis from within this document.
+
+resultScenario = "S13"
+resultSet ="cmult6ay1aSf1"
+figSet = c("diffslow","distScns","EVSI","power","EVPIall","powerLam","probViability")
+
+erSums = read.csv(here::here(paste0("tabs/",resultScenario,"/ErrorsByBandcmult6ay1aSf1.csv")))
+
+if(runScenarios){
+    source(here::here("analysis/scripts/0_nationalModelTrajectories.R"))
+    source(here::here("analysis/scripts/0_bayesianModelExamples.R"))
+    #create scenarios with sensitivity1.R and run in cloud
+    #run viewSensitivityResults.R for S8 and S7
+    #run EVSIPlots.R
+}
+
+for(f in figSet){
+  if(f =="EVPIall"){addBit=""}else{addBit=resultSet}
+  currentFile = here::here(paste0("figs/",resultScenario,"/",f,addBit,".png"))
+  newFile = here::here(paste0("figs/forPaper/",f,addBit,".png"))
+  file.copy(currentFile, newFile,overwrite=T)
+
+  currentFile = here::here(paste0("figs/",resultScenario,"/",f,addBit,".pdf"))
+  newFile = here::here(paste0("analysis/paper/figs_submit/",f,addBit,".pdf"))
+  file.copy(currentFile, newFile,overwrite=T)
+}
+```
+
+## Caribou population monitoring data and composition survey errors
 
@@
 aerial surveys of the composition of groups of animals that include
-collared animals [@nbckc_nbckc_2022; @eacker_web-based_2019; @dalgarno_bboutools_2024]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that
-different demographic groups are equally detectable
+collared animals [@nbckc_nbckc_2022]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that different demographic groups are equally detectable
 [@ellington_divergent_2020]. Errors that can bias composition survey
@@
 failing to detect calves [@ellington_divergent_2020;
-@decesare_estimating_2012].
-
-Survival and recruitment survey data can be combined in a simple
+@decesare_estimating_2012]. Survival and recruitment survey data can be combined in a simple
 recruitment-mortality population model [@hatter_moose_1991;@hatter_revisiting_2020] that serves
@@
 @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].
-Various efforts have been made to improve caribou population viability
-projections and growth rate estimates using Bayesian methods
-[@moeller_integrating_2021; @mcnay_demographic_2022;
-@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024], following broader trends in the analysis of population dynamics [@schaub_integrated_2021].
-An advantage of Bayesian demographic models is that they account for multiple sources of uncertainty
-[@schaub_integrated_2011; @eacker_assessing_2017;
-@schaub_integrated_2021]. However, there is very little monitoring data
-available for some caribou populations [@johnson_science_2020] and there
-is a risk that lack of local data will be used as a reason to delay
-conservation action [@bissonnette_controverses_2020; @boan_climate_2018]. Bayesian caribou population
-models [@moeller_integrating_2021; @mcnay_demographic_2022;
-@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not thus far included demographic-disturbance relationships, and we are concerned that uncertainty in models
-informed only by local demographic data could be used to delay conservation action for poorly monitored populations. We propose to reduce uncertainty in assessments and projections of local
-caribou population viability by integrating prior information from
-national analysis of demographic-disturbance relationships [@johnson_science_2020] with
-available local demographic data in a simple Bayesian population model
-for boreal caribou.
 
@@
 [@ellington_divergent_2020; @decesare_estimating_2012], none have thus
-far examined sensitivity of Bayesian caribou population models to these
-sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
-@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. We address this gap with a bias term derived from a simple model of the recruitment survey observation
-process.
-
-Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population viability
-[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
-accounts for uncertainties in local demographic data and
-demographic-disturbance relationships. We show that integrating the most
-informative predictor of outcomes for boreal caribou across their range
-can substantially reduce uncertainty in local demographic projections
-and status assessments, can improve viability projections on changing
-landscapes, and can help clarify where, why and when additional data is
-required to inform decisions and project outcomes on changing
-landscapes. We hope that this approach will help to identify where available
-information is sufficient for assessing and projecting population
-viability, and thereby reduce the risk that uncertainty is used as a
-reason to delay conservation action for imperiled populations [@bissonnette_controverses_2020; @boan_climate_2018].
-
-# Methods
-
-## Overview of methods and notation
-
-To investigate how the accuracy and precision of predictions vary with both landscape condition and monitoring effort we analyze simulated data using a Bayesian model with informative priors and a composition survey bias term (section 2.1).
-Steps in our analysis procedure include defining anthopogenic disturbance and monitoring scenarios (section 2.2), simulating population dynamics and monitoring (section 2.3) to produce example demographic trajectories for each scenario, and analyzing example demographic trajectories with a Bayesian model that enables integration of (simulated) local demographic data with prior knowledge of demographic-disturbance relationships (section 2.4).
-Outcomes of interest include expected population growth rate, the probability of correctly assessing population status, and the expected value of local information (section 2.5).
-We refer to the model used to simulate example trajectories as the "national" model because parameters are from a national analysis of demographic-disturbance relationships [@johnson_science_2020].
-We distinguish this national model from the Bayesian model that is used to analyze simulated example demographic trajectories.
-Where appropriate the national model and the Bayesian model share the same structure, and we have selected notation to highlight the similarities.
-Bayesian model estimates are distinguished from observed values, estimated expected values, true values, and true expected values using $\hat{\text{ }}$ $\tilde{\text{ }}$ $\dot{\text{ }}$ and $\bar{\text{ }}$ symbols respectively; for example, $S_t$ is the Bayesian estimate of survival in year $t$, $\hat{S}_t$ is observed survival, $\tilde{S}_t$ is the Bayesian estimate of expected survival (without interannual variation), $\dot{S}_t$ is true survival, and $\bar{S_t}$ is true expected survival (without interannual variation).
-For reference, all variables and parameters are defined in supplemental Table A1.
-
-```{r setup2}
-library(ggplot2)
-library(dplyr)
-library(stringr)
-library(tidyr)
-library(purrr)
-library(ggpubr)
-library(MetBrewer)
-library(here)
-library(tmap)
-library(sf)
-
-devtools::load_all(here::here())
-
-#set paths
-data_path_raw <- "analysis/data/raw_data/"
-data_path_drvd <- "analysis/data/derived_data"
-
-pal_nm <- "Redon"
-
-fig_widths <- c(min = 1.18, single = 3.543, mid = 5.51, two = 7.48)
-
-runScenarios = F #To do: set up to run full analysis from within this document.
-
-resultScenario = "S14"
-resultSet ="cmult6ay1aSf1"
-figSet = c("diffslow","distScns","EVSI","power","EVPIall","powerLam","probViability")
-
-if(runScenarios){
-    source(here::here("analysis/scripts/0_nationalModelTrajectories.R"))
-    source(here::here("analysis/scripts/0_bayesianModelExamples.R"))
-    #create scenarios with sensitivity1.R and run in cloud
-    #run viewSensitivityResults.R for S8 and S7
-    #run EVSIPlots.R
-}
-
-for(f in figSet){
-  if(f =="EVPIall"){addBit=""}else{addBit=resultSet}
-  currentFile = here::here(paste0("figs/",resultScenario,"/",f,addBit,".png"))
-  newFile = here::here(paste0("figs/forPaper/",f,addBit,".png"))
-  file.copy(currentFile, newFile,overwrite=T)
-
-  currentFile = here::here(paste0("figs/",resultScenario,"/",f,addBit,".pdf"))
-  newFile = here::here(paste0("analysis/paper/figs_submit/",f,addBit,".pdf"))
-  file.copy(currentFile, newFile,overwrite=T)
-}
-```
-
-## Caribou population monitoring data and composition survey errors
-
-We focus on analyzing survival of radio-collared females and the ratio of calves to cows in aerial
-surveys of groups that include collared females [@nbckc_nbckc_2022],
-because these remain the most widely available sources of demographic
-information for boreal caribou populations [@johnson_science_2020].
-GPS radio collars report animal locations at regular intervals, and cease moving when animals die.
-Aerial recruitment surveys are usually done in March or April, and radio-collared females are often used to locate groups of animals.
-Surveyors use visual cues to distinguish calves that have survived their first winter (~ 9 months old) from adult cows and bulls.
-Surveyed groups also contain uncollared adults so sample size is typically larger for recruitment than for survival estimates.
-Errors that can bias composition survey results include misidentifying young
-bulls as cows or vice versa, and failing to detect calves
-[@ellington_divergent_2020; @decesare_estimating_2012].
+far integrated these sources of error into Bayesian caribou population models [@moeller_integrating_2021; @mcnay_demographic_2022; @eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024].
+We address this gap with a bias term derived from a simple model of the recruitment survey observation process.
 
@@
 and $z$ (see Supplement B for details). For simplicity, we combine the
-probabilities into a bias term $c$ that is integrated into
-the Bayesian population model and simulations (below):
+probabilities into a bias term $c$ that is integrated into the national simulation model (section 2.4) and
+the Bayesian model (section 2.5):
 
@@
 To investigate the effectiveness of various monitoring strategies, and to show how this
-varies with landscape condition, we begin by simulating example demographic trajectories and observations of those trajectories.
-To simulated example trajectories we use a modified version of Johnson et al's [-@johnson_science_2020] demographic model described by Dyson et al.
+varies with landscape condition, we begin by simulating example demographic trajectories and observations of those trajectories using a modified version of Johnson et al's [-@johnson_science_2020] demographic model described by Dyson et al.
 [-@dyson_existing_2022].
@@
 Together, a set of example trajectories from the national model represent alternative hypotheses about the true state of a particular caribou population.
-To assess the accuracy of population status assessments and the expected value of alternative monitoring strategies [@runge_which_2011; @dunham_strategic_2023] we analyze each trajectory of simulated observations using a Bayesian model (section 2.4).
+To assess the accuracy of population status assessments and the expected value of alternative monitoring strategies [@runge_which_2011; @dunham_strategic_2023] we analyze each trajectory of simulated observations using the Bayesian model (section 2.4).
 We also use the sets of simulated trajectories to confirm that prior predictions from our Bayesian model are consistent with knowledge of the observed distribution of outcomes for caribou across the country [@johnson_science_2020].
@@
 The post-juvenile female population in the next year includes both
-survivors and new recruits:
-
-\begin{equation}
-\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t),
-  (\#eq:dotN)
-\end{equation}
-
-where the maximum potential population growth rate is $r_{max}=1.3$.
+survivors and new recruits: $\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t)$, where the maximum potential population growth rate is $r_{max}=1.3$.
 
@@
 
-In our simulated example populations, mortality risk does not vary over
-among seasons or individuals, so the number of collared cows at the time of the composition survey $\hat{T}_t$ depends on the true survival rate $\dot{S}_t$ and the number of collared cows at the beginning of the year $\hat{I}_t$ as follows:
-\begin{equation}
-\hat{T}_t \sim \text{Binomial}(\hat{I}_t,\dot{S}_t).
-  (\#eq:hatI)
-\end{equation}
-
-The number of cows in the composition survey $\hat{W}_t$ is given by the
-number of collared cows $\hat{T}_t$ and the apparent number of adult
-females per collared female observed in the composition survey $w$:
-\begin{equation}
-\hat{W}_t = w\hat{T}_t.
-  (\#eq:hatW)
-\end{equation}
-The number of observed calves $\hat{J}_t$ also depends on
-the unadjusted apparent recruitment for the population $\dot{R}_t$:
+In our simulated example populations the number of collared cows at the time of the composition survey $\hat{T}_t$ depends on the true survival rate $\dot{S}_t$ and the number of collared cows at the beginning of the year $\hat{I}_t$: $\hat{T}_t \sim \text{Binomial}(\hat{I}_t,\dot{S}_t)$.
+The number of cows in the composition survey $\hat{W}_t$ is given by the number of collared cows $\hat{T}_t$ and the apparent number of adult
+females per collared female observed in the composition survey $w$: $\hat{W}_t = w\hat{T}_t$.
+The number of observed calves $\hat{J}_t$ also depends on the unadjusted apparent recruitment for the population $\dot{R}_t$:
+
 \begin{equation}
@@
 
-Where appropriate, the Bayesian and national population models share the same structure, and we have selected notation to highlight the similarities. The difference between equations
-\@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an
+The difference between equations \@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an
 unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to
@@
 
-Survival $S_t$ is estimated from the observed number of collared cows at the beginning of each year $\hat{I}_t$ and the number of those that survive the year $\hat{T}_t$ assuming a Binomial distribution:
-\begin{equation}
-\hat{T}_t \sim \text{Binomial}(\hat{I}_t,S_t).
-  (\#eq:hatSEst)
-\end{equation}
-
-Note that for analysis of real known-fate radio collar data data one should use a survival analysis method that accounts for variation throughout the year in both survival rate and the number of collared animals [e.g.
+Survival $S_t$ is estimated from the observed number of collared cows at the beginning of each year $\hat{I}_t$ and the number of those that survive the year $\hat{T}_t$ assuming a Binomial distribution: $\hat{T}_t \sim \text{Binomial}(\hat{I}_t,S_t)$.
+Note that real known-fate radio collar data should be analyzed with a method that accounts for variation in survival rate and the number of collared animals throughout the year [e.g.
 @dalgarno_bboutools_2024; @eacker_web-based_2019; @lamb_effectiveness_2024].
@@
 \@ref(fig:examplePredictions) and \@ref(fig:figDiffs)).
-When local data is limited (e.g.
-4 years, 15 collars per year) the model with informative priors yields predictions that are consistent with knowledge of observed outcomes across the country, and the model with weakly informative priors does not (Figure
-\@ref(fig:examplePredictionsLowEffort)). As expected,
-uncertainty about the population growth rate decreases as sampling
-effort in each year (i.e. number of collars) or the number of years of
-monitoring (Figure \@ref(fig:figDiffs)) increase.
-Interannual variation makes reducing uncertainty challenging, and insufficiently informative priors can yield predictions that are both imprecise and inaccurate even in a case where there is 16 years of local monitoring data (Figure \@ref(fig:examplePredictions)).
-Increasing the forecast horizon [@petchey_ecological_2015] does not increase uncertainty about expected population growth (compare 5 and 20 year projections in
+When local data are limited (e.g.
+4 years, 15 collars per year) the model with informative priors (i.e.
+knowledge of national relationships) yields predictions that are consistent with knowledge of observed outcomes across the country, and the model with weakly informative priors does not (Figure \@ref(fig:examplePredictionsLowEffort)).
+When there is interannual variation increasing the duration of monitoring improves predictions of population growth rate substantially, and increasing the number of collars each year is less effective (Figure \@ref(fig:figDiffs)).
+Insufficiently informative priors can yield predictions that are both imprecise and inaccurate even in a case where there are 16 years of local monitoring data (Figure \@ref(fig:examplePredictions)), highlighting the benefits of integrating national relationships with these local data.
+Increasing the forecast horizon [@petchey_ecological_2015] does not substantially decrease the accuracy of expected population growth projections (compare 0, 5 and 20 year projections in
 Figure \@ref(fig:figDiffs)) in these scenarios.
@@
 When anthropogenic disturbance is high (right panels of Figures
-\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
-of population status informed only by priors from the national
-demographic-disturbance relationships are accurate, and adding
-additional local monitoring data does not improve accuracy.
-Interestingly, when the projection period is long and anthropogenic
-disturbance is high, we see a negative EVSI (lower right panel of Figure
-\@ref(fig:figEVSI)), but we do not see an associated decrease in the
-probability of correctly assessing population status (lower right panel
-of Figure \@ref(fig:figStatusErrors)). This means that the long-term
-projections informed by local monitoring data are less
-precise but not less accurate, which can happen if the prior viability predictions are more certain than the posterior predictions (e.g left side of right panels in Figure \@ref(fig:examplePredictions)).
-At
-lower levels of disturbance, local monitoring can increase accuracy, but
-there are diminishing returns as the number of years of monitoring increases (Figures
+\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy.
+At lower levels of disturbance, local monitoring can increase accuracy, but there are diminishing returns as the number of years of monitoring increases (Figures
 \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).
-Increasing accuracy is more difficult when interannual variation is high (Figures \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
-relationships between monitoring effort and the probability of correct
-status assessment depend on how anthropogenic disturbance is changing
-over time, but there are diminishing benefits of increased monitoring
-regardless of whether disturbance is increasing or decreasing
-(Supplement D). Acknowledging the possibility of composition survey errors increases uncertainty, but these increases are not large for the set of parameters we investigate
-(Supplement F).
-
-Averaging over the full set of example trajectories for each scenario (Figures \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)) obscures variation among them that is important for monitoring decisions.
-For example, where disturbance is low but increasing, in the near-term monitoring is most valuable for identifying populations with lower than expected population growth rates (row 1 of Figure \@(fig:figStatusErrorsLow)), but refining knowledge of the current population state also helps identify populations that are likely to remain viable as disturbance increases (row 3 of Figure \@(fig:figStatusErrorsLow)).
-Status assessment accuracy is much higher among examples where the 95% credible interval for expected population growth rate does not include one (xx %) than among examples where it does (xx %), and this information could be used to guide monitoring.
-A more nuanced option is to recognize that the posterior probability that the population is viable ($P(\psi|x,\tau)$ in eq \@ref(eq:EVSI)) is an indicator of error probability (Figure \@ref(fig:figProbViability)).
+Increasing accuracy is more difficult when interannual variation is high (Figures \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).
+The shapes of the
+relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring
+regardless of whether disturbance is increasing or decreasing (Supplement D).
+Acknowledging the possibility of composition survey errors reduces accuracy, but the reductions are not large for the set of parameters we investigate (Supplement F).
+
+Averaging over the full set of example trajectories for each scenario (Figures \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)) obscures variation among them that is also important for monitoring decisions.
+For example, where disturbance is low but increasing, monitoring helps identify populations that are faring less well than expected now (top left of Figure \@ref(fig:figStatusErrorsLow)), and also helps identify populations that are likely to remain viable in the future when disturbance is higher (bottom right of Figure \@ref(fig:figStatusErrorsLow)).
+Status assessment accuracy is much higher when the 95% credible interval for expected population growth rate does not include one (`r round(100-erSums$probWrong[1],2)` %) than when it does (`r round(100-erSums$probWrong[2])` %), and this information could be used to guide monitoring.
+A more nuanced option is to recognize that the posterior probability of population viability ($P(\psi|x,\tau)$ in eq \@ref(eq:EVSI)) is an indicator of error probability (Figure \@ref(fig:figProbViability)).
+The probability of making an error also depends on the true (unknown) level of bias in the composition survey, but as the estimated posterior probability of viability approaches 0 or 1 accuracy increases to 100% even when the survey is biased (Figure \@ref(fig:figProbViability)).
 One could inform a decision about whether more information is needed for a particular population by selecting an error tolerance (e.g.
-the probability of correct status assessment should be >0.98) and using that threshold to distinguish cases where the posterior probability of viability is sufficiently certain (i.e.
+the probability of correct status assessment should be >95 %) and using that threshold to distinguish cases where the posterior probability of viability is sufficiently certain (i.e.
 near enough to 0 or 1) from cases where more information is needed (Figure \@ref(fig:figProbViability)).
+A key advantage of accounting for composition bias uncertainty in the Bayesian model is that when the model indicates high confidence (i.e.
+the estimated probability of viability very close to 0 or 1) the answer is likely to be correct (high status assessment accuracy) even when the composition survey is biased (Figure \@ref(fig:figProbViability)).
 
@@
 will be used as a reason to delay conservation action for highly
-disturbed and poorly monitored populations [@bissonnette_controverses_2020; @boan_climate_2018]. Integrating
-demographic-disturbance relationships into a viability projection model
-clarifies that no additional information is needed to project outcomes
-where disturbance is high and actions to support recovery are not taken.
-
-There are a variety of possible actions intended to improve survival and
+disturbed and poorly monitored populations [@bissonnette_controverses_2020; @boan_climate_2018]. Accounting for
+demographic-disturbance relationships in a viability projection model
+clarifies that no additional local information is needed to project population decline
+where disturbance is high and actions to support recovery are not taken; the distribution of observed demographic rates from across the country (as summarized by the national analysis) is sufficient to predict problems for caribou on highly disturbed landscapes.
+
+Interventions such as wolf culls, maternal penning, and restoration of
+linear features are intended to improve survival and
 reproduction of populations that are not otherwise self-sustaining
-[@mcnay_demographic_2022; @serrouya_predicting_2020]. The simple models
-in this paper do not include these actions, and cannot be used to
-project outcomes for managed populations. Our intention is to reduce
+[@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a
+need to evaluate and compare effects of boreal caribou conservation
+actions with careful monitoring. Our focus in this paper is on assessing
+and projecting viability of populations in the absence of
+these interventions, to clarify when additional information is needed to
+reduce uncertainty about whether there is a need for action. The model we
+propose does not include effects of management interventions, and
+cannot be used to project outcomes for managed populations. Our intention is to reduce
 uncertainty that can be used to delay conservation action; there is
@@
 actions. Even when no additional monitoring is required for assessing
-self-sustainability, there are a variety of reasons to continue
+viability, there are a variety of reasons to continue
 monitoring highly disturbed boreal caribou populations. The point we wish to emphasize is
@@
 self-sustaining caribou populations at a particularly high level of
-disturbance, it is unreasonable to project or expect this outcome. These
-premises constrain the plausible demographic-disturbance priors, and
-clarify that no additional information is needed to predict outcomes for
-caribou where anthropogenic disturbance is high (Figure 5) and no
-meaningful actions are taken to support recovery.
+disturbance, it is unreasonable to project or expect this outcome.
 
@@
 
-Another novel aspect of this work is a term that acknowledges possible bias
-caused by misclassification of adults and missing calves in
-composition surveys [@ellington_divergent_2020;
-@decesare_estimating_2012]. Lacking information about the true magnitude
-of these errors, we examined sensitivity to a range of assumed values, and
-found that while acknowledging bias did increase uncertainty, these
-increases were not large (Supplement E). A key reason for this result is
-that we assumed the same range of values for the probability of
-misclassifying adults and the probability of missing calves. Presuming
-that surveyed groups contain more cows than bulls, and that there is an
-equal probability of misclassifying cows as young bulls and vice versa,
-then misclassification errors cause high estimated recruitment. Missing
-calves causes low estimated recruitment, and together these errors tend
-to cancel one another out.
-Given lack of knowledge about true composition survey error rates, the most appropriate interpretation of our results is that better information is required in order to assess the importance of these errors.
+Another novel aspect of this work is a term that acknowledges possible bias caused by misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020;@decesare_estimating_2012].
+Lacking information about the true magnitude of these errors, we examined the accuracy and precision of a Bayesian model that includes uncertainty about composition bias using examples with biased composition surveys.
+Acknowledging bias did decrease overall accuracy and precision, though these decreases were not large (Supplement E).
+Importantly, when the Bayesian model predictions were precise they were also accurate even if the true composition survey was biased (Figure 9).
+In other words, acknowledging composition bias uncertainty made predictions less confident, but in cases where predictions were confident they were also trustworthy.
+
+A potential limitation of our analysis is that we assumed the same range of values for the probability of misclassifying adults and the probability of missing calves.
+Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows as young bulls and vice versa, then misclassification errors cause high estimated recruitment.
+Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out.
+An interesting extension of this work would be to investigate model performance over a broader range of assumptions about composition survey error rates, and to examine the implications of mismatches between the true distribution of errors and the unknown distribution in the Bayesian model.
+Better information about true composition survey error rates would also help assess their importance.
 
@@
 collaring [@ellington_divergent_2020]. In an analysis of fecal pellets from 14 populations of boreal and southern mountain caribou, Jones et al. [-@jones_novel_2023] found that sampled
-groups contained more females than males, and that mothers tend to stay
-with their daughters and sons, while fathers do not. These results are
+groups contained more females than males, and that cows tend to stay
+with their male and female offspring, while bulls do not. These results are
 broadly consistent with our assumption that composition survey groups
-include mothers and younger animals of both sexes that can be difficult
+include cows and younger animals of both sexes that can be difficult
 to classify visually. Documenting animals that are difficult to classify
@@
 @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011;
-@johnson_science_2020; @dalgarno_bboutools_2024] will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is
-needed.
+@johnson_science_2020; @dalgarno_bboutools_2024] will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is needed.
 
@@
 
-The main objective of Johnson et al's [-@johnson_science_2020;
-@eccc_scientific_2011] national analysis was to inform a national
-disturbance policy that applies to all ranges except SK1, and to inform
+The main objectives of Johnson et al's [-@johnson_science_2020;
+@eccc_scientific_2011] national analysis were to inform a national
+disturbance policy that applies to all ranges except one in northern Saskatchewan (SK1) that is subject to very low anthropogenic disturbance and high fire disturbance, and to inform
 a different policy for SK1 [@eccc_scientific_2011; @eccc_boreal_2019]. A
@@
 suited for estimating variation in demographic model parameters among
-populations, and distinguishing that from variation among years.
+populations, and for distinguishing that from variation among years.
 We have not reanalyzed the national data using different methods because the data are owned by many jurisdictions, and obtaining permission from all of these jurisdictions would be difficult and time-consuming.
@@
 
-For this analysis we used Kaplan-Meier estimates of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ [as in @eacker_web-based_2019; @lamb_effectiveness_2024], and assumed Beta distributed interannual variability [as in @johnson_science_2020], but we acknowledge there are other possible approaches [e.g.
-@dalgarno_bboutools_2024].
-The proof-of-concept model presented in this paper is adequate for making general points about the integration of national
-demographic-disturbance relationships with local data, and accounting for
-composition survey errors.
+For this analysis we used a simple survival analysis method that does not account for variation in survival rate or the number of collared animals throughout the year.
+More sophisticated methods should be used for analysis of real survival data, but our implementation of a two-step non-parametric survival analysis method [following @eacker_web-based_2019] yielded slightly biased survival estimates.
+Dalgarno et al.’s [-@dalgarno_bboutools_2024] method may be a better alternative.
+More generally, the proof-of-concept model presented in this paper is adequate for making general points about the integration of national demographic-disturbance relationships with local data, and for accounting for composition survey errors.
 However, it is a research prototype, not a fully developed analysis tool.
 The more fully developed and documented *bboutools* R package [@dalgarno_bboutools_2024] is likely a better starting point for development of tools to enable local decision makers to inform decisions with their existing data.
-A possible next step is to extend that package to include demographic-disturbance relationships and composition survey errors.
+Possible next steps are to extend that package to include priors from demographic-disturbance relationships and composition survey errors.
 
@@
 among individuals [@armstrong_incorporating_2021], population size
-information from aerial or fecal dna surveys, and variation in demographic rates among age classes [@moeller_integrating_2021; @mcnay_demographic_2022] into Bayesian population models.
-However, in many contexts there is not enough to data to estimate more complex model parameters, and simple models remain relevant.
-
-Our proof-of-concept analysis shows the value of integrating national demographic-disturbance relationships with local data, accounting for
-composition survey errors, and assessing the value of local monitoring
+information from aerial or fecal DNA surveys, and variation in demographic rates among age classes [@moeller_integrating_2021; @mcnay_demographic_2022] into Bayesian population models.
+However, in many contexts there are not enough to data to estimate more complex model parameters, and simple models remain relevant.
+
+Our proof-of-concept analysis shows the value of integrating national demographic-disturbance relationships with local data, accounting for composition survey errors, and assessing the value of local monitoring
 information. We applied the framework proposed by Dunham et al. (2023) to show that
 no additional local monitoring information is needed to predict whether
-boreal caribou populations in highly disturbed landscapes are
-self-sustaining. There is more variation in caribou demographic rates
+boreal caribou populations in highly disturbed landscapes are viable in the absence of interventions such as predator control or maternal penning; self-sustainability is unlikely.
+There is more variation in caribou demographic rates
 among study areas with low to moderate disturbance, and thus more need
@@
 
-Figure 3: Prior (a) and posterior predictions from the Bayesian model for population trajectories with informative (a,b) and weakly informative (c) priors given lower interannual variation ($v_R=0.23,v_S=0.435$) and intensive monitoring.
+Figure 3: Prior (a) and posterior predictions from the Bayesian model for population trajectories with informative (a,b) and weakly informative (c) priors given lower interannual variation ($v_R=0.23,v_S=0.435$) and long-term monitoring.
 In this example scenario anthropogenic disturbance increases by 2% per year from 0 to 100%.
@@
 Figure 6: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
-When disturbance is high the probability of correct status assessment is high even without local monitoring.
+When disturbance is high the probability of correct status assessment is high even with one year of monitoring.
 Where disturbance is low or moderate, increasing the number of years of monitoring increases accuracy, but there are diminishing returns on additional monitoring as the number of years increases.
@@
 Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases.
-Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction.
 See Figure \@ref(fig:figStatusErrors) for additional scenario details.
 
-Figure 8: In the low anthropogenic disturbance scenario (Figure \@ref(fig:figDisturbance)) probability of a correct status assessment varies with the true growth rate of the population (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
-When anthropogenic disturbance is low and expected to increase (Figure \@ref(fig:figDisturbance)) increasing the number of years of monitoring helps to distinguish populations with current demographic rates substantially lower than 1 (top left), and to identify populations with higher than expected growth rates that fare better as disturbance increases (lower right).
+Figure 8: When anthropogenic disturbance is low and expected to increase (low scenario in Figure \@ref(fig:figDisturbance)) the probability of a correct status assessment varies with the true growth rate of the population (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
+Increasing the number of years of monitoring helps identify populations with currently declining populations (top left), and to identify populations with higher than expected growth rates that fare better as disturbance increases (lower right).
 Increasing the number of collars more effectively increases near-term accuracy when interannual variation is low (top left).
-There is little advantage to increasing the number of collars if the goal is long-term projection (bottom row).
-Increasing the accuracy of status assessment for populations with growth rates close to 1 is more difficult (middle columns).
+If the goal is long-term projection (bottom row) there is little advantage to increasing the number of collars.
+Increasing the accuracy of status assessment for populations with growth rates close to one is difficult (middle columns).
 In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year.
@@
 Figure 9: The posterior probability that the population is viable ($P(\psi|x,\tau)$ in eq \@ref(eq:EVSI)) predicts the probability of a correct status assessment (estimated from the 114,000 example cases in Figures 5 to 7 with a generalized additive model).
-Status assessment accuracy increases as the probability of viability nears 0 or 1.
-Cases in which there is no need for additional monitoring if one is willing to accept a 2% chance of error fall above the dashed line.
+The relationship between accuracy and the probability of viability depends on true composition survey bias (colours), but regardless of bias accuracy is high when the probability of viability is very near to 0 or 1.
+Cases in which there is no need for additional monitoring if one is willing to accept a 5% error rate fall above the dashed line.
 

### 2025-02-21 analysis/paper/supplement.Rmd
preparing EI submission

@@
 
-resultScenario = "S11"
+resultScenario = "S15"
 noBiasResultScenario = "S14"
@@
 
-Table A1. Variables and parameters in the Bayesian population model and the national demographic model.
+Table A1.
+Variables and parameters in the Bayesian population model and the national demographic model.
+Bayesian model estimates are distinguished from observed values, estimated expected values, true values, and true expected values using $\hat{\text{ }}$ $\tilde{\text{ }}$ $\dot{\text{ }}$ and $\bar{\text{ }}$ symbols respectively; for example, $S_t$ is the Bayesian estimate of survival in year $t$, $\hat{S}_t$ is observed survival, $\tilde{S}_t$ is the Bayesian estimate of expected survival (without stochastic interannual variation), $\dot{S}_t$ is true survival, and $\bar{S_t}$ is true expected survival (without stochastic interannual variation).
 
@@
 
+For this analysis we assume the same range of values for the probability of misclassifying adults, and the probability of missing calves (Figure \@ref{fig:biasDistribution}).
+Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment.
+Missing calves causes low estimated recruitment, and on average these errors tend to cancel one another out, but the set of example trajectories does include cases with high and low bias terms (Figure \@ref{fig:biasDistribution}).
+
 ```{r biasDistribution,fig.cap="In the Bayesian model, uncertainty about the value of the bias correction term $c$ (grey bars) is approximated with a Log-normal distribution (red).
@@
 ```{r figDisturbanceAnthroStatic, fig.cap="Static anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult6ay1aSf0.png")))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult9ay1aSf0.png")))
 
@@
 
-```{r powercmult6ay1aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios).
-In these examples there are 6 cows per collared cow in recruitment surveys.
-See Figure 4 for additional details.
-", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult6ay1aSf0.png")))
-```
-
 ```{r powercmult9ay1aSf0, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroStatic) for static anthropogenic disturbance scenarios).
@@
 ```{r figDisturbanceAnthroDeclining, fig.cap="Declining anthropogenic disturbance and monitoring scenarios.", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult6ay1aSf-1.png")))
+knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/distScnscmult9ay1aSf-1.png")))
 ```
@@
 knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult3ay1aSf-1.png")))
-```
-
-```{r powercmult6ay1aSfm1, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbanceAnthroDeclining) for declining anthropogenic disturbance scenarios).
-In these examples there are 6 cows per collared cow in recruitment surveys.
-See Figure 4 for additional details.
-", cache=TRUE}
-knitr::include_graphics(here::here(paste0("figs/",resultScenario,"/powercmult6ay1aSf-1.png")))
 ```
@@
 
-# Sensitivity to unknown bias in recruitment estimates from calf:cow surveys
+# Effects of unknown bias in recruitment estimates from calf:cow surveys
 
-Adding unknown unknown bias does increase uncertainty, but these increases are not large (compare Figs 6 and \@ref(fig:figStatusErrorsNoBias)).
-A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults, and the probability of missing calves (supplement B).
-Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment.
-Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out.
+Adding unknown bias does increase uncertainty and reduce accuracy, but these increases are not large (compare Figs 5 to 9 with Figs \@ref(fig:figDiffsNoBias) to \@ref(fig:figProbViabilityNoBias)) given our assumptions about the the range of plausible values for the probability of misclassifying adults, and the probability of missing calves (supplement B).
+In both versions of the model the estimated posterior probability of viability approaches 0 or 1 accuracy increases to 100% (compare Fig 9 and \@ref(fig:figProbViabilityNoBias)).
+Note that these conclusions are from a case in which assumed the same range of values for the probability of misclassifying adults, and the probability of missing calves (supplement B).
 
@@
 
-```{r figStatusErrorsLowNoBias, fig.cap=".", cache=TRUE}
+```{r figEVSIsNoBias, fig.cap="The expected value of sample information without bias in recruitment estimates from calf:cow surveys.
+In these examples there are 6 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
+See Figure 7 for additional details.", cache=TRUE}
+knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/EVSIcmult6ay1aSf1.png")))
+```
+
+```{r figStatusErrorsLowNoBias, fig.cap="Variation in the probability of a correct status assessment with the true growth rate of the population the low anthropogenic disturbance scenario without bias in recruitment estimates from calf:cow surveys.
+In these examples there are 6 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1).
+See Figure 8 for additional details.", cache=TRUE}
 knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powerLamcmult6ay1aSf1.png")))
@@
 
-```{r figProbViabilityNoBias, fig.cap=".", cache=TRUE}
+```{r figProbViabilityNoBias, fig.cap="Relationship between posterior probability that the population is viable ($P(\\psi|x,\\tau)$) and the probability of a correct status assessment (estimated from the 114,000 example cases in Figures \\@ref(fig:figDiffsNoBias) to Figures \\@ref(fig:figEVSIsNoBias) with a generalized additive model) without bias in recruitment estimates from calf:cow surveys.", cache=TRUE}
 knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/probViabilitycmult6ay1aSf1.png")))

### 2025-02-21 R/CaribouDemographicFunctions.R

@@ -6,8 +6,7 @@
 # functions for caribou demographics paper
 
 # run scenario - temporary wrapper
-runScenario<-function(scns,quants=NULL,Anthro=NULL,survAnalysisMethod="KaplanMeier",getKSDists=T){
-  #quants=NULL;Anthro=NULL;survAnalysisMethod="Exponential";getKSDists=F
+runScenario<-function(scns,quants=NULL,Anthro=NULL,survAnalysisMethod="Binomial",getKSDists=T){
   retdir <-getwd()
 
   dir.create("figs")

### 2025-02-21 analysis/scripts/0_PriorsFromPIs.R

@@ -19,7 +19,7 @@ johnsonCompare <- read.csv(paste0(baseDir,"/data/Johnson et al. figures5_6.csv")
 
 
 getAnswers <-function(scns,par,johnsonCompare,vName){
-  scResults = runScenario(scns,survAnalysisMethod = "Exponential")
+  scResults = runScenario(scns,survAnalysisMethod = "Binomial")
   rr = subset(scResults$rr.summary.all,Parameter == par)
   rr$ylow = rr[["Lower 95% CRI"]]
   rr$yhigh = rr[["Upper 95% CRI"]]

### 2025-02-21 analysis/scripts/0_calibratePriors.R

@@ -15,7 +15,7 @@ scnsSInt=expand.grid(obsAnthroSlope = 0, projAnthroSlope = 5,sIntSE=sIntSE,
 scnsSInt$NuMin = scnsSInt$sNuMin
 scns=merge(scnsSInt,scnsApplyAll);scns=merge(scns,scnsNoData)
 scns$projYears=20
-scResults = runScenario(scns,survAnalysisMethod = "Exponential")
+scResults = runScenario(scns,survAnalysisMethod = "Binomial")
 KSAll = subset(scResults$ksDists,(Year==2023+scnsNoData$projYears)&(Parameter=="Adult female survival"))
 addBit = paste0("InterceptAll")
 scResults$obs.all = NULL
@@ -33,7 +33,7 @@ scnsRInt=expand.grid(obsAnthroSlope = 0, projAnthroSlope = 5,rIntSE=rIntSE,
 scnsRInt$NuMin = scnsRInt$rNuMin
 scns=merge(scnsRInt,scnsApplyAll);scns=merge(scns,scnsNoData)
 scns$projYears=20
-scResults = runScenario(scns,survAnalysisMethod = "Exponential")
+scResults = runScenario(scns,survAnalysisMethod = "Binomial")
 KSAll = subset(scResults$ksDists,(Year==2023+scnsNoData$projYears)&(Parameter=="Recruitment"))
 addBit = paste0("InterceptAll")
 scResults$obs.all = NULL

### 2025-02-21 analysis/scripts/1_sensitivity1.R

@@ -54,9 +54,9 @@ monitoringScns = expand.grid(obsYears=c(1,2,4,8,16,24),collarCount=c(0,15,30,60)
                              assessmentYrs=c(1))
 #TO DO - in next iteration, remove multiple years of one collar
 monitoringScns = subset(monitoringScns, !((obsYears>1)&(collarCount==0))&!((collarCount==1)&(cowMult>3))&!((collarCount==0)&(collarInterval>1)))
-stateScns = data.frame(tA=c(0,20,40,60,60,40,20),
-                       obsAnthroSlope=c(0,0,0,0,0,0,0),
-                       projAnthroSlope=c(0,0,0,0,-1,-1,-1))
+stateScns = data.frame(tA=c(0,20,40,60,60,40,20,0,20,40,60),
+                       obsAnthroSlope=c(0,0,0,0,0,0,0,0,1,1,1),
+                       projAnthroSlope=c(0,0,0,0,-1,-1,-1,1,1,1,1))
 stateScns = merge(stateScns,data.frame(rep=seq(1:400)))
 
 stateScns$sQuantile=runif(nrow(stateScns),min=0.01,max=0.99)
@@ -71,6 +71,8 @@ scns$projYears = 20
 unique(scns$iAnthro)
 scns$repBatch = ceiling(scns$rep/12)
 table(scns$repBatch)
+table(unique(subset(scns,select=c(repBatch,rep)))$repBatch)
+
 scns$N0 = 5000
 
 scns$pageLab = paste0("cmult",scns$cowMult,"ay",scns$assessmentYrs,"aSf",scns$projAnthroSlope,"repBatch",scns$repBatch)

### 2025-02-21 analysis/scripts/2_viewSensitivityResults.R

@@ -6,7 +6,8 @@ library(RColorBrewer)
 library(scales)
 
 pal2 = brewer.pal(7,"RdBu")[c(2,6)]
-pal4 = brewer.pal(5,"RdBu")[c(1,2,4,5)]
+pal4 = brewer.pal(5,"RdBu")[c(2,1,5,4)]
+pal3 = pal4[2:4]
 
 pal4b = brewer.pal(5,"Purples")[c(2,3,4,5)]
 
@@ -14,12 +15,13 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 
 ########################
 #sensitivity
-setName = "s14" #need to do s13, s14, and s15
+setName = "s15" #need to do s13, s14, and s15
 dir.create(paste0("figs/",setName),recursive=T)
 dir.create(paste0("tabs/",setName),recursive=T)
 
 scns = read.csv(here::here(paste0("tabs/",setName,".csv")))
 
+scns = scns[order(scns$pageId),]
 
 numBatches= max(scns$repBatch)
 str(scns)
@@ -46,8 +48,10 @@ setLTYVar <-function(scns){
 scns = setLTYVar(scns)
 if(is.element("interannualVar",names(scns))){
   ltyLabel = "Interannual\nvariation"
+  ltySel="none"
 }else{
   ltyLabel = "Collar\nrenewal\ninterval"
+  ltySel=1
 }
 
 
@@ -71,7 +75,7 @@ addEV = F
 addProbs = F
 
 for(i in 1:length(pages)){
-  #combine=F;i=12
+  #combine=F;i=1
   cpageId=pages[i]
 
   if(i==length(pages)){
@@ -97,7 +101,7 @@ for(i in 1:length(pages)){
     scResults = readRDS(paste0("results/",setName,"/rTest",cpageId,".Rds"))
   }
 
-  if((as.numeric(strsplit(p,"repBatch")[[1]][2])!=numBatches)&(as.numeric(strsplit(nextP,"repBatch")[[1]][2])==1)){#<=as.numeric(strsplit(p,"repBatch")[[1]][2]))){
+  if((as.numeric(strsplit(nextP,"repBatch")[[1]][2])==1)){#<=as.numeric(strsplit(p,"repBatch")[[1]][2]))){
     #print(p,nextP)
     combine=F
   }else{combine=T;next}
@@ -109,7 +113,7 @@ for(i in 1:length(pages)){
   head(scResults$rr.summary.all)
   unique(scResults$rr.summary.all$collarInterval)
[…]

### 2025-02-21 analysis/scripts/3_EVSIPlots.R

@@ -8,7 +8,8 @@ library(caribouMetrics)
 library(RColorBrewer)
 
 pal2 = brewer.pal(7,"RdBu")[c(2,6)]
-pal4 = brewer.pal(5,"RdBu")[c(1,2,4,5)]
+pal4 = brewer.pal(5,"RdBu")[c(2,1,5,4)]
+pal3 = pal4[2:4]
 
 pal4b = brewer.pal(5,"Purples")[c(2,3,4,5)]
 
@@ -21,7 +22,7 @@ batchStrip<-function(l,batches=c(10,seq(1:9))){
   return(l)
 }
 
-setName = "s14"
+setName = "s15"
 probsSum = read.csv(here::here(paste0("tabs/EVsample",setName,".csv")),stringsAsFactors = F)
 
 probsSum$pageLabC = batchStrip(probsSum$pageLab)
@@ -85,15 +86,16 @@ probsSum$grp = paste(probsSum$collarCount,probsSum$ltyVariable)
 probsSum$RenewalInterval=as.factor(probsSum$ltyVariable)
 probsSum$NumCollars = as.factor(probsSum$collarCount)
 
+probsSum = probsSum %>% relocate(RenewalInterval,.after=NumCollars)
 
 probsSum$CollarYrs = as.numeric(as.character(probsSum$NumCollars))*probsSum$obsYears
 EVuncertainty$YearsOfProjection=as.factor(EVuncertainty$YearsOfProjection)
 for(pp in pagesC){
   #pp=pagesC[1]
-  base=ggplot(subset(probsSum,pageLabC==pp),aes(x=obsYears,y=EVSI,col=NumCollars,linetype=RenewalInterval,group=grp))+geom_line()+
-    facet_grid(YearsOfProjection~AnthroScn,labeller="label_both")+labs(color="Number of\n collars", linetype=ltyLabel)+
+  base=ggplot(subset(probsSum,pageLabC==pp),aes(x=obsYears,y=EVSI,linetype=RenewalInterval,col=NumCollars,group=grp))+geom_line()+
+    facet_grid(YearsOfProjection~AnthroScn,labeller="label_both")+labs(color="Number of\ncollars", linetype=ltyLabel)+
     theme_bw()+xlab("Years of monitoring")+ylab("Expected value of sample information EVSI")+
-    scale_color_discrete(type=(pal4))
+    scale_color_discrete(type=(pal3))+ scale_x_continuous(breaks=c(0,5,10,15,20))
 
[…]

### 2025-02-21 analysis/scripts/sensitivityMinimalTest.R

@@ -11,7 +11,7 @@ cDir = getwd()
 setName = args[2]
 
 library(caribouMetrics)
-# setName="s10"; cpageId <- 1;n_reps <- 1;cDir = "C:/Users/HughesJo/Documents/gitprojects/Caribou-Demographic-Projection-Paper"
+# setName="s15"; cpageId <- 1;n_reps <- 1;cDir = "C:/Users/HughesJo/Documents/gitprojects/Caribou-Demographic-Projection-Paper"
 
 
 #######################
@@ -51,7 +51,7 @@ if(n_reps=="all"){
   scResults = caribouMetrics:::runScnSet(scns,eParsIn,simBig,getKSDists=F,printProgress=F)
 }else{
   scResults = caribouMetrics:::runScnSet(scns[1:n_reps,],eParsIn,simBig,getKSDists=F,printProgress=F)
-  # scResults = runScnSet(scns[10:10,],eParsIn,simBig,getKSDists=F,printProgress=F)
+  # scResults = runScnSet(scns[950:950,],eParsIn,simBig,getKSDists=F,printProgress=F)
 }
 
 unique(scResults$rr.summary.all$Parameter)

### 2025-02-21 cloud/setup_azure.sh

@@ -94,13 +94,13 @@ az batch task show --job-id $jobName \
 --query "{state: state, executionInfo: executionInfo}" --output yaml
 
 # download output file for a task
-taskNum=12
+taskNum=1
 
 az batch task file download --task-id caribou-demog_sens_batch$taskNum \
 --job-id $jobName --file-path "wd/nohup_"$taskNum".out" \
 --destination "./nohup_"$taskNum".out"
 
-tail -n 20 "./nohup_"$taskNum".out"
+tail -n 30 "./nohup_"$taskNum".out"
 
 rm "./nohup_"$taskNum".out"
 
@@ -127,7 +127,6 @@ az storage remove -c jhughes --account-name ecdcwls --sas-token $sastoken --recu
 #### Delete pool and job ##########################
 az batch job delete --job-id $jobName
 az batch pool delete --pool-id $poolName
-
 
 # downloading and resizing by hand because upload failed
 for ((i=1;i<=30;i++))

### 2025-02-24 analysis/paper/paper.Rmd
submitted.

@@
 abstract: |
-  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed.
-There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but no populations inhabiting areas with high anthropogenic disturbance are considered self-sustaining.
+  Across fifty-eight boreal caribou study areas in Canada, survival and recruitment decrease with the percentage of the study area that is disturbed.
+There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but no populations inhabiting areas with high anthropogenic disturbance are considered viable.
 Demographic projections derived from local population-specific data are uncertain for populations with limited monitoring.
@@
 Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships.
-At lower disturbance levels where initial uncertainty is high, each additional year of monitoring provides less new information.
+At lower disturbance levels where initial uncertainty is high local data improve accuracy but each additional year of monitoring provides less new information.
 The estimated probability of viability indicates whether more information is needed to improve accuracy of population viability projections.
@@
 Resources to support conservation and recovery are limited [@buxton_balancing_2022; @mccarthy_financial_2012;@gerber_conservation_2016; @wintle_spending_2019], delays in action increase risk [@martin_acting_2012], and some populations have been monitored to extinction [@lindenmayer_counting_2013].
-A recent analysis of action plans for Canadian species at risk found that half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022].
+A recent analysis of action plans for Canadian species at risk found that half of recovery actions are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022].
 Analysis of the implications of monitoring outcomes can help to clarify whether additional monitoring is necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;@gregory_structured_2012].
@@
 @venter_sixteen_2016].
-Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the area that is disturbed, and no populations inhabiting areas with high anthropogenic disturbance are considered viable in the absence of interventions such as predator control or maternal penning (i.e.
-self-sustaining) [@eccc_scientific_2011; @johnson_science_2020].
+Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the area that is disturbed, and no populations inhabiting areas with high anthropogenic disturbance are considered viable, defined as a population  experiencing stable or positive growth (≥0.99) in the absence of interventions such as predator control or maternal penning [@eccc_scientific_2011; @johnson_science_2020].
+We acknowledge that other considerations such as population size are important, but our focus in this analysis is population growth rate.
 Identifying populations that are declining is important for clarifying where and when conservation action is needed.
@@
 demographic-disturbance relationships in @johnson_science_2020] with available demographic data to reduce uncertainty in population viability projections.
-Note that for the purposes of this analysis we consider a population to be viable if the population growth rate is > 0.99.
 We also offer a novel method of acknowledging and accounting for errors in composition surveys.
@@
 To investigate how the accuracy and precision of predictions vary with both landscape condition and monitoring effort, we analyze simulated data using a Bayesian model with informative priors (section 2.4) and a composition survey bias term (section 2.1).
-Steps in our analysis procedure include defining anthopogenic disturbance and monitoring scenarios (section 2.2), simulating population dynamics and monitoring (section 2.3) to produce example demographic trajectories for each scenario, and analyzing example demographic trajectories with a Bayesian model that enables integration of (simulated) local demographic data with prior knowledge of demographic-disturbance relationships (section 2.4).
+Steps in our analysis procedure include defining anthropogenic disturbance and monitoring scenarios (section 2.2), simulating population dynamics and monitoring (section 2.3) to produce example demographic trajectories for each scenario, and analyzing example demographic trajectories with a Bayesian model that enables integration of (simulated) local demographic data with prior knowledge of demographic-disturbance relationships (section 2.4).
 Outcomes of interest include expected population growth rate, the probability of correctly assessing population status, the posterior probability of population viability, and the expected value of local information (section 2.5).
@@
 [@ellington_divergent_2020]. Errors that can bias composition survey
-results include misidentifying young bulls as cows or vice versa , and
+results include misidentifying young bulls as cows or vice versa, and
 failing to detect calves [@ellington_divergent_2020;
@@
 
-We focus on exploring a set of scenarios in which buffered anthropogenic
+We focus on exploring a set of scenarios in which anthropogenic
 disturbance, measured as the percentage of area within 500 m of some
@@
 
-In both models recruitment probability is a function of anthropogenic
+In both models, recruitment probability is a function of anthropogenic
 disturbance and fire with a log link and Beta distributed interannual variation [@johnson_science_2020;
@@
 As done by Johnson et al. [-@johnson_science_2020], we model survival
-probability as an adjusted function of buffered anthropogenic
+probability as an adjusted function of anthropogenic
 disturbance $A_t$ with a log link and Beta distributed interannual variation with coefficient of variation $\nu_S$, mean $\tilde{S_t}$ and standard deviation $\sigma_{S,t}$ [@stewart_climate-informed_2023;
@@
 recruitment from the Bayesian model match the 95% prediction intervals
-from Johnson et al's regression model (Supplement C, Figure
+from Johnson et al's (2020) regression model (Supplement C, Figure
 \@ref(fig:examplePredictions)).
@@
 knowledge of national relationships) yields predictions that are consistent with knowledge of observed outcomes across the country, and the model with weakly informative priors does not (Figure \@ref(fig:examplePredictionsLowEffort)).
-When there is interannual variation increasing the duration of monitoring improves predictions of population growth rate substantially, and increasing the number of collars each year is less effective (Figure \@ref(fig:figDiffs)).
+When there is interannual variation, increasing the duration of monitoring improves predictions of population growth rate substantially, and increasing the number of collars each year is less effective (Figure \@ref(fig:figDiffs)).
 Insufficiently informative priors can yield predictions that are both imprecise and inaccurate even in a case where there are 16 years of local monitoring data (Figure \@ref(fig:examplePredictions)), highlighting the benefits of integrating national relationships with these local data.
@@
 linear features are intended to improve survival and
-reproduction of populations that are not otherwise self-sustaining
-[@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a
-need to evaluate and compare effects of boreal caribou conservation
-actions with careful monitoring. Our focus in this paper is on assessing
+reproduction of populations that are not otherwise viable
+[@mcnay_demographic_2022; @serrouya_predicting_2020]. Our focus in this paper is on assessing
 and projecting viability of populations in the absence of
@@
 propose does not include effects of management interventions, and
-cannot be used to project outcomes for managed populations. Our intention is to reduce
-uncertainty that can be used to delay conservation action; there is
+cannot be used to project outcomes for managed populations. There is
 nothing in our arguments, models or results that can or should be used
 to discourage monitoring of the effectiveness of conservation
-actions. Even when no additional monitoring is required for assessing
-viability, there are a variety of reasons to continue
-monitoring highly disturbed boreal caribou populations. The point we wish to emphasize is
+actions. The point we wish to emphasize is
 that action is also required, and lack of monitoring should not be used
@@
 implausible. In other words, if we have seen no examples of
-self-sustaining caribou populations at a particularly high level of
+viable caribou populations at a particularly high level of
 disturbance, it is unreasonable to project or expect this outcome.
@@
 information from aerial or fecal DNA surveys, and variation in demographic rates among age classes [@moeller_integrating_2021; @mcnay_demographic_2022] into Bayesian population models.
-However, in many contexts there are not enough to data to estimate more complex model parameters, and simple models remain relevant.
+However, in many contexts there are not enough data to estimate more complex model parameters, and simple models remain relevant.
 
@@
 no additional local monitoring information is needed to predict whether
-boreal caribou populations in highly disturbed landscapes are viable in the absence of interventions such as predator control or maternal penning; self-sustainability is unlikely.
+boreal caribou populations in highly disturbed landscapes are viable in the absence of interventions such as predator control or maternal penning; viability is unlikely.
 There is more variation in caribou demographic rates
@@
 adequacy and cost-effectiveness of monitoring scenarios in particular
-circumstances. Monitoring costs vary substantially with technologies
+circumstances.
+
+Monitoring costs vary substantially with technologies
 used, the remoteness of herds, and the difficulty of collaring animals,
@@
 Code required to create the data, run analyses and reproduce figures in this paper is available at https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper, and in the caribouMetrics R package available on GitHub at https://github.com/LandSciTech/caribouMetrics/tree/EI-paper-submission.
+Results are available at https://osf.io/ryn59/.
 
 # Acknowledgements
+We thank two anonymous reviewers for their insightful comments and helpful suggestions.
 This work was informed by a variety of discussions with more people than we can name, but here are a few.
@@
 Comments from Joseph Bennett and Daniel Eacker on early versions of this work motivated a variety of improvements.
-Note that our acknowledgement of these contributions does not imply endorsement; we also acknowledge disagreement on some points, and that all remaining errors are our own.
+Note that our acknowledgement of these contributions does not imply endorsement; we also acknowledge disagreement on some points, and that all remaining errors are our own.Funding for this work was provided by the Wildlife and Landscape Science Directorate of Environment and Climate Change Canada.
 
@@
 knitr::include_graphics(here::here("figs/forPaper/probViabilitycmult6ay1aSf1.png"))
-file.rename(here::here("analysis/paper/figs_submit/powerLamcmult6ay1aSf1.pdf"), 
+file.rename(here::here("analysis/paper/figs_submit/probViabilitycmult6ay1aSf1.pdf"), 
             here::here("analysis/paper/figs_submit/Figure9_probViabilitycmult6ay1aSf1.pdf"))

### 2025-02-24 analysis/paper/supplement.Rmd
submitted.

@@
 
-# (APPENDIX) Supplementary Material for Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions {-}
+# (APPENDIX) Supplementary Material for Integration of national demographic-disturbance relationships and local data can improve caribou population viability projections and inform monitoring decisions {-}
 

### 2025-02-24 analysis/scripts/0_bayesianModelExamplesA.R

@@ -262,6 +262,8 @@ ggsave(paste0(baseDir,"/analysis/paper/figs/bayesianModelBiasSensitivity.png"),
 
 ###################################
 #Low effort scenario
+simBig<-getSimsNational(replicates=3000,forceUpdate=T)
+
 monitoringScns = data.frame(obsYears=c(1,4),collarCount=c(0,15),cowMult=c(3),collarInterval=c(1),
                             assessmentYrs=c(1))
 stateScns = data.frame(obsAnthroSlope=c(2),projAnthroSlope=c(2))
@@ -341,5 +343,5 @@ ggsave(paste0(baseDir,"/analysis/paper/figs/bayesianModelExamplesLowEffort.png")
        dpi = 1200)
 
 ggsave(paste0(baseDir,"/analysis/paper/figs_submit/bayesianModelExamplesLowEffort.pdf"),
-       width = 9.6*0.779, height = 9.2, units = "in",
+       width = 9.6*0.779, height = 9.2*2/3, units = "in",
        dpi = 1200)

### 2025-02-24 analysis/scripts/formatScenarios.R

@@ -1,7 +1,7 @@
 
 library(caribouMetrics)
 
-rawScns<-read.csv(here::here("tabs/s10.csv"))
+rawScns<-read.csv(here::here("tabs/s13.csv"))
 nrow(rawScns)
 rawScns = subset(rawScns,collarCount!=1)
 

### 2025-09-16 analysis/scripts/2_viewSensitivityResults.R
Update in progress: main text figures have been updated. Associated text has not been revised.

@@ -15,7 +15,7 @@ scn_defaults <- eval(formals(getScenarioDefaults))
 
 ########################
 #sensitivity
-setName = "s2" #need to do s13, s14, and s15
+setName = "s3" #need to do s3, s14, and s15
 dir.create(paste0("figs/",setName),recursive=T)
 dir.create(paste0("tabs/",setName),recursive=T)
 
@@ -54,7 +54,7 @@ if(is.element("interannualVar",names(scns))&&(length(unique(scns$interannualVar)
 
 for (p in pagesa){
   #p=pagesa[1]
-  if(!file.exists(paste0("results/",setName,"/rTest",p,"all",p,".Rds"))){pages=pages[pages!=p]}
+  if(!file.exists(paste0("results/",setName,"/rTest",p,".Rds"))){pages=pages[pages!=p]}
 }
 
 
@@ -72,7 +72,7 @@ addEV = F
 addProbs = F
 
 for(i in 1:length(pages)){
-  #combine=F;i=20
+  #combine=F;i=1
   cpageId=pages[i]
 
   if(i==length(pages)){
@@ -86,7 +86,7 @@ for(i in 1:length(pages)){
   print(paste(i,p))
 
   if(combine&(cpageId>1)){
-    scNew = readRDS(paste0("results/",setName,"/rTest",cpageId,"all",cpageId,".Rds"))
+    scNew = readRDS(paste0("results/",setName,"/rTest",cpageId,".Rds"))
     for(n in names(scNew)){
       if(n=="errorLog"){
         scResults[[n]]=c(scResults[[n]],scNew[[n]])
@@ -95,7 +95,7 @@ for(i in 1:length(pages)){
       }
     }
   }else{
-    scResults = readRDS(paste0("results/",setName,"/rTest",cpageId,"all",cpageId,".Rds"))
+    scResults = readRDS(paste0("results/",setName,"/rTest",cpageId,".Rds"))
   }
 
   if((as.numeric(strsplit(nextP,"repBatch")[[1]][2])==1)){#<=as.numeric(strsplit(p,"repBatch")[[1]][2]))){
@@ -108,9 +108,9 @@ for(i in 1:length(pages)){
   scResults$obs.all = setLTYVar(scResults$obs.all)
   #figure out how to count out errors.
   head(scResults$rr.summary.all)
-  unique(scResults$rr.summary.all$collarInterval)
+  unique(scResults$rr.summary.all$collarCount)
   #show examples projections
[…]

### 2025-09-16 analysis/scripts/3_EVSIPlots.R

@@ -22,7 +22,7 @@ batchStrip<-function(l,batches=c(10,seq(1:9))){
   return(l)
 }
 
-setName = "s1"
+setName = "s3"
 probsSum = read.csv(here::here(paste0("tabs/EVsample",setName,".csv")),stringsAsFactors = F)
 
 probsSum$pageLabC = batchStrip(probsSum$pageLab)

### 2025-10-02 analysis/paper/paper.Rmd
Update in progress: main text figures have been updated. Associated text has not been revised.

@@
 
-resultScenario = "S13"
+resultScenario = "S3"
 resultSet ="cmult6ay1aSf1"
@@
 
-Figure 6: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
+Figure 6: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
 When disturbance is high the probability of correct status assessment is high even with one year of monitoring.
@@
 
-Figure 7: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort.
+Figure 7: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort.
 When disturbance is high EVSI is low, indicating that local monitoring does not change predicted status.
@@
 
-Figure 8: When anthropogenic disturbance is low and expected to increase (low scenario in Figure \@ref(fig:figDisturbance)) the probability of a correct status assessment varies with the true growth rate of the population (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
+Figure 8: When anthropogenic disturbance is low and expected to increase (low scenario in Figure \@ref(fig:figDisturbance)) the probability of a correct status assessment varies with the true growth rate of the population (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios).
 Increasing the number of years of monitoring helps identify populations with currently declining populations (top left), and to identify populations with higher than expected growth rates that fare better as disturbance increases (lower right).
@@
 
-Figure 9: The posterior probability that the population is viable ($P(\psi|x,\tau)$ in eq \@ref(eq:EVSI)) predicts the probability of a correct status assessment (estimated from the 114,000 example cases in Figures 5 to 7 with a generalized additive model).
+Figure 9: The posterior probability that the population is viable ($P(\psi|x,\tau)$ in eq \@ref(eq:EVSI)) predicts the probability of a correct status assessment (estimated from the 76,000 example cases in Figures 5 to 7 with a generalized additive model).
 The relationship between accuracy and the probability of viability depends on true composition survey bias (colours), but regardless of bias accuracy is high when the probability of viability is very near to 0 or 1.
